### 2015-07-29 ms.Rmd
initial commit

---
title: "Template for writing scientific papers in R markdown"
author: "Tom Harrop"
csl: csl/the-plant-journal.csl
output:
  html_document:
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
    theme: cerulean
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document:
    fig_caption: yes
bibliography: bib/test.bib
---

# Abstract

lorem ipsum :)

# Methods

Doing stuff [@Hrycay:2009ua]. Even more stuff [@Ren:2013jv]. @Harrop:2014dg disagrees. @Engstrom:2013cx wrote a nice paper about RNASeq mapping tools.

# References

### 2015-07-29 ms.Rmd
regenerating genome index for 50b SE reads

@@
 ---
-title: "Template for writing scientific papers in R markdown"
-author: "Tom Harrop"
+title: Genome-wide changes in expression during inflorescence architecture development
+  in rice
 csl: csl/the-plant-journal.csl
 output:
+  word_document:
+    fig_caption: yes
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: yes
   html_document:
@@
     theme: cerulean
-  pdf_document:
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
-  word_document:
-    fig_caption: yes
 bibliography: bib/test.bib
@@
 
-# Abstract
+Authors:
 
-lorem ipsum :)
+Addresses:
 
-# Methods
+### Running title
 
-Doing stuff [@Hrycay:2009ua]. Even more stuff [@Ren:2013jv]. @Harrop:2014dg disagrees. @Engstrom:2013cx wrote a nice paper about RNASeq mapping tools.
+### Keywords
+
+### SRA details for libaries
+
+Word count
+
+***
+
+# Summary
+
+# Significance statement
+
+# Introduction
+
+* Rice inflorescence development
++ branching & grain yield
++ sequential changes in meristem activity
++ few key genes
+
+# Results
+
+## Dissections in early stages of inflorescence development
++  **Figure**: laser microdissections
+
+## Transcriptome sequencing of meristem samples
++ Short (boring) section about sequencing stats
++ **Table** (or supplementary): # reads, mapping %, rRNA, introns, genes blah blah
+
+## Clustering of expression patterns
++ **Figure**: Gene expression clusters
++ GO enrichment?
++ Clusters including 'known' genes
+
+## Patterns of expression of transcription families
++ **Figure**: GSEA of transcription factor families
+
+## MADS genes: comparison of expression & homology in rice, tomato, arabidopsis
++ **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.
+
+# Discussion
+
+# Experimental procedures
+
+# Accession Numbers
+
+# Acknowledgements
+
+# Short legends for Supporting Information
 
 # References
+
+***
+
+# Tables
+
+***
+
+# Figure legends
+
+***
+
+# Figures
+
+***
+
+# Junk
+
+I think `inline code` should work?

### 2015-07-29 src/R/madsComp.R

@@ -1,0 +1,11 @@
+
+
+# On a phylogeny of MADS genes in At, Sl & Os, compare LFCs:
+
+# Arabidopsis (vero's paper): IM vs FM
+
+# Rice: RM vs SM
+
+# Tomato
+# (ftp://ftp.solgenomics.net/user_requests/LippmanZ/public_releases/by_species/Solanum_lycopersicum/transcripts/) :
+# SIM/TM vs. FM

### 2015-07-29 src/sh/downloadReads.sh

@@ -1,0 +1,56 @@
+#!/bin/bash
+
+#SBATCH --job-name dlLibs
+#SBATCH --ntasks 1
+#SBATCH --output /tmp/dl.%N.%j.out
+#SBATCH --open-mode=append
+#SBATCH --mail-type=ALL
+
+# script for downloading publicly available reads for tomato and arabidopsis
+
+# tomato
+
+outdirTom="data/reads/tomato"
+if [[ ! -d $outdirTom ]]; then
+	mkdir -p $outdirTom
+fi
+
+wget --no-parent --directory-prefix "$outdirTom" --recursive --no-directories \
+	--accept wt_{sim,tm,fm}*.tgz ftp://ftp.solgenomics.net/user_requests/LippmanZ/public_releases/by_species/Solanum_lycopersicum/transcripts/ &
+
+# arabidopsis
+
+outdirAt="data/reads/at"
+if [[ ! -d $outdirAt ]]; then
+	mkdir -p $outdirAt
+fi
+
+# IM_R1
+wget --output-document "$outdirAt"/IM_R1.fastq.gz ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR351/ERR351334/ERR351334.fastq.gz &
+
+# IM_R2
+
+wget --output-document "$outdirAt"/IM_R2.fastq.gz ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR351/ERR351333/ERR351333.fastq.gz &
+
+# FM_R1
+
+wget --output-document "$outdirAt"/FM_R1.fastq.gz ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR351/ERR351336/ERR351336.fastq.gz &
+
+# FM_R2
+
+wget --output-document "$outdirAt"/FM_R2.fastq.gz ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR351/ERR351331/ERR351331.fastq.gz &
+
+wait
+
+# email output
+NOW="$(date)"
+MESSAGE="$(cat /tmp/dl."$SLURM_JOB_NODELIST"."$SLURM_JOBID".out)"
+cat <<- _EOF_ | mail -s "[Tom@SLURM] Job "$SLURM_JOBID" finished" tom
+	Job $SLURM_JOBID submitted at $THEN is finished.
+	Concatenated stdout files:
+	$MESSAGE
+_EOF_
+
+rm /tmp/dl."$SLURM_JOB_NODELIST"."$SLURM_JOBID".out
+
+exit 0

### 2015-07-29 src/sh/starGenomeGenerate.sh

@@ -1,0 +1,77 @@
+#!/bin/bash
+
+#SBATCH --job-name stargg
+#SBATCH --ntasks 1
+#SBATCH --cpus-per-task=6
+#SBATCH --output /tmp/stargg.%N.%j.out
+#SBATCH --open-mode=append
+
+THEN="$(date)"
+echo -e "[ "$(date)": Genome generation with STAR ]"
+
+# make output directory
+outdir="output/star-index-"$(date +%F)""
+if [[ ! -d $outdir ]]; then
+	mkdir -p $outdir
+fi
+
+# catch sigkill
+clean_up() {
+	echo -e "[ "$(date)" : Script aborted ]"
+	# email output
+	NOW="$(date)"
+	MESSAGE="$(cat /tmp/stargg."$SLURM_JOB_NODELIST"."$SLURM_JOBID".out)"
+	cat <<- _EOF_ | mail -s "[Tom@SLURM] Job "$SLURM_JOBID" aborted" tom
+	Job $SLURM_JOBID submitted at $THEN was aborted.
+	Concatenated stdout files:
+	$MESSAGE
+_EOF_
+	mv /tmp/stargg."$SLURM_JOB_NODELIST"."$SLURM_JOBID".out "$outdir"/
+	exit 1
+}
+trap clean_up SIGHUP SIGINT SIGTERM
+
+# set parameters
+genomeFastaFiles="data/genome/Osativa_204_v7.0.fa"
+sjdbGTFfile="data/genome/Osativa_204_v7.0.gene_exons.cuffcomp.rRNAremoved.gtf"
+sjdbGTFtagExonParentTranscript="oId"
+sjdbGTFtagExonParentGene="gene_name"
+sjdbOverhang=49
+outFileNamePrefix="$outdir/"
+
+# log metadata
+version="$(STAR --version)"
+cat -t <<- _EOF_ > $outdir/METADATA.csv
+	Script,${0}
+	branch,$(git rev-parse --abbrev-ref HEAD)
+	hash,$(git rev-parse HEAD)
+	date,$(date +%F)
+	STAR version,$version
+	genomeFastaFiles,$genomeFastaFiles
+	sjdbGTFfile,$sjdbGTFfile
+	sjdbOverhang,$sjdbOverhang
+_EOF_
+
[…]

### 2015-07-30 ms.Rmd
MS updates. Tomato download server only allows one session.

@@
 output:
-  word_document:
-    fig_caption: yes
-  pdf_document:
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
   html_document:
@@
     theme: cerulean
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: yes
+  word_document:
+    fig_caption: yes
 bibliography: bib/test.bib
@@
 ## Transcriptome sequencing of meristem samples
-+ Short (boring) section about sequencing stats
++ Max. 1 paragraph about boring sequencing stats
 + **Table** (or supplementary): # reads, mapping %, rRNA, introns, genes blah blah
@@
 ## Clustering of expression patterns
+
+To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014cf], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
+To determine the optimum number of clusters (*c*), the minimum distance between cluster centroids of clusters produced with *c* values between `cmin` to `cmax`,
+
 + **Figure**: Gene expression clusters

### 2015-07-30 ms.Rmd
run DESeq2 script

@@
 output:
-  html_document:
-    fig_caption: yes
-    force_captions: yes
-    highlight: pygments
-    number_sections: yes
-    theme: cerulean
   pdf_document:
     fig_caption: yes
+    keep_md: yes
     keep_tex: yes
@@
     fig_caption: yes
+  BiocStyle::html_document:
+    fig_caption: yes
+    force_captions: yes
+    highlight: pygments
+    keep_md: yes
+    number_sections: yes
 bibliography: bib/test.bib
@@
 
-To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014cf], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
-To determine the optimum number of clusters (*c*), the minimum distance between cluster centroids of clusters produced with *c* values between `cmin` to `cmax`,
+To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
+To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (@SI).
+A cluster number of `{r} c` was chosen,
 

### 2015-07-30 src/R/runDeseq.R

@@ -1,0 +1,90 @@
+#!/usr/bin/Rscript
+
+# set variables
+scriptName <- 'runDeseq'
+outputBasename <- paste(
+  Sys.Date(),
+  scriptName,
+  sep = "-"
+)
+
+# 1. SETUP
+
+# find the most recent cutadapt output
+outputDirs <- list.dirs(path = 'output', full.names = TRUE, recursive = FALSE)
+cutadaptDir <- rev(sort(outputDirs[grep('cutadapt', outputDirs)]))[1]
+
+# find the most recent STAR output
+outputDirs <- dir(path = cutadaptDir, pattern = "STAR", full.names = TRUE)
+starDir <- rev(sort(outputDirs[grep('STAR', outputDirs)]))[1]
+
+# load the quant files
+starFiles <- list.files(starDir, pattern = "ReadsPerGene", full.names = TRUE)
+starCounts <- do.call(cbind,
+                      lapply(starFiles, read.table, header = FALSE, sep = "\t", row.names = 1,
+                             colClasses = c("character", "integer", rep("NULL", 2))))
+colnames(starCounts) <- gsub("\\..*", "", basename(starFiles))
+
+# remove n1r2 and n2r2 (degraded RNA)
+starCounts$n1r2 <- NULL
+starCounts$n2r2 <- NULL
+
+# generate colData
+sample <- colnames(starCounts)
+sampleToStage <- c(
+  n1 = "RM",
+  n2 = "PBM",
+  n3 = "ePBM/SBM",
+  n4 = "RM"
+)
+stage <- sampleToStage[substr(sample, 1, 2)]
+batch <- as.numeric(substr(sample, 4, 4))
+batch[batch <= 2] <- 'first'
+batch[!batch == 'first'] <- 'second'
+colData <- data.frame(row.names = sample, stage, batch)
+
+# generate DESeq2 object (exclude the STAR diagnostic lines starting with "N")
+dds <- DESeq2::DESeqDataSetFromMatrix(countData = subset(starCounts, !grepl("^N_", rownames(starCounts))),
+                                      colData = colData,
+                                      design = ~ stage + batch)
+
+
+
+# RUN DESEQ2
+
+dds <- DESeq2::estimateSizeFactors(dds)
[…]

### 2015-07-30 src/sh/downloadReads.sh

@@ -19,7 +19,7 @@ wget --no-parent --directory-prefix "$outdirTom" --recursive --no-directories \
 	--accept wt_sim*.tgz ftp://ftp.solgenomics.net/user_requests/LippmanZ/public_releases/by_species/Solanum_lycopersicum/transcripts/
 
 wget --no-parent --directory-prefix "$outdirTom" --recursive --no-directories \
-	--accept wt_tm*.tgz ftp://ftp.solgenomics.net/user_requests/LippmanZ/public_releases/by_species/Solanum_lycopersicum/transcripts/
+	--accept "wt_tm*.tgz" ftp://ftp.solgenomics.net/user_requests/LippmanZ/public_releases/by_species/Solanum_lycopersicum/transcripts/
 
 wget --no-parent --directory-prefix "$outdirTom" --recursive --no-directories \
 	--accept wt_fm*.tgz ftp://ftp.solgenomics.net/user_requests/LippmanZ/public_releases/by_species/Solanum_lycopersicum/transcripts/

### 2015-08-04 src/R/mfuzz.R
SHIT. Typo in DESeq script causing spikelet meristem samples to be treated as rachis meristem. Good one.

@@ -1,0 +1,294 @@
+#!/usr/bin/Rscript
+
+#SBATCH --job-name Rscript
+#SBATCH --ntasks=1
+#SBATCH --cpus-per-task=1
+#SBATCH --output log/shuffledTpm.%N.%j.out
+#SBATCH --open-mode=append
+#SBATCH --mail-type=ALL
+
+library(Mfuzz)
+
+# set variables
+scriptName <- 'mfuzz'
+outputBasename <- paste(
+  scriptName,
+  Sys.Date(),
+  sep = "-"
+)
+
+# find results
+outputDirs <- list.dirs(path = 'output', full.names = TRUE, recursive = FALSE)
+tpmDir <- rev(sort(outputDirs[grep('tpm-', outputDirs)]))[1]
+cutadaptDir <- rev(sort(outputDirs[grep('cutadapt', outputDirs)]))[1]
+outputDirs <- list.dirs(path = cutadaptDir, full.names = TRUE, recursive = FALSE)
+starDir <- rev(sort(outputDirs[grep('STAR', outputDirs)]))[1]
+outputDirs <- list.dirs(path = starDir, full.names = TRUE, recursive = FALSE)
+deseqDir <- rev(sort(outputDirs[grep('DESeq2', outputDirs)]))[1]
+outputDirs <- list.dirs(tpmDir, full.names = TRUE, recursive = FALSE)
+cutoffDir <- rev(sort(outputDirs[grep('calculateCutoffs', outputDirs)]))[1]
+
+# take the geometric mean of the vst expression values
+vst <- readRDS(paste0(deseqDir, "/vst.Rds"))
+gm_mean <- function(x, na.rm = TRUE) {
+  exp(sum(log(x[x > 0]), na.rm = na.rm)/length(x))
+}
+vstMeans.matrix <- sapply(levels(vst$stage), function(x)
+  apply(GenomicRanges::assay(vst[, vst$stage == x]), 1, gm_mean))
+
+
+# select genes that were significant in the LRT from DESeq2
+# sigLRT <- read.table('/home/tom/Desktop/laserdissect/finalAnalysis/agriGO/LRToutput/lrtSig.txt')$V1
+# varGenes <- vstMeans.matrix[sigLRT,]
+# rmean <- apply(varGenes, 1, mean)
+# rmax <- apply(varGenes, 1, max)
+# varGenes <- varGenes[rmean > 6 | rmax > 6,]
+
[…]

### 2015-08-04 src/R/runDeseq.R

@@ -35,7 +35,7 @@ sampleToStage <- c(
   n1 = "RM",
   n2 = "PBM",
   n3 = "ePBM/SBM",
-  n4 = "RM"
+  n4 = "SM"
 )
 stage <- sampleToStage[substr(sample, 1, 2)]
 batch <- as.numeric(substr(sample, 4, 4))

### 2015-08-05 ms.Rmd
SHIT. Typo in DESeq script causing spikelet meristem samples to be treated as rachis meristem. Good one.
mfuzz script

@@
 ---
-title: Genome-wide changes in expression during inflorescence architecture development
-  in rice
+title: "Genome-wide changes in expression during inflorescence architecture development
+  in rice"
 csl: csl/the-plant-journal.csl
@@
 ## Dissections in early stages of inflorescence development
+
 +  **Figure**: laser microdissections
@@
 ## Transcriptome sequencing of meristem samples
+
 + Max. 1 paragraph about boring sequencing stats

### 2015-08-05 si.Rmd
mfuzz script

---
title: "Supporting Information"
csl: csl/the-plant-journal.csl
output:
  word_document:
    fig_caption: yes
bibliography: bib/test.bib
---

### 2015-08-05 src/R/mfuzz.R

@@ -9,6 +9,8 @@
 
 library(Mfuzz)
 library(ggplot2)
+library(data.table)
+library(xlsx)
 
 # set variables
 scriptName <- 'mfuzz'
@@ -28,6 +30,12 @@ outputDirs <- list.dirs(path = starDir, full.names = TRUE, recursive = FALSE)
 deseqDir <- rev(sort(outputDirs[grep('DESeq2', outputDirs)]))[1]
 outputDirs <- list.dirs(tpmDir, full.names = TRUE, recursive = FALSE)
 cutoffDir <- rev(sort(outputDirs[grep('calculateCutoffs', outputDirs)]))[1]
+
+# make output folder
+outDir <- paste(deseqDir, outputBasename, sep = "/")
+if (!dir.exists(outDir)) {
+  dir.create(outDir)
+}
 
 # take the geometric mean of the vst expression values
 vst <- readRDS(paste0(deseqDir, "/vst.Rds"))
@@ -74,154 +82,29 @@ centPlot <- ggplot(centroids, aes(x = x, y = y)) +
   geom_point() +
   stat_smooth(method = loess, se = FALSE)
 
-# we will go with 6 clusters
-c <- 6
-
 # can try to find inflection points, doesn't work very well.
 points <- seq(2, maxClust, length.out = 10000)
 pred <- predict(loess(centroids$y ~ centroids$x), points)
 infl <- c(FALSE, diff(diff(pred) > 0) != 0)
-centPlot + geom_point(data = data.frame(x = points[infl], y = pred[infl]), colour = 'red')
+centPlotWithPoints <- centPlot + geom_point(data = data.frame(x = points[infl], y = pred[infl]), colour = 'red')
 
+# we will go with 6 clusters
+c <- 6
+memCutoff <- 0.5
+
+# run the clustering
 set.seed(1)
 c1 <- mfuzz(vg.s, c = c, m = m1)
-clusters <- acore(vg.s, c1, min.acore = 0.5)
+clusters <- acore(vg.s, c1, min.acore = memCutoff)
 print(ggplotClusters(clusters = c1, expressionMatrix = exprs(vg.s), memCutoff = 0.5, pointsize = 10, ncol = 3))
 
-clusteredGenes <- oryzr::LocToGeneName(unique(unlist(sapply(clusters, rownames))))$names
-clusteredGenes[!is.na(clusteredGenes)]
-clusteredGenes[!is.na(clusteredGenes) & substr(clusteredGenes, 1, 4) == "MADS"]
-
[…]

### 2015-08-06 ms.Rmd
final commit before moving to simpler struct and pipeline with ruffus

@@
 ---
-title: "Genome-wide changes in expression during inflorescence architecture development
-  in rice"
+title: Genome-wide changes in expression during inflorescence architecture development
+  in rice
 csl: csl/the-plant-journal.csl
+header-includes: \usepackage{graphicx}
 output:
+  word_document:
+    fig_caption: yes
   pdf_document:
     fig_caption: yes
-    keep_md: yes
     keep_tex: yes
     number_sections: yes
-  word_document:
-    fig_caption: yes
   BiocStyle::html_document:
@@
 ---
+
+```{r Setup, include=FALSE, results="hide", warning=FALSE}
+
+library(data.table)
+library(ggplot2)
+extrafont::loadfonts()
+
+knitr::opts_chunk$set(fig.path='figures/',
+               cache.path='cache/',
+               dev=c("cairo_pdf", "svg", "png"),
+               fig.width=6.614,
+               fig.height=6.614,
+               fig.align = 'center',
+               dpi=300,
+               fig.show='hold',
+               cache=TRUE,
+               echo=FALSE,
+               results="hide",
+               message=FALSE,
+               warning=FALSE)
+
+# https://gist.githubusercontent.com/rmflight/3858973/raw/591ee603ce0996ac407902f7dd59520b35b3702a/tableFigureFuncs.r
+pasteLabel <- function(preText, inObj, objName, insLink=TRUE){
+	objNum <- inObj[objName]
+
+	useText <- paste(preText, objNum, sep=" ")
+	if (insLink){
+		useText <- paste("[", useText, "](#", objName, ")", sep="")
+	}
+	useText
+}
+
+# this increments the counter, and gives a name to the number so we can reference it later
+incCount <- function(inObj, useName){
+	nObj <- length(inObj)
+	useNum <- max(inObj) + 1
+	inObj <- c(inObj, useNum)
+	names(inObj)[nObj+1] <- useName
+	inObj
+}
+figCount <- c("_"=0)
+tableCount <- c("_"=0)
+
+tableCat <- function(inFrame){
+	outText <- paste(names(inFrame), collapse=" | ")
+	outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
+	invisible(apply(inFrame, 1, function(inRow){
+		outText <<- c(outText, paste(inRow, collapse=" | "))
+	}))
+	return(outText)
+}
+```
 
@@
 
+```{r mfuzz}
+expressionMatrix <- readRDS('output/cutadapt-2015-07-29/STAR-2015-07-29/DESeq2-2015-08-04/mfuzz-2015-08-05/expressionMatrix.Rds')
+c1 <- readRDS('output/cutadapt-2015-07-29/STAR-2015-07-29/DESeq2-2015-08-04/mfuzz-2015-08-05/c1.Rds')
+memCutoff <- 0.5
+
+# get clusters and membership
+cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster, Membership = apply(c1$membership, 1, max), key = "id")
+
+# get standardised VST counts
+exprs <- data.table(Biobase::exprs(expressionMatrix), keep.rownames = TRUE, key = "rn")
+
+# set up cluster labels
+numGenes <- cluster[Membership > 0.5, .(number = length(id)), by = Cluster]
+setkey(numGenes, 'Cluster')
+clustLabels <- numGenes[, paste0("Cluster ", Cluster, "\n(", number, " genes)")]
+
+# merge clusters with expression values
+plotData.wide <- cluster[exprs]
+
+# make long
+plotData <- reshape2::melt(plotData.wide[ Membership > memCutoff ], id.vars = c('id', 'Cluster', 'Membership'), variable.name = 'Stage', value.name = "Normalised, transformed read counts")
+
+# add cluster labels
+plotData[, Cluster := plyr::mapvalues(as.factor(Cluster), seq(1, length(unique(Cluster))), clustLabels)]
+
+# fix stage label
+plotData[, Stage := plyr::mapvalues(Stage, "ePBM.SBM", "ePBM/SBM")]
+
+# set up heatscale
+heatscale <- RColorBrewer::brewer.pal(n = 6, name = "YlOrRd")
+
+# add centres
+centres.wide <- data.table(c1$centers, Cluster = clustLabels)
+centres <- reshape2::melt(centres.wide, id.vars = 'Cluster', variable.name = 'Stage', value.name = "Normalised, transformed read counts")
+centres[, Stage := plyr::mapvalues(Stage, "ePBM.SBM", "ePBM/SBM")]
+
+f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed read counts`,
+                     colour = Membership, group = id)) +
+  theme_minimal(base_size = 8, base_family = "Helvetica") +
+  xlab(NULL) +
+  scale_colour_gradientn(colours = heatscale, limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
+  geom_line() +
+  geom_line(data = centres, mapping = aes(group = 1), colour = "black", alpha = 0.5) +
+  facet_wrap(~ Cluster)
+
+figCount <- incCount(figCount, "f_mfuzzClusters")
+```
+
 To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (@SI).
-A cluster number of `{r} c` was chosen,
+A cluster number of `{r} c` was chosen (**`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`**).
+
+<a id="f_mfuzzClusters"></a>
+```{r f_mfuzzClusters, fig.cap="Fuzzy \\emph{c}-means clustering of normalised, transformed read counts"}
+print(f_mfuzzClusters)
+```
 
@@
 ## Patterns of expression of transcription families
+
 + **Figure**: GSEA of transcription factor families
@@
 ## MADS genes: comparison of expression & homology in rice, tomato, arabidopsis
+
 + **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.

### 2015-08-06 ms.Rmd
minor change to MS

@@
                cache.path='cache/',
-               dev=c("cairo_pdf", "svg", "png"),
+               dev=c("cairo_pdf"),
                fig.width=6.614,
@@
 }
+
+capwords <- function(s, strict = FALSE) {
+    cap <- function(s) paste(toupper(substring(s, 1, 1)),
+                  {s <- substring(s, 2); if(strict) tolower(s) else s},
+                             sep = "", collapse = " " )
+    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
+
 ```
@@
 
+# number of clusters (for text)
+c <- length(unique(c1$cluster))
+
 f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed read counts`,
@@
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (@SI).
-A cluster number of `{r} c` was chosen (**`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`**).
+A cluster number of `r as.character(english::english(c))` *justify chosing this?
+double diffential was closest to zero?* (**`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`**).
 
 <a id="f_mfuzzClusters"></a>
-```{r f_mfuzzClusters, fig.cap="Fuzzy \\emph{c}-means clustering of normalised, transformed read counts"}
+```{r f_mfuzzClusters, fig.height=4.40933}
 print(f_mfuzzClusters)
@@
 
+`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`
+**Fuzzy *c*-means clustering of normalised, transformed read counts**
+
+`r capwords(as.character(english::english(c)))` common patterns of expression were recovered from the normalised, variance-stabilised read counts. Genes expressed in blah blah
+
 ***

### 2015-08-13 ms.Rmd
can't execute scripts if they are not executable

@@
 To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
-To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (@SI).
+To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (**SI**).
 A cluster number of `r as.character(english::english(c))` *justify chosing this?
 double diffential was closest to zero?* (**`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`**).
+
+To understand the contents of the clusters, **z-score for each tf family by cluster**.
 
@@
 + GO enrichment?
+NO!
+z-score enrichment of transcription factor genes per cluster.
+Use dhyper in R (see http://jura.wi.mit.edu/bio/education/hot_topics/enrichment/Gene_list_enrichment_Mar10.pdf or http://stats.stackexchange.com/a/16259)
 + Clusters including 'known' genes
@@
 + **Figure**: GSEA of transcription factor families
+
+Because of the enrichment of TF family members in some of the gene expression clusters, geneset-level enrichment analysis (GSEA) was used to compare the levels of expression of TF family members in the meristem samples.
+**Something like: Consistent with their enrichment in cluster 1 and 6, MADS genes were significantly enriched in the SM/FM samples**.
+TF families involved in branching were enriched in **which families, which samples**.
 
@@
 
-`r capwords(as.character(english::english(c)))` common patterns of expression were recovered from the normalised, variance-stabilised read counts. Genes expressed in blah blah
+`r capwords(as.character(english::english(c)))` common patterns of expression were recovered from normalised, variance-stabilised read counts of expressed genes.
+Each line describes the expression pattern of one gene, with the gene's membership to the cluster mapped to the colour of the line.
+The core values for each cluster are drawn in black.
 

### 2015-08-13 src/R/calculateTpm.R

@@ -32,7 +32,7 @@ gtfLength <- data.frame(Length = output, row.names = names(output))
 starDir <- "output/STAR"
 if (!dir.exists(starDir)) {
   cat("starDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # parse STAR files for the average mapped fragment length
@@ -60,7 +60,7 @@ names(mu) <- gsub('.Log.final.out', '', basename(names(mu)), fixed = TRUE)
 deseqDir <- "output/DESeq2"
 if (!dir.exists(deseqDir)) {
   cat("deseqDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # load DESeq2 output
@@ -107,3 +107,5 @@ sInf <- c(paste("git branch:",system("git rev-parse --abbrev-ref HEAD", intern =
           capture.output(sessionInfo()))
 logLocation <- paste0(outDir, "/SessionInfo.txt")
 writeLines(sInf, logLocation)
+
+quit(save = "no", status = 0)

### 2015-08-13 src/R/mfuzz.R

@@ -9,27 +9,27 @@ library(xlsx)
 deseqDir <- "output/DESeq2"
 if (!dir.exists(deseqDir)) {
   cat("deseqDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # check for expressed genes output
 cutoffDir <- "output/expressedGenes"
 if (!dir.exists(cutoffDir)) {
   cat("cutoffDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # check for msu annotation file
 msuAnn.file <- "data/genome/os/all.locus_brief_info.7.0.tab"
 if (!file.exists(msuAnn.file)) {
   cat("MSU annotation not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # make output folder
 outDir <- "output/mfuzz"
 if (!dir.exists(outDir)) {
-  dir.create(outDir)
+  quit(save = "no", status = 1)
 }
 
 # take the geometric mean of the vst expression values
@@ -149,3 +149,5 @@ sInf <- c(paste("git branch:",system("git rev-parse --abbrev-ref HEAD", intern =
           capture.output(sessionInfo()))
 logLocation <- paste0(outDir, "/SessionInfo.txt")
 writeLines(sInf, logLocation)
+
+quit(save = "no", status = 0)

### 2015-08-13 src/R/shuffledTpm.R

@@ -6,35 +6,35 @@ library(rtracklayer)
 shuffleDir <- "output/shuffle"
 if (!dir.exists(shuffleDir)) {
   cat("shuffleDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # check for htseq counts
 htseqDir <- "output/dnaTpm/shuffledCounts"
 if (!dir.exists(htseqDir)) {
   cat("htseqDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # check for STAR output
 starDir <- "output/STAR"
 if (!dir.exists(starDir)) {
   cat("starDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # check for DESeq2 output
 deseqDir <- "output/DESeq2"
 if (!dir.exists(deseqDir)) {
   cat("deseqDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # check for tpm output
 tpmDir <- "output/tpm"
 if (!dir.exists(tpmDir)) {
   cat("tpmDir not found, exiting\n", file = stderr())
-  quit(status = 1)
+  quit(save = "no", status = 1)
 }
 
 # FEATURE LENGTHS FROM GTF
@@ -162,3 +162,5 @@ sInf <- c(paste("git branch:",system("git rev-parse --abbrev-ref HEAD", intern =
           capture.output(sessionInfo()))
 logLocation <- paste0(outDir, "/SessionInfo.txt")
 writeLines(sInf, logLocation)
+
+quit(save = "no", status = 0)

### 2015-08-13 ms.Rmd
rtracklayer needed update???

@@
 + **Table** (or supplementary): # reads, mapping %, rRNA, introns, genes blah blah
++ Description of the dataset
 
@@
 
+The presence of MADS genes was enriched in **which clusters** and their overall expression is enriched in the spikelet meristem sample.
+These results correspond with previous studies that have established the importance of MADS genes in inflorescence development.
+The evolution of their expression patterns was
+
 + **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.

### 2015-08-13 src/R/calculateTpm.R

@@ -23,7 +23,7 @@ elementMetadata(reducedGtf)$widths <- width(reducedGtf)
 calc_length <- function(x) {
   sum(elementMetadata(x)$widths)
 }
-output <- OpenRepGrid::sapply_pb(split(reducedGtf, elementMetadata(reducedGtf)$gene_name), calc_length)
+output <- sapply(split(reducedGtf, elementMetadata(reducedGtf)$gene_name), calc_length)
 gtfLength <- data.frame(Length = output, row.names = names(output))
 
 # PARSE STAR FILES

### 2015-08-14 ms.Rmd
fixed typo in shuffledTpm script; switched to dplyr for gtfLength calcs (much faster)

@@
 
-To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do], and the standardised, VST-transformed read counts were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
+To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
+After determining an expression cutoff for each library, genes that were not above the cutoff in any sample were excluded, leaving (**how many genes remaining, link to methods**).
+These genes were ranked by variance, and genes with variance higher than the 90th percentile were used for clustering (**how many genes**).
+The standardised, VST-transformed read counts for the expressed, variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (**SI**).
@@
 `r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`
-**Fuzzy *c*-means clustering of normalised, transformed read counts**
+**Common patterns of expression in the meristem samples.**
 
-`r capwords(as.character(english::english(c)))` common patterns of expression were recovered from normalised, variance-stabilised read counts of expressed genes.
+Fuzzy *c*-means clustering of normalised, variance-stabilised read counts was used to recover `r capwords(as.character(english::english(c)))` common patterns of expression.
 Each line describes the expression pattern of one gene, with the gene's membership to the cluster mapped to the colour of the line.
-The core values for each cluster are drawn in black.
+The core values for each cluster are plotted in black.
 

### 2015-08-14 src/R/calculateTpm.R

@@ -1,6 +1,7 @@
 #!/usr/bin/Rscript
 
 library(rtracklayer)
+library(dplyr)
 
 # FEATURE LENGTHS FROM GTF
 
@@ -19,12 +20,10 @@ reducedGtf <- unlist(grl, use.names = TRUE)
 elementMetadata(reducedGtf)$gene_name <- rep(names(grl), elementLengths(grl))
 elementMetadata(reducedGtf)$widths <- width(reducedGtf)
 
-# calculate feature lengths 
-calc_length <- function(x) {
-  sum(elementMetadata(x)$widths)
-}
-output <- sapply(split(reducedGtf, elementMetadata(reducedGtf)$gene_name), calc_length)
-gtfLength <- data.frame(Length = output, row.names = names(output))
+# calculate feature lengths with dplyr
+output <- group_by(as.data.frame(reducedGtf), gene_name) %>%
+  summarize(length = sum(widths))
+gtfLength <- data.frame(Length = output$length, row.names = output$gene_name)
 
 # PARSE STAR FILES
 

### 2015-08-14 src/R/shuffledTpm.R

@@ -1,6 +1,7 @@
 #!/usr/bin/Rscript
 
 library(rtracklayer)
+library(dplyr)
 
 # check for shuffled GTF
 shuffleDir <- "output/shuffle"
@@ -54,12 +55,10 @@ reducedGtf <- unlist(grl, use.names = TRUE)
 elementMetadata(reducedGtf)$gene_name <- rep(names(grl), elementLengths(grl))
 elementMetadata(reducedGtf)$widths <- width(reducedGtf)
 
-# calculate feature lengths 
-calc_length <- function(x) {
-  sum(elementMetadata(x)$widths)
-}
-output <- OpenRepGrid::sapply_pb(split(reducedGtf, elementMetadata(reducedGtf)$gene_name), calc_length)
-gtfLength <- data.frame(Length = output, row.names = names(output))
+# calculate feature lengths with dplyr
+output <- group_by(as.data.frame(reducedGtf), gene_name) %>%
+  summarize(length = sum(widths))
+gtfLength <- data.frame(Length = output$length, row.names = output$gene_name)
 
 # PARSE STAR FILES
 
@@ -84,10 +83,6 @@ mu <- sapply(files, parseStarFile)
 names(mu) <- gsub('.Log.final.out', '', basename(names(mu)), fixed = TRUE)
 
 # CALCULATE TPM
-
-# read shuffled htseq-count results into DESeq to get normalised counts
-outputDirs <- list.dirs(shuffleDir, full.names = TRUE, recursive = FALSE)
-htseqDir <- rev(sort(outputDirs[grep('htseq-count', outputDirs)]))[1]
 
 countFiles <- list.files(htseqDir, pattern = 'htseq-count$', full.names = TRUE)
 countMatrix <- do.call(cbind, lapply(countFiles, read.table, header = FALSE, sep = "\t", row.names = 1))

### 2015-08-14 ms.Rmd
re-run with maxClust = 25; output maxClust

@@
 ```{r Setup, include=FALSE, results="hide", warning=FALSE}
-
-library(data.table)
-library(ggplot2)
-extrafont::loadfonts()
 
@@
 
+source('src/R/figures.R')
+
 ```
@@
 ```{r mfuzz}
-expressionMatrix <- readRDS('output/cutadapt-2015-07-29/STAR-2015-07-29/DESeq2-2015-08-04/mfuzz-2015-08-05/expressionMatrix.Rds')
-c1 <- readRDS('output/cutadapt-2015-07-29/STAR-2015-07-29/DESeq2-2015-08-04/mfuzz-2015-08-05/c1.Rds')
-memCutoff <- 0.5
-
-# get clusters and membership
-cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster, Membership = apply(c1$membership, 1, max), key = "id")
-
-# get standardised VST counts
-exprs <- data.table(Biobase::exprs(expressionMatrix), keep.rownames = TRUE, key = "rn")
-
-# set up cluster labels
-numGenes <- cluster[Membership > 0.5, .(number = length(id)), by = Cluster]
-setkey(numGenes, 'Cluster')
-clustLabels <- numGenes[, paste0("Cluster ", Cluster, "\n(", number, " genes)")]
-
-# merge clusters with expression values
-plotData.wide <- cluster[exprs]
-
-# make long
-plotData <- reshape2::melt(plotData.wide[ Membership > memCutoff ], id.vars = c('id', 'Cluster', 'Membership'), variable.name = 'Stage', value.name = "Normalised, transformed read counts")
-
-# add cluster labels
-plotData[, Cluster := plyr::mapvalues(as.factor(Cluster), seq(1, length(unique(Cluster))), clustLabels)]
-
-# fix stage label
-plotData[, Stage := plyr::mapvalues(Stage, "ePBM.SBM", "ePBM/SBM")]
-
-# set up heatscale
-heatscale <- RColorBrewer::brewer.pal(n = 6, name = "YlOrRd")
-
-# add centres
-centres.wide <- data.table(c1$centers, Cluster = clustLabels)
-centres <- reshape2::melt(centres.wide, id.vars = 'Cluster', variable.name = 'Stage', value.name = "Normalised, transformed read counts")
-centres[, Stage := plyr::mapvalues(Stage, "ePBM.SBM", "ePBM/SBM")]
-
-# number of clusters (for text)
-c <- length(unique(c1$cluster))
-
-f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed read counts`,
-                     colour = Membership, group = id)) +
-  theme_minimal(base_size = 8, base_family = "Helvetica") +
-  xlab(NULL) +
-  scale_colour_gradientn(colours = heatscale, limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
-  geom_line() +
-  geom_line(data = centres, mapping = aes(group = 1), colour = "black", alpha = 0.5) +
-  facet_wrap(~ Cluster)
-
 figCount <- incCount(figCount, "f_mfuzzClusters")

### 2015-08-14 src/R/mfuzz.R

@@ -63,10 +63,10 @@ vg.s <- standardise(vg.e)
 m1 <- mestimate(vg.s)
 
 # estimate the cluster number
-maxClust <- 15
+maxClust <- 25
 centroids <- data.frame(
   x = 2:(maxClust),
-  y = Dmin(vg.s, m = m1, crange = seq(2, maxClust, 1), repeats = 1, visu = FALSE)
+  y = Dmin(vg.s, m = m1, crange = seq(2, maxClust, 1), repeats = 3, visu = FALSE)
 )
 
 # visualise dmin vs cluster number
@@ -143,6 +143,7 @@ saveRDS(c1, paste0(outDir, "/c1.Rds"))
 saveRDS(clusters, paste0(outDir, "/clusters.Rds"))
 saveRDS(annotatedClusters, paste0(outDir, "/annotatedClusters.Rds"))
 saveRDS(centPlot, paste0(outDir, "/centPlot.Rds"))
+saveRDS(maxClust, paste0(outDir, "/maxClust.Rds"))
 
 # SAVE LOGS
 sInf <- c(paste("git branch:",system("git rev-parse --abbrev-ref HEAD", intern = TRUE)),

### 2015-08-14 ms.Rmd
split code for generating figures into separate script; just source and print in MS

@@
 The standardised, VST-transformed read counts for the expressed, variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
-To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between `cmin` and `cmax` (**SI**).
+To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust`.
 A cluster number of `r as.character(english::english(c))` *justify chosing this?
@@
 <a id="f_mfuzzClusters"></a>
-```{r f_mfuzzClusters, fig.height=4.40933}
+```{r f_mfuzzClusters, fig.height=4.40933, cache = FALSE}
 print(f_mfuzzClusters)

### 2015-08-14 si.Rmd
split code for generating figures into separate script; just source and print in MS

@@
 csl: csl/the-plant-journal.csl
+header-includes: \usepackage{graphicx}
 output:
@@
 ---
+
+```{r Setup, include=FALSE, results="hide", warning=FALSE}
+
+knitr::opts_chunk$set(fig.path='figures/',
+               cache.path='cache/',
+               dev=c("cairo_pdf"),
+               fig.width=6.614,
+               fig.height=6.614,
+               fig.align = 'center',
+               dpi=300,
+               fig.show='hold',
+               cache=TRUE,
+               echo=FALSE,
+               results="hide",
+               message=FALSE,
+               warning=FALSE)
+
+# https://gist.githubusercontent.com/rmflight/3858973/raw/591ee603ce0996ac407902f7dd59520b35b3702a/tableFigureFuncs.r
+pasteLabel <- function(preText, inObj, objName, insLink=TRUE){
+	objNum <- inObj[objName]
+
+	useText <- paste(preText, objNum, sep=" ")
+	if (insLink){
+		useText <- paste("[", useText, "](#", objName, ")", sep="")
+	}
+	useText
+}
+
+# this increments the counter, and gives a name to the number so we can reference it later
+incCount <- function(inObj, useName){
+	nObj <- length(inObj)
+	useNum <- max(inObj) + 1
+	inObj <- c(inObj, useNum)
+	names(inObj)[nObj+1] <- useName
+	inObj
+}
+figCount <- c("_"=0)
+tableCount <- c("_"=0)
+
+tableCat <- function(inFrame){
+	outText <- paste(names(inFrame), collapse=" | ")
+	outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
+	invisible(apply(inFrame, 1, function(inRow){
+		outText <<- c(outText, paste(inRow, collapse=" | "))
+	}))
+	return(outText)
+}
+
+capwords <- function(s, strict = FALSE) {
+    cap <- function(s) paste(toupper(substring(s, 1, 1)),
+                  {s <- substring(s, 2); if(strict) tolower(s) else s},
+                             sep = "", collapse = " " )
+    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
+
+source('src/R/figures.R')
+
+```
+
+### Fuzzy *c*-means clustering
+
+```{r mfuzz}
+figCount <- incCount(figCount, "f_mfuzzCentroids")
+print(f_mfuzzCentroids)
+
+figCount <- incCount(figCount, "f_mfuzzPca")
+f_mfuzzPca
+```

### 2015-08-14 src/R/mfuzz.R

@@ -144,6 +144,7 @@ saveRDS(clusters, paste0(outDir, "/clusters.Rds"))
 saveRDS(annotatedClusters, paste0(outDir, "/annotatedClusters.Rds"))
 saveRDS(centPlot, paste0(outDir, "/centPlot.Rds"))
 saveRDS(maxClust, paste0(outDir, "/maxClust.Rds"))
+saveRDS(vstFiltered, paste0(outDir, "/vstFiltered.Rds"))
 
 # SAVE LOGS
 sInf <- c(paste("git branch:",system("git rev-parse --abbrev-ref HEAD", intern = TRUE)),

### 2015-08-14 src/R/tfHypergeom.R

@@ -1,0 +1,10 @@
+#!/usr/bin/Rscript
+
+# load tfdb
+
+tfdbFile <- 'data/tfdb/os.Rds'
+
+if (!file.exists(tfdbFile)) {
+cat("tfdb file not found, exiting\n", file = stderr())
+quit(save = "no", status = 1)
+}

### 2015-08-14 src/py/pipeline.py

@@ -188,6 +188,21 @@ def download_at_reads(outputFiles):
     touch(outputFiles)
 
 #---------------------------------------------------------------
+# download TFDB
+#
+@originate(['ruffus/tfdb.data'])
+
+def download_tfdb(outputFiles):
+    jobScript = 'src/sh/downloadTfdb.sh'
+    ntasks = '1'
+    cpus_per_task = '1'
+    job_name = 'tfdb'
+    jobId = submit_job(jobScript, ntasks, cpus_per_task, job_name)
+    # update ruffus flag
+    print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
+    touch(outputFiles)
+
+#---------------------------------------------------------------
 # define rice reads
 #
 @originate(['ruffus/os.reads'])

### 2015-08-14 src/sh/downloadTfdb.sh

@@ -1,0 +1,60 @@
+#!/bin/bash
+
+set -eu
+
+# bash traceback code from https://docwhat.org/tracebacks-in-bash/
+
+_showed_traceback=f
+
+_exit_trap () {
+  local _ec="$?"
+  if [[ $_ec != 0 && "${_showed_traceback}" != t ]]; then
+    traceback 1
+  fi
+}
+
+_err_trap() {
+  local _ec="$?"
+  local _cmd="${BASH_COMMAND:-unknown}"
+  traceback 1
+  _showed_traceback=t
+  echo "The command ${_cmd} exited with exit code ${_ec}." 1>&2
+}
+
+traceback() {
+  # Hide the traceback() call.
+  local -i start=$(( ${1:-0} + 1 ))
+  local -i end=${#BASH_SOURCE[@]}
+  local -i i=0
+  local -i j=0
+
+  echo "Traceback (last called is first):" 1>&2
+  for ((i=${start}; i < ${end}; i++)); do
+    j=$(( $i - 1 ))
+    local function="${FUNCNAME[$i]}"
+    local file="${BASH_SOURCE[$i]}"
+    local line="${BASH_LINENO[$j]}"
+    echo "     ${function}() in ${file}:${line}" 1>&2
+  done
+}
+
+# traps
+trap _err_trap SIGHUP SIGINT SIGTERM
+trap _exit_trap EXIT
+trap _err_trap ERR
+
+# output directory
+outdir="data/tfdb"
+if [[ ! -d "$outdir" ]]; then
+	mkdir "$outdir"
+fi
+
+# download the tfdb
+curl --data-urlencode "_submit_check=1" --data-urlencode "DOWNLOAD=list of transcription factors" \
+	--data-urlencode "SUBMIT=Download" http://plntfdb.bio.uni-potsdam.de/v3.0/export.php > "$outdir"/tfdb.tab
+
+# call R processing script
+
+src/R/processTfdb.R
+
+exit 0

### 2015-08-14 ms.Rmd
fix formatting on hypergeom table

@@
     fig_caption: yes
-  pdf_document:
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
   BiocStyle::html_document:
@@
     keep_md: yes
+    number_sections: yes
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
     number_sections: yes
@@
 figCount <- incCount(figCount, "f_mfuzzClusters")
+tableCount <- incCount(tableCount, "t_hypergeom")
+t_hypergeom <- readRDS('output/mfuzz/sigFamilies.Rds')
+# fix p-values
+t_hypergeom[,grepl("^\\*p", colnames(t_hypergeom))] <- gsub("e-(\\d+)", "x10^-\\1^",
+format.pval(t_hypergeom[,grepl("^\\*p", colnames(t_hypergeom))], digits = 2,
+            na.form = "-"))
+t_hypergeom[is.na(t_hypergeom)] <- "-"
 ```
@@
 
-To understand the contents of the clusters, **z-score for each tf family by cluster**.
+After examining the lists of genes in each cluster it was evident that most clusters were enriched for transcription factors (**tables in SI**).
+To determine whether particular TF families were associated with particular patterns of expression, the enrichment TF family members in each cluster was calculated using hypergeometric tests.
+These calculations were carried out for families that had two or more members in any given cluster (**table**).
+
+```{r hypergeom, results = "markup"}
+knitr::kable(t_hypergeom, caption = 'Transcription factor family enrichment per cluster', align = 'r')
+```
+
+Hypergeometric tests were performed for each cluster that contained more than one member of a particular TF family to calculate the probability of observing the corresponding
 

### 2015-08-14 src/R/tfHypergeom.R

@@ -64,12 +64,12 @@ resultsLong$padj <- p.adjust(resultsLong$`p-value`, method = "BH",
 padj <- reshape2::dcast(resultsLong, Family ~ Cluster, value.var = 'padj')
 rownames(padj) <- padj$Family
 padj$Family <- NULL
-colnames(padj) <- paste0("padj(", colnames(padj), ")")
+colnames(padj) <- paste0("*p*~adj,\\ ", colnames(padj), "~")
 
 Number <- reshape2::dcast(resultsLong, Family ~ Cluster, value.var = 'Number')
 rownames(Number) <- Number$Family
 Number$Family <- NULL
-colnames(Number) <- paste0("n(", colnames(Number), ")")
+colnames(Number) <- paste0("*n*~", colnames(Number), "~")
 
 # remove columns that don't have p-values
 keep <- apply(padj, 1, function(x) !all(is.na(x)))
@@ -78,16 +78,19 @@ sigFamilies <- cbind(Number[keep, ], padj[keep,])
 
 # add the number of genes per family
 genesPerFamily <- sapply(tfdbByFamily, function(x) dim(x)[1])
-sigFamilies$`Number of genes` <- genesPerFamily[rownames(sigFamilies)]
+sigFamilies$`*n*` <- genesPerFamily[rownames(sigFamilies)]
 
 # add the number of expresed genes per family
-sigFamilies$`Number of expressed genes` <- exprPerFamily[rownames(sigFamilies)]
+sigFamilies$`*n*~expr~` <- exprPerFamily[rownames(sigFamilies)]
 
 # set the column order
-sigFamilies <- sigFamilies[,c('Number of genes', 'Number of expressed genes',
-                              'n(C1)', 'padj(C1)', 'n(C2)', 'padj(C2)', 'n(C3)',
-                              'padj(C3)', 'n(C4)', 'padj(C4)', 'n(C5)',
-                              'padj(C5)', 'n(C6)', 'padj(C6)')]
-
+#paste(colnames(sigFamilies)[c(13,14,1,7,2,8,3,9,4,10,5,11,6,12)], collapse = "', '")
+sigFamilies <- sigFamilies[,c('*n*', '*n*~expr~', '*n*~C1~', '*p*~adj,\\ C1~',
+                              '*n*~C2~', '*p*~adj,\\ C2~', '*n*~C3~',
[…]

### 2015-08-14 ms.Rmd
test effect of increasing # of genes sent for clustering on hypergeom tests

--- 2015-08-14 ms.Rmd
@@
 output:
-  word_document:
-    fig_caption: yes
-  pdf_document:
-    fig_caption: yes
-    keep_tex: yes
-    number_sections: yes
   BiocStyle::html_document:
@@
     number_sections: yes
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: yes
+  word_document:
+    fig_caption: yes
 bibliography: bib/test.bib
@@
                fig.show='hold',
-               cache=TRUE,
+               cache=FALSE,
                echo=FALSE,
@@
 figCount <- incCount(figCount, "f_mfuzzClusters")
+tableCount <- incCount(tableCount, "t_hypergeom")
+t_hypergeom <- readRDS('output/mfuzz/sigFamilies.Rds')
+# fix p-values
+t_hypergeom[,grepl("^\\*p", colnames(t_hypergeom))] <- gsub("e-(\\d+)", "x10^-\\1^",
+format.pval(t_hypergeom[,grepl("^\\*p", colnames(t_hypergeom))], digits = 2,
+            na.form = ""))
+t_hypergeom[is.na(t_hypergeom)] <- ""
 ```
@@
 
-To understand the contents of the clusters, **z-score for each tf family by cluster**.
-
 <a id="f_mfuzzClusters"></a>
-```{r f_mfuzzClusters, fig.height=4.40933, cache = FALSE}
+```{r f_mfuzzClusters, fig.height=4.40933}
 print(f_mfuzzClusters)
 ```
+
+After examining the lists of genes in each cluster it was evident that most clusters had an enrichment of transcription factors (**tables in SI**).
+To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
+These calculations were carried out for families that had two or more members in any given cluster (**`r I(pasteLabel("Table", tableCount, "t_hypergeom"))`**).
+After adjustment for multiple comparisons,   MADS genes were significantly over-represented in clusters 1 and 4, both of which contain genes that have higher expression in later meristem samples.
+ABI3VP transcription factors are
+
+<a id="t_hypergeom"></a>
+```{r hypergeom, results = "markup"}
+knitr::kable(t_hypergeom, caption = 'Transcription factor family enrichment per cluster', align = 'r')
+```
+
+Hypergeometric tests were performed for each cluster that contained more than one member of a particular TF family to calculate the probability of observing the corresponding
 

### 2015-08-14 si.Rmd
test effect of increasing # of genes sent for clustering on hypergeom tests

@@
 output:
+  BiocStyle::html_document:
+    fig_caption: yes
+    force_captions: yes
+    highlight: pygments
+    keep_md: yes
+    number_sections: yes
   word_document:

### 2015-08-14 src/R/mfuzz.R

@@ -47,9 +47,11 @@ expressedGenes <- readRDS(paste0(cutoffDir, '/expressedGenesAll.Rds'))
 vstMeans <- data.frame(vstMeans.matrix)
 vstFiltered <- vstMeans[expressedGenes,]
 
+varGenes <- vstFiltered
+
 # get the most variable genes
-vstByVar <- vstFiltered[(rev(order(apply(vstFiltered, 1, var)))),]
-varGenes <- vstByVar[1:(0.1 * dim(vstByVar)[1]),]
+#vstByVar <- vstFiltered[(rev(order(apply(vstFiltered, 1, var)))),]
+#varGenes <- vstByVar[1:(0.1 * dim(vstByVar)[1]),]
 
 # set up the expressionSet
 pData <- data.frame(Stage = as.factor(colnames(varGenes)), row.names = colnames(varGenes))
@@ -86,7 +88,7 @@ centPlotWithPoints <- centPlot +
 
 # we will go with 6 clusters
 c <- 6
-memCutoff <- 0.5
+memCutoff <- 0.7
 
 # run the clustering
 set.seed(1)

### 2015-08-14 src/R/mfuzz.R
uh oh, broke mfuzz results

@@ -88,7 +88,7 @@ centPlotWithPoints <- centPlot +
 
 # we will go with 6 clusters
 c <- 6
-memCutoff <- 0.7
+memCutoff <- 0.5
 
 # run the clustering
 set.seed(1)

### 2015-08-14 ms.Rmd
uh oh, broke mfuzz results
could be ok like this...

@@
     number_sections: yes
+  word_document:
+    fig_caption: yes
   pdf_document:
@@
     number_sections: yes
-  word_document:
-    fig_caption: yes
 bibliography: bib/test.bib
@@
 <a id="f_mfuzzClusters"></a>
-```{r f_mfuzzClusters, fig.height=4.40933}
+```{r f_mfuzzClusters, fig.height=4.40933, cache = TRUE}
 print(f_mfuzzClusters)

### 2015-08-14 src/R/figures.R

@@ -57,21 +57,21 @@ f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed 
   geom_line(data = centres, mapping = aes(group = 1), colour = "black", alpha = 0.5) +
   facet_wrap(~ Cluster)
 
-# centroid dis vs. c (for SI)
-f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds')
-
-# PCA (for SI)
-clustered <- unique(names(c1$cluster))
-
-vg.d <- dist(Biobase::exprs(expressionMatrix[clustered,]))
-vg.mds <- cmdscale(vg.d, 2)
-vg.mds <- data.frame(MDS1 = vg.mds[,1], MDS2 = vg.mds[,2], cluster = c1$cluster,
-                     max.membership = apply(c1$membership,1,max))
-
-f_mfuzzPca <- ggplot(vg.mds, aes(x = MDS1, y=MDS2, colour = factor(cluster))) +
-  theme_minimal(base_size = 8, base_family = "Helvetica") +
-  coord_fixed(ratio = 1) +
-  geom_point(aes(size=max.membership),alpha=0.5, shape = 16) +
-  scale_colour_brewer(palette = "Set1", name = "Cluster") +
-  scale_size_area(guide = FALSE)
+# # centroid dis vs. c (for SI)
+# f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds')
+# 
+# # PCA (for SI)
+# clustered <- unique(names(c1$cluster))
+# 
+# vg.d <- dist(Biobase::exprs(expressionMatrix[clustered,]))
+# vg.mds <- cmdscale(vg.d, 2)
+# vg.mds <- data.frame(MDS1 = vg.mds[,1], MDS2 = vg.mds[,2], cluster = c1$cluster,
+#                      max.membership = apply(c1$membership,1,max))
+# 
+# f_mfuzzPca <- ggplot(vg.mds, aes(x = MDS1, y=MDS2, colour = factor(cluster))) +
+#   theme_minimal(base_size = 8, base_family = "Helvetica") +
+#   coord_fixed(ratio = 1) +
+#   geom_point(aes(size=max.membership),alpha=0.5, shape = 16) +
+#   scale_colour_brewer(palette = "Set1", name = "Cluster") +
+#   scale_size_area(guide = FALSE)
 

### 2015-08-17 ms.Rmd
tried using lrt result for mfuzz but it doesn't work

--- 2015-08-14 ms.Rmd
@@
     number_sections: yes
+    dev: svg
+  word_document:
+    fig_caption: yes
   pdf_document:
@@
                cache.path='cache/',
-               dev=c("cairo_pdf"),
-               fig.width=6.614,
-               fig.height=6.614,
+               dev=c("svg", "cairo_pdf"),
+               fig.width=3.150,
+               fig.height=6.3,
                fig.align = 'center',
@@
 
+```{r lmdFigure}
+figCount <- incCount(figCount, "f_lmdFigure")
+f_lmdFigure <- "data/lmdFigure.pdf"
+```
+
 ## Dissections in early stages of inflorescence development
@@
 +  **Figure**: laser microdissections
+
+![`r I(pasteLabel("Figure", figCount, "f_lmdFigure"))`](data/lmdFigure.pdf)
 
@@
 <a id="f_mfuzzClusters"></a>
-```{r f_mfuzzClusters, fig.height=4.40933}
+```{r f_mfuzzClusters, fig.height=6.3, cache = TRUE}
 print(f_mfuzzClusters)

### 2015-08-17 src/R/figures.R

@@ -10,7 +10,7 @@ extrafont::loadfonts()
 
 expressionMatrix <- readRDS('output/mfuzz/expressionMatrix.Rds')
 c1 <- readRDS('output/mfuzz/c1.Rds')
-memCutoff <- 0.5
+memCutoff <- 0.7
 
 # get clusters and membership
 cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster, Membership = apply(c1$membership, 1, max), key = "id")
@@ -19,7 +19,7 @@ cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster, Membership =
 exprs <- data.table(Biobase::exprs(expressionMatrix), keep.rownames = TRUE, key = "rn")
 
 # set up cluster labels
-numGenes <- cluster[Membership > 0.5, .(number = length(id)), by = Cluster]
+numGenes <- cluster[Membership > memCutoff, .(number = length(id)), by = Cluster]
 setkey(numGenes, 'Cluster')
 clustLabels <- numGenes[, paste0("Cluster ", Cluster, "\n(", number, " genes)")]
 
@@ -51,11 +51,12 @@ maxClust <- readRDS('output/mfuzz/maxClust.Rds')
 f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed read counts`,
                                         colour = Membership, group = id)) +
   theme_minimal(base_size = 8, base_family = "Helvetica") +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
   xlab(NULL) +
   scale_colour_gradientn(colours = heatscale, limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
   geom_line(alpha = 0.8) +
   geom_line(data = centres, mapping = aes(group = 1), colour = "black", alpha = 0.5) +
-  facet_wrap(~ Cluster)
+  facet_wrap(~ Cluster, nrow = 3)
 
 # # centroid dis vs. c (for SI)
 # f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds')

### 2015-08-17 src/R/mfuzz.R

@@ -80,7 +80,7 @@ centPlot <- ggplot(centroids, aes(x = x, y = y)) +
 # can try to find inflection points, doesn't work very well.
 points <- seq(2, maxClust, length.out = 10000)
 pred <- predict(loess(centroids$y ~ centroids$x), points)
-infl <- c(FALSE, diff(diff(diff(pred)) > 0) != 0)
+infl <- c(FALSE, diff(diff(pred) > 0) != 0)
 centPlotWithPoints <- centPlot +
   geom_point(data = data.frame(x = points[infl], y = pred[infl]), colour = 'red')
 

### 2015-08-17 ms.Rmd
Flip, capitalization error in shuffle script was excluding rRNA regions. start again

@@
 
-# Results
+# Results and Discussion
 
@@
 
-![`r I(pasteLabel("Figure", figCount, "f_lmdFigure"))`](data/lmdFigure.pdf)
+To investigate gene expression during inflorescence development, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) (`r I(pasteLabel("Figure", figCount, "f_lmdFigure"))`).
+RNA from the meristem samples was isolated, amplified and used for the production of a sequencing library.
+**Details of the procedure, yield, etc.**.
+
+<a id="f_lmdFigure"></a>
+![](data/lmdFigure.pdf)
 
@@
 <a id="f_mfuzzClusters"></a>
-```{r f_mfuzzClusters, fig.height=6.3, cache = TRUE}
+```{r f_mfuzzClusters, fig.height=6.3}
 print(f_mfuzzClusters)
@@
 
-Hypergeometric tests were performed for each cluster that contained more than one member of a particular TF family to calculate the probability of observing the corresponding
-
-+ **Figure**: Gene expression clusters
-+ GO enrichment?
-NO!
-z-score enrichment of transcription factor genes per cluster.
-Use dhyper in R (see http://jura.wi.mit.edu/bio/education/hot_topics/enrichment/Gene_list_enrichment_Mar10.pdf or http://stats.stackexchange.com/a/16259)
 + Clusters including 'known' genes
@@
 + **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.
-
-# Discussion
 
@@
 
+`r I(pasteLabel("Figure", figCount, "f_lmdFigure"))`
+**LASER microdissection of meristem samples in the early stages of inflorescences development.** Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and secondary branch meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
+Images show the samples before (**a**–**d**) and after (**e**–**h**) dissection.
+
 `r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`
-**Common patterns of expression in the meristem samples.**
-
-Fuzzy *c*-means clustering of normalised, variance-stabilised read counts was used to recover `r capwords(as.character(english::english(c)))` common patterns of expression.
+**Common patterns of expression in the meristem samples.** Fuzzy *c*-means clustering of normalised, variance-stabilised read counts was used to recover `r capwords(as.character(english::english(c)))` common patterns of expression.
 Each line describes the expression pattern of one gene, with the gene's membership to the cluster mapped to the colour of the line.

### 2015-08-17 src/R/fixGffBed.R

@@ -4,16 +4,21 @@ args <- commandArgs(TRUE)
 outdir <- args[1]
 
 bed9to6 <- function(bed9file) {
-  bed9 <- read.table(bed9file, sep = "\t")
+  bed9 <- read.table(bed9file, sep = "\t", stringsAsFactors = FALSE)
   bed9[, 5] <- 0
   return(bed9[, 1:6])
 }
 
-bed9files <- list.files(outdir, pattern = "bed", full.names = TRUE)
+bed9files <- list.files(outdir, pattern = "bed$", full.names = TRUE)
 
 bed6 <- lapply(bed9files, bed9to6)
 names(bed6) <- paste0(gsub("\\.bed.*", "", sapply(bed9files, basename)), '.bed6')
 
+bed6[["irgsp1_rRNA_tRNA.bed6"]]$V1 <- gsub("0(\\d)", "\\1",
+                                         gsub("chr", "Chr",
+                                              bed6[["irgsp1_rRNA_tRNA.bed6"]]$V1,
+                                              fixed = TRUE))
+  
 for (i in 1:length(bed6)) {
   outpath <- paste0(outdir, "/", names(bed6)[i])
   write.table(bed6[i], file = outpath, quote = FALSE, sep = "\t", row.names = FALSE, col.names = FALSE)

### 2015-08-18 ms.Rmd
possible problem with setting IFS for grep query. Need to unset afterwards?

@@
 
-# https://gist.githubusercontent.com/rmflight/3858973/raw/591ee603ce0996ac407902f7dd59520b35b3702a/tableFigureFuncs.r
-pasteLabel <- function(preText, inObj, objName, insLink=TRUE){
-	objNum <- inObj[objName]
-
-	useText <- paste(preText, objNum, sep=" ")
-	if (insLink){
-		useText <- paste("[", useText, "](#", objName, ")", sep="")
-	}
-	useText
-}
-
-# this increments the counter, and gives a name to the number so we can reference it later
-incCount <- function(inObj, useName){
-	nObj <- length(inObj)
-	useNum <- max(inObj) + 1
-	inObj <- c(inObj, useNum)
-	names(inObj)[nObj+1] <- useName
-	inObj
-}
-figCount <- c("_"=0)
-tableCount <- c("_"=0)
-
-tableCat <- function(inFrame){
-	outText <- paste(names(inFrame), collapse=" | ")
-	outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
-	invisible(apply(inFrame, 1, function(inRow){
-		outText <<- c(outText, paste(inRow, collapse=" | "))
-	}))
-	return(outText)
-}
-
-capwords <- function(s, strict = FALSE) {
-    cap <- function(s) paste(toupper(substring(s, 1, 1)),
-                  {s <- substring(s, 2); if(strict) tolower(s) else s},
-                             sep = "", collapse = " " )
-    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
-}
-
 source('src/R/figures.R')
@@
 figCount <- incCount(figCount, "f_lmdFigure")
-f_lmdFigure <- "data/lmdFigure.pdf"
 ```

### 2015-08-18 si.Rmd
possible problem with setting IFS for grep query. Need to unset afterwards?

--- 2015-08-05 si.Rmd
@@
 csl: csl/the-plant-journal.csl
+header-includes: \usepackage{graphicx}
 output:
+  BiocStyle::html_document:
+    fig_caption: yes
+    force_captions: yes
+    highlight: pygments
+    keep_md: yes
+    number_sections: yes
   word_document:
@@
 ---
+
+```{r Setup, include=FALSE, results="hide", warning=FALSE}
+
+knitr::opts_chunk$set(fig.path='figures/',
+               cache.path='cache/',
+               dev=c("cairo_pdf"),
+               fig.width=6.614,
+               fig.height=6.614,
+               fig.align = 'center',
+               dpi=300,
+               fig.show='hold',
+               cache=FALSE,
+               echo=FALSE,
+               results="hide",
+               message=FALSE,
+               warning=FALSE)
+
+source('src/R/figures.R')
+
+```
+
+### Fuzzy *c*-means clustering
+
+```{r mfuzz}
+figCount <- incCount(figCount, "f_mfuzzCentroids")
+print(f_mfuzzCentroids)
+
+figCount <- incCount(figCount, "f_mfuzzPca")
+f_mfuzzPca
+```

### 2015-08-18 src/R/figures.R

@@ -4,6 +4,49 @@ library(data.table)
 library(ggplot2)
 extrafont::loadfonts()
 
+#####################
+### SETUP FOR RMD ###
+#####################
+
+# https://gist.githubusercontent.com/rmflight/3858973/raw/591ee603ce0996ac407902f7dd59520b35b3702a/tableFigureFuncs.r
+pasteLabel <- function(preText, inObj, objName, insLink=TRUE){
+  objNum <- inObj[objName]
+  
+  useText <- paste(preText, objNum, sep=" ")
+  if (insLink){
+    useText <- paste("[", useText, "](#", objName, ")", sep="")
+  }
+  useText
+}
+
+# this increments the counter, and gives a name to the number so we can reference it later
+incCount <- function(inObj, useName){
+  nObj <- length(inObj)
+  useNum <- max(inObj) + 1
+  inObj <- c(inObj, useNum)
+  names(inObj)[nObj+1] <- useName
+  inObj
+}
+figCount <- c("_"=0)
+tableCount <- c("_"=0)
+
+tableCat <- function(inFrame){
+  outText <- paste(names(inFrame), collapse=" | ")
+  outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
+  invisible(apply(inFrame, 1, function(inRow){
+    outText <<- c(outText, paste(inRow, collapse=" | "))
+  }))
+  return(outText)
+}
+
+capwords <- function(s, strict = FALSE) {
+  cap <- function(s) paste(toupper(substring(s, 1, 1)),
+                           {s <- substring(s, 2); if(strict) tolower(s) else s},
+                           sep = "", collapse = " " )
+  sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
+
+
 #############
 ### Mfuzz ###
 #############

### 2015-08-18 src/sh/countRTrna.sh

@@ -96,6 +96,7 @@ IFS="
 "
 headers=( $(grep "rRNA" "$outdir"/TIGR_Oryza_Repeats.v3.3_0_0.fsa) )
 for header in "${headers[@]}"; do b=$(echo ${header#>}); echo "${b%%\ *}"; done > "$outdir"/rrnaHeaders.tmp
+unset IFS
 
 # extract rRNA sequences from repeat database
 faidx "$outdir"/TIGR_Oryza_Repeats.v3.3_0_0.fsa $(cat "$outdir"/rrnaHeaders.tmp) > "$outdir"/rRna.fasta
@@ -141,9 +142,9 @@ for bam in $bamfiles; do
 	library=$(basename "$bam" ".Aligned.out.bam")
 	echo -e "[ $(date): Counting rRNA and tRNA for "$library" ]"
 	cmdR="samtools view -c -F0x100 -L "$outdir"/rRna.combined.bed "$bam""
-	srun --ntasks=1 --exclusive --output="$outdir"/"$library".rrna.tmp $cmdR &
+	srun --ntasks=1 --cpus-per-task=1 --exclusive --output="$outdir"/"$library".rrna.tmp $cmdR &
 	cmdT="samtools view -c -F0x100 -L "$outdir"/rap_tRNA.bed9 "$bam""
-	srun --ntasks=1 --exclusive --output="$outdir"/"$library".trna.tmp $cmdT &
+	srun --ntasks=1 --cpus-per-task=1 --exclusive --output="$outdir"/"$library".trna.tmp $cmdT &
 done
 echo -e "[ $(date): Waiting for jobs to finish ]"
 

### 2015-08-18 si.Rmd
try 8 clusters

@@
     number_sections: yes
+    dev: svg
   word_document:
     fig_caption: yes
+  pdf_document:
+    fig_caption: yes
+    keep_tex: yes
+    number_sections: yes
 bibliography: bib/test.bib
@@
                cache.path='cache/',
-               dev=c("cairo_pdf"),
+               dev=c("svg", "cairo_pdf"),
                fig.width=6.614,
@@
 
+``` {r libStats}
+tableCount <- incCount(tableCount, "t_libStats")
+```
+
+### `r I(pasteLabel("Table", tableCount, "t_libStats"))`
+
+``` {r t_libStats, results = "markup"}
+knitr::kable(t_libStats, caption = "Library and sequencing statistics")
+```
+
 ### Fuzzy *c*-means clustering
 
-```{r mfuzz}
+```{r mfuzz, cache = TRUE}
 figCount <- incCount(figCount, "f_mfuzzCentroids")

### 2015-08-18 src/R/figures.R

@@ -46,6 +46,11 @@ capwords <- function(s, strict = FALSE) {
   sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
 }
 
+###################
+### STATS TABLE ###
+###################
+
+t_libStats <- readRDS('output/quantStats/libStats.Rds')
 
 #############
 ### Mfuzz ###
@@ -102,8 +107,8 @@ f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed 
   geom_line(data = centres, mapping = aes(group = 1), colour = "black", alpha = 0.5) +
   facet_wrap(~ Cluster, nrow = 3)
 
-# # centroid dis vs. c (for SI)
-# f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds')
+# centroid dis vs. c (for SI)
+f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds')
 # 
 # # PCA (for SI)
 # clustered <- unique(names(c1$cluster))

### 2015-08-18 src/R/figures.R
started PCA figure, need to operate on all libraries

@@ -52,6 +52,36 @@ capwords <- function(s, strict = FALSE) {
 
 t_libStats <- readRDS('output/quantStats/libStats.Rds')
 
+################
+### PCA PLOT ###
+################
+
+# make this for all libraries, not just the final ones
+
+expressedGenes <- readRDS('output/expressedGenes/expressedGenesAll.Rds')
+vst <- readRDS('output/DESeq2/vst.Rds')
+
+exprVst <- GenomicRanges::assay(vst)[expressedGenes,]
+pca <- prcomp(t(exprVst))
+percentVar <- pca$sdev^2/sum(pca$sdev^2)
+
+pcaPlotData <- data.frame(
+  label = toupper(rownames(pca$x)),
+  PCA1 = pca$x[,1],
+  PCA2 = pca$x[,2],
+  Stage = GenomicRanges::colData(vst)$stage
+)
+
+f_pca <- ggplot(data = pcaPlotData, aes(x = PCA1, y = PCA2, colour = Stage)) +
+  theme_minimal(base_size = 8, base_family = "Helvetica") +
+  scale_colour_brewer(palette = "Set1", name = 'Stage') +
+  guides(colour=guide_legend(title=NULL)) +
+  scale_fill_brewer(palette = "Set1", guide = FALSE) +
+  xlab(paste0("PC1: ", round(percentVar[1] * 100), "% variance")) +
+  ylab(paste0("PC2: ", round(percentVar[2] * 100), "% variance")) +
+  geom_point(shape = 16, size = 5, alpha = 0.7) +
+  geom_text(aes(x = PCA1, y = PCA2, label = label), hjust = 1.2, show.legend = FALSE)
+
 #############
 ### Mfuzz ###
 #############
@@ -108,7 +138,8 @@ f_mfuzzClusters <- ggplot(plotData, aes(x = Stage, y = `Normalised, transformed 
   facet_wrap(~ Cluster, ncol = 2)
 
 # centroid dis vs. c (for SI)
-f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds')
+f_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds') +
+  theme_minimal(base_size = 8, base_family = "Helvetica")
 
 # MDS for SI
 vg.mds <- readRDS('output/mfuzz/vg.mds.Rds')

### 2015-08-19 ms.Rmd
started PCA figure, need to operate on all libraries
minor changes; need to rerun pipeline (overnight).

@@
     number_sections: yes
-    dev: svg
+    dev: "CairoPDF"
   word_document:
@@
 * Rice inflorescence development
-+ branching & grain yield
-+ sequential changes in meristem activity
-+ few key genes
++ Branching & grain yield
++ Sequential changes in meristem activity
++ Discuss a few key genes
 
@@
 
-```{r lmdFigure}
-figCount <- incCount(figCount, "f_lmdFigure")
-```
+To investigate gene expression during inflorescence development, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
+RNA from the meristem samples was isolated, amplified and used for the production of a sequencing library.
+**need area cut numbers**.
+RNA input for library preparation was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
+Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values two libraries that were made from degraded RNA were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
+Following this, further samples were prepared to provide a total of three biological replicates with intact RNA (RIN ≥ 7) for each meristem type.
 
-## Dissections in early stages of inflorescence development
-
-+  **Figure**: laser microdissections
-
-To investigate gene expression during inflorescence development, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) (`r I(pasteLabel("Figure", figCount, "f_lmdFigure"))`).
-RNA from the meristem samples was isolated, amplified and used for the production of a sequencing library.
-**Details of the procedure, yield, etc.**.
-
-<a id="f_lmdFigure"></a>
 ![](data/lmdFigure.pdf)
 
-## Transcriptome sequencing of meristem samples
+At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after removal of reads likely to have originated from rRNA and tRNA.
+Using cutoffs calculated from the 95^th^ percentile of distributions of expression values estimated from intergenic regions for each library (*i.e*.
+reads that were not generated from genes), `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
 
-+ Max. 1 paragraph about boring sequencing stats
-+ **Table** (or supplementary): # reads, mapping %, rRNA, introns, genes blah blah
-+ Description of the dataset
+##  Patterns of gene expression during inflorescence development
 
-## Clustering of expression patterns
+To recover common expression patterns in genes detected in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
+Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
+To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
+Clusters were recovered that contain genes that gradually decrease or increase in expression during development (clusters **2**, **3**, **6** and **8**), or that are apparently turned on or off in primary and secondary branch meristems (clusters **7** and **4** respectively) along with genes with more complex expression patterns (clusters **1** and **5**).
 
-```{r mfuzz}
-figCount <- incCount(figCount, "f_mfuzzClusters")
-tableCount <- incCount(tableCount, "t_hypergeom")
-t_hypergeom <- readRDS('output/mfuzz/sigFamilies.Rds')
-# fix p-values
-t_hypergeom[,grepl("^\\*p", colnames(t_hypergeom))] <- gsub("e-(\\d+)", "x10^-\\1^",
-format.pval(t_hypergeom[,grepl("^\\*p", colnames(t_hypergeom))], digits = 2,
-            na.form = ""))
-t_hypergeom[is.na(t_hypergeom)] <- ""
-```
-
-To recover common expression patterns in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
-After determining an expression cutoff for each library, genes that were not above the cutoff in any sample were excluded, leaving (**how many genes remaining, link to methods**).
-These genes were ranked by variance, and genes with variance higher than the 90th percentile were used for clustering (**how many genes**).
-The standardised, VST-transformed read counts for the expressed, variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
-To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust`.
-A cluster number of `r as.character(english::english(c))` *justify chosing this?
-double diffential was closest to zero?* (**`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`**).
-
-<a id="f_mfuzzClusters"></a>
 ```{r f_mfuzzClusters, fig.height=6.3}
@@
 
-After examining the lists of genes in each cluster it was evident that most clusters had an enrichment of transcription factors (**tables in SI**).
+Examination of the clustered genes suggested that transcription factor genes were overrepresented in most clusters (**annotated cluster tables in SI**).
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-These calculations were carried out for families that had two or more members in any given cluster (**`r I(pasteLabel("Table", tableCount, "t_hypergeom"))`**).
-After adjustment for multiple comparisons,   MADS genes were significantly over-represented in clusters 1 and 4, both of which contain genes that have higher expression in later meristem samples.
-ABI3VP transcription factors are
+Using the list of transcription factor families provided by the Plant TFDB (**reference**), a hypergeometric test was carried out if there were at least two member of TF family in a given cluster.
+After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one TF family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
+For example, members of the ABI3VP1 family were overrepresented in clusters **2** and **4**, both of which contain genes with higher transformed read counts in the rachis meristem than in samples from later stages of development.
+The MADS family was overrepresented in clusters **6** and **8**, suggesting that they could be involved in both early and late stages of inflorescence development.
 
-<a id="t_hypergeom"></a>
-```{r hypergeom, results = "markup"}
-knitr::kable(t_hypergeom, caption = 'Transcription factor family enrichment per cluster', align = 'r')
-```
-
-+ Clusters including 'known' genes
++ Clusters including characterised genes
 
@@
 
-The presence of MADS genes was enriched in **which clusters** and their overall expression is enriched in the spikelet meristem sample.
+The presence of MADS genes was enriched in **some clusters** and their overall expression is enriched in the spikelet meristem sample.
 These results correspond with previous studies that have established the importance of MADS genes in inflorescence development.
@@
 
+**`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling of Euclidean distances between transformed read counts for clustered genes.
+
+**`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
+
 # References
@@
 
+```{r hypergeom, results = "markup"}
+knitr::kable(t_hypergeom, caption = paste0(
+  I(pasteLabel("**Table ", tableCount, "t_hypergeom")),
+  '**: Transcription factor family enrichment per cluster'))
+```
+
 ***
@@
 
-`r I(pasteLabel("Figure", figCount, "f_lmdFigure"))`
-**LASER microdissection of meristem samples in the early stages of inflorescences development.** Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and secondary branch meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
+**`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`**.
+LASER microdissection of meristem samples in the early stages of inflorescences development.
+Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and secondary branch meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
 Images show the samples before (**a**–**d**) and after (**e**–**h**) dissection.
 
-`r I(pasteLabel("Figure", figCount, "f_mfuzzClusters"))`
-**Common patterns of expression in the meristem samples.** Fuzzy *c*-means clustering of normalised, variance-stabilised read counts was used to recover `r capwords(as.character(english::english(c)))` common patterns of expression.
-Each line describes the expression pattern of one gene, with the gene's membership to the cluster mapped to the colour of the line.
+**`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`**.
+Common patterns of expression in the meristem samples.
+Fuzzy *c*-means clustering of normalised, variance-stabilised read counts was used to recover `r as.character(english::english(c))` common patterns of expression.
+Each line describes the expression pattern of one gene, with the gene's membership to the cluster represented by the colour of the line.
 The core values for each cluster are plotted in black.
-
-***
-
-# Figures
-
-***
-
-# Junk
-
-I think `inline code` should work?

### 2015-08-19 si.Rmd
started PCA figure, need to operate on all libraries
minor changes; need to rerun pipeline (overnight).

@@
 
-``` {r libStats}
-tableCount <- incCount(tableCount, "t_libStats")
+``` {r st_libStats, results = "markup"}
+knitr::kable(st_libStats, caption = paste0(
+  I(pasteLabel("**Table S", s_tableCount, "st_libStats")),
+    "**: Library and sequencing statistics"))
 ```
 
-### `r I(pasteLabel("Table", tableCount, "t_libStats"))`
-
-``` {r t_libStats, results = "markup"}
-knitr::kable(t_libStats, caption = "Library and sequencing statistics")
+```{r sf_pca}
+print(sf_pca)
 ```
@@
 
-```{r mfuzz, cache = TRUE}
-figCount <- incCount(figCount, "f_mfuzzCentroids")
-print(f_mfuzzCentroids)
+```{r mfuzz}
+print(sf_mfuzzCentroids)
 
-figCount <- incCount(figCount, "f_mfuzzPca")
-f_mfuzzPca
+print(sf_mfuzzPca)
 ```

### 2015-08-19 src/R/figures.R

@@ -9,10 +9,10 @@ extrafont::loadfonts()
 #####################
 
 # https://gist.githubusercontent.com/rmflight/3858973/raw/591ee603ce0996ac407902f7dd59520b35b3702a/tableFigureFuncs.r
-pasteLabel <- function(preText, inObj, objName, insLink=TRUE){
+pasteLabel <- function(preText, inObj, objName, insLink=FALSE){
   objNum <- inObj[objName]
   
-  useText <- paste(preText, objNum, sep=" ")
+  useText <- paste0(preText, objNum)
   if (insLink){
     useText <- paste("[", useText, "](#", objName, ")", sep="")
   }
@@ -27,17 +27,11 @@ incCount <- function(inObj, useName){
   names(inObj)[nObj+1] <- useName
   inObj
 }
+
 figCount <- c("_"=0)
 tableCount <- c("_"=0)
-
-tableCat <- function(inFrame){
-  outText <- paste(names(inFrame), collapse=" | ")
-  outText <- c(outText, paste(rep("---", ncol(inFrame)), collapse=" | "))
-  invisible(apply(inFrame, 1, function(inRow){
-    outText <<- c(outText, paste(inRow, collapse=" | "))
-  }))
-  return(outText)
-}
+s_figCount <- c("_"=0)
+s_tableCount <- c("_"=0)
 
 capwords <- function(s, strict = FALSE) {
   cap <- function(s) paste(toupper(substring(s, 1, 1)),
@@ -50,18 +44,24 @@ capwords <- function(s, strict = FALSE) {
 ### STATS TABLE ###
 ###################
 
-t_libStats <- readRDS('output/quantStats/libStats.Rds')
+st_libStats <- readRDS('output/quantStats/libStats.Rds')
+s_tableCount <- incCount(s_tableCount, "st_libStats")
+
+##################
+### LMD FIGURE ###
+##################
+
+figCount <- incCount(figCount, "f_lmdFigure")
 
 ################
 ### PCA PLOT ###
 ################
 
-# make this for all libraries, not just the final ones
-
 expressedGenes <- readRDS('output/expressedGenes/expressedGenesAll.Rds')
-vst <- readRDS('output/DESeq2/vst.Rds')
+vst <- readRDS('output/DESeq2/vstAll.Rds')
 
[…]

### 2015-08-19 src/R/mfuzz.R

@@ -74,8 +74,8 @@ centPlot <- ggplot(centroids, aes(x = x, y = y)) +
   theme_minimal() +
   xlab(expression(Cluster~number~"("*italic(c)*")")) +
   ylab("Minimum centroid distance") +
-  geom_point() +
-  stat_smooth(method = loess, se = FALSE)
+  stat_smooth(method = loess, se = FALSE) +
+  geom_point()
 
 # can try to find inflection points, doesn't work very well.
 points <- seq(2, maxClust, length.out = 10000)

### 2015-08-19 src/R/parseQuantStats.R

@@ -81,7 +81,7 @@ if (!file.exists(ddsFile)) {
   quit(save = "no", status = 1)
 }
 dds <- readRDS(ddsFile)
-readsPerLib <- as.data.table(round(colSums(DESeq2::counts(dds))/1000000, 2),
+readsPerLib <- as.data.table(round(colSums(DESeq2::counts(dds))/1000000, 1),
                              keep.rownames = TRUE)
 setnames(readsPerLib, c('lib', 'Reads in genes (M)'))
 

### 2015-08-19 src/py/pipeline.py

@@ -117,6 +117,11 @@ def touch(fname, mode=0o666, dir_fd=None, **kwargs):
         os.utime(f.fileno() if os.utime in os.supports_fd else fname,
             dir_fd=None if os.supports_fd else dir_fd, **kwargs)
 
+
+#---------------------------------------------------------------
+# SETUP TASKS
+#---------------------------------------------------------------
+
 #---------------------------------------------------------------
 # download rice genomes
 #
@@ -217,6 +222,10 @@ def define_os_reads(outputFiles):
         assert os.path.isfile(qualName), "Error: read file " + qualName + " missing"
         print(qualName)
     touch(outputFiles)
+
+#---------------------------------------------------------------
+# ANALYSIS TASKS
+#---------------------------------------------------------------
 
 #---------------------------------------------------------------
 # trim rice reads
@@ -436,6 +445,46 @@ def libStats(inputFiles, outputFiles):
     # update ruffus flag
     print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
     touch(outputFiles)
+    
+
+#---------------------------------------------------------------
+# FIGURES AND TABLES
+#---------------------------------------------------------------
+
+#---------------------------------------------------------------
+# library stats table
+#
+@merge(libStats, "ruffus/table.st_libStats")
+def st_libStats(inputFiles, outputFiles):
+    touch(outputFiles)
+
+#---------------------------------------------------------------
+# Hypergeometric tests table
+#
+@merge(tf_hypergeom, "ruffus/table.t_hypergeom")
+def t_hypergeom(inputFiles, outputFiles):
+    touch(outputFiles)
+
+#---------------------------------------------------------------
+# PCA plot
+#
+@merge([run_deseq2_os, detect_expressed_genes], "ruffus/figure.sf_pca")
+def sf_pca(inputFiles, outputFiles):
+    touch(outputFiles)
+
+#---------------------------------------------------------------
+# Clustering figures
+#
+@merge(clustering, "ruffus/figure.f_mfuzzClusters")
+def f_mfuzzClusters(inputFiles, outputFiles):
+    touch(outputFiles)
+@merge(clustering, "ruffus/figure.sf_mfuzzCentroids")
+def sf_mfuzzCentroids(inputFiles, outputFiles):
+    touch(outputFiles)
+@merge(clustering, "ruffus/figure.sf_mfuzzPca")
[…]

### 2015-08-19 ms.Rmd
rerun pipeline from DESeq2

@@
                cache.path='cache/',
-               dev=c("svg", "cairo_pdf"),
+               dev=c("png", "cairo_pdf"),
                fig.width=3.150,
@@
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling of Euclidean distances between transformed read counts for clustered genes.
+**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
 
 **`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
-
-# References
 
@@
 The core values for each cluster are plotted in black.
+
+***
+
+# References (move to before *Tables*)

### 2015-08-19 si.Rmd
rerun pipeline from DESeq2

@@
                cache.path='cache/',
-               dev=c("svg", "cairo_pdf"),
+               dev=c("png", "cairo_pdf"),
                fig.width=6.614,
@@
 
+**`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
+
 ``` {r st_libStats, results = "markup"}
@@
 
-### Fuzzy *c*-means clustering
+**`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
+The two libraries made from degraded RNA did not appear to cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA had an effect on the quantification of read counts.
+For this reason these two libraries were not included in subsequent analyses.
 
-```{r mfuzz}
+```{r sf_mfuzzCentroids}
 print(sf_mfuzzCentroids)
+```
 
+**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
+Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after *c* = 8.
+
+``` {r sf_mfuzzPca}
 print(sf_mfuzzPca)
 ```
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
+Genes are coloured by the cluster to which they were assigned.
+The separation of genes by cluster on the MDS plot supports the use of eight clusters.

### 2015-08-19 src/py/pipeline.py

@@ -408,7 +408,7 @@ def clustering(inputFiles, outputFiles):
 # hypergeometric tests for enrichment of TFs
 #
 @merge([clustering, download_tfdb], 'ruffus/os.hypergeom')
-def tf_hypergeom(inputFiles, outputFiles):
+def hypergeom(inputFiles, outputFiles):
     jobScript = 'src/R/tfHypergeom.R'
     ntasks = '1'
     cpus_per_task = '1'
@@ -461,7 +461,7 @@ def st_libStats(inputFiles, outputFiles):
 #---------------------------------------------------------------
 # Hypergeometric tests table
 #
-@merge(tf_hypergeom, "ruffus/table.t_hypergeom")
+@merge(hypergeom, "ruffus/table.t_hypergeom")
 def t_hypergeom(inputFiles, outputFiles):
     touch(outputFiles)
 

### 2015-08-20 ms.Rmd
implemented gsea and gsea figure

@@
 
-Examination of the clustered genes suggested that transcription factor genes were overrepresented in most clusters (**annotated cluster tables in SI**).
+Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**).
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor families provided by the Plant TFDB (**reference**), a hypergeometric test was carried out if there were at least two member of TF family in a given cluster.
+Using the list of transcription factor families available on the Plant Transcription Factor Database (**reference**), a hypergeometric test was carried out if there were at least two members of TF family in a given cluster.
 After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one TF family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
@@
 
-+ **Figure**: GSEA of transcription factor families
-
 Because of the enrichment of TF family members in some of the gene expression clusters, geneset-level enrichment analysis (GSEA) was used to compare the levels of expression of TF family members in the meristem samples.
@@
 TF families involved in branching were enriched in **which families, which samples**.
+
+```{r f_gsea}
+print(f_gsea)
+```
 

### 2015-08-20 src/R/gsea.R

@@ -1,0 +1,112 @@
+#!/usr/bin/Rscript
+
+library(gage)
+library(DESeq2)
+library(data.table)
+
+# check for DESeq2 output
+deseqDir <- "output/DESeq2"
+if (!dir.exists(deseqDir)) {
+  cat("deseqDir not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+ddsFile <- paste0(deseqDir, "/ddsWald.Rds")
+if (!file.exists(ddsFile)) {
+  cat("ddsFile file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+dds <- readRDS(ddsFile)
+
+# find tfdb
+tfdbFile <- 'data/tfdb/os.Rds'
+if (!file.exists(tfdbFile)) {
+  cat("tfdb file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+tfdb <- readRDS(tfdbFile)
+
+# one-by-one comparisons of each stage to all other stages using contrast
+# vectors like this: cv <- c(0,1,1,-1,-1,0,0). Repeat for each level of stage
+stage <- levels(dds$stage)
+
+getRes <- function(x) {
+  x <- make.names(x)
+  # contrast vector based on results names
+  cv <- resultsNames(dds)
+  # resultsNames that don't start with stage aren't in the comparison
+  cv[!grepl("^stage", cv)] <- 0
+  # resultsNames for our stage of interest are +1.
+  cv[grepl(x, cv)] <- 1
+  # all other stages are for comparison. Set them to -1.
+  cv[!(cv == 1 | cv == 0)] <- -1
+  # make numeric for results call
+  cv <- as.numeric(cv)
+  return(results(dds, contrast = cv)$log2FoldChange)
+}
+
+lfcFrame <- sapply(stage, getRes)
+rownames(lfcFrame) <- rownames(counts(dds))
+
+# run gage on the data.frame and extract test statistics
+res <- gage(lfcFrame, gsets = split(tfdb$Protein.ID, tfdb$Family),
+            samp = NULL, ref = NULL)
+gageFrame <- res$stats
+gageFrame <- gageFrame[, stage]
+gageFrame <- gageFrame[complete.cases(gageFrame), ]
+
+# get the family order for the plot
+famOrder <- rownames(gageFrame)[hclust(dist(gageFrame))$order]
+
+# make p-value data.frames
+greaterP <- res$greater
[…]

### 2015-08-20 src/R/tfHypergeom.R

@@ -47,6 +47,12 @@ hyperGeomResults <- sapply(1:length(tfdbByFamily), function(i)
          clusterSize, lower.tail = FALSE))
 colnames(hyperGeomResults) <- names(tfdbByFamily)
 
+# what about ALOGs (not TFs)
+ALOG <- c('LOC_Os07g04670', 'LOC_Os02g07030', 'LOC_Os06g46030',
+          'LOC_Os02g41460', 'LOC_Os04g43580', 'LOC_Os10g33780',
+          'LOC_Os02g56610', 'LOC_Os01g61310', 'LOC_Os05g39500', 'LOC_Os05g28040')
+
+
 # combine the results for adjusting p-values
 resultsLong <- merge(reshape2::melt(inClusterPerFamily,
                                     varnames = c("Cluster", "Family"),

### 2015-08-20 src/py/pipeline.py

@@ -446,6 +446,19 @@ def libStats(inputFiles, outputFiles):
     print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
     touch(outputFiles)
     
+#---------------------------------------------------------------
+# Geneset enrichment analysis
+#
+@merge([download_tfdb, run_deseq2_os], 'ruffus/os.gsea')
+def gsea(inputFiles, outputFiles):
+    jobScript = 'src/R/gsea.R'
+    ntasks = '1'
+    cpus_per_task = '1'
+    job_name = 'libStats'
+    jobId = submit_job(jobScript, ntasks, cpus_per_task, job_name)
+    # update ruffus flag
+    print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
+    touch(outputFiles)
 
 #---------------------------------------------------------------
 # FIGURES AND TABLES
@@ -485,6 +498,12 @@ def sf_mfuzzCentroids(inputFiles, outputFiles):
 def sf_mfuzzPca(inputFiles, outputFiles):
     touch(outputFiles)
 
+#---------------------------------------------------------------
+# GSEA plot
+#
+@merge(gsea, "ruffus/figure.f_gsea")
+def f_gsea(inputFiles, outputFiles):
+    touch(outputFiles)
 
 # options for visualising
 pipeline_printout()

### 2015-08-20 src/R/figures.R
use adjusted p-values for gsea figure. nothing significant now :(

@@ -202,7 +202,7 @@ gsea[, Stage := plyr::mapvalues(Stage, "ePBM/SBM", "ePBM/\nSBM")]
 
 heatscale <- rev(RColorBrewer::brewer.pal(6, "RdBu"))
 
-f_gsea <- ggplot(gsea, aes(x = Stage, y = rn, label = pval, fill = `Test\nstatistic`)) +
+f_gsea <- ggplot(gsea, aes(x = Stage, y = rn, label = padj, fill = `Test\nstatistic`)) +
   theme_minimal(base_size = 8, base_family = "Helvetica") +
   xlab(NULL) + ylab(NULL) +
   theme(legend.key.size = grid::unit(1, "lines"),

### 2015-08-20 src/R/gsea.R

@@ -84,10 +84,15 @@ plotData <- lpTable.melt[gpTable.melt][gageTable.melt]
 # pick which p-value to use
 plotData[`Test statistic` < 0, pval := pl]
 plotData[`Test statistic` > 0, pval := pg]
-plotData[, showPval := pval < 0.1]
+
+# adjust p-values, need un-data-table syntax to avoid segfault. yuck.
+plotData[!is.na(pval), padj := p.adjust(pval, "BH")]
+
+# factor for lazy ggplotting
+plotData[, showPval := padj < 0.1]
 
 # make the pvals pretty
-plotData[, pval := format.pval(c(0.12, pval), digits = 1, eps = 0.001, na.form = '')[-1]]
+plotData[, padj := format.pval(c(0.12, padj), digits = 2, eps = 0.001, na.form = '')[-1]]
 
 # order the y-axis
 plotData[, rn := factor(rn, levels = famOrder)]

### 2015-08-20 ms.Rmd
use adjusted p-values for gsea figure. nothing significant now :(
writing

@@
 
-+ Clusters including characterised genes
+* Clusters including characterised genes
 
@@
 
-Because of the enrichment of TF family members in some of the gene expression clusters, geneset-level enrichment analysis (GSEA) was used to compare the levels of expression of TF family members in the meristem samples.
-**Something like: Consistent with their enrichment in cluster 1 and 6, MADS genes were significantly enriched in the SM/FM samples**.
+Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of TF family expression in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
+In general, the GSEA results correspond to the hypergeometric tests for transcription factor enrichment in the clusters.
+For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster **6**, which contains genes with higher expression in those meristem types.
+
 TF families involved in branching were enriched in **which families, which samples**.
@@
 
-The presence of MADS genes was enriched in **some clusters** and their overall expression is enriched in the spikelet meristem sample.
-These results correspond with previous studies that have established the importance of MADS genes in inflorescence development.
-The evolution of their expression patterns was
+There was an overrepresentation of MADS genes in **some clusters** and their overall expression is enriched in the spikelet meristem sample.
+These results correspond with previous studies that have established the importance of MADS genes in inflorescence development in rice and other plants (**expand**).
 
-+ **Figure**: Similarity tree of MADS proteins with expression values and dN/dS values for each one.
+To investigate the evolution of the MADS family in Angiosperms, their expression values in the dicots *A.
+thaliana* and *Solanum lycopersicum* and the monocot *O.
+sativa* were compared to their homology in these species.
+
+* **Figure**: Similarity tree of MADS proteins with L2FC values between comparable samples (and dN/dS values?) for each one:
++ **rice**: rachis meristem vs. spikelet merisem
++ **arapidospis**: inflorescence meristem vs. floret meristem
++ **tomato**: simpodial meristem vs. floret meristem (need to find list of tomato MADS genes)
++ (could I map the l2fc to the background of the geom_label rectangle? "The fill aesthetic controls the background colour of the label.")
+
+## Comparison with characterised genes
+
+To compare previously reported expression patterns with the results presented here, which were produced with high-throughput sequencing, a selection of studies of corresponding stages of inflorescence development were searched for qualitative *in situ* hyrbridisation data.
+
+* **Table**: our expression calls (yes/no by stage) *vs* in situ reports. Fun times.
+* our in situ results could go here
 
@@
 
+**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**.
+Geneset enrichment analysis (GSEA) for transcription factor families.
+Test statistics were calculated on a matrix of Log~2~(fold change) values between each sample and the other three samples.
+A positive test statistic indicates enrichment and a negative statistic indicates depletion.
+Although some test statistics had low *p*-values none were less than 0.10 after Benjamini-Hochberg correction.
+
 ***

### 2015-08-21 ms.Rmd
writing

@@
     fig_caption: yes
+    reference_docx: csl/wordStyles.docx
   pdf_document:
@@
                fig.width=3.150,
-               fig.height=6.3,
+               fig.height=8.661,
                fig.align = 'center',
@@
 
-### Running title
+## Running title
 
-### Keywords
+## Keywords
 
-### SRA details for libaries
+## SRA details for libaries
 
 Word count
-
-***
 
@@
 
-```{r f_mfuzzClusters, fig.height=6.3}
+```{r f_mfuzzClusters, fig.height = 8.661 / 2}
 print(f_mfuzzClusters)
@@
 
-```{r f_gsea}
+```{r f_gsea, fig.height = 8.661 / 2}
 print(f_gsea)
@@
 
-***
-
 # Tables
@@
 
-***
+```{r pander, results = "asis"}
+pander::set.alignment('left', row.names = 'right')
+pander::set.caption(paste0(
+  I(pasteLabel("**Table ", tableCount, "t_hypergeom")),
+  '**: Transcription factor family enrichment per cluster'))
+pander::pandoc.table(t_hypergeom, split.tables = Inf, emphasize.rownames = FALSE)
+```
 
@@
 
-***
+# `R sessionInfo()` (*hide*)
+
+```{r sessionInfo, results = "asis"}
+pander::pander(sessionInfo())
+git <- c(paste("**git branch:**",system("git rev-parse --abbrev-ref HEAD", intern = TRUE)),
+          paste("**git hash:**", system("git rev-parse HEAD", intern = TRUE)))
+
+for (x in git){cat(x, "\n")}
+
+```
 

### 2015-08-24 ms.Rmd
run GSEA only on expressed genes

@@
 
+The patterns are not completely representative of all members of each gene family.
+For example, there are several MADS genes that have lower expression in the SM than in the RM, but expressed MADS genes were detected at a higher level overall in SM.
+Nevertheless, there are some notable modes of expression shared between different TF families.
+Several TF families, including bZIP and ABI3VP1, are strongly expressed in RM and lower in the other meristem types.
+Another group incluiding Squamosa binding proteins and Jumonji transcription factors is more weakly expressed in the RM but enriched in the PBM, suggesting a switch of expression between these two meristem types.
+Correspondingly, the difference between the RM and PBM samples was the most remarkable in the GSEA.
+
 TF families involved in branching were enriched in **which families, which samples**.
@@
 Geneset enrichment analysis (GSEA) for transcription factor families.
-Test statistics were calculated on a matrix of Log~2~(fold change) values between each sample and the other three samples.
+Test statistics were calculated on a matrix of Log~2~(fold change) values for expressed TF family members between each sample and the other three samples.
 A positive test statistic indicates enrichment and a negative statistic indicates depletion.
-Although some test statistics had low *p*-values none were less than 0.10 after Benjamini-Hochberg correction.
+Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.
 

### 2015-08-24 si.Rmd
implemented gsea and gsea figure
run GSEA only on expressed genes

@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
-The two libraries made from degraded RNA did not appear to cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA had an effect on the quantification of read counts.
+The two libraries made from degraded RNA did not cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA had an effect on the quantification of read counts.
 For this reason these two libraries were not included in subsequent analyses.
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
-Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after *c* = 8.
+Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after $c = 8$.
 

### 2015-08-24 src/R/gsea.R

@@ -25,6 +25,14 @@ if (!file.exists(tfdbFile)) {
 }
 tfdb <- readRDS(tfdbFile)
 
+# load expressed genes
+exprGenFile <- "output/expressedGenes/expressedGenesAll.Rds"
+if (!file.exists(exprGenFile)) {
+  cat("exprGenFile file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+expressedGenes <- readRDS(exprGenFile)
+
 # one-by-one comparisons of each stage to all other stages using contrast
 # vectors like this: cv <- c(0,1,1,-1,-1,0,0). Repeat for each level of stage
 stage <- levels(dds$stage)
@@ -47,8 +55,10 @@ getRes <- function(x) {
 lfcFrame <- sapply(stage, getRes)
 rownames(lfcFrame) <- rownames(counts(dds))
 
-# run gage on the data.frame and extract test statistics
-res <- gage(lfcFrame, gsets = split(tfdb$Protein.ID, tfdb$Family),
+# run gage on the expressed TFDB genes and extract test statistics
+tfdb.expressed <- tfdb[Protein.ID %in% expressedGenes]
+
+res <- gage(lfcFrame, gsets = split(tfdb.expressed$Protein.ID, tfdb.expressed$Family),
             samp = NULL, ref = NULL)
 gageFrame <- res$stats
 gageFrame <- gageFrame[, stage]

### 2015-08-24 src/py/pipeline.py

@@ -449,7 +449,7 @@ def libStats(inputFiles, outputFiles):
 #---------------------------------------------------------------
 # Geneset enrichment analysis
 #
-@merge([download_tfdb, run_deseq2_os], 'ruffus/os.gsea')
+@merge([download_tfdb, run_deseq2_os, detect_expressed_genes], 'ruffus/os.gsea')
 def gsea(inputFiles, outputFiles):
     jobScript = 'src/R/gsea.R'
     ntasks = '1'

### 2015-08-24 src/R/compareVsInSitu.R
started nasty code for comparing published in situ results

@@ -1,0 +1,87 @@
+#!/usr/bin/Rscript
+
+library(data.table)
+
+
+# load zhang file
+zhangFile <- "data/zhangGenes.tab"
+if (!file.exists(zhangFile)) {
+  cat("zhangFile file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+zhangGenes <- data.table(read.table(zhangFile, sep = "\t", header = TRUE,
+                                    stringsAsFactors = FALSE))
+setkey(zhangGenes, 'msuId')
+
+# load expressed genes
+exprGenFile <- "output/expressedGenes/expressedGenesByLibrary.Rds"
+if (!file.exists(exprGenFile)) {
+  cat("exprGenFile file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+expressedGenesByLibrary <- readRDS(exprGenFile)
+
+# split the zhang genes expression field
+trim <- function(x) {gsub("^\\s+|\\s+$", "", x)}
+idVsExp <- zhangGenes[, .(
+  meristem = toupper(trim(unlist(strsplit(geneExp, "[,]"))))
+  ), by = msuId]
+setkey(idVsExp, 'msuId')
+
+# get the LOCs that have expression we're interested in
+RM <- c("IM", "RM")
+PBM <- c("BM", "PBM")
+SBM <- c("BM", "SBM")
+SM <- c("SM", "FM")
+
+RMt <- idVsExp[meristem %in% RM, .(msuId = unique(msuId), RM = TRUE)]
+PBMt <- idVsExp[meristem %in% PBM, .(msuId = unique(msuId), PBM = TRUE)]
+SBMt <- idVsExp[meristem %in% SBM, .(msuId = unique(msuId), SBM = TRUE)]
+SMt <- idVsExp[meristem %in% SM, .(msuId = unique(msuId), SM = TRUE)]
+
+# this is nasty
+merge(RMt, merge(PBMt, merge(SBMt, SMt, all = TRUE), all = TRUE), all = TRUE)
+
+
+
+inSituResults <- idVsExp[meristem %in% tissues]
+
+newtab <- inSituResults[, .(msuId = unique(msuId))]
+
+
+inSituResults[,expressed := TRUE]
+
+
+
+# cast to wide
+reshape2::dcast(inSituResults, msuId ~ meristem, value.var = "expressed")
+
+
+# how is this going to be plotted?
[…]

### 2015-08-25 src/R/compareVsInSitu.R
bug in data.table split leading to lost data points'

@@ -10,7 +10,6 @@ if (!file.exists(zhangFile)) {
 }
 zhangGenes <- data.table(read.table(zhangFile, sep = "\t", header = TRUE,
                                     stringsAsFactors = FALSE))
-setkey(zhangGenes, 'msuId')
 
 # load expressed genes
 exprGenFile <- "output/expressedGenes/expressedGenesByLibrary.Rds"
@@ -22,11 +21,9 @@ expressedGenesByLibrary <- readRDS(exprGenFile)
 
 # split the zhang genes expression field.
 trim <- function(x) {gsub("^\\s+|\\s+$", "", x)}
-zhangGenes[, geneExp := toupper(trim(unlist(strsplit(geneExp, ",", fixed = TRUE)))),
-           by = msuId]
-
-# remove some columns that we don't need for this plot
-zhangGenes[,c("geneName", "geneFunction", "geneEvidence"):= NULL]
+zhangGenes <- zhangGenes[,.(
+  geneExp = trim(unlist(strsplit(geneExp, "[,]")))
+  ), by = c('msuId', 'zhangRef', 'plotLabel')]
 
 # get the LOCs that have expression we're interested in
 RM <- c("IM", "RM")
@@ -48,7 +45,7 @@ expZhangGenes[,FM := geneExp %in% FM]
 setkey(expZhangGenes, "msuId", "zhangRef")
 expZhangGenes <- expZhangGenes[, .(
   plotLabel = unique(plotLabel),
-  geneExp = paste(geneExp, collapse = " | "),
+  geneExp = paste(unique(geneExp), collapse = " | "),
   RM = any(RM),
   PBM = any(PBM),
   SBM = any(SBM),

### 2015-08-25 src/R/figures.R

@@ -219,25 +219,31 @@ figCount <- incCount(figCount, "f_gsea")
 ### COMPARE IN SITU ###
 #######################
 
-compare <- readRDS('output/compare/')
+compare <- readRDS('output/compare/compare.Rds')
 
 # test plot
 colours <- RColorBrewer::brewer.pal(3, "Set1")[c(2,1,3)]
 labels <- c("1" = "Detected\nin both", "2" = "Only detected\nby in situ", "3" = "Only detected by\nRNA-sequencing")
-ggplot(plotData.long, aes(x = stage, y = id, colour = as.factor(compare))) +
+f_reviewInSitu <- ggplot(compare, aes(x = stage, y = id, colour = as.factor(compare))) +
   theme_minimal(base_size = 8, base_family = "Helvetica") +
   theme(axis.text.y = element_blank(),
+        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
         panel.grid = element_blank()) +
   xlab(NULL) + ylab(NULL) +
+  coord_fixed(ratio = 1) +
   scale_colour_manual(values = colours, na.value = NA,
-                      labels = labels) +
+                      labels = labels, name = NULL) +
   geom_point(size = 1) +
   scale_x_discrete(labels = c("", "RM", "PBM", "SBM", "SM", "FM", ""),
                    limits = c("id", "RM", "PBM", "SBM", "SM", "FM", "zhangRef"),
-                   expand = c(0,-0.5)) +
+                   expand = c(2,0)) +
   # map one label to 1.5 and hjust to the left
-  geom_text(mapping = aes(x = 1.9, y = id, label = plotLabel),
-            hjust = 1, colour = "black", size = 2, fontface = "italic") +
+  geom_text(mapping = aes(x = 1.5, y = id, label = plotLabel),
+            hjust = 1, colour = "black", size = 2, fontface = "italic",
+            family = "Helvetica") +
   # map the other to max + 0.5 and hjust to the right
-  geom_text(mapping = aes(x = 6.1, y = id, label = zhangRef),
-            hjust = 0, colour = "black", size = 2, fontface = "plain") 
+  geom_text(mapping = aes(x = 6.5, y = id, label = zhangRef),
+            hjust = 0, colour = "black", size = 2, fontface = "plain",
+            family = "Helvetica") 
+f_reviewInSitu
+figCount <- incCount(figCount, "f_reviewInSitu")

### 2015-08-25 src/R/compareVsInSitu.R
stupid error calling expressed genes

@@ -75,7 +75,7 @@ egbl.melt[, stage := sub("r\\d+", "", sample)]
 egbl.melt[, stage := plyr::mapvalues(stage, from = c("n1", "n2", "n3", "n4"),
                                      to = c("RM", "PBM", "SBM", "SM"))]
 geneCalls <- data.table(reshape2::dcast(egbl.melt, msuId ~ stage, value.var = "call",
-                                        fun.aggregate = function(x) sum(x) > 2), key = "msuId")
+                                        fun.aggregate = function(x) sum(x) > 1), key = "msuId")
 # add a fake "FM" column to check if we are getting a lot of genes in the SM sample
 geneCalls[, FM := SM]
 

### 2015-08-25 ms.Rmd
started nasty code for comparing published in situ results
bug in data.table split leading to lost data points'
stupid error calling expressed genes
added factor for facetting GSEA plot on TF category

@@
 
-## Patterns of expression of transcription families
+## Dynamic expression of transcription factor families
 
@@
 Another group incluiding Squamosa binding proteins and Jumonji transcription factors is more weakly expressed in the RM but enriched in the PBM, suggesting a switch of expression between these two meristem types.
-Correspondingly, the difference between the RM and PBM samples was the most remarkable in the GSEA.
+Correspondingly, the most pronounced difference in the GSEA was that between the RM and PBM samples.
 
@@
 To compare previously reported expression patterns with the results presented here, which were produced with high-throughput sequencing, a selection of studies of corresponding stages of inflorescence development were searched for qualitative *in situ* hyrbridisation data.
+
+```{r f_reviewInSitu, fig.height = 8.661 / 4}
+print(f_reviewInSitu)
+```
 

### 2015-08-25 src/R/figures.R

@@ -195,10 +195,19 @@ tableCount <- incCount(tableCount, "t_hypergeom")
 ############
 
 gsea <- readRDS('output/gsea/gseaTable.Rds')
+famCat <- readRDS("data/tfdb/famCat.Rds")
 
 # format some labels
 setnames(gsea, old = "Test statistic", new = "Test\nstatistic")
 gsea[, Stage := plyr::mapvalues(Stage, "ePBM/SBM", "ePBM/\nSBM")]
+
+# separate by TF / other proteins
+setkey(famCat, 'Family')
+setkey(gsea, 'rn')
+gsea[, Category := famCat[gsea][,Category]]
+gsea[, Category := plyr::mapvalues(Category, from = c("TF", "Other"),
+                       to = c("Transcription factors", "Other regulators"))]
+gsea[, Category := factor(Category, levels = c("Transcription factors", "Other regulators"))]
 
 heatscale <- rev(RColorBrewer::brewer.pal(6, "RdBu"))
 
@@ -210,6 +219,7 @@ f_gsea <- ggplot(gsea, aes(x = Stage, y = rn, label = padj, fill = `Test\nstatis
   scale_y_discrete(expand = c(0,0)) +
   scale_x_discrete(expand = c(0,0)) +
   scale_fill_gradientn(colours = heatscale) +
+  facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
   geom_raster() +
   geom_text(data = gsea[showPval == TRUE], size = 2)
 
@@ -230,7 +240,7 @@ f_reviewInSitu <- ggplot(compare, aes(x = stage, y = id, colour = as.factor(comp
         axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
         panel.grid = element_blank()) +
   xlab(NULL) + ylab(NULL) +
-  coord_fixed(ratio = 1) +
+#  coord_fixed(ratio = 1) +
   scale_colour_manual(values = colours, na.value = NA,
                       labels = labels, name = NULL) +
   geom_point(size = 1) +

### 2015-08-25 src/R/processTfdb.R

@@ -27,8 +27,12 @@ at <- unique(tfdbRaw[Species == "Arabidopsis thaliana" , .(
   Family
 )])
 
+# list of family vs. category
+famCat <- unique(tfdbRaw["Oryza sativa subsp. japonica",.(Family, Category)])
+
 # output
 saveRDS(os, 'data/tfdb/os.Rds')
 saveRDS(at, 'data/tfdb/at.Rds')
+saveRDS(famCat, 'data/tfdb/famCat.Rds')
 
 quit(save = "no", status = 0)

### 2015-08-26 ms.Rmd
switched table type from kable to pander

@@
 RNA from the meristem samples was isolated, amplified and used for the production of a sequencing library.
-**need area cut numbers**.
+**Need area cut +/- number of panicles used for each library from Israr**.
 RNA input for library preparation was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
@@
 reads that were not generated from genes), `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
+As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (currently `r I(pasteLabel("Figure ", figCount, "f_reviewInSitu"))`, **move figure to supporting table**).
+
+```{r f_reviewInSitu, fig.height = 8.661 / 4}
+print(f_reviewInSitu)
+```
+
+(**our in situ results could go here or in SI**)
 
@@
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
-Clusters were recovered that contain genes that gradually decrease or increase in expression during development (clusters **2**, **3**, **6** and **8**), or that are apparently turned on or off in primary and secondary branch meristems (clusters **7** and **4** respectively) along with genes with more complex expression patterns (clusters **1** and **5**).
+Clusters were recovered that contain genes that gradually decrease or increase in expression during development (clusters **x**, **x**, **x** and **x**), or that are apparently turned on or off in primary and secondary branch meristems (clusters **x** and **x** respectively) along with genes with more complex expression patterns (clusters **x** and **x**).
 
@@
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor families available on the Plant Transcription Factor Database (**reference**), a hypergeometric test was carried out if there were at least two members of TF family in a given cluster.
-After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one TF family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-For example, members of the ABI3VP1 family were overrepresented in clusters **2** and **4**, both of which contain genes with higher transformed read counts in the rachis meristem than in samples from later stages of development.
-The MADS family was overrepresented in clusters **6** and **8**, suggesting that they could be involved in both early and late stages of inflorescence development.
+Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database (**reference**), a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
+After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
+For example, members of the ABI3VP1 family were overrepresented in clusters **x** and **x**, both of which contain genes with higher transformed read counts in the rachis meristem than in samples from later stages of development.
+The MADS family was overrepresented in clusters **x** and **x**, suggesting that they could be involved in both early and late stages of inflorescence development.
 
-* Clusters including characterised genes
+**Discussion: which clusters have genes that are already candidates for panicle branching?**
 
@@
 
-Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of TF family expression in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
+Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
 In general, the GSEA results correspond to the hypergeometric tests for transcription factor enrichment in the clusters.
-For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster **6**, which contains genes with higher expression in those meristem types.
+For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster **x**, which contains genes with higher expression in those meristem types.
 
-The patterns are not completely representative of all members of each gene family.
+**cluster y-axis of GSEA plot w/in protein type**
+
+These patterns are not intended to represent each member of each gene family.
 For example, there are several MADS genes that have lower expression in the SM than in the RM, but expressed MADS genes were detected at a higher level overall in SM.
-Nevertheless, there are some notable modes of expression shared between different TF families.
-Several TF families, including bZIP and ABI3VP1, are strongly expressed in RM and lower in the other meristem types.
-Another group incluiding Squamosa binding proteins and Jumonji transcription factors is more weakly expressed in the RM but enriched in the PBM, suggesting a switch of expression between these two meristem types.
-Correspondingly, the most pronounced difference in the GSEA was that between the RM and PBM samples.
+Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
+In the GSEA, the most pronounced change in expression appears to occur between the RM and branch meristem samples.
+Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
+Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enrinched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
 
-TF families involved in branching were enriched in **which families, which samples**.
+**Discussion: In which samples were TF families involved in branching enriched?**
 
@@
 
-## Comparison with characterised genes
-
-To compare previously reported expression patterns with the results presented here, which were produced with high-throughput sequencing, a selection of studies of corresponding stages of inflorescence development were searched for qualitative *in situ* hyrbridisation data.
-
-```{r f_reviewInSitu, fig.height = 8.661 / 4}
-print(f_reviewInSitu)
-```
-
-* **Table**: our expression calls (yes/no by stage) *vs* in situ reports. Fun times.
-* our in situ results could go here
-
 # Experimental procedures
@@
 
-```{r hypergeom, results = "markup"}
-knitr::kable(t_hypergeom, caption = paste0(
-  I(pasteLabel("**Table ", tableCount, "t_hypergeom")),
-  '**: Transcription factor family enrichment per cluster'))
-```
-
 ```{r pander, results = "asis"}
@@
   I(pasteLabel("**Table ", tableCount, "t_hypergeom")),
-  '**: Transcription factor family enrichment per cluster'))
+  "**: Transcription factor and transcriptional regulator family enrichment per cluster, including the number of family members (*n*), number expressed in the whole dataset (*n*~expr~) and number and adjusted *p*-value for each cluster, for families that had at least two members in any one cluster."))
 pander::pandoc.table(t_hypergeom, split.tables = Inf, emphasize.rownames = FALSE)

### 2015-08-26 ms.Rmd
MM

@@
 output:
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
   BiocStyle::html_document:
@@
     dev: "CairoPDF"
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
   pdf_document:
@@
 
-(**our in situ results could go here or in SI**)
+**Our in situ results could go here or in SI**
 
@@
 
-**cluster y-axis of GSEA plot w/in protein type**
+```{r f_gsea, fig.height = 8.661 / 2}
+print(f_gsea)
+```
 
@@
 
-```{r f_gsea, fig.height = 8.661 / 2}
-print(f_gsea)
-```
-
 ## MADS genes: comparison of expression & homology in rice, tomato, arabidopsis
 
-There was an overrepresentation of MADS genes in **some clusters** and their overall expression is enriched in the spikelet meristem sample.
+There was an overrepresentation of MADS genes in clusters **xyz** and their overall expression is enriched in the spikelet meristem sample.
 These results correspond with previous studies that have established the importance of MADS genes in inflorescence development in rice and other plants (**expand**).
@@
 # Experimental procedures
+
+## Plant material and sampling
+
+*Oryza sativa* ssp.
+japonica cv.
+Nipponbare plants were grown in a growth chamber at 70% relative humidity at 30°C during the day and 26°C at night.
+After 7 weeks in long day conditions (LD; 14 hours light and 10 hours darkness) plants were transferred to short day conditions (SD; 10 hours light and 14 hours darkness) to induce floral transition.
+Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod **under which microscope?**.
+
+## Tissue embedding, LASER microdissection and sequencing
+
+Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated **filtered?** twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C.
+The following steps of tissue embedding for dissection were performed as described by @Mantegazza:2014di.
+Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 LASER dissector (Leica).
+RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN).
+Library preparation with **which kit** and 50-base single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
+
+## Data analyses
+
+Analysis of the sequenced libraries was carried out with custom `bash`, `R` (**ref**) and `python` (**ref**) scripts.
+All of the code used for the analysis and to generate figures and tables for this report, along with the versions and paramaters of the software used, are available on GitHub (https://github.com/evoreprice/lmdPaper).
+Briefly, adaptor trimming was performed with `cutadapt` (**ref**) and reads were mapped to the MSUv7 genome (**ref**) using `rna-star` (**ref**) in 2-pass mode with annotations.
+Differential expression analysis to calculate L~2~FC values and transformed read counts was performed with the `DESeq2` package (**ref**).
 

### 2015-08-27 ms.Rmd
code for re-ordering mfuzz plot. need to run clustering and hypergeom again to verify

--- 2015-08-26 ms.Rmd
@@
 
-At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after removal of reads likely to have originated from rRNA and tRNA.
-Using cutoffs calculated from the 95^th^ percentile of distributions of expression values estimated from intergenic regions for each library (*i.e*.
-reads that were not generated from genes), `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
+At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
+Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
 As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (currently `r I(pasteLabel("Figure ", figCount, "f_reviewInSitu"))`, **move figure to supporting table**).
@@
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
-Clusters were recovered that contain genes that gradually decrease or increase in expression during development (clusters **x**, **x**, **x** and **x**), or that are apparently turned on or off in primary and secondary branch meristems (clusters **x** and **x** respectively) along with genes with more complex expression patterns (clusters **x** and **x**).
+There are several expression patterns descibed by the clusters: genes that increase or decrease steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that are constant between RM and PBM and then change between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
+There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).
 
@@
 The MADS family was overrepresented in clusters **x** and **x**, suggesting that they could be involved in both early and late stages of inflorescence development.
+**run clustering script again and check these statements!!!***
 
@@
 Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
-Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enrinched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
+Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
 
@@
 # Experimental procedures
+
+## Plant material and sampling
+
+*Oryza sativa* ssp.
+japonica cv.
+Nipponbare plants were grown in a growth chamber at 70% relative humidity at 30°C during the day and 26°C at night.
+After 7 weeks in long day conditions (LD; 14 hours light and 10 hours darkness) plants were transferred to short day conditions (SD; 10 hours light and 14 hours darkness) to induce floral transition.
+Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.
+
+## Tissue embedding, LASER microdissection and sequencing
+
+Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (=**filtered**?) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C.
+Embedding for dissection was performed as described by @Mantegazza:2014di.
+Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 LASER dissector (Leica).
+RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN).
+Library preparation with **which kit** and 50-base single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
+
+## Data analyses
+
+Analysis of the sequenced libraries was carried out with custom `bash` and `R` (**ref**) scripts, which were arranged into a pipeline using the `ruffus` package for `python3` (**ref**).
+All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.
+
+Briefly, adaptor trimming was performed with `cutadapt` (**ref**) and reads were mapped with the MSUv7 genome and annotation (**ref**) downloaded from Phytozome 10.3 (**ref**) using `rna-star` (**ref**) in 2-pass mode.
+The `quantMode` argument of `rna-star` was used for read counting.
+rRNA and tRNA contamination was estimated using `htseq-count` (**ref**) to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with `wgsim` (**ref**) from rRNA sequences in the Plant Repeats Database (**ref**) also mapped.
+Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package (**ref**).
+To determine a strict cutoff for unexpressed genes, expression values in transcripts per million (TPM) were calculated for each gene (**2 refs**).
+These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values.
+For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values.
+Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package (**ref**) and geneset enrichment analysis with the `gage` package (**ref**) respectively.
+All plots were produced with the `ggplot2` package (**ref**).
 

### 2015-08-27 si.Rmd
code for re-ordering mfuzz plot. need to run clustering and hypergeom again to verify

@@
                dev=c("png", "cairo_pdf"),
-               fig.width=6.614,
-               fig.height=6.614,
                fig.align = 'center',

### 2015-08-27 src/R/figures.R

@@ -77,7 +77,7 @@ pcaPlotData$rIdx <- 1:dim(pcaPlotData)[1]
 opIdx <- pcaPlotData[c('n2r3', 'n4r1'),'rIdx']
 
 sf_pca <- ggplot(data = pcaPlotData,
-                aes(x = PCA1, y = PCA2, colour = Stage, label = label)) +
+                 aes(x = PCA1, y = PCA2, colour = Stage, label = label)) +
   theme_minimal(base_size = 8, base_family = "Helvetica") +
   scale_colour_brewer(palette = "Set1", name = 'Stage') +
   guides(colour=guide_legend(title=NULL)) +
@@ -103,24 +103,66 @@ c1 <- readRDS('output/mfuzz/c1.Rds')
 memCutoff <- 0.5
 
 # get clusters and membership
-cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster, Membership = apply(c1$membership, 1, max), key = "id")
+cluster <- data.table(id = names(c1$cluster), Cluster = c1$cluster,
+                      Membership = apply(c1$membership, 1, max), key = "id")
 
 # get standardised VST counts
-exprs <- data.table(Biobase::exprs(expressionMatrix), keep.rownames = TRUE, key = "rn")
+exprs <- data.table(Biobase::exprs(expressionMatrix), keep.rownames = TRUE)
+exprs[,id := rn][, rn := NULL]
+setkey(exprs, "id")
 
-# set up cluster labels
-numGenes <- cluster[Membership > memCutoff, .(number = length(id)), by = Cluster]
-setkey(numGenes, 'Cluster')
-clustLabels <- numGenes[, paste0("Cluster ", Cluster, "\n(", number, " genes)")]
+plotData.wide <- exprs[cluster[Membership > memCutoff]]
+plotData.wide[, number := length(id), by = Cluster]
+plotData.wide[, label := factor(paste0("Cluster ", Cluster, "\n(", number, " genes)"))]
 
-# merge clusters with expression values
-plotData.wide <- cluster[exprs]
+# relevel clusters
+centres.wide <- data.table(c1$centers)
+
+# re-order the cluster for the plot
+centres.wide[, Cluster := paste("Cluster", 1:nrow(centres.wide))]
+
+# find the changes between RM and PBM and PBM and SM for each cluster
+centres.wide[, c("n1n2", "n2n4") :=
+               list(PBM - RM,
+                    SM - PBM)]
+# divide these changes into categories
[…]

### 2015-08-28 ms.Rmd
fixed gene name error in zhang in situ db. need to re-run table generator

@@
     number_sections: yes
-bibliography: bib/test.bib
+bibliography: bib/papersLib.bib
 ---
@@
 To recover common expression patterns in genes detected in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
-Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo].
+Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
@@
 
-Analysis of the sequenced libraries was carried out with custom `bash` and `R` (**ref**) scripts, which were arranged into a pipeline using the `ruffus` package for `python3` (**ref**).
+Analysis of the sequenced libraries was carried out with custom `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp].
 All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.
 
-Briefly, adaptor trimming was performed with `cutadapt` (**ref**) and reads were mapped with the MSUv7 genome and annotation (**ref**) downloaded from Phytozome 10.3 (**ref**) using `rna-star` (**ref**) in 2-pass mode.
-The `quantMode` argument of `rna-star` was used for read counting.
-rRNA and tRNA contamination was estimated using `htseq-count` (**ref**) to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with `wgsim` (**ref**) from rRNA sequences in the Plant Repeats Database (**ref**) also mapped.
-Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package (**ref**).
-To determine a strict cutoff for unexpressed genes, expression values in transcripts per million (TPM) were calculated for each gene (**2 refs**).
+Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped with the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode .
+The `'quantMode'` argument of `STAR` was used for read counting.
+rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with [`wgsim`](https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
+Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
+To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf].
 These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values.
 For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values.
-Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package (**ref**) and geneset enrichment analysis with the `gage` package (**ref**) respectively.
-All plots were produced with the `ggplot2` package (**ref**).
+Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively.
+All plots were produced with the `ggplot2` package [@Wickham:2009tn].
 

### 2015-08-28 si.Rmd
fixed gene name error in zhang in situ db. need to re-run table generator

@@
 The separation of genes by cluster on the MDS plot supports the use of eight clusters.
+
+# References for SI (citations will be included in *in situ* table)
+
+`r pander::pander(list("85: @Li:2011es", "118: @Ren:2013jv", "129: @Suzaki:2004ib", 
+    "21: @Chu:2006fw", "82: @Lee:2012hj", "154: @Yoshida:2013ff", "148: @Xue:2008ki", 
+    "3: @Ashikari:2005eg", "150: @Yan:2011hw", "86: @Li:2013iq", "153: @Yoshida:2012fg", 
+    "78: @Kurakawa:2007go", "82: @Lee:2012hj", "83: @Lee:2007cj", "48: @Gao:2010iz", 
+    "59: @IkedaKawakatsu:2012co", "54: @Horigome:2009gt", "83: @Lee:2007cj", 
+    "68: @Jiao:2010ft", "100: @Miura:2010it"))`

### 2015-08-28 src/R/compareVsInSitu.R

@@ -34,6 +34,12 @@ FM <- c("FM")
 tissues <- unique(c(RM, PBM, SBM, SM, FM))
 expZhangGenes <- zhangGenes[geneExp %in% tissues]
 
+############
+### TODO ###
+############
+
+# add genes not in zhang review here
+
 # make a truth table for the zhang genes
 expZhangGenes[,RM := geneExp %in% RM]
 expZhangGenes[,PBM := geneExp %in% PBM]

### 2015-08-28 si.Rmd
switch in situ comparison figure to table

@@
 
-**`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
+``` {r st_libStats, results = "asis"}
+pander::set.alignment('left', row.names = 'right')
+pander::set.caption(paste0(
+  I(pasteLabel("**Table S", s_tableCount, "st_libStats")),
+  "**: Library and sequencing statistics."))
+pander::pandoc.table(st_libStats, split.tables = Inf, emphasize.rownames = FALSE, missing = "")
+```
 
-``` {r st_libStats, results = "markup"}
-knitr::kable(st_libStats, caption = paste0(
-  I(pasteLabel("**Table S", s_tableCount, "st_libStats")),
-    "**: Library and sequencing statistics"))
+```{r st_reviewInSitu, results = "asis"}
+pander::set.alignment('left', row.names = 'right')
+pander::set.caption(paste0(
+  I(pasteLabel("**Table S", s_tableCount, "st_reviewInSitu")),
+  "**: Comparison of the dataset with published *in situ* expression profiles.
+For each tissue, a '1' indicates that the gene was detected by both methods, a '2' indicates that the gene was detected by *in situ* hybridisation but not by RNA sequencing and a '3' indicates that the gene was detected by RNA sequencing but not reported as expressed by *in situ*.
+The last category is not informative because *in situ* analysis for the stage in question may have been omitted from the publication.
+A blank cell indicates that the gene was not detected by either method."))
+pander::pandoc.table(st_reviewInSitu, split.tables = Inf, emphasize.rownames = FALSE,
+                     emphasize.cols = c(1,2), missing = "")
 ```
@@
 
-# References for SI (citations will be included in *in situ* table)
+# References for supporting information
 
-`r pander::pander(list("85: @Li:2011es", "118: @Ren:2013jv", "129: @Suzaki:2004ib", 
-    "21: @Chu:2006fw", "82: @Lee:2012hj", "154: @Yoshida:2013ff", "148: @Xue:2008ki", 
-    "3: @Ashikari:2005eg", "150: @Yan:2011hw", "86: @Li:2013iq", "153: @Yoshida:2012fg", 
-    "78: @Kurakawa:2007go", "82: @Lee:2012hj", "83: @Lee:2007cj", "48: @Gao:2010iz", 
-    "59: @IkedaKawakatsu:2012co", "54: @Horigome:2009gt", "83: @Lee:2007cj", 
-    "68: @Jiao:2010ft", "100: @Miura:2010it"))`
+labels <- c("1" = "Detected\nin both", "2" = "Only detected\nby in situ", "3" = "Only detected by\nRNA-sequencing")

### 2015-08-28 ms.Rmd
re-arranged supp figs and tables. removed nasty in situ figure from ms

--- 2015-08-28 ms.Rmd
@@
 Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
-As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (currently `r I(pasteLabel("Figure ", figCount, "f_reviewInSitu"))`, **move figure to supporting table**).
+As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))`).
 
-```{r f_reviewInSitu, fig.height = 8.661 / 4}
-print(f_reviewInSitu)
-```
-
-**Our in situ results could go here or in SI**
+**Our in situ results could go here (or in SI).**
 
@@
 After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-For example, members of the ABI3VP1 family were overrepresented in clusters **x** and **x**, both of which contain genes with higher transformed read counts in the rachis meristem than in samples from later stages of development.
-The MADS family was overrepresented in clusters **x** and **x**, suggesting that they could be involved in both early and late stages of inflorescence development.
-**run clustering script again and check these statements!!!***
+Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 and G2-like TFs.
+MADS TFs and TRAF genes were overreprsented among genes that have low expression in the SM and PBM but higher expression in the RM (cluster `r getClusterName('small', 'inc')`) and bHLH, SBP and C2C2-GATA TFs were overreprented among genes with the inverse pattern (cluster `r getClusterName('small', 'dec')`)
 
+**Discussion: what might these patterns mean biologically?**
 **Discussion: which clusters have genes that are already candidates for panicle branching?**
@@
 In general, the GSEA results correspond to the hypergeometric tests for transcription factor enrichment in the clusters.
-For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster **x**, which contains genes with higher expression in those meristem types.
+For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster `r getClusterName('small', 'dec')`, which contains genes with higher expression in those meristem types.
 
@@
 
+**`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
+
+**`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))`**: Comparison of the dataset with published *in situ* expression profiles.
+
 **`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
-
-**`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
 

### 2015-08-31 ms.Rmd
download Peking Uni TFDB for Tomato MADS genes

@@
 
-To recover common expression patterns in genes detected in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
+To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
 Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
@@
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database (**reference**), a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
+Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database (**reference for http://plntfdb.bio.uni-potsdam.de/v3.0/**), a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
 After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
@@
 * **Figure**: Similarity tree of MADS proteins with L2FC values between comparable samples (and dN/dS values?) for each one:
-+ **rice**: rachis meristem vs. spikelet merisem
-+ **arapidospis**: inflorescence meristem vs. floret meristem
-+ **tomato**: simpodial meristem vs. floret meristem (need to find list of tomato MADS genes)
++ **rice**: rachis meristem vs. spikelet merisem.
++ **arapidospis**: inflorescence meristem vs. floret meristem.
++ **tomato**: simpodial meristem vs. floret meristem. (Need to find list of tomato MADS genes. List of ~32 MIKC domain proteins on the other plant TFDB, http://planttfdb.cbi.pku.edu.cn/)
 + (could I map the l2fc to the background of the geom_label rectangle? "The fill aesthetic controls the background colour of the label.")
@@
 The `'quantMode'` argument of `STAR` was used for read counting.
-rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with [`wgsim`](https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
+rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
 Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
-To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf].
+To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculathged for each gene [@Li:2010cl; @Wagner:2012jf].
 These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values.
@@
 **`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`**.
-Common patterns of expression in the meristem samples.
+Common patterns of gene expression in developing inflorescences.
 Fuzzy *c*-means clustering of normalised, variance-stabilised read counts was used to recover `r as.character(english::english(c))` common patterns of expression.
@@
 The core values for each cluster are plotted in black.
+The ordering of the panels in the plot was rearranged to enable side-by-side comparison of complementary clusters.
 
 **`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**.
-Geneset enrichment analysis (GSEA) for transcription factor families.
-Test statistics were calculated on a matrix of Log~2~(fold change) values for expressed TF family members between each sample and the other three samples.
+Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families.
+Test statistics were calculated on a matrix of Log~2~(fold change) values between each sample and the other three samples for expressed family members.
 A positive test statistic indicates enrichment and a negative statistic indicates depletion.

### 2015-08-31 src/R/tomatoTFDB.R

@@ -1,0 +1,11 @@
+#!/usr/bin/Rscript
+
+temp <- tempfile()
+download.file("http://planttfdb.cbi.pku.edu.cn/download/gene_model_family/Sly",
+              temp)
+
+puTfdb <- read.table(temp, header = TRUE, stringsAsFactors = FALSE)
+
+saveRDS(puTfdb, 'data/tfdb/puTfdbSl.Rds')
+
+quit(save = "no", status = 0)

### 2015-08-31 src/R/tomatoTFDB.R
use phytozome for MADS IDs & peptides

@@ -1,11 +1,0 @@
-#!/usr/bin/Rscript
-
-temp <- tempfile()
-download.file("http://planttfdb.cbi.pku.edu.cn/download/gene_model_family/Sly",
-              temp)
-
-puTfdb <- read.table(temp, header = TRUE, stringsAsFactors = FALSE)
-
-saveRDS(puTfdb, 'data/tfdb/puTfdbSl.Rds')
-
-quit(save = "no", status = 0)

### 2015-08-31 src/py/pipeline.py

@@ -208,15 +208,15 @@ def download_tfdb(outputFiles):
     touch(outputFiles)
 
 #---------------------------------------------------------------
-# download Peking Uni TFDB
+# download MADS peptides
 #
-@originate(['ruffus/puTfdb.data'])
+@originate(['ruffus/madsPeptides.data'])
 
-def download_pu_tfdb(outputFiles):
-    jobScript = 'src/R/tomatoTFDB.R'
+def get_mads_peptides(outputFiles):
+    jobScript = 'src/R/getMADSsequences.R'
     ntasks = '1'
     cpus_per_task = '1'
-    job_name = 'puTFDB'
+    job_name = 'madsPept'
     jobId = submit_job(jobScript, ntasks, cpus_per_task, job_name)
     # update ruffus flag
     print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")

### 2015-09-01 ms.Rmd
use phytozome for MADS IDs & peptides
output gene expression truth table from calculateCutoffs

@@
 
-Authors:
+Thomas W.R. Harrop^§^[^1], Israr Ud Din^¶^[^1], Veronica Gregis^¶^, Michela Osnato^¶^, Stefan Jouannic^§^, Hélène Adam^§^ and Martin Kater^¶^
 
-Addresses:
+[^1]: These authors contributed equally to this work.
+
+| ^§^ Institut de Recherche pour le Développement (IRD), 911 Avenue Agropolis, 34394 Montpellier, France.
+| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
 
@@
 
-Word count
+## Word count
 
@@
 
-**Discussion: what might these patterns mean biologically?**
-**Discussion: which clusters have genes that are already candidates for panicle branching?**
+|**Discussion: what might these patterns mean biologically?**
+|**Discussion: which clusters have genes that are already candidates for panicle branching?**
 
@@
 
+### Comparative analysis of MADS family
+
+The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface (**refs: PANTHER/BIOMART**).
+
 # Accession Numbers

### 2015-09-01 src/R/figures.R

@@ -79,6 +79,16 @@ setcolorder(st_reviewInSitu, neworder =
 setkey(st_reviewInSitu, "Gene symbol")
 s_tableCount <- incCount(s_tableCount, "st_reviewInSitu")
 
+# tpm plots of some/all of these genes
+tpm <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
+setnames(tpm, 'rn', "MSU identifier")
+setkey(tpm, "MSU identifier")
+
+genes <- st_reviewInSitu[,.(`MSU identifier`, `Gene symbol`)]
+setkey(genes, "MSU identifier")
+
+plotData <- tpm[genes]
+
 ##################
 ### LMD FIGURE ###
 ##################

### 2015-09-01 si.Rmd
re-arranged supp figs and tables. removed nasty in situ figure from ms
started code for sf_isGenesTpm

@@
 
+```{r sf_isGenesTpm, fig.height = 14}
+print(sf_isGenesTpm)
+```
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Expression patterns of genes compared to *in situ* results.
+
 ```{r sf_mfuzzCentroids}
@@
 # References for supporting information
-
-labels <- c("1" = "Detected\nin both", "2" = "Only detected\nby in situ", "3" = "Only detected by\nRNA-sequencing")

### 2015-09-01 src/R/figures.R

@@ -86,8 +86,46 @@ setkey(tpm, "MSU identifier")
 
 genes <- st_reviewInSitu[,.(`MSU identifier`, `Gene symbol`)]
 setkey(genes, "MSU identifier")
+genes <- unique(genes)
+genTpm <- tpm[genes]
+genTpm.long <- reshape2::melt(genTpm, id.vars = c('MSU identifier', 'Gene symbol'),
+                                variable.name = "lib", value.name = "Expression (TPM)")
 
-plotData <- tpm[genes]
+# expression calls for these genes by library
+expGenTT <- data.table(readRDS('output/expressedGenes/expGenTT.Rds'), key = "id")
+expGenTT.wide <- expGenTT[genTpm[,`MSU identifier`]]
+expGenTT.long <- reshape2::melt(expGenTT.wide, id.vars = "id",
+                                variable.name = "lib", value.name = "expressed")
+
+setkey(genTpm.long, "MSU identifier", "lib")
+setkey(expGenTT.long, "id", "lib")
+
+plotData <- expGenTT.long[genTpm.long]
+
+# add stage names from deseq2
+colData <- data.table(as.data.frame(GenomicRanges::colData(
+  readRDS('output/DESeq2/ddsLrt.Rds'))), keep.rownames = TRUE)
+plotData[, stage := colData[rn == as.character(lib), stage]]
+
+# add replicate number to colour points
+plotData[, replicate := sub(".*(\\d+)", "\\1", lib)]
+
+# format gene name for the plot
+plotData[, name:= oryzr::LocToGeneName(id)$symbols, by = id]
+plotData[!is.na(name), label := paste(name, "·", id)]
+plotData[is.na(name), label := id]
+
+# draw it
+sf_isGenesTpm <- ggplot(plotData, aes(x = stage, y = `Expression (TPM)`, colour = replicate, shape = expressed)) +
+  theme_minimal(base_size = 8, base_family = "Helvetica") +
+  theme(strip.text = element_text(face = "italic")) +
+  xlab(NULL) +
+  scale_color_brewer(palette = "Set1") + 
+  stat_smooth(aes(group = label), method = "loess", se = FALSE, colour = "grey") + 
+  geom_point() +
+  facet_wrap(~ label, scales = "free_y", ncol = 3)
+#sf_isGenesTpm
+figCount <- incCount(s_figCount, "sf_isGenesTpm")
 
 ##################
 ### LMD FIGURE ###

### 2015-09-02 ms.Rmd
try changing table 1 headers

@@
 output:
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
   BiocStyle::html_document:
@@
     dev: "CairoPDF"
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
   pdf_document:
@@
                warning=FALSE)
+
+pander::panderOptions("table.caption.prefix", "")
 
@@
 To investigate gene expression during inflorescence development, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
-RNA from the meristem samples was isolated, amplified and used for the production of a sequencing library.
+RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
 **Need area cut +/- number of panicles used for each library from Israr**.
 RNA input for library preparation was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
-Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values two libraries that were made from degraded RNA were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
+Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
 Following this, further samples were prepared to provide a total of three biological replicates with intact RNA (RIN ≥ 7) for each meristem type.
@@
 Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
-As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))`).
+As a preliminary quality check, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`).
 
@@
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database (**reference for http://plntfdb.bio.uni-potsdam.de/v3.0/**), a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
+Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
 After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
@@
 japonica cv.
-Nipponbare plants were grown in a growth chamber at 70% relative humidity at 30°C during the day and 26°C at night.
+Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night.
 After 7 weeks in long day conditions (LD; 14 hours light and 10 hours darkness) plants were transferred to short day conditions (SD; 10 hours light and 14 hours darkness) to induce floral transition.
@@
 
-Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped with the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode .
+Briefly, adaptor trimming was performed with `cutadapt` [@Martin:2011eu] and reads were mapped against the MSUv7 genome and annotation [@Kawahara:2013gh; @Ouyang:2007il] downloaded from Phytozome 10.3 [@Goodstein:2012ho] using `STAR` [@Dobin:2012fg] in 2-pass mode.
 The `'quantMode'` argument of `STAR` was used for read counting.
@@
 Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
-To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculathged for each gene [@Li:2010cl; @Wagner:2012jf].
+To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf].
 These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values.
@@
 
+**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Quantitative expression values for the genes that were compared to published *in situ* results.
+
 **`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
@@
 ```{r pander, results = "asis"}
-pander::set.alignment('left', row.names = 'right')
+pander::set.alignment('center', row.names = 'right')
 pander::set.caption(paste0(
   I(pasteLabel("**Table ", tableCount, "t_hypergeom")),
-  "**: Transcription factor and transcriptional regulator family enrichment per cluster, including the number of family members (*n*), number expressed in the whole dataset (*n*~expr~) and number and adjusted *p*-value for each cluster, for families that had at least two members in any one cluster."))
+  "**: Transcription factor and transcriptional regulator family enrichment for families that had at least two members in any one cluster, including the number of family members (*n*), number expressed in the whole dataset (*n*~expr~) and number and adjusted *p*-value for each cluster."))
 pander::pandoc.table(t_hypergeom, split.tables = Inf, emphasize.rownames = FALSE)
@@
 **`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`**.
-LASER microdissection of meristem samples in the early stages of inflorescences development.
+LASER microdissection of meristem samples in the early stages of inflorescence development.
 Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and secondary branch meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
@@
 Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families.
-Test statistics were calculated on a matrix of Log~2~(fold change) values between each sample and the other three samples for expressed family members.
+Test statistics were calculated on a matrix of Log~2~ fold change values between each sample and the other three samples for expressed family members.
 A positive test statistic indicates enrichment and a negative statistic indicates depletion.

### 2015-09-02 si.Rmd
try changing table 1 headers

--- 2015-08-28 si.Rmd
@@
     dev: "CairoPDF"
-  word_document:
-    fig_caption: yes
   pdf_document:
@@
     number_sections: yes
+    dev: "CairoPDF"
+  word_document:
+    fig_caption: yes
 bibliography: bib/papersLib.bib
@@
                fig.show='hold',
+               fig.height = 3.150,
+               fig.width = 3.150,
                cache=FALSE,
@@
 
+pander::panderOptions("table.caption.prefix", "")
+
 source('src/R/figures.R')
@@
 
-**`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`**: Library and sequencing statistics.
+``` {r st_libStats, results = "asis"}
+pander::set.caption(paste0(
+  I(pasteLabel("**Table S", s_tableCount, "st_libStats")),
+  "**: Library and sequencing statistics."))
+pander::pandoc.table(st_libStats, split.tables = Inf, missing = "", round = 1, emphasize.strong.cols = 1,
+                     justify = c("right", rep("center", length(names(st_libStats)) - 1)))
+```
 
-``` {r st_libStats, results = "markup"}
-knitr::kable(st_libStats, caption = paste0(
-  I(pasteLabel("**Table S", s_tableCount, "st_libStats")),
-    "**: Library and sequencing statistics"))
+```{r st_reviewInSitu, results = "asis"}
+pander::set.caption(paste0(
+  I(pasteLabel("**Table S", s_tableCount, "st_reviewInSitu")),
+  "**: Comparison of the dataset with published *in situ* expression profiles.
+'×' indicates that the gene was detected by both methods, '+' indicates that the gene was detected by *in situ* hybridisation but not by RNA sequencing and '·' indicates that the gene was detected by RNA sequencing but expression was not reported by *in situ*.
+The last category is not intended to be informative, because it relies on negative or unreported results.
+A blank cell indicates that the gene was not detected by either method."))
+pander::pandoc.table(st_reviewInSitu, split.tables = Inf,
+                     justify = c("right", rep("centre", length(names(st_reviewInSitu)) - 2), "left"),
+                     emphasize.cols = c(1,2), missing = "")
 ```
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
-The two libraries made from degraded RNA did not cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA had an effect on the quantification of read counts.
+The two libraries made from degraded RNA did not cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA affected quantification of expression.
 For this reason these two libraries were not included in subsequent analyses.
+
+```{r sf_isGenesTpm, fig.height = 8.661, fig.width = 6.614}
+print(sf_isGenesTpm)
+```
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Quantitative expression values for the genes that were compared to published *in situ* results.
+The expression value in transcripts per million is shown on the *x*-axis.
+Points are coloured by replicate, and plotted as a triangle if the gene was above the expression cutoff or a circle otherwise.
+Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).
 
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
-Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after $c = 8$.
+Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after *c* = 8.
 
@@
 
-# References for SI (citations will be included in *in situ* table)
-
-`r pander::pander(list("85: @Li:2011es", "118: @Ren:2013jv", "129: @Suzaki:2004ib", 
-    "21: @Chu:2006fw", "82: @Lee:2012hj", "154: @Yoshida:2013ff", "148: @Xue:2008ki", 
-    "3: @Ashikari:2005eg", "150: @Yan:2011hw", "86: @Li:2013iq", "153: @Yoshida:2012fg", 
-    "78: @Kurakawa:2007go", "82: @Lee:2012hj", "83: @Lee:2007cj", "48: @Gao:2010iz", 
-    "59: @IkedaKawakatsu:2012co", "54: @Horigome:2009gt", "83: @Lee:2007cj", 
-    "68: @Jiao:2010ft", "100: @Miura:2010it"))`
+# References for supporting information

### 2015-09-02 src/R/figures.R

@@ -47,102 +47,6 @@ capwords <- function(s, strict = FALSE) {
 st_libStats <- readRDS('output/quantStats/libStats.Rds')
 s_tableCount <- incCount(s_tableCount, "st_libStats")
 
-#######################
-### COMPARE IN SITU ###
-#######################
-
-compare <- readRDS('output/compare/compare.Rds')
-setkey(compare, "zhangRef")
-
-# convert zhangRef to citekey (these papers were mined from zhang paper manually)
-citekeys <- structure(c("@Ikeda:2007ja", "@Li:2011es", "@Ren:2013jv", "@Suzaki:2004ib", 
-                        "@Chu:2006fw", "@Lee:2012hj", "@Yoshida:2013ff", "@Xue:2008ki", "@Ashikari:2005eg", 
-                        "@Yan:2011hw", "@Li:2013iq", "@Yoshida:2012fg", "@Kurakawa:2007go", 
-                        "@Lee:2012hj", "@Lee:2007cj", "@Gao:2010iz", "@IkedaKawakatsu:2012co", 
-                        "@Horigome:2009gt", "@Lee:2007cj", "@Jiao:2010ft", "@Miura:2010it", 
-                        "**zhang55**", "**zhang73**", "@Komatsu:2003iu", "@Li:2010ks", "**zhang108**", 
-                        "**zhang131**", "@Zhu:2010je", "**m.Lopez-Dee1999**", "**m.Yadav2007**", 
-                        "**m.Nagasawa2003**", "**m.Yun2013**", "**m.Duan2012**", "**m.Ohmori2009**", 
-                        "**m.Dreni2007**", "**m.Cui2010**", "**m.Fornara2004**"),
-                      .Names = c("56", "85", "118", "129", "21", "82", "154", "148", "3", "150", "86", "153", 
-                                 "78", "82", "83", "48", "59", "54", "83", "68", "100", "55", "73", 
-                                 "75", "84", "108", "131", "165", "991", "992", "993", "994", "995", 
-                                 "996", "997", "998", "999"))
-compare[, Reference := citekeys[as.character(zhangRef)]]
-
-# SI table
-st_reviewInSitu <- data.table(reshape2::dcast(compare, msuId + Reference ~ stage,
-                                              value.var = 'compare'))
-
[…]

### 2015-09-02 src/R/tfHypergeom.R

@@ -71,12 +71,12 @@ resultsLong$padj <- p.adjust(resultsLong$`p-value`, method = "BH")
 padj <- reshape2::dcast(resultsLong, Family ~ Cluster, value.var = 'padj')
 rownames(padj) <- padj$Family
 padj$Family <- NULL
-colnames(padj) <- paste0("*p*~adj,\\ ", colnames(padj), "~")
+colnames(padj) <- paste0("C~", sub(".*(\\d+).*", "\\1", colnames(padj)),"~\n*p*~adj~")
 
 Number <- reshape2::dcast(resultsLong, Family ~ Cluster, value.var = 'Number')
 rownames(Number) <- Number$Family
 Number$Family <- NULL
-colnames(Number) <- paste0("*n*~", colnames(Number), "~")
+colnames(Number) <- paste0("C~", sub(".*(\\d+).*", "\\1", colnames(Number)),"~\n*n*")
 
 # remove columns that don't have p-values
 keep <- apply(padj, 1, function(x) !all(is.na(x)))

### 2015-09-02 src/py/pipeline.py

@@ -558,11 +558,15 @@ def f_gsea(inputFiles, outputFiles):
     touch(outputFiles)
 
 #---------------------------------------------------------------
-# in situ comparison plot
+# in situ comparisons
 #
 @merge(compare_inSitus, "ruffus/table.st_reviewInSitu")
 def st_reviewInSitu(inputFiles, outputFiles):
     touch(outputFiles)
+@merge([compare_inSitus, calculate_tpm, run_deseq2_os], "ruffus/figure.sf_isGenesTpm")
+def sf_isGenesTpm(inputFiles, outputFiles):
+    touch(outputFiles)
+
 
 
 

### 2015-09-04 ms.Rmd
how many genes per cluster for each family?

@@
     number_sections: yes
-bibliography: bib/papersLib.bib
+bibliography: bib/papersLib.bibtex
 ---
@@
 *Oryza sativa* ssp.
-japonica cv.
+*japonica* cv.
 Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night.
@@
 The `'quantMode'` argument of `STAR` was used for read counting.
-rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
+rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
 Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
@@
 
-The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface (**refs: PANTHER/BIOMART**).
+The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db, @Durinck:2009ki].
 

### 2015-09-04 si.Rmd
how many genes per cluster for each family?

@@
     fig_caption: yes
-bibliography: bib/papersLib.bib
+bibliography: bib/papersLib.bibtex
 ---

### 2015-09-04 src/R/familyGenesPerCluster.R

@@ -1,0 +1,51 @@
+#!/usr/bin/Rscript
+
+library(data.table)
+
+# find tfdb
+tfdbFile <- 'data/tfdb/os.Rds'
+if (!file.exists(tfdbFile)) {
+  cat("tfdb file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+tfdb <- readRDS(tfdbFile)
+
+# check for mfuzz output
+clusterFile <- "output/mfuzz/clusters.Rds"
+if (!file.exists(clusterFile)) {
+  cat("clusterFile not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+clusters <- readRDS(clusterFile)
+
+# find expressed genes
+expGenFile <- "output/expressedGenes/expressedGenesAll.Rds"
+if (!file.exists(expGenFile)) {
+  cat("tfdb file not found, exiting\n", file = stderr())
+  quit(save = "no", status = 1)
+}
+expressedGenesAll <- readRDS(expGenFile)
+
+# think we need a long table of cluster vs gene vs family
+clusters.wide <- data.table(msuId = unique(unlist(sapply(clusters, rownames))))
+n <- paste("Cluster", 1:length(clusters))
+clusters.wide[, (n) := lapply(clusters, function(x) msuId %in% rownames(x)), by = msuId]
+clusters.wide[msuId %in% tfdb$Protein.ID, Family := tfdb[Protein.ID == msuId, Family], by = msuId]
+clusters.long <- reshape2::melt(clusters.wide, id.vars = c("msuId", "Family"), variable.name = "Cluster")
+
+
+reshape2::dcast(clusters.long[!is.na(Family) & value], msuId + Family ~ Cluster, fun.aggregate = sum, value.var = "value")
+
+
+# cluster table
+genesPerCluster <- tfdb[,.("*n*" = length(Protein.ID),
+        "*n*~exp~" = sum(Protein.ID %in% expressedGenesAll)), by = Family]  
+
+paste("Cluster", 1:length(clusters))
+genesPerCluster
+
+
+x <- 'ABI3VP1'
+tfdb[Family == x, sum(Protein.ID %in% expressedGenesAll), by = Family]
+
+

### 2015-09-04 src/R/figures.R

@@ -101,12 +101,13 @@ citekeys <- structure(c("@Ikeda:2007ja", "@Li:2011es", "@Ren:2013jv", "@Suzaki:2
                         "@Ashikari:2005eg", "@Yan:2011hw", "@Li:2013iq", "@Yoshida:2012fg", 
                         "@Kurakawa:2007go", "@Lee:2012hj", "@Lee:2007cj", "@Gao:2010iz", 
                         "@IkedaKawakatsu:2012co", "@Horigome:2009gt", "@Lee:2007cj", 
-                        "@Jiao:2010ft", "@Miura:2010it", "**zhang55**", "**zhang73**", 
-                        "@Komatsu:2003iu", "@Li:2010ks", "**zhang108**", "**zhang131**", 
-                        "@Zhu:2010je", "**m.Lopez-Dee1999**", "**m.Yadav2007**", "**m.Nagasawa2003**", 
-                        "**m.Yun2013**", "**m.Duan2012**", "**m.Ohmori2009**", "**m.Dreni2007**", 
-                        "**m.Cui2010**", "**m.Fornara2004**", "**m.Komatsu2009**"),
-                      .Names = c("56","85", "118", "129", "21", "82", "154", "148", "3", "150", "86", 
+                        "@Jiao:2010ft", "@Miura:2010it", "@Huang:2009kf", "@Kobayashi:2010hn", 
+                        "@Komatsu:2003iu", "@Li:2010ks", "@Oikawa:2009cs", "@Tabuchi:2011gx", 
+                        "@Zhu:2010je", "@LopezDee:1999ds", "@Yadav:2007im", "@Nagasawa:2003gu", 
+                        "@Yun:2013id", "@Duan:2012bt", "@Ohmori:2009kj", "@Dreni:2007jd", 
+                        "@Cui:2010kt", "@Fornara:2004bi", "**m.Komatsu2009**"),
+                      .Names = c("56", 
+                                 "85", "118", "129", "21", "82", "154", "148", "3", "150", "86", 
                                  "153", "78", "82", "83", "48", "59", "54", "83", "68", "100", 
                                  "55", "73", "75", "84", "108", "131", "165", "991", "992", "993", 
                                  "994", "995", "996", "997", "998", "999", "990"))
@@ -142,7 +143,7 @@ setkey(genes, "MSU identifier")
 genes <- unique(genes)
 genTpm <- tpm[genes]
 genTpm.long <- reshape2::melt(genTpm, id.vars = c('MSU identifier', 'Gene symbol'),
[…]

### 2015-09-04 ms.Rmd
added second cluster table

@@
 
-```{r pander, results = "asis"}
+```{r t_familiesPerCluster, results = "asis"}
+
+pander::set.caption(paste0(
+  I(pasteLabel("**Table ", tableCount, "t_familiesPerCluster")),
+  "**: Transcription factor and transcriptional regulator family members in each cluster.
+*n*, total number of family members, *n*~expr~, number of family members expressed in the whole dataset, C~1~–C~", c, "~, number of family members in each cluster, and the total number of family members in all clusters."))
+t_familiesPerCluster <- as.data.frame(t_familiesPerCluster[!Total == 0])
+rownames(t_familiesPerCluster) <- t_familiesPerCluster$Family
+t_familiesPerCluster$Family <- NULL
+pander::pandoc.table(t_familiesPerCluster, split.tables = Inf, emphasize.rownames = FALSE,
+                     justify = c("right", rep("centre", length(names(t_familiesPerCluster)) - 1), "left"))
+
+```
+
+```{r t_hypergeom, results = "asis"}
 pander::set.alignment('center', row.names = 'right')

### 2015-09-04 src/R/tfHypergeom.R
fixed figure legends and results for hypergeom fixes #1

@@ -64,7 +64,7 @@ familyClusters <- data.table(reshape2::dcast(familyStats[clusterStats],
                                              Fam + n + nexpr ~ Cluster,
                                              value.var = "nclust"), key = "Fam")
 cc <- grep("^C\\d+", names(familyClusters), value = TRUE)
-familyClusters[, total := rowSums(.SD), .SDcols = nc]
+familyClusters[, total := rowSums(.SD), .SDcols = cc]
 
 # combine the results for adjusting p-values
 resultsLong <- merge(reshape2::melt(inClusterPerFamily,

### 2015-09-04 ms.Rmd
typo in hypergeom code

@@
 Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
-In general, the GSEA results correspond to the hypergeometric tests for transcription factor enrichment in the clusters.
-For example, SBP transcription factors are enriched in RM and PBM, consistent with their overrepresentation in cluster `r getClusterName('small', 'dec')`, which contains genes with higher expression in those meristem types.
+Unlike soft clustering, which groups genes based on shared expression dynamics, this GSEA was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in the expression of their members in each stage.
+These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
+Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
+The most pronounced change in expression appears to occur between the RM and branch meristem samples.
+Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
+Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
 
@@
 ```
-
-These patterns are not intended to represent each member of each gene family.
-For example, there are several MADS genes that have lower expression in the SM than in the RM, but expressed MADS genes were detected at a higher level overall in SM.
-Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
-In the GSEA, the most pronounced change in expression appears to occur between the RM and branch meristem samples.
-Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
-Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
 
@@
 
-Analysis of the sequenced libraries was carried out with custom `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp].
+Analysis of the sequenced libraries was carried out with `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp].
 All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.
@@
 rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
-Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
+Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
 To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf].
@@
 Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families.
-Test statistics were calculated on a matrix of Log~2~ fold change values between each sample and the other three samples for expressed family members.
-A positive test statistic indicates enrichment and a negative statistic indicates depletion.
+Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset.
+The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.
 Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.

### 2015-09-04 ms.Rmd
Added t_familiesPerCluster but decided to hide it because it's basically the same as t_hypergeom

@@
 
-```{r t_familiesPerCluster, results = "asis"}
-
+```{r t_familiesPerCluster, results = "hide"}
 pander::set.caption(paste0(
@@
                      justify = c("right", rep("centre", length(names(t_familiesPerCluster)) - 1), "left"))
-
 ```
@@
 for (x in git){cat(x, "\n")}
-
 ```

### 2015-09-04 ms.Rmd
Sent draft

@@
     highlight: pygments
-    keep_md: yes
     number_sections: yes
     dev: "CairoPDF"
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
   pdf_document:
     fig_caption: yes
-    keep_tex: yes
     number_sections: yes
@@
 
-## SRA details for libaries
+## SRA details for libraries
 
@@
 
-* Rice inflorescence development
-+ Branching & grain yield
-+ Sequential changes in meristem activity
-+ Discuss a few key genes
+## Rice inflorescence development
+### Branching & grain yield
+### Sequential changes in meristem activity
+### Discuss a few key genes?
 
@@
 Following this, further samples were prepared to provide a total of three biological replicates with intact RNA (RIN ≥ 7) for each meristem type.
-
-![](data/lmdFigure.pdf)
 
@@
 
-```{r f_mfuzzClusters, fig.height = 8.661 / 2}
-print(f_mfuzzClusters)
-```
-
 Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**).
@@
 
-|**Discussion: what might these patterns mean biologically?**
-|**Discussion: which clusters have genes that are already candidates for panicle branching?**
+**Discussion: what might these patterns mean biologically? Which clusters have genes that are already candidates for panicle branching?**
 
@@
 Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
-Unlike soft clustering, which groups genes based on shared expression dynamics, this GSEA was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in the expression of their members in each stage.
+Unlike soft clustering, which groups genes based on shared expression dynamics, this GSEA was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in expression of their members in each stage.
 These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
@@
 Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
-
-```{r f_gsea, fig.height = 8.661 / 2}
-print(f_gsea)
-```
 
@@
 
-To investigate the evolution of the MADS family in Angiosperms, their expression values in the dicots *A.
+To investigate the evolution of the MADS family in Angiosperms, their expression profiles in the dicots *A.
 thaliana* and *Solanum lycopersicum* and the monocot *O.
@@
 
-* **Figure**: Similarity tree of MADS proteins with L2FC values between comparable samples (and dN/dS values?) for each one:
-+ **rice**: rachis meristem vs. spikelet merisem.
-+ **arapidospis**: inflorescence meristem vs. floret meristem.
-+ **tomato**: simpodial meristem vs. floret meristem. (Need to find list of tomato MADS genes. List of ~32 MIKC domain proteins on the other plant TFDB, http://planttfdb.cbi.pku.edu.cn/)
-+ (could I map the l2fc to the background of the geom_label rectangle? "The fill aesthetic controls the background colour of the label.")
+| **Figure**: Similarity tree of MADS proteins with L2FC values between comparable samples (and dN/dS values?) for each one:
+| **rice**: rachis meristem vs. spikelet merisem.
+| **arapidospis**: inflorescence meristem vs. floret meristem.
+| **tomato**: simpodial meristem vs. floret meristem.
 
@@
 
-The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db, @Durinck:2009ki].
+The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
 
@@
 
+----
+
+![](data/lmdFigure.pdf)
+
 **`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`**.
@@
 Images show the samples before (**a**–**d**) and after (**e**–**h**) dissection.
+
+----
+
+```{r f_mfuzzClusters, fig.height = 8.661 / 2}
+print(f_mfuzzClusters)
+```
 
@@
 
+----
+
+```{r f_gsea, fig.height = 8.661 / 2}
+print(f_gsea)
+```
+
 **`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**.
@@
 
-# `R sessionInfo()` (*hide*)
+----
+
+# `sessionInfo` (*hide*)
 
@@
           paste("**git hash:**", system("git rev-parse HEAD", intern = TRUE)))
-
+cat("\n")
 for (x in git){cat(x, "\n")}
@@
 
-# References (move to before *Tables*)
+# References (move to before **Tables**)

### 2015-09-04 si.Rmd
Sent draft

--- 2015-09-02 si.Rmd
@@
     highlight: pygments
-    keep_md: yes
     number_sections: yes
@@
     fig_caption: yes
-    keep_tex: yes
     number_sections: yes
@@
 
+----
+
 ```{r st_reviewInSitu, results = "asis"}
@@
 
+----
+
 ```{r sf_pca}
@@
 For this reason these two libraries were not included in subsequent analyses.
+
+----
 
@@
 
+----
+
 ```{r sf_mfuzzCentroids}
@@
 Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after *c* = 8.
+
+----
 
@@
 
+----
+
 # References for supporting information

### 2015-09-14 ms.Rmd
Cleaned up script for selection by clade

@@
 The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
+An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix (**ref clustalo**).
+To clean the alignment, poorly conserved regions where greater than `minpcnongap` percent of sequences had gaps or less than `minpcident` perc ent of pairwise comparisons between sequences had identical residues were removed.
+To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server (**ref for AGRIS**), with as few type I proteins as possible.
+Protein sequences in this clade that were longer than `minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above, used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) and rooted to a type I MADS sequence (`og`).
+The tree was plotted using the `ggtree` extension for `ggplot2` (**ref ggtree**).
 

### 2015-09-14 src/R/atTfdb.R

@@ -11,7 +11,6 @@ setnames(atTfdb, old = names(atTfdb),
          new = c("FamilyID", "LocusName", "GeneName", "Description",
                  "RNIntegration", "SubFamily", "BindingSite", "SpecialInfo"))
 
-
 # MAKE OUTPUT FOLDER
 outDir <- "data/attfdb"
 if (!dir.exists(outDir)) {

### 2015-09-14 ms.Rmd
Added compMadsTree figure to ms Found some errors in figures.R

@@
 
-There was an overrepresentation of MADS genes in clusters **xyz** and their overall expression is enriched in the spikelet meristem sample.
+There was an overrepresentation of MADS genes in cluster `r getClusterName('small', 'inc')` and their overall expression is enriched in the spikelet meristem sample.
 These results correspond with previous studies that have established the importance of MADS genes in inflorescence development in rice and other plants (**expand**).
 
-To investigate the evolution of the MADS family in Angiosperms, their expression profiles in the dicots *A.
-thaliana* and *Solanum lycopersicum* and the monocot *O.
-sativa* were compared to their homology in these species.
+The change in expression of MIKC proteins between undifferentiated and differentiated meristems was extracted from the dataset reported here and two other studies that used LMD and RNA-sequencing in early stages of inflorescence development in Arabidopsis and tomato [@Mantegazza:2014di; @Park:2012dg].
+To investigate the evolution of the MADS family in Angiosperms, these changes were compared to homology of MIKC proteins in the three species (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
 
@@
 An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix (**ref clustalo**).
-To clean the alignment, poorly conserved regions where greater than `minpcnongap` percent of sequences had gaps or less than `minpcident` perc ent of pairwise comparisons between sequences had identical residues were removed.
+To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed.
 To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server (**ref for AGRIS**), with as few type I proteins as possible.
-Protein sequences in this clade that were longer than `minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above, used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) and rooted to a type I MADS sequence (`og`).
+Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) rooted to a type I MADS sequence (`r og`).
+L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
 The tree was plotted using the `ggtree` extension for `ggplot2` (**ref ggtree**).
@@
 
+```{r f_madsTree}
+print(f_madsTree)
+```
+
+**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**.
+Neighbour-joining similarity tree of putatative MIKC proteins in *A.
+thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp.
+*japonica*.
+Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, and other proteins are labelled with their unique identifiers.
+The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
+A white background indicates that no results were returned by `DESeq2` for the gene in question, *e.g.* because of low read counts or an extreme count outlier.
+For rice, L~2~FC was estimated from the RM and SM samples reported here.
+For *A.
+thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di.
+The simpodial meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.
+Clades are labelled following the nomenclature presented by (**ref: Arora paper**).
+
+----
+
 # `sessionInfo` (*hide*)

### 2015-09-14 src/R/figures.R

@@ -2,7 +2,6 @@
 
 library(data.table)
 library(ggplot2)
-library(ggtree)
 extrafont::loadfonts()
 
 #####################
@@ -302,8 +301,13 @@ getClusterName <- function(deln1n2, deln2n4){
 }
 
 # centroid dis vs. c (for SI)
-sf_mfuzzCentroids <- readRDS('output/mfuzz/centPlot.Rds') +
-  theme_minimal(base_size = 8, base_family = "Helvetica")
+centroids <- readRDS('output/mfuzz/centroids.Rds')
+sf_mfuzzCentroids <- ggplot(centroids, aes(x = x, y = y)) +
+  theme_minimal(base_size = 8, base_family = "Helvetica") +
+  xlab(expression(Cluster~number~"("*italic(c)*")")) +
+  ylab("Minimum centroid distance") +
+  stat_smooth(method = loess, se = FALSE) +
+  geom_point()
 
 s_figCount <- incCount(s_figCount, "sf_mfuzzCentroids")
 
@@ -380,14 +384,18 @@ f_gsea <- ggplot(gsea, aes(x = Stage, y = rn, label = padj, fill = `Test\nstatis
   scale_x_discrete(expand = c(0,0)) +
   scale_fill_gradientn(colours = heatscale) +
   facet_grid(Category ~ ., scales = "free_y", space = "free_y") +
-  geom_raster() +
-  geom_text(data = gsea[showPval == TRUE], size = 2)
+  geom_raster()
+if (gsea[,any(showPval)]){
+  f_gsea <- f_gsea + geom_text(data = gsea[which(showPval),], size = 2)
+}
 
 figCount <- incCount(figCount, "f_gsea")
 
 #################
 ### MADS TREE ###
 #################
+
+library(ggtree)
 
 njTree <- readRDS('output/madsComp/clustal/njTree.Rds')
 madsPeptides <- readRDS('output/madsComp/clustal/madsPeptides.Rds')
@@ -396,26 +404,82 @@ minpcident <- readRDS('output/madsComp/clustal/minpcident.Rds')
 minProtLength <- readRDS('output/madsComp/clustal/minProtLength.Rds')
 og <- readRDS('output/madsComp/clustal/og.Rds')
 
-# draw a tree (move to figures)
-heatscale <- rev(RColorBrewer::brewer.pal(5, "PuOr"))
-f_madsTree <- ggplot(njTree, aes(x = x, y = y, label = label)) +
[…]

### 2015-09-16 ms.Rmd
adjusted figures for new version of oryza::LocToGeneNames()

@@
     dev: "CairoPDF"
+    toc: true
   word_document:
@@
 
-## MADS genes: comparison of expression & homology in rice, tomato, arabidopsis
+## Evolution of MADS genes in monocots and dicots
 

### 2015-09-16 src/R/figures.R

@@ -124,7 +124,7 @@ st_reviewInSitu <- st_reviewInSitu[, .(
   , by = c("msuId", "RM", "PBM", "SBM", "SM", "FM")]
 
 st_reviewInSitu[, `Gene symbol` :=
-                  oryzr::LocToGeneName(msuId, plotLabels = FALSE)$symbols,
+                  oryzr::LocToGeneName(msuId)$symbols,
                 by = msuId]
 setnames(st_reviewInSitu, old = "msuId", new = "MSU identifier")
 setcolorder(st_reviewInSitu, neworder =

### 2015-09-16 ms.Rmd
fixes #8

@@
 *japonica*.
-Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, and other proteins are labelled with their unique identifiers.
+Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.
 The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.

### 2015-09-16 src/R/figures.R

@@ -436,31 +436,31 @@ f_madsTree <- f_madsTree +
   ggplot2::geom_text(size= 1.5, na.rm = TRUE, hjust = -0.01)
 
 # annotate clades
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 146, "AGL2-like",
-                                       offset = 0.08, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 160, "AGL6-like",
-                                       offset = 0.09, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 133, "TM3-like",
-                                       offset = 0.1, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 169, "AG-like",
-                                       offset = 0.09, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 166, "AGL12-like",
-                                       offset = 0.09, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 179, "SQUA-like",
-                                       offset = 0.07, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 193, "STMADS11-like",
-                                       offset = 0.1, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 201, "AGL17-like",
-                                       offset = 0.06, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 209, "FLC-like",
-                                       offset = 0.1, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 215, "GLO-like",
-                                       offset = 0.06, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 219, "DEF-like",
-                                       offset = 0.07, font.size = 1.5, bar.size = 0.5)
-f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 115, "MIKC*",
+f_madsTree <- ggtree::annotation_clade(f_madsTree, node = 145, "AGL2-like",
[…]

### 2015-09-25 ms.Rmd
try MADS tree with genes with padj = NA removed

@@
 | **arapidospis**: inflorescence meristem vs. floret meristem.
-| **tomato**: simpodial meristem vs. floret meristem.
+| **tomato**: sympodial IM vs. floret meristem.
 
@@
 
+```{r f_alogFamily, fig.width = 6.614}
+print(f_alogFamily)
+```
+
+**`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`**.
+Expression of the rice G1 (ALOG) family in inflorescence meristems.
+Points for are plotted in red if the gene was detected in that sample or in blue otherwise.
+Three G1 genes, *G1L1*, *G1L2* and *G1L5* (*TAWAWA*) are expressed, and they share a similar pattern of expression.
+
+----
+
 ```{r f_gsea, fig.height = 8.661 / 2}
@@
 thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di.
-The simpodial meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.
+The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.
 Clades are labelled following the nomenclature presented by (**ref: Arora paper**).

### 2015-09-25 src/R/figures.R

@@ -355,6 +355,68 @@ t_hypergeom[is.na(t_hypergeom)] <- ""
 tableCount <- incCount(tableCount, "t_hypergeom")
 
 ############
+### ALOG ###
+############
+
+# set up biomaRt for phytozome
+phytozome <- biomaRt::useMart(biomart = 'phytozome_mart', dataset = "phytozome")
+
+# query biomaRt
+filters <- c("pfam_id_list", "organism_id")
+values <- list("PF04852", "204")
+atts <- c("gene_name1", "organism_name")
+res <- biomaRt::getBM(atts, filters, values, phytozome, uniqueRows = TRUE)
+martResults <- data.table(res, key = c("gene_name1"))
+rm(phytozome)
+
+# add labels
+alogTable <- martResults[, oryzr::LocToGeneName(gene_name1, shortLabels = TRUE), by = gene_name1]
+
+# add expression values
+tpm <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
+setkey(tpm, "rn")
+plotData.wide <- tpm[alogTable]
+
+# convert to long
+plotData <- reshape2::melt(plotData.wide, id.vars = c("rn", "RapID", "symbols", "names", "labels"),
+                           variable.name = "library", value.name = "Expression (TPM)")
+setkey(plotData, "rn", "library")
+
+# add expression calls
+expGenTT.wide <- data.table(readRDS('output/expressedGenes/expGenTT.Rds'), key = "id")
+expGenTT <- reshape2::melt(expGenTT.wide, id.vars = "id", variable = "library", value = "isExpr")
+setkey(expGenTT, "id", "library")
+plotData <- expGenTT[plotData, .(
+  id,
+  library,
+  labels,
+  `Expression (TPM)`,
+  isExpr)]
+
+# add stage
+plotData[, stage := substr(library, start = 1, stop = 2)]
+old <- c("n1", "n2", "n3", "n4")
+new <- c("RM", "PBM", "ePBM/\nSBM", "SM")
+plotData[, stage := factor(plyr::mapvalues(stage, from = old, to = new), levels = new)]
+
+# put g1 first on the plot
+plotData[, labels := relevel(factor(labels), "G1 (lng/ELE)")]
+
+# make a plot
+cols <- RColorBrewer::brewer.pal(3, "Set1")[c(2,1)]
[…]

### 2015-09-25 src/R/madsCompTree.R

@@ -47,6 +47,8 @@ resSl <- data.table(as.data.frame(
 
 # make DESeq results data.table
 resAll <- rbind(resOs, resAt, resSl)
+# remove genes that failed filtering
+resAll[is.na(padj), log2FoldChange := NA]
 
 # set up biomaRt for phytozome
 phytozome <- biomaRt::useMart(biomart = 'phytozome_mart', dataset = "phytozome")

### 2015-09-29 ms.Rmd
Added code to exclude unexpressed ALOG genes but this is disabled for now.

@@
 To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server (**ref for AGRIS**), with as few type I proteins as possible.
-Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) rooted to a type I MADS sequence (`r og`).
+Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) rooted to type I MADS sequences (`r og`).
 L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
@@
 
-```{r f_alogFamily, fig.width = 6.614}
+```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 2}
 print(f_alogFamily)
@@
 
-**`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`**.
-Expression of the rice G1 (ALOG) family in inflorescence meristems.
-Points for are plotted in red if the gene was detected in that sample or in blue otherwise.
-Three G1 genes, *G1L1*, *G1L2* and *G1L5* (*TAWAWA*) are expressed, and they share a similar pattern of expression.
+**`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
+Expression of rice G1/ALOG family genes in inflorescence meristems.
+Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho] (**ref: Pfam**).
+Red or blue points indicate whether the gene was detected or not detected respectively for each library.
+Only three G1 genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA*) were detected, but they share a similar pattern of expression.
 
@@
 The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-A white background indicates that no results were returned by `DESeq2` for the gene in question, *e.g.* because of low read counts or an extreme count outlier.
+Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do].
 For rice, L~2~FC was estimated from the RM and SM samples reported here.
@@
 ```{r sessionInfo, results = "asis"}
+cat(paste0("**branch:** _", system("git rev-parse --abbrev-ref HEAD", intern = TRUE), "_\n\n"))
+cat(paste0("**hash:** _", system("git rev-parse HEAD", intern = TRUE), "_\n\n"))
 pander::pander(sessionInfo())
-git <- c(paste("**git branch:**",system("git rev-parse --abbrev-ref HEAD", intern = TRUE)),
-          paste("**git hash:**", system("git rev-parse HEAD", intern = TRUE)))
-cat("\n")
-for (x in git){cat(x, "\n")}
 ```

### 2015-09-29 src/R/figures.R

@@ -2,7 +2,7 @@
 
 library(data.table)
 library(ggplot2)
-extrafont::loadfonts()
+#extrafont::loadfonts()
 
 #####################
 ### SETUP FOR RMD ###
@@ -159,7 +159,7 @@ plotData <- expGenTT.long[genTpm.long]
 # add stage names from deseq2
 colData <- data.table(as.data.frame(GenomicRanges::colData(
   readRDS('output/DESeq2/ddsLrt.Rds'))), keep.rownames = TRUE)
-plotData[, stage := colData[rn == as.character(lib), stage]]
+plotData[, stage := colData[rn == as.character(lib), stage], by = lib]
 
 # add replicate number to colour points
 plotData[, Replicate := sub(".*(\\d+)", "\\1", lib)]
@@ -370,7 +370,7 @@ martResults <- data.table(res, key = c("gene_name1"))
 rm(phytozome)
 
 # add labels
-alogTable <- martResults[, oryzr::LocToGeneName(gene_name1, shortLabels = TRUE), by = gene_name1]
+alogTable <- martResults[, oryzr::LocToGeneName(gene_name1), by = gene_name1]
 
 # add expression values
 tpm <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
@@ -378,7 +378,7 @@ setkey(tpm, "rn")
 plotData.wide <- tpm[alogTable]
 
 # convert to long
-plotData <- reshape2::melt(plotData.wide, id.vars = c("rn", "RapID", "symbols", "names", "labels"),
+plotData <- reshape2::melt(plotData.wide, id.vars = c("rn", "RapID", "symbols", "names"),
                            variable.name = "library", value.name = "Expression (TPM)")
 setkey(plotData, "rn", "library")
 
@@ -389,7 +389,7 @@ setkey(expGenTT, "id", "library")
 plotData <- expGenTT[plotData, .(
   id,
   library,
-  labels,
+  symbols,
   `Expression (TPM)`,
   isExpr)]
 
@@ -399,17 +399,34 @@ old <- c("n1", "n2", "n3", "n4")
 new <- c("RM", "PBM", "ePBM/\nSBM", "SM")
 plotData[, stage := factor(plyr::mapvalues(stage, from = old, to = new), levels = new)]
 
+# set up labels
+plotData[!is.na(symbols), symbols := paste(symbols, "·", id)]
+plotData[is.na(symbols), symbols := id]
[…]

### 2015-09-30 ms.Rmd
Add manual list of "interesting" genes in clusters Add SPL/ALOG sections to discussion Reformat legend on sf_pca (fixes #5) Jitter expression (tpm) plots Escape unicode error when exporting to PDF Output SI as tufte pdf?

@@
 csl: csl/the-plant-journal.csl
-header-includes: \usepackage{graphicx}
+header-includes:
+  \usepackage{graphicx}
+  \usepackage[utf8]{inputenc}
+  \DeclareUnicodeCharacter{2265}{$\geq$}
 output:
@@
     number_sections: yes
+  rmarkdown::tufte_handout:
+    default
 bibliography: bib/papersLib.bibtex
@@
 ## Rice inflorescence development
-### Branching & grain yield
-### Sequential changes in meristem activity
-### Discuss a few key genes?
+## Branching & grain yield
+## Sequential changes in meristem activity
+## Discuss a few key genes?
 
@@
 
+Two G1-Like (ALOG) genes, *G1L2* and *G1L5* (*TAWAWA*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively.
+Only three ALOG genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
+*G1L5* mutants **do what, incl.
+ref**, and the co-expression of other ALOG genes suggests redundancy within the family.
+
 **Discussion: what might these patterns mean biologically? Which clusters have genes that are already candidates for panicle branching?**
+
+1. Cluster 1
+  + FZP
+  + WRKY81
+  + OsFBX127, OsFBX291
+  + MADS5, MADS56
+  + BZIP46, BZIP62
+  + YAB6
+  + LOC_Os07g02800 (MYB family?), LOC_Os07g12910 (PHD protein?)
+2. Cluster 2
+  + G1L2
+  + SPL19
+  + MADS47
+  + HOX13
+  + OsFBX30, OsFBO13, OsFBX258
+3. Cluster 3
+  + SNB
+  + G1L5 (TAWAWA)
+  + EP3 (OsFBK5), OsFBX257, OsFBX263, OsFBX333, OsFBX334, OsFBX421
+  + MADS57
+  + BZIP79
+  + ARF9, SAUR25, OsSAUR48
+  + WRKY36
+  + BC10
+  + LOC_Os12g39640 (MYB family?)
+4. Cluster 4
+  + PLT9 (AP2-like)
+  + MADS61
+  + ARF6A, OsSAUR33, LOC_Os08g44750 (auxin-induced protein 5NG4)
+  + OsFBX18
+  + BZIP23
+  + ROC1, ROC3, LOC_Os09g35760 (homeobox/START)
+5. Cluster 5
+  + IDS1, ERF61 (AP2)
+  + OsFBX109
+  + LOC_Os08g25799 (MYB family?)
+6. Cluster 6
+  + SPL3, SPL12
+  + WRKY123
+  + MIR319A (leaf morphology / cold tolerance in indica: 10.1111/pce.12130. Nothing about inflorescence. Strange results, check.)
+  + OsFBX14, OsFBL4, OsFBLD1
+  + ARF2
+  + C3H22
+  + LOC_Os04g35430 (PHD domain?)
+7. Cluster 7
+  + LOGL1, LOGL6
+  + BZIP47
+  + HOX14
+  + WRKY88
+  + SHR1
+  + SCR (GRAS)
+  + NAC41 (No apical meristem protein)
+  + OsFBO2, OsFBX125, OsFBX349
+  + LOC_Os03g53700 (PHD domain?)
+  + LOC_Os08g06240 (MYB domain?)
+8. Cluster 8
+  + PAP2 (MADS34)
+  + DEP1
+  + OsFBK1
+  + SAUR6
+  + LOC_Os02g13310 (homeobox?)
 
@@
 Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
+
+The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
+Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
+*WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
+These genes are all targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).
 

### 2015-09-30 si.Rmd
Add manual list of "interesting" genes in clusters Add SPL/ALOG sections to discussion Reformat legend on sf_pca (fixes #5) Jitter expression (tpm) plots Escape unicode error when exporting to PDF Output SI as tufte pdf?

@@
 csl: csl/the-plant-journal.csl
-header-includes: \usepackage{graphicx}
+header-includes:
+  \usepackage{graphicx}
+  \usepackage{caption}
+  \captionsetup[figure]{labelfont=bf}
+  \renewcommand{\thefigure}{S\arabic{figure}}
 output:
+  rmarkdown::tufte_handout:
+    dev: "CairoPDF"
+  pdf_document:
+    fig_caption: yes
+    number_sections: yes
+    dev: "CairoPDF"
   BiocStyle::html_document:
@@
     highlight: pygments
-    number_sections: yes
-    dev: "CairoPDF"
-  pdf_document:
-    fig_caption: yes
     number_sections: yes
@@
                cache.path='cache/',
-               dev=c("png", "cairo_pdf"),
-               fig.align = 'center',
+#               fig.align = 'center', 
+               dev = "cairo_pdf",
                dpi=300,
@@
                fig.width = 3.150,
-               cache=FALSE,
+               out.width = 3.150,
+               out.height = 3.150,
+#               cache=FALSE,
                echo=FALSE,
@@
 pander::panderOptions("table.caption.prefix", "")
-
 source('src/R/figures.R')
@@
 
-``` {r st_libStats, results = "asis"}
+``` {r st_libStats, results = "asis", fig.cap = I(pasteLabel("**Table S", s_tableCount, "st_libStats"))}
 pander::set.caption(paste0(
@@
 
-```{r sf_pca}
+```{r sf_pca, fig.cap = paste0("Principal components analysis of transformed read counts for each library.
+The two libraries made from degraded RNA did not cluster with other libraries (see ", I(pasteLabel("Table S", s_tableCount, "st_libStats")), "), suggesting that the use of degraded RNA affected quantification of expression.
+For this reason these two libraries were not included in subsequent analyses.")}
 print(sf_pca)

### 2015-09-30 src/R/figures.R

@@ -2,6 +2,7 @@
 
 library(data.table)
 library(ggplot2)
+set.seed(1)
 #extrafont::loadfonts()
 
 #####################
@@ -70,6 +71,10 @@ pcaPlotData <- data.frame(
 pcaPlotData$rIdx <- 1:dim(pcaPlotData)[1]
 opIdx <- pcaPlotData[c('n2r3', 'n4r1'),'rIdx']
 
+# format legend
+pcaPlotData$Stage <- plyr::mapvalues(pcaPlotData$Stage, from = "ePBM/SBM",
+                                     to = "ePBM/\nSBM")
+
 sf_pca <- ggplot(data = pcaPlotData,
                  aes(x = PCA1, y = PCA2, colour = Stage, label = label)) +
   theme_minimal(base_size = 8, base_family = "Helvetica") +
@@ -86,7 +91,6 @@ sf_pca <- ggplot(data = pcaPlotData,
             hjust = -0.2, show.legend = FALSE)
 
 s_figCount <- incCount(s_figCount, "sf_pca")
-
 
 #######################
 ### COMPARE IN SITU ###
@@ -182,7 +186,7 @@ sf_isGenesTpm <- ggplot(plotData, aes(x = stage, y = `Expression (transcripts pe
   scale_color_brewer(palette = "Set1") +
   guides(shape = FALSE, size = FALSE) +
   stat_smooth(aes(group = label), method = "loess", se = FALSE, colour = "grey", size = 0.5) + 
-  geom_point(size = 1, alpha = 0.8) +
+  geom_point(size = 1, alpha = 0.8, position = position_jitter(width = 0.2)) +
   facet_wrap(~ label, scales = "free_y", ncol = 4)
 s_figCount <- incCount(s_figCount, "sf_isGenesTpm")
 
@@ -429,7 +433,7 @@ f_alogFamily <- ggplot(plotData, aes(x = stage, y = `Expression (TPM)`, group = 
   facet_wrap(~symbols, ncol = 2) +
   xlab(NULL) +
   stat_smooth(se = FALSE, colour = "grey", size = 0.5) +
-  geom_point(alpha = 0.7)
+  geom_point(shape = 16, alpha = 0.7, position = position_jitter(height = 0, width = 0.3))
 
 figCount <- incCount(figCount, "f_alogFamily")
 
[…]

### 2015-10-02 ms.Rmd
Added discussion on clusters, GSEA and MADS.

@@
                fig.show='hold',
-               cache=FALSE,
+               cache=TRUE,
                echo=FALSE,
@@
 
-| ^§^ Institut de Recherche pour le Développement (IRD), 911 Avenue Agropolis, 34394 Montpellier, France.
+| ^§^ Institut de Recherche pour le Développement, 911 Avenue Agropolis, 34394 Montpellier, France.
 | ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
@@
 
-At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
+At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
 Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
-As a preliminary quality check, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`).
-
-**Our in situ results could go here (or in SI).**
+As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**).
 
@@
 
-Two G1-Like (ALOG) genes, *G1L2* and *G1L5* (*TAWAWA*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively.
+The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes.
+Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively.
 Only three ALOG genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
-*G1L5* mutants **do what, incl.
-ref**, and the co-expression of other ALOG genes suggests redundancy within the family.
+Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
+*G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff].
+This might be explained by redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other ALOG genes, but I'm not sure how much is required for off-target effects.]
 
-**Discussion: what might these patterns mean biologically? Which clusters have genes that are already candidates for panicle branching?**
+The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
+Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [**Komatsu 2003 FZP paper**], were found in clusters containing genes that are highly expressed in the SM.
+
+| **What could any of these patterns mean biologically?**
+| **I picked out some genes from the clusters (list directly below), are there any others in here I should discuss?**
 
@@
 Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
-Unlike soft clustering, which groups genes based on shared expression dynamics, this GSEA was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in expression of their members in each stage.
-These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
-Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
+Unlike soft clustering, which groups genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
+There are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
 The most pronounced change in expression appears to occur between the RM and branch meristem samples.
@@
 *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
-These genes are all targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).
 
-**Discussion: In which samples were TF families involved in branching enriched?**
+These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
+An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
+*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs, and mutants of the former gene produce fewer spikelets than wild-type plants [@Gao:2010iz] (**check role of PAP2**).
+In contrast to *PAP2*, *LHS1* was detected only in the SM samples, indicative of the overall pattern for other MADS genes.
 
-## Evolution of MADS genes in monocots and dicots
+## Comparative analysis of MADS gene expression in eudicots and monocots
 
-There was an overrepresentation of MADS genes in cluster `r getClusterName('small', 'inc')` and their overall expression is enriched in the spikelet meristem sample.
-These results correspond with previous studies that have established the importance of MADS genes in inflorescence development in rice and other plants (**expand**).
+The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
+In the eudicot *A.
+thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by **ref: Prunet/Jack review**).
+In rice, functions homologous to those provided by *A.
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on?
+check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
+[**Add info about tomato MADS**].
 
-The change in expression of MIKC proteins between undifferentiated and differentiated meristems was extracted from the dataset reported here and two other studies that used LMD and RNA-sequencing in early stages of inflorescence development in Arabidopsis and tomato [@Mantegazza:2014di; @Park:2012dg].
-To investigate the evolution of the MADS family in Angiosperms, these changes were compared to homology of MIKC proteins in the three species (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
+* Something about MADS genes involved earlier in development?
+* Expression of MADS genes that interact with *G1L5*? (Maybe this is too much detail?)
 
-| **Figure**: Similarity tree of MADS proteins with L2FC values between comparable samples (and dN/dS values?) for each one:
-| **rice**: rachis meristem vs. spikelet merisem.
-| **arapidospis**: inflorescence meristem vs. floret meristem.
-| **tomato**: sympodial IM vs. floret meristem.
+A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper.
+The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A.
+thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di].
+A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
+[**Note about relative genome size (number of genes) and MADS family size between species**.]
+
+As expected based on their functions in floral organ specification in rice and *A.
+thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
+However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
+
+* *AGL6*-like: AGL6 has a defined function but it doesn't look like other genes in the clade
+* *STMADS11*-like:
+* *SQUA*-like:
+* *DEF*-like:
+* *GLO*-like:
+* *FLC*-like:
 
@@
 Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night.
-After 7 weeks in long day conditions (LD; 14 hours light and 10 hours darkness) plants were transferred to short day conditions (SD; 10 hours light and 14 hours darkness) to induce floral transition.
+After 7 weeks in long day conditions (14 hours light and 10 hours darkness) plants were transferred to short day conditions (10 hours light and 14 hours darkness) to induce floral transition.
 Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.
@@
 
-Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (=**filtered**?) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C.
+Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (**filtered?**) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C.
 Embedding for dissection was performed as described by @Mantegazza:2014di.
@@
 RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN).
-Library preparation with **which kit** and 50-base single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
+Library preparation with **which kit** and 50-base, single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
 

### 2015-10-02 ms.Rmd
No actual changes :)

@@
     toc: true
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
   pdf_document:

### 2015-10-09 ms.Rmd
Fix table numbering Hormones, other writing

@@
 csl: csl/the-plant-journal.csl
-header-includes:
-  \usepackage{graphicx}
-  \usepackage[utf8]{inputenc}
-  \DeclareUnicodeCharacter{2265}{$\geq$}
+header-includes: \usepackage{graphicx} \usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2265}{$\geq$}
 output:
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
   BiocStyle::html_document:
+    dev: CairoPDF
     fig_caption: yes
@@
     number_sections: yes
-    dev: "CairoPDF"
-    toc: true
+    toc: yes
   pdf_document:
@@
     number_sections: yes
-  rmarkdown::tufte_handout:
-    default
+  rmarkdown::tufte_handout: default
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
 bibliography: bib/papersLib.bibtex
@@
 
+Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). The latter,
+
+Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')`.
+The LOG enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go].
+In contrast, the LABA1 protein present in *O.
+rufipogon*, which has an identical primary sequence to the protein from *O.
+sativa* ssp.
+*japonica* cv.
+Nipponbare, also activates cytokinin *in vitro* [**ref Gu 2015; Hua et all 2015**], and introduction of the *Laba1* allele from *O.
+rufipogon* into an *O.
+sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [**Hua et all 2015**], but reduces the number of grains per panicle and the number of tillers per plant [**ref Gu 2015**], suggesting a link between increased cytokinin signalling and reduced panicle branching.
+*laba1* is a domestication allele found at high frequencies in cultivated accessions, suggesting that it was affected by artificial selection either for awn phenotype or grain yield [**ref Gu 2015; Hua et all 2015**].
+No reports of *LOGL1* function in rice were found.
+An uncharecterized gene annotated as a cytokinin dehydrogenase, *LOC_Os05g31040*, is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), which suggests a model of increasing cytokinin activation and decreasing inactivation from RM–SM, but it is currently not clear how such a mechanism might influence meristem identity.
+There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A response regulator (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex.
+**Grain number 1a?**.
+**Effect of cytokinin on meristem size/activity?**
+
 The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes.
 Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively.
-Only three ALOG genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
+Only three G1L genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
 Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
 *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff].
-This might be explained by redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other ALOG genes, but I'm not sure how much is required for off-target effects.]
+This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other G1L genes, but I'm not sure how much is required for off-target effects.]
 
 The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
-Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [**Komatsu 2003 FZP paper**], were found in clusters containing genes that are highly expressed in the SM.
+Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], were found in clusters containing genes that are highly expressed in the SM.
 
@@
 *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
-These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed (**ref: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1557610/**).
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].
 
-These patterns are not intended to represent each member of each gene family: for example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
+These patterns are not intended to represent each member of each gene family.
+For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
 An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
-*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs, and mutants of the former gene produce fewer spikelets than wild-type plants [@Gao:2010iz] (**check role of PAP2**).
-In contrast to *PAP2*, *LHS1* was detected only in the SM samples, indicative of the overall pattern for other MADS genes.
+*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *PAP2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz, @Kobayashi:2010hn].
+In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples [**ref Jeon 2000, Agrawal 2005**].
 
@@
 In the eudicot *A.
-thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by **ref: Prunet/Jack review**).
+thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by @Prunet:2013ki).
 In rice, functions homologous to those provided by *A.
@@
 The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
-An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix (**ref clustalo**).
+An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn].
 To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed.
-To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server (**ref for AGRIS**), with as few type I proteins as possible.
-Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package (**ref:ape***) rooted to type I MADS sequences (`r og`).
+To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible.
+Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`).
 L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
@@
 Expression of rice G1/ALOG family genes in inflorescence meristems.
-Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho] (**ref: Pfam**).
+Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj].
 Red or blue points indicate whether the gene was detected or not detected respectively for each library.

### 2015-10-09 src/R/figures.R

@@ -326,15 +326,15 @@ sf_mfuzzPca <- ggplot(vg.mds, aes(x = MDS1, y=MDS2, colour = factor(cluster))) +
 
 s_figCount <- incCount(s_figCount, "sf_mfuzzPca")
 
-# table of families/cluster
-t_familiesPerCluster <- readRDS('output/mfuzz/familyClusters.Rds')
-t_familiesPerCluster[!total == 0]
-from <- grep("^C\\d+", names(t_familiesPerCluster), value = TRUE)
-to <- sub("(\\d+)", "~\\1~", from)
-setnames(t_familiesPerCluster, old = c("Fam", "n", "nexpr", "total", from),
-         new = c("Family", "*n*", "*n*~expr~", "Total", to))
-
-tableCount <- incCount(tableCount, "t_familiesPerCluster")
+# # table of families/cluster
+# t_familiesPerCluster <- readRDS('output/mfuzz/familyClusters.Rds')
+# t_familiesPerCluster[!total == 0]
+# from <- grep("^C\\d+", names(t_familiesPerCluster), value = TRUE)
+# to <- sub("(\\d+)", "~\\1~", from)
+# setnames(t_familiesPerCluster, old = c("Fam", "n", "nexpr", "total", from),
+#          new = c("Family", "*n*", "*n*~expr~", "Total", to))
+# 
+# tableCount <- incCount(tableCount, "t_familiesPerCluster")
 
 #################
 ### HYPERGEOM ###

### 2015-10-13 ms.Rmd
Stefan's comments 2015-10-13

@@
     toc: yes
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
   pdf_document:
@@
   rmarkdown::tufte_handout: default
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
 bibliography: bib/papersLib.bibtex
@@
 At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
-Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
+Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
 As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**).
+[**How do in situ / published results compare to transcriptome?**]
 
@@
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
-There are several expression patterns descibed by the clusters: genes that increase or decrease steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that are constant between RM and PBM and then change between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
+Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that have constant expression between RM and PBM and then increase or decrease in expression between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
 There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).
+
+Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
+Cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene *GH3-8*.
+[**Role of auxin influx / efflux in axillary / apical meristem specification**].
+In *A.
+thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv].
+Transgenic plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* (*i.e.* PBM and SBM; [@IkedaKawakatsu:2012co]), have smaller panicles with fewer branches [@Yadav:2011db].
+GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated to reduced auxin signalling.
+[**Conclusion about auxin signalling**.]
+
+**Auxin-related genes in clusters 2 & 4:**
+
++ *GH3-8*, *IAA10*, *ABCB27*, *LOC_Os01g63770* (AUX/LAX transporter) in cluster 2
++ *ARF6A*, *SAUR33*, *MDR12* (ABC), *ALS1* (ABC) in cluster 4
+
+Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that are increasingly expressed between RM and SM.
+The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go].
+Reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg].
+These results suggest that an increase in active cytokinin is associated with grain yield via an effect on panicle branching, but the function of *LABA1* suggests the opposite.
+The LABA1 protein present in *O.
+rufipogon*, which has an identical primary sequence to the protein from *O.
+sativa* ssp.
+*japonica* cv.
+Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O.
+rufipogon* into an *O.
+sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy].
+*laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs].
+No reports were found of *LOGL1* function, or the function of the gene annotated as a cytokinin dehydrogenase (*LOC_Os05g31040*) that is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), but the expression pattern of *LABA1* and these two genes suggests increasing cytokinin activation and decreasing inactivation from RM to SM.
+It is not currently clear how such a mechanism might influence meristem identity or panicle branching at the molecular level.
+[**Check for at/tomato papers**].
+There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex.
 
@@
 Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 and G2-like TFs.
-MADS TFs and TRAF genes were overreprsented among genes that have low expression in the SM and PBM but higher expression in the RM (cluster `r getClusterName('small', 'inc')`) and bHLH, SBP and C2C2-GATA TFs were overreprented among genes with the inverse pattern (cluster `r getClusterName('small', 'dec')`)
+MADS TFs and TRAF genes were overrepresented among genes that have low expression in the SM and PBM but higher expression in the RM (cluster `r getClusterName('small', 'inc')`) and bHLH, SBP and C2C2-GATA TFs were overreprented among genes with the inverse pattern (cluster `r getClusterName('small', 'dec')`)
 
-Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). The latter,
-
-Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')`.
-The LOG enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go].
-In contrast, the LABA1 protein present in *O.
-rufipogon*, which has an identical primary sequence to the protein from *O.
-sativa* ssp.
-*japonica* cv.
-Nipponbare, also activates cytokinin *in vitro* [**ref Gu 2015; Hua et all 2015**], and introduction of the *Laba1* allele from *O.
-rufipogon* into an *O.
-sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [**Hua et all 2015**], but reduces the number of grains per panicle and the number of tillers per plant [**ref Gu 2015**], suggesting a link between increased cytokinin signalling and reduced panicle branching.
-*laba1* is a domestication allele found at high frequencies in cultivated accessions, suggesting that it was affected by artificial selection either for awn phenotype or grain yield [**ref Gu 2015; Hua et all 2015**].
-No reports of *LOGL1* function in rice were found.
-An uncharecterized gene annotated as a cytokinin dehydrogenase, *LOC_Os05g31040*, is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), which suggests a model of increasing cytokinin activation and decreasing inactivation from RM–SM, but it is currently not clear how such a mechanism might influence meristem identity.
-There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A response regulator (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex.
-**Grain number 1a?**.
-**Effect of cytokinin on meristem size/activity?**
-
-The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes suggests that the clusters capture biologically meaningful groups of genes.
-Two G1-Like (G1L/ALOG) genes, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively.
-Only three G1L genes are expressed in the meristem RNA sequencing dataset, but they share a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
+The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
+Only three G1-Like (G1L/ALOG) genes were detected in the dataset.
+Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
 Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
 *G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff].
-This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5* [**This is just speculation!** I checked and there is some homology between the RNAi construct and other G1L genes, but I'm not sure how much is required for off-target effects.]
+This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*.
+[**Downstream targets of G1L5 (MADS22/MADS47/MADS55)**]
 
-The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in flower development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
+The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
 Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], were found in clusters containing genes that are highly expressed in the SM.
 
-| **What could any of these patterns mean biologically?**
 | **I picked out some genes from the clusters (list directly below), are there any others in here I should discuss?**
@@
 
-The strong expression of SPL genes (which possess the SBP-box) in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
-Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
+The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
+Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
+**Downstream targets of WFP**.
 *WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
@@
 An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
-*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *PAP2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz, @Kobayashi:2010hn].
-In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples [**ref Jeon 2000, Agrawal 2005**].
+*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
+In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.
 
@@
 Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
-To determine a strict cutoff for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf].
-These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values.
-For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values.
+To determine a strict cut-off for unexpressed genes, expression values in transcripts per million were calculated for each gene [@Li:2010cl; @Wagner:2012jf].
+These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distribution to those used for calculating the 'genic' expression values.
+For each library, the cut-off was placed at the 95^th^ percentile of the distribution of intergenic expression values.
 Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively.
@@
 The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.
-Clades are labelled following the nomenclature presented by (**ref: Arora paper**).
+Clades are labelled following the nomenclature presented by [@Arora:2007fa].
 

### 2015-10-14 ms.Rmd
Just plot the interesting G1-like genes

@@
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
-Cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene *GH3-8*.
-[**Role of auxin influx / efflux in axillary / apical meristem specification**].
+Axillary meristem initiation requires auxin synthesis and transport in *A.
+thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice.
+Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*.
 In *A.
-thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv].
-Transgenic plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* (*i.e.* PBM and SBM; [@IkedaKawakatsu:2012co]), have smaller panicles with fewer branches [@Yadav:2011db].
-GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated to reduced auxin signalling.
-[**Conclusion about auxin signalling**.]
-
-**Auxin-related genes in clusters 2 & 4:**
-
-+ *GH3-8*, *IAA10*, *ABCB27*, *LOC_Os01g63770* (AUX/LAX transporter) in cluster 2
-+ *ARF6A*, *SAUR33*, *MDR12* (ABC), *ALS1* (ABC) in cluster 4
+thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db].
+GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated with reduced auxin signalling.
+The presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
 
@@
 The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
-Only three G1-Like (G1L/ALOG) genes were detected in the dataset.
-Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`).
+The G1-Like (G1L/ALOG) genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj].
+Only three were detected in the dataset.
+Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
 Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
@@
 This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*.
-[**Downstream targets of G1L5 (MADS22/MADS47/MADS55)**]
+Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55* [@Yoshida:2013ff].
+*MADS22* and *MADS55* are also highly expresed in the RM, and the third rice *STMADS11*-like gene, *MADS47*, is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
 
@@
 
-```{r t_familiesPerCluster, results = "hide"}
-pander::set.caption(paste0(
-  I(pasteLabel("**Table ", tableCount, "t_familiesPerCluster")),
-  "**: Transcription factor and transcriptional regulator family members in each cluster.
-*n*, total number of family members, *n*~expr~, number of family members expressed in the whole dataset, C~1~–C~", c, "~, number of family members in each cluster, and the total number of family members in all clusters."))
-t_familiesPerCluster <- as.data.frame(t_familiesPerCluster[!Total == 0])
-rownames(t_familiesPerCluster) <- t_familiesPerCluster$Family
-t_familiesPerCluster$Family <- NULL
-pander::pandoc.table(t_familiesPerCluster, split.tables = Inf, emphasize.rownames = FALSE,
-                     justify = c("right", rep("centre", length(names(t_familiesPerCluster)) - 1), "left"))
-```
-
 ```{r t_hypergeom, results = "asis"}
@@
 
-```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 2}
-print(f_alogFamily)
+```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 3/2}
+grid::grid.draw(f_alogFamily)
 ```
@@
 **`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
-Expression of rice G1/ALOG family genes in inflorescence meristems.
-Genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj].
+Expression of detected G1-Like (G1L/ALOG)  (**a**) and *STMADS11*-like MADS (**b**) genes in inflorescence meristems.
 Red or blue points indicate whether the gene was detected or not detected respectively for each library.
-Only three G1 genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA*) were detected, but they share a similar pattern of expression.
+Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression.
+Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
 
@@
 The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.
-Clades are labelled following the nomenclature presented by [@Arora:2007fa].
+Clades are labelled following the nomenclature presented by @Arora:2007fa.
 

### 2015-10-14 src/R/figures.R

@@ -362,27 +362,21 @@ tableCount <- incCount(tableCount, "t_hypergeom")
 ### ALOG ###
 ############
 
-# set up biomaRt for phytozome
-phytozome <- biomaRt::useMart(biomart = 'phytozome_mart', dataset = "phytozome")
+svpMads <- c("LOC_Os02g52340", "LOC_Os03g08754", "LOC_Os06g11330")
+expG1s <- c("LOC_Os02g07030", "LOC_Os06g46030", "LOC_Os10g33780")
 
-# query biomaRt
-filters <- c("pfam_id_list", "organism_id")
-values <- list("PF04852", "204")
-atts <- c("gene_name1", "organism_name")
-res <- biomaRt::getBM(atts, filters, values, phytozome, uniqueRows = TRUE)
-martResults <- data.table(res, key = c("gene_name1"))
-rm(phytozome)
+expTable <- rbind(data.table(msuId = svpMads, type = "mads"),
+data.table(msuId = expG1s, type = "g1l"))
 
-# add labels
-alogTable <- martResults[, oryzr::LocToGeneName(gene_name1), by = gene_name1]
+expTable[, symbol := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
 
 # add expression values
 tpm <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
 setkey(tpm, "rn")
-plotData.wide <- tpm[alogTable]
+plotData.wide <- tpm[expTable]
 
 # convert to long
-plotData <- reshape2::melt(plotData.wide, id.vars = c("rn", "RapID", "symbols", "names"),
+plotData <- reshape2::melt(plotData.wide, id.vars = c("rn", "type", "symbol"),
                            variable.name = "library", value.name = "Expression (TPM)")
 setkey(plotData, "rn", "library")
 
@@ -392,8 +386,9 @@ expGenTT <- reshape2::melt(expGenTT.wide, id.vars = "id", variable = "library", 
 setkey(expGenTT, "id", "library")
 plotData <- expGenTT[plotData, .(
   id,
+  type,
   library,
-  symbols,
+  symbol,
   `Expression (TPM)`,
   isExpr)]
 
@@ -404,36 +399,31 @@ new <- c("RM", "PBM", "ePBM/\nSBM", "SM")
 plotData[, stage := factor(plyr::mapvalues(stage, from = old, to = new), levels = new)]
 
 # set up labels
[…]

### 2015-10-14 src/R/madsCompTree.R
Try without the threshold requirement

@@ -34,15 +34,15 @@ if (!dir.exists(outDir)) {
 # get DESeq2 results
 ddsOs <- readRDS(ddsOsFile)
 resOs <- data.table(as.data.frame(
-  DESeq2::results(ddsOs, contrast = c("stage", "SM", "RM"), lfcThreshold = log2(1.5), alpha = 0.1)),
+  DESeq2::results(ddsOs, contrast = c("stage", "SM", "RM"), alpha = 0.1)),
   keep.rownames = TRUE, key = "rn")
 ddsAt <- readRDS(ddsAtFile)
 resAt <- data.table(as.data.frame(
-  DESeq2::results(ddsAt, contrast = c("stage", "FM", "IM"), lfcThreshold = log2(1.5), alpha = 0.1)),
+  DESeq2::results(ddsAt, contrast = c("stage", "FM", "IM"), alpha = 0.1)),
   keep.rownames = TRUE, key = "rn")
 ddsSl <- readRDS(ddsSlFile)
 resSl <- data.table(as.data.frame(
-  DESeq2::results(ddsSl, contrast = c("stage", "fm", "sim"), lfcThreshold = log2(1.5), alpha = 0.1)),
+  DESeq2::results(ddsSl, contrast = c("stage", "fm", "sim"), alpha = 0.1)),
   keep.rownames = TRUE, key = "rn")
 
 # make DESeq results data.table

### 2015-10-16 ms.Rmd
final commit before changing gene expression cutoff

@@
 
-Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that are increasingly expressed between RM and SM.
-The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants produce small panicles with small SAMs and branching defects, and have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes [@Kurakawa:2007go].
-Reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg].
-These results suggest that an increase in active cytokinin is associated with grain yield via an effect on panicle branching, but the function of *LABA1* suggests the opposite.
+Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM.
+In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**].
+The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go].
+Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg].
+These results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LABA1* gene provides evidence for an opposite function.
 The LABA1 protein present in *O.
@@
 *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs].
-No reports were found of *LOGL1* function, or the function of the gene annotated as a cytokinin dehydrogenase (*LOC_Os05g31040*) that is present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`), but the expression pattern of *LABA1* and these two genes suggests increasing cytokinin activation and decreasing inactivation from RM to SM.
-It is not currently clear how such a mechanism might influence meristem identity or panicle branching at the molecular level.
-[**Check for at/tomato papers**].
-There are also genes that may be involved in cytokinin signalling in other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')`, containing genes more highly expressed in RM, suggesting that the role of cytokinins in inflorescence development may be complex.
+The function of *LOGL1* has not been reported.
+There is also an uncharacterized gene, *LOC_Os05g31040*, which is annotated as a cytokinin dehydrogenase, present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`).
+The expression patterns of *LABA1*, *LOGL1* and *LOC_Os05g31040* suggest increasing cytokinin activation and decreasing inactivation from RM to SM.
+Such a mechanism would indicate a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level.
+Genes that may be involved in cytokinin signalling are also present other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.
 
@@
 This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*.
-Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55* [@Yoshida:2013ff].
-*MADS22* and *MADS55* are also highly expresed in the RM, and the third rice *STMADS11*-like gene, *MADS47*, is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, is not reported [@Yoshida:2013ff].
+*MADS22* and *MADS55* are also enriched in the RM, and *MADS47* is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
 
@@
 
+| **Shorten MADS section and insert here. MADS DEG figure goes to SI?**
+| **GRAS TFs (SHR1/SCR)?**
+
 The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
-Overexpression of *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
-**Downstream targets of WFP**.
-*WFP*, *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
+*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
 These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].
+Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
+WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O.
+sativa* ssp.
+*japonica* varieties.
+Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf].
+WFP also binds to the promoters of *LOG1*, *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].
+
+| **[Re]move following MADS section and insert section on Homeobox. Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007)**
 
@@
 In the eudicot *A.
-thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen (recently reviewed by @Prunet:2013ki).
+thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
 In rice, functions homologous to those provided by *A.
@@
 **`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**.
-Neighbour-joining similarity tree of putatative MIKC proteins in *A.
+Differential expression of putatative MIKC proteins in *A.
 thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp.
 *japonica*.
+The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences.
 Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.
 The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-Where there is no coloured label, the gene for the protein in question was removed from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do].
+Where there is no coloured label, the adjusted *p*-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do].
 For rice, L~2~FC was estimated from the RM and SM samples reported here.

### 2015-10-16 src/R/calculateCutoffs.R

@@ -100,10 +100,13 @@ names(expressedGenesByLibrary) <- levels(q95.plot$lib)
 # expressed genes truth table
 tt <- data.table(dnaTpm.plot)[type == 'Genic']
 tt[, expressed := tpm > cutoffs[lib], by = id]
+tt[, stage := substr(lib, 1, 2)]
+tt[, expByStage := sum(expressed) > 1, by = .(id, stage)]
 expGenTT <- reshape2::dcast(tt, id ~ lib, value.var = "expressed")
 
 # list of expressed genes
-expressedGenes <- unique(do.call(c, expressedGenesByLibrary))
+expressedGenes <- unique(do.call(c, expressedGenesByLibrary)) # expressed in any lib
+#expressedGenes <- tt[expByStage == TRUE, unique(id)] # expressed in at least 2 libs in 1 stage
 
 # what are the real cutoffs in the tpm calculations performed *without* the shuffled GTF?
 realTpm <- readRDS(paste0(tpmDir, '/tpm.Rds'))

### 2015-10-19 ms.Rmd
Handled changes caused by stricter detection of expressed genes

@@
     toc: yes
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
   pdf_document:
@@
   rmarkdown::tufte_handout: default
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
 bibliography: bib/papersLib.bibtex
@@
 At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
-Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries.
+Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage.
 As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**).
@@
 thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice.
-Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains three uncharacterized genes with annotations relating to auxin (*IAA10*, an auxin-responsive transcriptional regulator; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*.
+Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*.
 In *A.
@@
 GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated with reduced auxin signalling.
-The presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
+Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (see **ref next section**), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
 
-Two homologs of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*) and *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*), are present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM.
+A homolog of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM.
 In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**].
 The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go].
-Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield rice varieties [@Ashikari:2005eg].
-These results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LABA1* gene provides evidence for an opposite function.
+Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O.
+sativa* ssp.
+*indica* varieties [@Ashikari:2005eg].
+Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role.
 The LABA1 protein present in *O.
@@
 *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs].
-The function of *LOGL1* has not been reported.
-There is also an uncharacterized gene, *LOC_Os05g31040*, which is annotated as a cytokinin dehydrogenase, present in the complementary cluster (cluster `r getClusterName('dec', 'dec')`).
-The expression patterns of *LABA1*, *LOGL1* and *LOC_Os05g31040* suggest increasing cytokinin activation and decreasing inactivation from RM to SM.
-Such a mechanism would indicate a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level.
-Genes that may be involved in cytokinin signalling are also present other clusters, including a type-A *RESPONSE REGULATOR* (*RR3*) and a cytokinin receptor (*PHP2*) in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.
+*LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples.
+Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation from RM to SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level.
+A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.
 
@@
 Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
-After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 and G2-like TFs.
-MADS TFs and TRAF genes were overrepresented among genes that have low expression in the SM and PBM but higher expression in the RM (cluster `r getClusterName('small', 'inc')`) and bHLH, SBP and C2C2-GATA TFs were overreprented among genes with the inverse pattern (cluster `r getClusterName('small', 'dec')`)
+After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
+Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
+Auxin response factors are enriched in cluster `r getClusterName('dec', 'dec')`.
 
 The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
-The G1-Like (G1L/ALOG) genes were retrieved by querying Phytozome 10.3 for the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj].
-Only three were detected in the dataset.
-Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+Of the G1-Like (G1L/ALOG) genes, which are annotated with the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj, only three were detected in the dataset.
+Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
 Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
@@
 Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, is not reported [@Yoshida:2013ff].
-*MADS22* and *MADS55* are also enriched in the RM, and *MADS47* is present in cluster `r getClusterName('dec', 'small')` with *G1L2*, although it is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+*MADS22* and *MADS55* are also enriched in the RM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
 
 The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
-Two other AP2 genes, *INDETERMINATE SPIKELET*, which acts redundantly with *SNB* to promote transition from SM to FM [@Lee:2012hj], and *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], were found in clusters containing genes that are highly expressed in the SM.
-
-| **I picked out some genes from the clusters (list directly below), are there any others in here I should discuss?**
-
-1. Cluster 1
-  + FZP
-  + WRKY81
-  + OsFBX127, OsFBX291
-  + MADS5, MADS56
-  + BZIP46, BZIP62
-  + YAB6
-  + LOC_Os07g02800 (MYB family?), LOC_Os07g12910 (PHD protein?)
-2. Cluster 2
-  + G1L2
-  + SPL19
-  + MADS47
-  + HOX13
-  + OsFBX30, OsFBO13, OsFBX258
-3. Cluster 3
-  + SNB
-  + G1L5 (TAWAWA)
-  + EP3 (OsFBK5), OsFBX257, OsFBX263, OsFBX333, OsFBX334, OsFBX421
-  + MADS57
-  + BZIP79
-  + ARF9, SAUR25, OsSAUR48
-  + WRKY36
-  + BC10
-  + LOC_Os12g39640 (MYB family?)
-4. Cluster 4
-  + PLT9 (AP2-like)
-  + MADS61
-  + ARF6A, OsSAUR33, LOC_Os08g44750 (auxin-induced protein 5NG4)
-  + OsFBX18
-  + BZIP23
-  + ROC1, ROC3, LOC_Os09g35760 (homeobox/START)
-5. Cluster 5
-  + IDS1, ERF61 (AP2)
-  + OsFBX109
-  + LOC_Os08g25799 (MYB family?)
-6. Cluster 6
-  + SPL3, SPL12
-  + WRKY123
-  + MIR319A (leaf morphology / cold tolerance in indica: 10.1111/pce.12130. Nothing about inflorescence. Strange results, check.)
-  + OsFBX14, OsFBL4, OsFBLD1
-  + ARF2
-  + C3H22
-  + LOC_Os04g35430 (PHD domain?)
-7. Cluster 7
-  + LOGL1, LOGL6
-  + BZIP47
-  + HOX14
-  + WRKY88
-  + SHR1
-  + SCR (GRAS)
-  + NAC41 (No apical meristem protein)
-  + OsFBO2, OsFBX125, OsFBX349
-  + LOC_Os03g53700 (PHD domain?)
-  + LOC_Os08g06240 (MYB domain?)
-8. Cluster 8
-  + PAP2 (MADS34)
-  + DEP1
-  + OsFBK1
-  + SAUR6
-  + LOC_Os02g13310 (homeobox?)
+Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
 
@@
 The most pronounced change in expression appears to occur between the RM and branch meristem samples.
-Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
-Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
+Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
+Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
 
 | **Shorten MADS section and insert here. MADS DEG figure goes to SI?**
-| **GRAS TFs (SHR1/SCR)?**
+| **GRAS TFs (SHR1/SCR)? I think I will skip this because they are not highly expressed.**
 
@@
 However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
-
-* *AGL6*-like: AGL6 has a defined function but it doesn't look like other genes in the clade
-* *STMADS11*-like:
-* *SQUA*-like:
-* *DEF*-like:
-* *GLO*-like:
-* *FLC*-like:
 
@@
 
-```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 3/2}
-grid::grid.draw(f_alogFamily)
+```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2}
+#grid::grid.draw(f_alogFamily)
+
+# set layout
+grid::grid.newpage()
+grid::pushViewport(grid::viewport(layout = grid::grid.layout(5, 2)))
+
+# print the plots
+print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:2, layout.pos.col = 1:2))
+print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 1))
+print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 2))
 ```
@@
 **`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
-Expression of detected G1-Like (G1L/ALOG)  (**a**) and *STMADS11*-like MADS (**b**) genes in inflorescence meristems.
-Red or blue points indicate whether the gene was detected or not detected respectively for each library.
+Expression of selected cytokinin-related genes (**a**), detected G1-Like (G1L/ALOG) genes (**b**) and *STMADS11*-like MADS genes (**c**) in inflorescence meristems in transcripts per million (TPM).
+A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
+*GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/SBM and SM than in RM and PBM.
 Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression.

### 2015-10-19 src/R/figures.R

@@ -364,9 +364,11 @@ tableCount <- incCount(tableCount, "t_hypergeom")
 
 svpMads <- c("LOC_Os02g52340", "LOC_Os03g08754", "LOC_Os06g11330")
 expG1s <- c("LOC_Os02g07030", "LOC_Os06g46030", "LOC_Os10g33780")
+ck <- c("LOC_Os01g51210","LOC_Os04g43840","LOC_Os01g40630", "LOC_Os01g10110")
 
 expTable <- rbind(data.table(msuId = svpMads, type = "mads"),
-data.table(msuId = expG1s, type = "g1l"))
+                  data.table(msuId = expG1s, type = "g1l"),
+                  data.table(msuId = ck, type = 'ck'))
 
 expTable[, symbol := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
 
@@ -411,21 +413,21 @@ alogPlot <- function(plotData) {
       theme_minimal(base_size = 8, base_family = "Helvetica") +
       theme(axis.text.x = element_text(vjust = 0.5),
             strip.text = element_text(face = "italic"),
-            plot.title = element_text(hjust = 0, face = "bold")
-#            , plot.background = element_rect(colour = "grey50")
-            ) +
+            plot.title = element_text(hjust = 0, face = "bold"),
+            plot.background = element_rect(colour = "grey50")) +
       scale_colour_manual(values = cols, guide = FALSE) +
-      facet_wrap(~symbol, ncol = 1) +
       xlab(NULL) +
       stat_smooth(se = FALSE, colour = "grey", size = 0.5) +
       geom_point(shape = 16, alpha = 0.7, position = position_jitter(height = 0, width = 0.3))
   )
 }
-  
-f_alogFamily_a <- alogPlot(plotData[type == 'g1l']) + ggtitle("a")
-f_alogFamily_b <- alogPlot(plotData[type == 'mads']) + ggtitle("b") + ylab(NULL)
 
-f_alogFamily <- gridExtra::grid.arrange(f_alogFamily_a, f_alogFamily_b, ncol = 2)
+
+f_alogFamily_a <- alogPlot(plotData[type == 'ck']) + ggtitle("a") + facet_wrap(~symbol, ncol = 2)
+f_alogFamily_b <- alogPlot(plotData[type == 'g1l']) + ggtitle("b") + facet_wrap(~symbol, ncol = 1)
[…]

### 2015-10-19 src/R/shuffledTpm.R

@@ -45,7 +45,8 @@ if (!dir.exists(tpmDir)) {
 
 # import GTF
 gtfFile <- list.files(shuffleDir, pattern = 'shuffled.gff3', full.names = TRUE)
-gtf <- import.gff(gtfFile, format = 'gff3', genome = 'Osativa_204_v7.0', asRangedData=F, feature.type="CDS")
+gtf <- import.gff(gtfFile, format = 'gff3', genome = 'Osativa_204_v7.0',
+                  feature.type="CDS")
 
 # reduce ranges by gene_name (MSU ID), i.e. merge overlapping exons
 grl <- reduce(split(gtf, elementMetadata(gtf)$ID))

### 2015-10-21 si.Rmd
Match expressed genes to DESeq2 coldata by library rather than for the whole table (fixes # 14)

@@
   rmarkdown::tufte_handout:
-    dev: "CairoPDF"
-  pdf_document:
-    fig_caption: yes
-    number_sections: yes
-    dev: "CairoPDF"
+    dev: CairoPDF
   BiocStyle::html_document:
+    dev: CairoPDF
     fig_caption: yes
@@
     highlight: pygments
+    number_sections: yes
+    toc: yes
+  pdf_document:
+    fig_caption: yes
     number_sections: yes
@@
                cache.path='cache/',
-#               fig.align = 'center', 
-               dev = "cairo_pdf",
+               fig.align = 'center', 
+#               dev=c("png", "cairo_pdf"),
                dpi=300,
@@
 
-``` {r st_libStats, results = "asis", fig.cap = I(pasteLabel("**Table S", s_tableCount, "st_libStats"))}
+``` {r st_libStats, results = "asis", fig.fullwidth = TRUE, fig.cap = I(pasteLabel("**Table S", s_tableCount, "st_libStats"))}
 pander::set.caption(paste0(
@@
   "**: Library and sequencing statistics."))
-pander::pandoc.table(st_libStats, split.tables = Inf, missing = "", round = 1, emphasize.strong.cols = 1,
+pander::pandoc.table(st_libStats,
+                     missing = "",
+                     round = 1,
+                     emphasize.strong.cols = 1,
+                     split.tables = Inf,
                      justify = c("right", rep("center", length(names(st_libStats)) - 1)))

### 2015-10-21 src/R/compareVsInSitu.R

@@ -89,7 +89,7 @@ expGenTT <- reshape2::melt(expGenTT.wide, id.vars = "id", variable.name = "lib",
 # add stage names from deseq2, but map "ePBM/SBM" to "SBM"
 colData <- data.table(as.data.frame(GenomicRanges::colData(
   readRDS('output/DESeq2/ddsLrt.Rds'))), keep.rownames = TRUE)
-expGenTT[, stage := colData[rn %in% as.character(lib), stage]]
+expGenTT[, stage := colData[rn %in% as.character(lib), stage], by = lib]
 expGenTT[, stage := plyr::mapvalues(stage, "ePBM/SBM", "SBM")]
 
 # call a gene expressed if it's detected in > 1 library per stage

### 2015-10-21 src/R/hbClasses.R

@@ -1,0 +1,61 @@
+#!/usr/bin/Rscript
+
+library(data.table)
+
+# HB annotations from Jain et al in doc format, of course
+fileUrl <- "http://onlinelibrary.wiley.com/store/10.1111/j.1742-4658.2008.06424.x/asset/supinfo/FEBS_6424_sm_Table%20S1.doc?v=1&s=e522a4da4ba65322a6ef5bfd597e4e09dbfdb977"
+temp <- tempfile()
+download.file(fileUrl, temp)
+
+# need a dictionary of class names
+hbClasses <- c("HD-ZIP I", "HD-ZIP II", "HD-ZIP III", "HD-ZIP IV", "BLH", 
+               "KNOX I", "KNOX II", "WOX", "ZF-HD", "PHD", "Unclassified")
+
+# read the data using catdoc (http://freecode.com/projects/catdoc)
+dirtyData <- system(paste("catdoc", temp), intern = TRUE)
+
+# remove blank lines
+dirtyData <- dirtyData[!dirtyData == ""]
+
+# find lines that match each class
+classHits <- sapply(hbClasses, grep, dirtyData)
+matchIndex <- data.table(
+  class = names(classHits),
+cStart = sapply(classHits, function(x) x[1])
+)
+matchIndex[, cStop := c(cStart[-1] - 1, NA)]
+matchIndex[class == "Unclassified", cStop := grep("References", dirtyData) - 1]
+
+# find genes in each set of lines
+retrieveGeneIds <- function(cStart, cStop) {
+  genes <- grep("Os\\d+g\\d+", dirtyData[cStart:cStop], value = TRUE)
+  return(unique(gsub(".*(Os\\d+g\\d+).*", "\\1", genes)))
+}
+
+# retrieve gene IDs for each class
+hbGenes <- matchIndex[, .(msuId = retrieveGeneIds(cStart, cStop)), by = class]
+
+# Jain et al. are confused about identifiers, fix
+hbGenes[, msuId := paste0("LOC_", msuId)]
+
+# add oryzabase gene names
+hbGenes[, symbol := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
+
+# make output folder
+outDir <- "output/homeobox"
+if (!dir.exists(outDir)) {
+  dir.create(outDir)
+}
+
+# SAVE OUTPUT
+saveRDS(hbGenes, paste0(outDir, "/hbGenes.Rds"))
+
+# SAVE LOGS
[…]

### 2015-10-23 ms.Rmd
Moved MADS figure to SI. Needs cleaning.

@@
 
-# Results and Discussion
+# Results
 
@@
 
+Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**).
+To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
+Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
+After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
+Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
+Auxin response factors are enriched in cluster `r getClusterName('dec', 'dec')`.
+
+## Dynamic expression of transcription factor families
+
+Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
+Unlike soft clustering, which groups genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
+There are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
+The most pronounced change in expression appears to occur between the RM and branch meristem samples.
+Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
+Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
+
+# Discussion
+
+## Changes between apical and axillary meristems
+
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
@@
 Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (see **ref next section**), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
+
+## Gradual changes in expression during axillary meristem transitions
+
+The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
+Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
+
+The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
+Of the G1-Like (G1L/ALOG) genes, which are annotated with the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj, only three were detected in the dataset.
+Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
+*G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff].
+This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*.
+Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, is not reported [@Yoshida:2013ff].
+*MADS22* and *MADS55* are also enriched in the RM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
+
+The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
+*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].
+Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
+WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O.
+sativa* ssp.
+*japonica* varieties.
+Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf].
+WFP also binds to the promoters of *LOG1*, *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].
+
+| **Shorten MADS section and insert here. MADS DEG figure goes to SI?**
+
+These patterns are not intended to represent each member of each gene family.
+For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
+An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
+*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
+In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.
+
+The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
+In the eudicot *A.
+thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
+In rice, functions homologous to those provided by *A.
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on?
+check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
+[**Add info about tomato MADS**].
+
+A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper.
+The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A.
+thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di].
+A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
+[**Note about relative genome size (number of genes) and MADS family size between species**.]
+
+As expected based on their functions in floral organ specification in rice and *A.
+thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
+However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
 
@@
 
-Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**).
-To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
-After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
-Auxin response factors are enriched in cluster `r getClusterName('dec', 'dec')`.
+## Dynamic expression of homeodomain genes
 
-The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
-Of the G1-Like (G1L/ALOG) genes, which are annotated with the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj, only three were detected in the dataset.
-Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
-Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
-*G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff].
-This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*.
-Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, is not reported [@Yoshida:2013ff].
-*MADS22* and *MADS55* are also enriched in the RM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
-
-The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
-Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
-
-## Dynamic expression of transcription factor families
-
-Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
-Unlike soft clustering, which groups genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
-There are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
-The most pronounced change in expression appears to occur between the RM and branch meristem samples.
-Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
-Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
-
-| **Shorten MADS section and insert here. MADS DEG figure goes to SI?**
-| **GRAS TFs (SHR1/SCR)? I think I will skip this because they are not highly expressed.**
-
-The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
-*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
-These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].
-Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
-WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O.
-sativa* ssp.
-*japonica* varieties.
-Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf].
-WFP also binds to the promoters of *LOG1*, *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].
-
-| **[Re]move following MADS section and insert section on Homeobox. Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007)**
-
-These patterns are not intended to represent each member of each gene family.
-For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
-An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
-*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
-In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.
-
-## Comparative analysis of MADS gene expression in eudicots and monocots
-
-The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
-In the eudicot *A.
-thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
-In rice, functions homologous to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on?
-check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
-[**Add info about tomato MADS**].
-
-* Something about MADS genes involved earlier in development?
-* Expression of MADS genes that interact with *G1L5*? (Maybe this is too much detail?)
-
-A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper.
-The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A.
-thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di].
-A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
-[**Note about relative genome size (number of genes) and MADS family size between species**.]
-
-As expected based on their functions in floral organ specification in rice and *A.
-thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
-However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
+Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007).
 
@@
 ----
-
-```{r f_madsTree}
-print(f_madsTree)
-```
-
-**`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`**.
-Differential expression of putatative MIKC proteins in *A.
-thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp.
-*japonica*.
-The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences.
-Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.
-The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-Where there is no coloured label, the adjusted *p*-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do].
-For rice, L~2~FC was estimated from the RM and SM samples reported here.
-For *A.
-thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di.
-The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.
-Clades are labelled following the nomenclature presented by @Arora:2007fa.
-
-----
-
 # `sessionInfo` (*hide*)

### 2015-10-23 si.Rmd
Moved MADS figure to SI. Needs cleaning.

--- 2015-09-30 si.Rmd
@@
   \usepackage{caption}
+  \usepackage{fixltx2e}
+  \usepackage{lscape}
+  \usepackage{rotating}
   \captionsetup[figure]{labelfont=bf}
   \renewcommand{\thefigure}{S\arabic{figure}}
+  \renewcommand{\tablename}{ \hfill \newline Table}
+  \renewcommand{\thetable}{S\arabic{table}}
+documentclass: tufte-handout
+classoption: a4paper
 output:
   rmarkdown::tufte_handout:
-    dev: "CairoPDF"
-  pdf_document:
-    fig_caption: yes
-    number_sections: yes
-    dev: "CairoPDF"
+    dev: CairoPDF
+    keep_tex: yes
   BiocStyle::html_document:
+    dev: CairoPDF
     fig_caption: yes
@@
                cache.path='cache/',
-#               fig.align = 'center', 
-               dev = "cairo_pdf",
+               fig.align = 'center', 
+#               dev=c("png", "cairo_pdf"),
                dpi=300,
                fig.show='hold',
-               fig.height = 3.150,
-               fig.width = 3.150,
-               out.width = 3.150,
-               out.height = 3.150,
+               fig.height = 4.213,
+               fig.width = 4.213,
+               crop = FALSE,
 #               cache=FALSE,
@@
 
-pander::panderOptions("table.caption.prefix", "")
+#pander::panderOptions("table.caption.prefix", "")
 source('src/R/figures.R')
@@
 
-``` {r st_libStats, results = "asis", fig.cap = I(pasteLabel("**Table S", s_tableCount, "st_libStats"))}
+``` {r st_libStats, results = "hide", fig.cap = I(pasteLabel("**Table S", s_tableCount, "st_libStats"))}
+
 pander::set.caption(paste0(
@@
   "**: Library and sequencing statistics."))
-pander::pandoc.table(st_libStats, split.tables = Inf, missing = "", round = 1, emphasize.strong.cols = 1,
+pander::pandoc.table(st_libStats,
+                     missing = "",
+                     round = 1,
+                     emphasize.strong.cols = 1,
+                     split.tables = Inf,
                      justify = c("right", rep("center", length(names(st_libStats)) - 1)))
+
 ```
 
-----
+\hfill\newline
+\hfill\newline
+\begin{longtable}[c]{@{}cc@{}}
+\caption{\protect\label{tab:st_libstats}Library and sequencing statistics.}
+\end{longtable}
 
-```{r st_reviewInSitu, results = "asis"}
+``` {r st_libstats2, results = "asis"}
+
+st_libStats_x <- as.data.frame(st_libStats)
+rownames(st_libStats_x) <- st_libStats_x$Library
+st_libStats_x$Library <- NULL
+st_libStats_x <- xtable::xtable(st_libStats_x)
+xtable::digits(st_libStats_x) <- c(0, rep(1, ncol(st_libStats_x) - 1), 0)
+#xtable::caption(st_libStats_x) <- "Library and sequencing statistics."
+
+st_libStats_x_print <- print(st_libStats_x,
+      caption.placement = "top",
+      comment = FALSE,
+      rotate.colnames = TRUE,
+      floating = TRUE,
+      floating.environment = "table*",
+      print.results = FALSE)
+
+cat(st_libStats_x_print)
+
+#cat(sub("\\begin{table*}[ht]\n", "\\begin{table*}[ht]\n\\caption{Library and sequencing statistics.} \n", st_libStats_x_print, fixed = TRUE))
+
+```
+
+\pagebreak
+
+```{r st_reviewInSitu, results = "hide"}
 pander::set.caption(paste0(
@@
 
-----
-
 ```{r sf_pca, fig.cap = paste0("Principal components analysis of transformed read counts for each library.
-The two libraries made from degraded RNA did not cluster with other libraries (see ", I(pasteLabel("Table S", s_tableCount, "st_libStats")), "), suggesting that the use of degraded RNA affected quantification of expression.
+The two libraries made from degraded RNA did not cluster with other libraries (see Table \\ref{tab:st_libstats}), suggesting that the use of degraded RNA affected quantification of expression.
 For this reason these two libraries were not included in subsequent analyses.")}
@@
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
-The two libraries made from degraded RNA did not cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA affected quantification of expression.
-For this reason these two libraries were not included in subsequent analyses.
+```{r sf_isGenesTpm, fig.height = 9.535, fig.cap = paste0("\\protect\\label{fig:sf_isGenesTpm}","Comparison of the LMD RNA-sequencing dataset with published \\emph{in situ} results.
+Expression value in transcripts per million (TPM) is shown on the \\emph{y}-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below.
+Grey shading indicates a meristem type for which expression was detected by \\emph{in situ} hybridisation.
+See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
+[\\textbf{As this is now detable-ized include a separate supplemantary table of references}] Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).")}
 
-----
+print(sf_isGenesTpm)
 
-```{r sf_isGenesTpm, fig.height = 8.661, fig.width = 6.614}
-print(sf_isGenesTpm)
 ```
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Quantitative expression values for the genes that were compared to published *in situ* results.
-The expression value in transcripts per million is shown on the *x*-axis.
-Points are coloured by replicate, and plotted as a triangle if the gene was above the expression cutoff or a circle otherwise.
-Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).
+\clearpage
 
-----
+```{r insituRefs, results = "asis"}
 
-```{r sf_mfuzzCentroids}
-print(sf_mfuzzCentroids)
+# refFrame_x <- print(xtable::xtable(refFrame,
+#                                    caption = "References for Figure \\ref{fig:sf_isGenesTpm}",
+#                                    label = "tab:refFrame_x"),
+#       type = "latex",
+#       floating = TRUE,
+#       floating.environment = "table",
+#       rotate.colnames = FALSE,
+#       print.results = FALSE)
+
+setnames(refTable, c("msuId", "plotLabel"), c("MSU name", "CSGNL symbol"))
+pander::set.caption("\\protect\\label{tab:refFrame_x}References for Figure \\ref{fig:sf_isGenesTpm}.")
+pander::pandoc.table(refTable, emphasize.cols = c(1,2))
 ```
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
-Eight cluster cores were used because the rate of change of the minimum centroid distance over *c* decreases after *c* = 8.
+\clearpage
 
-----
+```{r sf_mfuzzCentroids, fig.cap = paste0("Minimum centroid distance for 2–", maxClust, " cluster cores (\\emph{c}).
+Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8."), fig.scap = "Something", fig.scap = c("One plot", "Another plot")}
 
-``` {r sf_mfuzzPca}
+print(sf_mfuzzCentroids)
+
+```
+
+``` {r sf_mfuzzPca, fig.cap = paste0("Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
+Genes are coloured by the cluster to which they were assigned.
+The separation of genes by cluster on the MDS plot supports the use of eight clusters.")}
 print(sf_mfuzzPca)
@@
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
-Genes are coloured by the cluster to which they were assigned.
-The separation of genes by cluster on the MDS plot supports the use of eight clusters.
+```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of putatative MIKC proteins in \\emph{A.
+thaliana}, \\emph{Solanum lycopersicum} and \\emph{Oryza sativa} ssp.
+\\emph{japonica}.
+The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences.
+Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
+Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering \\citep{@Love:2014do}.", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
+For \\emph{A.
+thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by \\citep{Mantegazza:2014di}.
+The sympodial inflorescence meristem and floret meristem samples generated by \\citep{Park:2012dg} were used to estimate L\\textsubscript{2}FC values for tomato genes.
+Clades are labelled following the nomenclature presented by \\citep{Arora:2007fa}.")}
 
-----
+print(sf_madsTree)
+```
+
+\clearpage
 

### 2015-10-23 src/R/compareVsInSitu.R

@@ -10,6 +10,7 @@ if (!file.exists(zhangFile)) {
 }
 zhangGenes <- data.table(read.table(zhangFile, sep = "\t", header = TRUE,
                                     stringsAsFactors = FALSE))
+zhangGenes[, plotLabel := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
 
 # load expressed genes
 exprGenFile <- "output/expressedGenes/expressedGenesByLibrary.Rds"
@@ -28,7 +29,7 @@ if (!file.exists(ddsFile)) {
 dds <- readRDS(ddsFile)
 
 # split the zhang genes expression field.
- trim <- function(x) {gsub("^\\s+|\\s+$", "", x)}
+trim <- function(x) {gsub("^\\s+|\\s+$", "", x)}
 zhangGenes <- zhangGenes[,.(
   geneExp = trim(unlist(strsplit(geneExp, "[,]")))
   ), by = c('msuId', 'zhangRef', 'plotLabel')]

### 2015-10-23 src/R/figures.R

@@ -137,57 +137,118 @@ setcolorder(st_reviewInSitu, neworder =
 setkey(st_reviewInSitu, "Gene symbol")
 s_tableCount <- incCount(s_tableCount, "st_reviewInSitu")
 
-# tpm plots of some/all of these genes
-tpm <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
-setnames(tpm, 'rn', "MSU identifier")
-setkey(tpm, "MSU identifier")
-
-genes <- st_reviewInSitu[,.(`MSU identifier`, `Gene symbol`)]
-setkey(genes, "MSU identifier")
-genes <- unique(genes)
-genTpm <- tpm[genes]
-genTpm.long <- reshape2::melt(genTpm, id.vars = c('MSU identifier', 'Gene symbol'),
-                              variable.name = "lib", value.name = "Expression (transcripts per million)")
-
-# expression calls for these genes by library
-expGenTT <- data.table(readRDS('output/expressedGenes/expGenTT.Rds'), key = "id")
-expGenTT.wide <- expGenTT[genTpm[,`MSU identifier`]]
-expGenTT.long <- reshape2::melt(expGenTT.wide, id.vars = "id",
-                                variable.name = "lib", value.name = "expressed")
-
-setkey(genTpm.long, "MSU identifier", "lib")
-setkey(expGenTT.long, "id", "lib")
-
-plotData <- expGenTT.long[genTpm.long]
-
-# add stage names from deseq2
+# get long tpm data with stage
+tpm.wide <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
+setnames(tpm.wide, 'rn', "MSU identifier")
+tpm <- reshape2::melt(tpm.wide, id.vars = "MSU identifier",
+                      variable.name = "library", value.name = "tpm")
 colData <- data.table(as.data.frame(GenomicRanges::colData(
   readRDS('output/DESeq2/ddsLrt.Rds'))), keep.rownames = TRUE)
-plotData[, stage := colData[rn == as.character(lib), stage], by = lib]
+tpm[, stage := colData[rn %in% as.character(library), stage], by = library]
+
+setkey(tpm, "MSU identifier", "stage")
+
+# merge tpm data with compare calls
+genes <- compare[, .(msuId, stage, call.z)]
+setkey(genes, "msuId", "stage")
[…]

### 2015-10-23 si.Rmd
SI decrufted. Table of references for in situ comp figure (fixes #2)

@@
   \usepackage{fixltx2e}
-  \usepackage{lscape}
   \usepackage{rotating}
-  \captionsetup[figure]{labelfont=bf}
   \renewcommand{\thefigure}{S\arabic{figure}}
-  \renewcommand{\tablename}{ \hfill \newline Table}
   \renewcommand{\thetable}{S\arabic{table}}
@@
     keep_tex: yes
-  BiocStyle::html_document:
-    dev: CairoPDF
-    fig_caption: yes
-    force_captions: yes
-    highlight: pygments
-    number_sections: yes
-    toc: yes
-  pdf_document:
-    fig_caption: yes
-    number_sections: yes
-    dev: "CairoPDF"
-  word_document:
-    fig_caption: yes
 bibliography: bib/papersLib.bibtex
@@
                fig.align = 'center', 
-#               dev=c("png", "cairo_pdf"),
-               dpi=300,
+               dpi=600,
                fig.show='hold',
@@
                crop = FALSE,
-#               cache=FALSE,
                echo=FALSE,
@@
 
-#pander::panderOptions("table.caption.prefix", "")
 source('src/R/figures.R')
-
-```
-
-``` {r st_libStats, results = "hide", fig.cap = I(pasteLabel("**Table S", s_tableCount, "st_libStats"))}
-
-pander::set.caption(paste0(
-  I(pasteLabel("**Table S", s_tableCount, "st_libStats")),
-  "**: Library and sequencing statistics."))
-pander::pandoc.table(st_libStats,
-                     missing = "",
-                     round = 1,
-                     emphasize.strong.cols = 1,
-                     split.tables = Inf,
-                     justify = c("right", rep("center", length(names(st_libStats)) - 1)))
 
@@
 
-``` {r st_libstats2, results = "asis"}
-
+``` {r st_libstats, results = "asis"}
 st_libStats_x <- as.data.frame(st_libStats)
@@
 xtable::digits(st_libStats_x) <- c(0, rep(1, ncol(st_libStats_x) - 1), 0)
-#xtable::caption(st_libStats_x) <- "Library and sequencing statistics."
 
@@
 cat(st_libStats_x_print)
-
-#cat(sub("\\begin{table*}[ht]\n", "\\begin{table*}[ht]\n\\caption{Library and sequencing statistics.} \n", st_libStats_x_print, fixed = TRUE))
-
 ```
@@
 \pagebreak
-
-```{r st_reviewInSitu, results = "hide"}
-pander::set.caption(paste0(
-  I(pasteLabel("**Table S", s_tableCount, "st_reviewInSitu")),
-  "**: Comparison of the dataset with published *in situ* expression profiles.
-'×' indicates that the gene was detected by both methods, '+' indicates that the gene was detected by *in situ* hybridisation but not by RNA sequencing and '·' indicates that the gene was detected by RNA sequencing but expression was not reported by *in situ*.
-The last category is not intended to be informative, because it relies on negative or unreported results.
-A blank cell indicates that the gene was not detected by either method."))
-pander::pandoc.table(st_reviewInSitu, split.tables = Inf,
-                     justify = c("right", rep("centre", length(names(st_reviewInSitu)) - 2), "left"),
-                     emphasize.cols = c(1,2), missing = "")
-```
 
@@
 [\\textbf{As this is now detable-ized include a separate supplemantary table of references}] Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).")}
-
 print(sf_isGenesTpm)
-
 ```
@@
 ```{r insituRefs, results = "asis"}
-
-# refFrame_x <- print(xtable::xtable(refFrame,
-#                                    caption = "References for Figure \\ref{fig:sf_isGenesTpm}",
-#                                    label = "tab:refFrame_x"),
-#       type = "latex",
-#       floating = TRUE,
-#       floating.environment = "table",
-#       rotate.colnames = FALSE,
-#       print.results = FALSE)
-
 setnames(refTable, c("msuId", "plotLabel"), c("MSU name", "CSGNL symbol"))
@@
 Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8."), fig.scap = "Something", fig.scap = c("One plot", "Another plot")}
-
 print(sf_mfuzzCentroids)
-
 ```
@@
 Clades are labelled following the nomenclature presented by \\citep{Arora:2007fa}.")}
-
 print(sf_madsTree)

### 2015-10-26 ms.Rmd
Revert changes to MADS; want more changes visible

@@
 csl: csl/the-plant-journal.csl
-header-includes: \usepackage{graphicx} \usepackage[utf8]{inputenc} \DeclareUnicodeCharacter{2265}{$\geq$}
+header-includes:
+  \usepackage{graphicx}
+  \usepackage[utf8]{inputenc}
+  \DeclareUnicodeCharacter{2265}{$\geq$}
 output:
@@
     number_sections: yes
-    toc: yes
-  pdf_document:
-    fig_caption: yes
-    number_sections: yes
-  rmarkdown::tufte_handout: default
   word_document:
@@
     reference_docx: csl/wordStyles.docx
+    toc: yes
 bibliography: bib/papersLib.bibtex
@@
 
-Thomas W.R. Harrop^§^[^1], Israr Ud Din^¶^[^1], Veronica Gregis^¶^, Michela Osnato^¶^, Stefan Jouannic^§^, Hélène Adam^§^ and Martin Kater^¶^
+Thomas W.R. Harrop^§,1^, Israr Ud Din^¶,^[^1], Veronica Gregis^¶^, Michela Osnato^¶^, Stefan Jouannic^§^, Hélène Adam^§,2^ and Martin Kater^¶,^[^2]
 
 [^1]: These authors contributed equally to this work.
+[^2]: To whom correspondence should be addressed: martin.kater@unimi.it or helene.adam@ird.fr.
 
@@
 
-To investigate gene expression during inflorescence development, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
+## Genome-wide expression analysis of inflorescence development in rice
+
+To investigate gene expression during the development of the rice inflorescence, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
 RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
-**Need area cut +/- number of panicles used for each library from Israr**.
+**Number of panicles used for each library?**.
 RNA input for library preparation was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
@@
 Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage.
-As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported (`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))` and `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`), and four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (**Figure Sx**).
+As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported.
+Although there were several exceptions, most genes that were detected by *in situ* hybridisation were also detected in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
+Four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
 [**How do in situ / published results compare to transcriptome?**]
@@
 To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
-Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
-To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
-Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that have constant expression between RM and PBM and then increase or decrease in expression between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
+Detected genes were ranked by variance, and standardised, VST-transformed read counts for the 33% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
+To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
+Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between the apical meristem (RM) and axillary meristem (PBM–SM) samples (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`) and genes that change expression gradually over the course of changes in axillary meristem identity (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
 There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).
 
-Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**).
+Examination of the clustered genes suggested that transcription factor (TF) genes were overrepresented in most clusters (Data S1).
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
@@
 After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-Cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the PBM–SM samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TFs whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
-Auxin response factors are enriched in cluster `r getClusterName('dec', 'dec')`.
+For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TF genes whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
+The enrichment of different TF families in different clusters suggests complex control of the landscape of gene expression during inflorescence development.
 
@@
 Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
-Unlike soft clustering, which groups genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
-There are some notable modes of expression shared between different families that may indicate roles in inflorescence development.
-The most pronounced change in expression appears to occur between the RM and branch meristem samples.
-Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types.
+Unlike soft clustering, which was used to group genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
+Some  modes of expression shared between different families may suggest roles in inflorescence development.
+The most pronounced change in expression appears to occur between the RM and branch meristem samples (*i.e.* the change from apical to axillary meristem).
+Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are more highly expressed in RM than in the other meristem types.
 Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
@@
 
+Two results are presented here that suggest differing landscapes of expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
@@
 
+Overexpression of the G1-Like (G1L/ALOG) transcription factor *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
+Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
+*MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
+The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
+Of the G1L genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+*g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff].
+This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*.
+Although the increase in secondary branches linked to *GIL5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.
+
 ## Gradual changes in expression during axillary meristem transitions
 
-The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
-Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
+Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from axillary mersitem to spikelet meristem.
+The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
+In the eudicot *A.
+thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
+In rice, functions homologous to those provided by *A.
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on?
+check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
 
-The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
-Of the G1-Like (G1L/ALOG) genes, which are annotated with the Pfam domain PF04852 [@Goodstein:2012ho; @Finn:2014kj, only three were detected in the dataset.
-Two of them, *G1L2* and *G1L5* (*TAWAWA1*), were present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` respectively, and the third, *G1L1*, was not clustered but appears to have a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
-Overexpression of *G1L5* increases the production of secondary branches and leads to the formation of tertiary branches, potentially by suppressing the transition from BM to SM identity, resulting in higher grain yield [@Yoshida:2013ff].
-*G1L5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNAi [@Yoshida:2013ff].
-This might be explained by redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with all three of these genes being affected by the RNAi construct used against *G1L5*.
-Yeast two-hybid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, is not reported [@Yoshida:2013ff].
-*MADS22* and *MADS55* are also enriched in the RM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
+A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper.
+The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A.
+thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di].
+A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
+[**Note about relative genome size (number of genes) and MADS family size between species**.]
+
+As expected based on their functions in floral organ specification in rice and *A.
+thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
+However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
+
+These patterns are not intended to represent each member of each gene family.
+For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
+An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
+*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
+In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.
 
@@
 
-| **Shorten MADS section and insert here. MADS DEG figure goes to SI?**
-
-These patterns are not intended to represent each member of each gene family.
-For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
-An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
-*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
-In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.
-
-The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
-In the eudicot *A.
-thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
-In rice, functions homologous to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on?
-check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
-[**Add info about tomato MADS**].
-
-A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper.
-The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A.
-thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di].
-A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
-[**Note about relative genome size (number of genes) and MADS family size between species**.]
-
-As expected based on their functions in floral organ specification in rice and *A.
-thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
-However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
+The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
+Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
 
@@
 
-# Short legends for Supporting Information
+# Supporting Information
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`**: Placeholder for the *in situ* results figure.
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Comparison of the LMD RNA-sequencing dataset with published *in situ* results.
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`**: Assessment of clustering results.
+
+**`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`**: Differential expression
+of genes encoding putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*.
 
@@
 
-**`r I(pasteLabel("Table S", s_tableCount, "st_reviewInSitu"))`**: Comparison of the dataset with published *in situ* expression profiles.
+**`r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`**: References for `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`.
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`**: Principal components analysis of transformed read counts for each library.
-
-**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Quantitative expression values for the genes that were compared to published *in situ* results.
-
-**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzCentroids"))`**: Minimum centroid distance for 2–`r maxClust` cluster cores (*c*).
-
-**`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPca"))`**: Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
+**Data S1**: List of clustered genes with annotations.
 

### 2015-10-26 si.Rmd
Revert changes to MADS; want more changes visible

@@
     keep_tex: yes
+    fig_crop: no
 bibliography: bib/papersLib.bibtex
@@
 ```{r Setup, include=FALSE, results="hide", warning=FALSE}
-
 knitr::opts_chunk$set(fig.path='figures/',
@@
                fig.width = 4.213,
-               crop = FALSE,
                echo=FALSE,
@@
 source('src/R/figures.R')
-
 ```
 
-\hfill\newline
-\hfill\newline
-\begin{longtable}[c]{@{}cc@{}}
-\caption{\protect\label{tab:st_libstats}Library and sequencing statistics.}
-\end{longtable}
+```{r sf_pca, fig.cap = paste0("Principal components analysis of transformed read counts for each library.
+The two libraries made from degraded RNA did not cluster with other libraries (see Table \\ref{tab:st_libstats}), suggesting that the use of degraded RNA affected quantification of expression.
+For this reason these two libraries were not included in subsequent analyses.")}
+print(sf_pca)
+```
+
+\clearpage
+\captionof{table}{\protect\label{tab:st_libstats}Library and sequencing statistics.}
 
@@
 st_libStats_x <- as.data.frame(st_libStats)
+# colnames(st_libStats_x) <- gsub("Uniquely mapped reads (M)",
+#                                 "Uniquely mapped reads (M)",
+#                                 colnames(st_libStats_x),
+#                                 fixed = TRUE)
 rownames(st_libStats_x) <- st_libStats_x$Library
 st_libStats_x$Library <- NULL
+st_libStats_x$Sample <- sub("ePBM/SBM", "SBM¹", st_libStats_x$Sample, fixed = TRUE)
 st_libStats_x <- xtable::xtable(st_libStats_x)
-xtable::digits(st_libStats_x) <- c(0, rep(1, ncol(st_libStats_x) - 1), 0)
-
+xtable::align(st_libStats_x) <- rep("r", ncol(st_libStats_x) + 1)
+xtable::digits(st_libStats_x) <- c(rep(0, 3), 3, rep(1, 11), 0)
 st_libStats_x_print <- print(st_libStats_x,
@@
 
+# add more space to the longest colname
+#st_libStats_x_print <- sub("Uniquely mapped reads (M)", "Uniquely mapped reads (M) \\hspace{3 mm}", st_libStats_x_print, fixed = TRUE)
+# OR remove the first hline?
+st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
 cat(st_libStats_x_print)
+cat("¹ ePBM/SBM")
 ```
 
-\pagebreak
-
-```{r sf_pca, fig.cap = paste0("Principal components analysis of transformed read counts for each library.
-The two libraries made from degraded RNA did not cluster with other libraries (see Table \\ref{tab:st_libstats}), suggesting that the use of degraded RNA affected quantification of expression.
-For this reason these two libraries were not included in subsequent analyses.")}
-print(sf_pca)
-```
+\begin{figure*}
+ \includegraphics{data/lmdFigure.pdf}
+\caption{Placeholder for the \emph{in situ} results figure.}
+\end{figure*}
 
@@
 See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
-[\\textbf{As this is now detable-ized include a separate supplemantary table of references}] Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).")}
+Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).", "\\newline ", "\\textbf{Do we need to discuss the results of individual genes?}")}
 print(sf_isGenesTpm)
@@
 \clearpage
+\captionof{table}{\protect\label{tab:refFrame_x}References for Figure \ref{fig:sf_isGenesTpm}.}
 
 ```{r insituRefs, results = "asis"}
-setnames(refTable, c("msuId", "plotLabel"), c("MSU name", "CSGNL symbol"))
-pander::set.caption("\\protect\\label{tab:refFrame_x}References for Figure \\ref{fig:sf_isGenesTpm}.")
-pander::pandoc.table(refTable, emphasize.cols = c(1,2))
+pander::pandoc.table(st_refTable, emphasize.cols = c(1,2))
 ```
 
-\clearpage
-
-```{r sf_mfuzzCentroids, fig.cap = paste0("Minimum centroid distance for 2–", maxClust, " cluster cores (\\emph{c}).
-Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8."), fig.scap = "Something", fig.scap = c("One plot", "Another plot")}
-print(sf_mfuzzCentroids)
+```{r sf_mfuzzPcaCentroids, fig.height = 6.741, fig.cap = paste0("Assessment of clustering results.
+\\textbf{a}.
+Minimum centroid distance for 2–", maxClust, " cluster cores (\\emph{c}).
+Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8.
+\\textbf{b}.
+Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
+Genes are coloured by the cluster to which they were assigned.
+The separation of genes by cluster on the MDS plot supports the use of eight clusters.")}
+grid::grid.draw(sf_mfuzzPcaCentroids)
 ```
 
-``` {r sf_mfuzzPca, fig.cap = paste0("Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
-Genes are coloured by the cluster to which they were assigned.
-The separation of genes by cluster on the MDS plot supports the use of eight clusters.")}
-print(sf_mfuzzPca)
-```
-
-```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of putatative MIKC proteins in \\emph{A.
+```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putatative MIKC proteins in \\emph{A.
 thaliana}, \\emph{Solanum lycopersicum} and \\emph{Oryza sativa} ssp.
@@
 Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering \\citep{@Love:2014do}.", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
+Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering \\citep{@Love:2014do}.", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
 For \\emph{A.

### 2015-10-26 src/R/figures.R

@@ -406,7 +406,7 @@ sf_mfuzzPca <- ggplot(vg.mds, aes(x = MDS1, y=MDS2, colour = factor(cluster))) +
 sf_mfuzzPcaCentroids <- gridExtra::arrangeGrob(
   sf_mfuzzCentroids,
   sf_mfuzzPca,
-  padding = unit(0, "lines"), ncol = 2)
+  padding = unit(0, "lines"), ncol = 1)
 
 s_figCount <- incCount(s_figCount, "sf_mfuzzPcaCentroids")
 

### 2015-10-26 src/R/madsCompTree.R

@@ -34,15 +34,15 @@ if (!dir.exists(outDir)) {
 # get DESeq2 results
 ddsOs <- readRDS(ddsOsFile)
 resOs <- data.table(as.data.frame(
-  DESeq2::results(ddsOs, contrast = c("stage", "SM", "RM"), lfcThreshold = 0.5, alpha = 0.1)),
+  DESeq2::results(ddsOs, contrast = c("stage", "SM", "RM"))),
   keep.rownames = TRUE, key = "rn")
 ddsAt <- readRDS(ddsAtFile)
 resAt <- data.table(as.data.frame(
-  DESeq2::results(ddsAt, contrast = c("stage", "FM", "IM"), lfcThreshold = 0.5, alpha = 0.1)),
+  DESeq2::results(ddsAt, contrast = c("stage", "FM", "IM"))),
   keep.rownames = TRUE, key = "rn")
 ddsSl <- readRDS(ddsSlFile)
 resSl <- data.table(as.data.frame(
-  DESeq2::results(ddsSl, contrast = c("stage", "fm", "sim"), lfcThreshold = 0.5, alpha = 0.1)),
+  DESeq2::results(ddsSl, contrast = c("stage", "fm", "sim"))),
   keep.rownames = TRUE, key = "rn")
 
 # make DESeq results data.table

### 2015-10-26 src/R/parseQuantStats.R

@@ -39,7 +39,10 @@ RIN <- c(7.08,6.1,8,7.6, # N1
          7.05,5.4,8,8.7, # N2
          7,7.05,8.1, # N3
          7.08,7.05,7.0) # N4
-
+area <- c(727647,511191,300886,848070, # N1
+          600176,2735798,394641,439534, # N2
+          1563411,1574719,1109344, # N3
+          975315,758919,2347647) # N4
 
 # function to get required info from STAR output
 parseStarInfo <- function(x){
@@ -94,7 +97,7 @@ if (!file.exists(expGenFile)) {
 expGen <- readRDS(expGenFile)
 expGenPerLib <- as.data.table(sapply(expGen, function(x) length(unique(x))),
                               keep.rownames = TRUE)
-setnames(expGenPerLib, c('lib', "Expressed genes"))
+setnames(expGenPerLib, c('lib', "Detected genes"))
 
 # merge
 setkey(starInfo, 'lib')
@@ -103,6 +106,7 @@ setkey(readsPerLib, 'lib')
 setkey(rnaStats, 'lib')
 libStats <- expGenPerLib[readsPerLib][rnaStats[starInfo]]
 libStats[, "Yield (ng)" := round(inputRNA, 1)][, RIN := RIN ]
+libStats[, "Dissected area (mm²)" := area/1000000] 
 
 # make all columns except 'lib' numeric
 libStats <- libStats[, lapply(.SD, as.numeric), by = lib]
@@ -114,9 +118,18 @@ libStats[, "tRNA (%)" := round(`tRNA reads (M)`/`Reads (M)`*100, 1)]
 # library names in upper case!
 libStats[, lib := toupper(lib)]
 
+# stage and replicate number
+libStats[, Sample := levels(GenomicRanges::colData(dds)$stage)[
+  as.integer(gsub("N([[:digit:]+]).*", "\\1", lib))]]
+libStats[, Replicate := as.integer(gsub(".*R(\\d+)", "\\1", lib))]
+
 # arrange table
-paste(colnames(libStats)[c(1, 10, 6, 4,11,5,12,7:9,3,2)], collapse = "', '")
-setcolorder(libStats, c('lib', 'Yield (ng)', 'RIN', 'Reads (M)', 'rRNA reads (M)', 'rRNA (%)', 'tRNA reads (M)', 'tRNA (%)', 'Uniquely mapped reads (M)', 'Multimapped reads (M)', 'Total mapping percentage', 'Reads in genes (M)', 'Expressed genes'))
[…]

### 2015-10-27 ms.Rmd
Add homeobox figure

@@
 
-Two results are presented here that suggest differing landscapes of expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
+Two results are presented here that suggest significant changes in expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
@@
 thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db].
-GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated with reduced auxin signalling.
-Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (see **ref next section**), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
+GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling.
+Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
 
-Overexpression of the G1-Like (G1L/ALOG) transcription factor *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
+Overexpression of the G1-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
 Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
@@
 This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*.
-Although the increase in secondary branches linked to *GIL5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.
+Although the increase in secondary branches linked to *G1L5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.
 
@@
 
-Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from axillary mersitem to spikelet meristem.
+The patterns of differential expression of TF genes presented in `r I(pasteLabel("Figure ", figCount, "f_gsea"))` are not intended to represent each member of each gene family.
+For example,  the test statistic for the whole MADS family was higher in the SM than in the RM, but there are several MADS genes that have lower expression in the SM than in the RM, such as *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
+*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
+In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of MADS genes in the SM.
+
+Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM.
 The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
@@
 check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
+Two recent studies in *A.
+thaliana* and the domesticated tomato, *Solanum lycopersicum*, used LMD and RNA sequencing to study inflorescence development [@Park:2012dg; @Mantegazza:2014di].
+Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
+*A.
+thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems.
+*AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [**ref Huang 2012**].
+The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems.
+Expression of the putatative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage.
+In contrast, consistent with their functions in floral organ specification in rice and *A.
+thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
 
-A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper.
-The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A.
-thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di].
-A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`).
-[**Note about relative genome size (number of genes) and MADS family size between species**.]
-
-As expected based on their functions in floral organ specification in rice and *A.
-thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
-However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].
-
-These patterns are not intended to represent each member of each gene family.
-For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM.
-An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
-*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
-In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.
-
-The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development.
-*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM.
+The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development.
+*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM, suggesting a redundant function in indeterminate axillary meristems.
 These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].
@@
 Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf].
-WFP also binds to the promoters of *LOG1*, *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].
+WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].
 
-The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
-Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
-
-A homolog of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM.
+A homolog of *LOG1*, *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM.
 In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**].
@@
 *LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples.
-Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation from RM to SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
 Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level.
@@
 Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007).
+
+## Orphan text (don't know where to put, consider deleting)
+
+The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
+Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
 
@@
 ----
+
+```{r f_hb, fig.height = 8.661}
+print(f_hb)
+```
+
+**`r I(pasteLabel("Figure ", figCount, "f_hb"))`**.
+Patterns of expression of homeobox genes in the meristem samples.
+The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [**ref Jain et al.
+2008**].
+There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples.
+The former group contains *OSH1* (**function?**), whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes.
+
+----
 # `sessionInfo` (*hide*)

### 2015-10-27 src/R/homeobox.R

@@ -1,0 +1,119 @@
+#!/usr/bin/Rscript
+
+library(data.table)
+
+tfdbFile <- 'data/tfdb/os.Rds'
+hbGenesFile <- 'output/homeobox/hbGenes.Rds'
+expressedGenesAllFile <- 'output/expressedGenes/expressedGenesAll.Rds'
+vstFile <- 'output/DESeq2/vst.Rds'
+
+# check files exist
+lapply(list(tfdbFile, expressedGenesAllFile, vstFile, hbGenesFile), function(x)
+  if(!file.exists(x)) {
+    cat(x, " not found, exiting\n", file = stderr())
+    quit(save = "no", status = 1)
+  })
+
+# vst geometric means
+vst <- readRDS(vstFile)
+gm_mean <- function(x, na.rm = TRUE) {
+  exp(sum(log(x[x > 0]), na.rm = na.rm)/length(x))
+}
+vstMeans.matrix <- sapply(levels(vst$stage), function(x)
+  apply(GenomicRanges::assay(vst[, vst$stage == x]), 1, gm_mean))
+
+# homeobox classes
+hbGenes <- readRDS(hbGenesFile)
+
+# tfdb
+tfdb <- readRDS(tfdbFile)
+
+# add tfdb Hbs missing from Jain et. al data
+missingHbs <- tfdb[Family == "HB", Protein.ID[which(!(Protein.ID %in% hbGenes[, msuId]))]]
+oryzr::LocToGeneName(missingHbs)
+if (length(missingHbs) >0 ){
+  warning(paste(
+    "The following msuIds were not found in the Jain et. al data:\n",
+    paste(missingHbs, collapse = ", "),
+    "\nThey will be categorized as \"Not annotated\""))
+  notAnn <- data.table(
+    class = "Not annotated",
+    msuId = missingHbs
+  )
+  notAnn[, symbol := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
+  hbGenes <- rbind(hbGenes, notAnn)
+}
+
+# expressed genes
+expressedGenesAll <- readRDS(expressedGenesAllFile)
+
+# expressed homeobox genes
+expHb <- hbGenes[msuId %in% expressedGenesAll] 
+
+# get vst read counts for expHb
+hbVst <- vstMeans.matrix[expHb[, msuId],]
+
+# scale the VST read counts
+hbVstScaled <- t(scale(t(hbVst), center = TRUE))
+
+# choose a cluster method
[…]

### 2015-10-27 src/py/pipeline.py

@@ -541,6 +541,20 @@ def madsCompTree(inputFiles, outputFiles):
     touch(outputFiles)
 
 #---------------------------------------------------------------
+# make dendogram of MADS genes
+#
+@merge([hbClasses_R, download_tfdb, detect_expressed_genes, run_deseq2_os], 'ruffus/hb.heatmap')
+def homeobox_R(inputFiles, outputFiles):
+    jobScript = 'src/R/homeobox.R'
+    ntasks = '1'
+    cpus_per_task = '1'
+    job_name = 'homeobox'
+    jobId = submit_job(jobScript, ntasks, cpus_per_task, job_name)
+    # update ruffus flag
+    print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
+    touch(outputFiles)
+
+#---------------------------------------------------------------
 # FIGURES AND TABLES
 #---------------------------------------------------------------
 
@@ -611,6 +625,13 @@ def sf_isGenesTpm(inputFiles, outputFiles):
 #
 @merge([calculate_tpm, detect_expressed_genes], "ruffus/figure.f_alogFamily")
 def f_alogFamily(inputFiles, outputFiles):
+    touch(outputFiles)
+
+#---------------------------------------------------------------
+# homeobox figure
+#
+@merge(homeobox_R, "ruffus/figure.f_hb")
+def f_hb(inputFiles, outputFiles):
     touch(outputFiles)
 
 

### 2015-10-27 ms.Rmd
Added intro

@@
 
-## Rice inflorescence development
-## Branching & grain yield
-## Sequential changes in meristem activity
-## Discuss a few key genes?
+Agricultural development is essential to ensure food production and security for a growing population, particularly in developing countries [@Borlaug:2007cp].
+Rice is a staple food for over half the world's population, including a large number of developing countries.
+A sustainable increase in the production of rice against a backdrop of climate change and diminishing water and land availability will require plants with improved grain output, and the establishment of high-yield varieties is a major goal of modern breeding programs [@Peng:2008jb].
+Rice yield is a complex trait influenced by genetic factors, as well as environmental and other epigenetic factors, and is progressively defined during the life cycle of the plant, first during the vegetative phase, where the number of fertile tillers is established, and then during the reproductive phase and grain-filling phase.
+Yield is directly dependent on the architecture of the inflorescence [@Ikeda:2004tk], which is the flower-bearing structure.
+The rice inflorescence (or panicle) is a branched structure composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
+Spikelets are short branches that form on other branches of the panicle and bear a single flower.
+They are the primary unit of grass inflorescence, with one spikelet bearing one floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo].
+
+During panicle development, the undifferentiated cells in the primary axis (i.e.
+the rachis meristem, RM) undergo transitions through a series of meristem identities.
+After the establishment of the rachis some RM cells differentiate into branch meristem (BM) **PBM/SBM** for the formation and elongation of primary branches and the establishment of higher-order branches.
+The axillary BM cells differentiate into spikelet meristems (SM), which develop into floret meristems (FM) and produce flowers.
+After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].
+
+The landscape of gene expression is a characteristic factor that differentiates meristem cell types, and is involved in the control of transitions between meristem identities and therefore panicle phenotype.
+Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
+Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq].
+A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], and some of them may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].
+
+Understanding of the events that occur during development that determine panicle characteristics such as branching complexity and its plasticity will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
+Despite advances in characterization at the level of individual genes or interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia].
+One step towards this goal is to describe the changes in gene expression that occur during development.
+In this report, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of rice panicle development, using a sensitive LASER microdissection and RNA sequencing approach.
 
@@
 
-## Dynamic expression of homeodomain genes
+## Patterns of expression of homeodomain genes
 
 Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007).
+
+## Summary
+
+This report describes the changes in gene expression that occur during the early stages of rice panicle development.
+A switch in expression between apical and axillary meristems followed by gradual changes between PBM and SM were observed, based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling.
+These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
 
@@
 
+**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Comparison of the LMD RNA-sequencing dataset with published *in situ* results.
+
 **`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`**: Placeholder for the *in situ* results figure.
-
-**`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Comparison of the LMD RNA-sequencing dataset with published *in situ* results.
 
@@
 
+```{r f_gsea, fig.height = 8.661 / 2}
+print(f_gsea)
+```
+
+**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**.
+Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families.
+Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset.
+The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.
+Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.
+
+----
+
 ```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2}
@@
 Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
-
-----
-
-```{r f_gsea, fig.height = 8.661 / 2}
-print(f_gsea)
-```
-
-**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**.
-Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families.
-Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset.
-The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.
-Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.
 

### 2015-10-27 src/R/figures.R

@@ -93,12 +93,6 @@ sf_pca <- ggplot(data = pcaPlotData,
 
 s_figCount <- incCount(s_figCount, "sf_pca")
 
-######################
-### IN SITU FIGURE ###
-######################
-
-s_figCount <- incCount(s_figCount, "sf_inSitu")
-
 #######################
 ### COMPARE IN SITU ###
 #######################
@@ -258,6 +252,12 @@ sf_isGenesTpm <- ggplot() +
   facet_wrap(~ plotLabel, scales = "free_y", ncol = 2)
 
 s_figCount <- incCount(s_figCount, "sf_isGenesTpm")
+
+######################
+### IN SITU FIGURE ###
+######################
+
+s_figCount <- incCount(s_figCount, "sf_inSitu")
 
 ##################
 ### LMD FIGURE ###
@@ -433,6 +433,43 @@ t_hypergeom[is.na(t_hypergeom)] <- ""
 tableCount <- incCount(tableCount, "t_hypergeom")
 
 ############
+### GSEA ###
+############
+
+gsea <- readRDS('output/gsea/gseaTable.Rds')
+famCat <- readRDS("data/tfdb/famCat.Rds")
+
+# format some labels
+setnames(gsea, old = "Test statistic", new = "Test\nstatistic")
+gsea[, Stage := plyr::mapvalues(Stage, "ePBM/SBM", "ePBM/\nSBM")]
+
+# separate by TF / other proteins
+setkey(famCat, 'Family')
+setkey(gsea, 'rn')
+gsea[, Category := famCat[gsea][,Category]]
+gsea[, Category := plyr::mapvalues(Category, from = c("TF", "Other"),
+                                   to = c("Transcription factors", "Other regulators"))]
+gsea[, Category := factor(Category, levels = c("Transcription factors", "Other regulators"))]
+
+heatscale <- rev(RColorBrewer::brewer.pal(6, "RdBu"))
+
+f_gsea <- ggplot(gsea, aes(x = Stage, y = rn, label = padj, fill = `Test\nstatistic`)) +
+  theme_minimal(base_size = 8, base_family = "Helvetica") +
+  xlab(NULL) + ylab(NULL) +
+  theme(legend.key.size = grid::unit(8, "point"),
+        axis.text.x = element_text(vjust = 0.5)) +
+  scale_y_discrete(expand = c(0,0)) +
+  scale_x_discrete(expand = c(0,0)) +
[…]

### 2015-10-28 figures.Rmd
Compile figures into single pdf

---
title: "Figures"
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage{caption}
  \usepackage{fixltx2e}
  \usepackage{rotating}
  \renewcommand{\thefigure}{S\arabic{figure}}
  \renewcommand{\thetable}{S\arabic{table}}
documentclass: tufte-handout
classoption: a4paper
output:
  rmarkdown::tufte_handout:
    dev: CairoPDF
    keep_tex: yes
    fig_crop: no
bibliography: bib/papersLib.bibtex
---

```{r Setup, include=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               fig.align = 'center', 
               dpi=600,
               fig.show='hold',
               fig.width=3.150,
               fig.height=8.661,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

source('src/R/figures.R')
```

\begin{figure*}
 \includegraphics{data/lmdFigure.pdf}
\caption{Laser microdissection of meristem samples in the early stages of inflorescence development. Samples were collected from the rachis meristem (\textbf{a} and \textbf{e}), the primary branch meristem (\textbf{b} and \textbf{f}), the elongating primary branch meristem and secondary branch meristem (\textbf{c} and \textbf{g}) and the spikelet meristem (\textbf{d} and \textbf{h}). Images show the samples before (\textbf{a}–\textbf{d}) and after (\textbf{e}–\textbf{h}) dissection.}
\end{figure*}

```{r f_mfuzzClusters, fig.height = 8.661 / 2, fig.cap = paste0("Common patterns of gene expression in developing inflorescences. Fuzzy \\emph{c}-means clustering of normalised, variance-stabilised read counts was used to recover ", as.character(english::english(c)), " common patterns of expression. Each line describes the expression pattern of one gene, with the gene's membership to the cluster represented by the colour of the line. The core values for each cluster are plotted in black. The ordering of the panels in the plot was rearranged to enable side-by-side comparison of complementary clusters.")}
print(f_mfuzzClusters)
```

```{r f_gsea, fig.height = 8.661 / 2, fig.cap = "Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families. Test statistics were calculated from the L\\textsubscript{2}FC values between the stage of interest and all other stages for the expressed genes in each geneset. The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion. Although some test statistics had low \\emph{p}-values none were lower than 0.10 after Benjamini-Hochberg correction."}
print(f_gsea)
```

```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2, fig.cap = "Expression of selected cytokinin-related genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}) and \\emph{STMADS11}-like MADS genes (\\textbf{c}) in inflorescence meristems in transcripts per million (TPM). A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected. \\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/SBM and SM than in RM and PBM. Only three G1L genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, but they share a similar pattern of expression. Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library."}
# set layout
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(5, 2)))

# print the plots
print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:2, layout.pos.col = 1:2))
print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 1))
print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 2))
```

```{r f_hb, fig.height = 8.661, fig.cap = "Patterns of expression of homeobox genes in the meristem samples. The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the \\emph{y}-axis shows the homeobox clade for each gene [@Jain:2008kg]. There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples. The former group contains \\emph{OSH1}, whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes."}
print(f_hb)
```

### 2015-10-28 ms.Rmd
Send draft

--- 2015-10-27 ms.Rmd
@@
 output:
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
+    toc: yes
   BiocStyle::html_document:
@@
     number_sections: yes
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
-    toc: yes
 bibliography: bib/papersLib.bibtex
@@
 
-## Rice inflorescence development
-## Branching & grain yield
-## Sequential changes in meristem activity
-## Discuss a few key genes?
+Agricultural development is essential to ensure food production and security for a growing population [@Borlaug:2007cp].
+Rice is a staple food for over half the world's population, including many developing countries, and a sustainable increase in the production of rice under the constraints of a changing climate and diminishing water and land availability will require plants with improved grain output, making the establishment of high-yield rice varieties a goal of modern breeding programs [@Peng:2008jb].
+Rice yield is a complex trait influenced by genetic and epigenetic factors, and is progressively defined during the life cycle of the plant, first during the vegetative phase, where the number of fertile tillers is established, and then during the reproductive phase and grain-filling phase.
+Yield is directly dependent on the architecture of the inflorescence [@Ikeda:2004tk].
+The rice inflorescence (or panicle) is a branched structure composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
+Spikelets are short branches that form on other branches of the panicle and bear a single flower.
+They are the primary unit of grass inflorescence, with one spikelet bearing one floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo].
+
+During panicle development, the undifferentiated cells in the primary axis (*i.e.* the rachis meristem, RM) undergo transitions through a series of meristem identities.
+After the establishment of the rachis some RM cells differentiate into branch meristem (BM) for the formation and elongation of primary branches and the establishment of higher-order branches.
+The axillary BM cells differentiate into spikelet meristems (SM), which develop into floret meristems (FM) and produce flowers.
+After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].
+
+The landscape of gene expression is a characteristic factor that differentiates meristem cell types, and is involved in the control of transitions between meristem identities and therefore panicle phenotype.
+Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
+Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq].
+A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], some of which may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].
+
+Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
+Despite advances in characterization at the level of individual genes and interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia].
+One step towards this goal is to describe the changes in gene expression between different cell types during development.
+In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of rice panicle development, using a sensitive laser microdissection and RNA sequencing approach.
 
@@
 
-To investigate gene expression during the development of the rice inflorescence, LASER microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
+To investigate gene expression during the development of the rice inflorescence, laser microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
 RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
-**Number of panicles used for each library?**.
-RNA input for library preparation was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
+RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`; **Number of panicles used for each library?**).
 Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
@@
 Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage.
-As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes were compared with patterns that had previously been reported.
+As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes was compared with patterns that had previously been reported.
 Although there were several exceptions, most genes that were detected by *in situ* hybridisation were also detected in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
 Four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
-[**How do in situ / published results compare to transcriptome?**]
+**How do in situ results compare to transcriptome?**
 
@@
 
-To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do].
-Detected genes were ranked by variance, and standardised, VST-transformed read counts for the 33% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
+To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do].
+Detected genes were ranked by variance, and standardised, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` most-variable genes (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
@@
 After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TF genes whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
+For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TF genes, whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
 The enrichment of different TF families in different clusters suggests complex control of the landscape of gene expression during inflorescence development.
@@
 
-## Changes between apical and axillary meristems
+## Switch in gene expression between apical and axillary meristems
 
-Two results are presented here that suggest significant changes in expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
+Two results are presented here that suggest significant changes in gene expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
@@
 
-Overexpression of the G1-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
+Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
 Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
@@
 
-## Gradual changes in expression during axillary meristem transitions
+## Gradual changes in expression during transitions in axillary meristem identity
 
@@
 thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
-In rice, functions homologous to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on?
-check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].
+In rice, functions similar to those provided by *A.
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**ref Nagasawa 2003**; **ref Yao 2008**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**ref kyozuka 2002**;  **ref Yamaguchi 2006**].
 Two recent studies in *A.
@@
 thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems.
-*AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [**ref Huang 2012**].
+*AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de].
 The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems.
@@
 Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
-WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O.
+WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O.
 sativa* ssp.
@@
 Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf].
-WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].
+WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].
 
 A homolog of *LOG1*, *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM.
-In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**].
+In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by @Han:2014dt].
 The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go].
@@
 
-## Dynamic expression of homeodomain genes
+## Patterns of expression of homeodomain genes
 
-Can link via WUS (HB, target of cytokinin, e.g. Han mini-review) and/or OSH1 (HB, downregulated in LOG mutant, e.g. Kurakawa 2007).
+*log1* mutants also appear to have reduced expression of the *HOMEOBOX 1* (*OSH1*) gene [@Kurakawa:2007go], which encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor.
+*osh1* mutants have defects in SAM maintenance during vegetative growth and lower expression of other *KNOX* genes, and expression of *KNOX* genes can be induced by cytokinin treatment during shoot regeneration [@Tsuda:2011gz].
+In *A.
+thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [**ref Leibfried 2005**; **ref Lindsay, 2006**; @Gordon:2009fp].
+**Can expand this section**.
 
-## Orphan text (don't know where to put, consider deleting)
+Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were found in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems.
+Examining the dynamics of expression of all homeobox genes revealed that a prominent group follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1*.
+**Anything about functions of these genes?** There are also smaller groups of homeobox genes that peak in expression in the SM or RM samples, suggesting multiple roles and possible redundancy for homeodomain proteins during inflorescence development.
+**Can expand this section**.
 
-The AP2 gene *SUPERNUMERARY BRACT* (*SNB*) is involved in the transition between SM and floret meristem (FM) that occurs later in panicle development [@Lee:2007cj], but is coexpressed with *G1L5* in cluster `r getClusterName('dec', 'dec')` suggesting that it may also be involved in earlier transitions.
-Another AP2 gene, *FRIZZY PANICLE*, which is required for determination of SM [@Komatsu:2003ko], was found in clusters containing genes that are highly expressed in the SM.
+## Summary
+
+This report describes the changes in gene expression that occur during the early stages of rice panicle development.
+A switch in expression between apical and axillary meristems followed by gradual changes between PBM and SM were observed, based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling.
+These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
 
@@
 
-## Tissue embedding, LASER microdissection and sequencing
+## Tissue embedding, laser microdissection and sequencing
 
@@
 Embedding for dissection was performed as described by @Mantegazza:2014di.
-Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 LASER dissector (Leica).
+Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 laser dissector (Leica).
 RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN).
@@
 **`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`**.
-LASER microdissection of meristem samples in the early stages of inflorescence development.
+Laser microdissection of meristem samples in the early stages of inflorescence development.
 Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and secondary branch meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
@@
 **`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
-Expression of selected cytokinin-related genes (**a**), detected G1-Like (G1L/ALOG) genes (**b**) and *STMADS11*-like MADS genes (**c**) in inflorescence meristems in transcripts per million (TPM).
+Expression of selected cytokinin-related genes (**a**), detected *G1*-Like (G1L/ALOG) genes (**b**) and *STMADS11*-like MADS genes (**c**) in inflorescence meristems in transcripts per million (TPM).
 A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
@@
 Patterns of expression of homeobox genes in the meristem samples.
-The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [**ref Jain et al.
-2008**].
+The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [@Jain:2008kg].
 There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples.

### 2015-12-09 ms.Rmd
Hb disco (fixes #23)

@@
 output:
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
-    toc: yes
   BiocStyle::html_document:
@@
     number_sections: yes
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
+    toc: yes
 bibliography: bib/papersLib.bibtex
@@
 
-## Patterns of expression of homeodomain genes
+## Homeodomain genes and meristem identity
 
-*log1* mutants also appear to have reduced expression of the *HOMEOBOX 1* (*OSH1*) gene [@Kurakawa:2007go], which encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor.
+*log1* mutants also appear to have reduced *HOMEOBOX 1* (*OSH1*) expression [@Kurakawa:2007go].
+*OSH1* encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor.
+The expression of class I *KNOX* genes in young inflorescences has been detected by *in situ* hybridisation [**ref Sentoku 1999**].
 *osh1* mutants have defects in SAM maintenance during vegetative growth and lower expression of other *KNOX* genes, and expression of *KNOX* genes can be induced by cytokinin treatment during shoot regeneration [@Tsuda:2011gz].
+Another class I *KNOX* gene, *HOMEOBOX 15* (*OSH15*) is expressed in a similar pattern to *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`), as previously reported in *O.
+sativa* cv.
+Nipponbare [**ref Sato 1998**].
+OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone, and the higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity.
 In *A.
 thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [**ref Leibfried 2005**; **ref Lindsay, 2006**; @Gordon:2009fp].
-**Can expand this section**.
+The rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [**ref Nardmann Werr 2006**], was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+Although these results and previous studies support a role of *KNOX* and *WOX* genes in the maintenance of indeterminate meristem identity, possibly coordinated by phytohormone signalling, more work is required to elucidate this mechanism.
 
-Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were found in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems.
-Examining the dynamics of expression of all homeobox genes revealed that a prominent group follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1*.
-**Anything about functions of these genes?** There are also smaller groups of homeobox genes that peak in expression in the SM or RM samples, suggesting multiple roles and possible redundancy for homeodomain proteins during inflorescence development.
-**Can expand this section**.
+Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were recovered in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems.
+A prominent group of homeobox genes follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+*Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [**ref Javelle 2011, Depege-Fargeix 2011**].
+In rice, several class IV HD-Zip genes have been detected in the vegetative SAM by *in situ* hybridisation [**ref Ito 2003**], but their role in axillary meristems in the developing inflorescence has not been explained.
+There are also smaller groups of homeobox genes that peak in expression in the SM or RM samples, suggesting multiple roles and possible redundancy for homeodomain proteins during inflorescence development.
 

### 2015-12-09 ms.Rmd
Changed title (issue #25)

@@
 ---
-title: Genome-wide changes in expression during inflorescence architecture development
-  in rice
+title: Changes in gene expression associated with transitions in reproductive meristem identity during early development of rice inflorescence.
 csl: csl/the-plant-journal.csl

### 2015-12-10 ms.Rmd
References

@@
 In rice, functions similar to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**ref Nagasawa 2003**; **ref Yao 2008**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**ref kyozuka 2002**; **ref Yamaguchi 2006**].
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br].
 Two recent studies in *A.
@@
 *OSH1* encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor.
-The expression of class I *KNOX* genes in young inflorescences has been detected by *in situ* hybridisation [**ref Sentoku 1999**].
+The expression of class I *KNOX* genes in young inflorescences has been detected by *in situ* hybridisation [@Sentoku:1999ti].
 *osh1* mutants have defects in SAM maintenance during vegetative growth and lower expression of other *KNOX* genes, and expression of *KNOX* genes can be induced by cytokinin treatment during shoot regeneration [@Tsuda:2011gz].
@@
 sativa* cv.
-Nipponbare [**ref Sato 1998**].
+Nipponbare [@Sato:1998tp].
 OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone, and the higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity.
 In *A.
-thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [**ref Leibfried 2005**; **ref Lindsay, 2006**; @Gordon:2009fp].
-The rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [**ref Nardmann Werr 2006**], was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [@Leibfried:2005hw; @Lindsay:2006bh; @Gordon:2009fp].
+The rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [@Nardmann:2006fd], was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
 Although these results and previous studies support a role of *KNOX* and *WOX* genes in the maintenance of indeterminate meristem identity, possibly coordinated by phytohormone signalling, more work is required to elucidate this mechanism.
@@
 A prominent group of homeobox genes follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
-*Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [**ref Javelle 2011, Depege-Fargeix 2011**].
-In rice, several class IV HD-Zip genes have been detected in the vegetative SAM by *in situ* hybridisation [**ref Ito 2003**], but their role in axillary meristems in the developing inflorescence has not been explained.
+*Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [@Javelle:2011dz; @DepegeFargeix:2011ds].
+In rice, several class IV HD-Zip genes have been detected in the vegetative SAM by *in situ* hybridisation [@Ito:2003ta], but their role in axillary meristems in the developing inflorescence has not been explained.
 There are also smaller groups of homeobox genes that peak in expression in the SM or RM samples, suggesting multiple roles and possible redundancy for homeodomain proteins during inflorescence development.

### 2015-12-10 figures.Rmd
Added NEWS for emailing updates Fixed figure 4 (alog) (fixes #21) More SPL discussion (fixes #17)

@@
 title: "Figures"
+date: "\\vspace{0 mm}"
 csl: csl/the-plant-journal.csl
@@
   \usepackage{rotating}
-  \renewcommand{\thefigure}{S\arabic{figure}}
-  \renewcommand{\thetable}{S\arabic{table}}
+  \renewcommand{\thefigure}{\arabic{figure}}
+  \renewcommand{\thetable}{\arabic{table}}
 documentclass: tufte-handout
@@
     dev: CairoPDF
-    keep_tex: yes
     fig_crop: no
@@
 ```{r Setup, include=FALSE, results="hide", warning=FALSE}
+## grep queries:
+  # bold: \*\*(\w*)\*\* -> \\\\textbf{\1}
+  # italic: \*(\w*)\* -> \\\\emph{\1}
 knitr::opts_chunk$set(fig.path='figures/',
@@
 
-```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2, fig.cap = "Expression of selected cytokinin-related genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}) and \\emph{STMADS11}-like MADS genes (\\textbf{c}) in inflorescence meristems in transcripts per million (TPM).
+```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614, fig.cap = paste0("Expression of selected \\emph{STMADS11}-like MADS genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}), \\emph{SPL} genes and a microRNA precursor (\\textbf{c}) and cytokinin-related genes (\\textbf{d}) in inflorescence meristems in transcripts per million (TPM).
 A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
-\\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/SBM and SM than in RM and PBM.
+Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library.
 Only three G1L genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, but they share a similar pattern of expression.
-Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library."}
+Several \\emph{SPL} genes and a co-regulated microRNA precursor, \\emph{MIR319A}, recovered from cluster ", getClusterName('small', 'dec'), ", are highly expressed in RM and PBM before decreasing in expression in ePBM/SBM and SM.
+\\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/SBM and SM than in RM and PBM.")}
 # set layout
 grid::grid.newpage()
-grid::pushViewport(grid::viewport(layout = grid::grid.layout(5, 2)))
+grid::pushViewport(grid::viewport(layout = grid::grid.layout(4, 4)))
 
 # print the plots
-print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:2, layout.pos.col = 1:2))
-print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 1))
-print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 2))
+print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 1))
+print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 2))
+print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 3:4))
+print(f_alogFamily_d, vp = grid::viewport(layout.pos.row = 4, layout.pos.col = 1:4))
 ```

### 2015-12-10 ms.Rmd
Added NEWS for emailing updates Fixed figure 4 (alog) (fixes #21) More SPL discussion (fixes #17)

@@
 ```{r Setup, include=FALSE, results="hide", warning=FALSE}
+
+# max width (2 col): 6.614
+# max width (1 col): 3.150
+# max height : 8.661
 
@@
 Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
-*MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
+*MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
 The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
@@
 The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development.
-*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM, suggesting a redundant function in indeterminate axillary meristems.
-These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc].
+*WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems.
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [**ref Jeong 2011**; @Wang:2015jx].
+The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
 Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
@@
 *indica* varieties [@Ashikari:2005eg].
-Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role.
+Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role.
 The LABA1 protein present in *O.
@@
 
-```{r f_alogFamily, fig.width = 3.150, fig.height = 3.150 * 5/2}
-#grid::grid.draw(f_alogFamily)
-
+```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614}
 # set layout
 grid::grid.newpage()
-grid::pushViewport(grid::viewport(layout = grid::grid.layout(5, 2)))
+grid::pushViewport(grid::viewport(layout = grid::grid.layout(4, 4)))
 
 # print the plots
-print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:2, layout.pos.col = 1:2))
-print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 1))
-print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 3:5, layout.pos.col = 2))
+print(f_alogFamily_a, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 1))
+print(f_alogFamily_b, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 2))
+print(f_alogFamily_c, vp = grid::viewport(layout.pos.row = 1:3, layout.pos.col = 3:4))
+print(f_alogFamily_d, vp = grid::viewport(layout.pos.row = 4, layout.pos.col = 1:4))
 ```
@@
 **`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
-Expression of selected cytokinin-related genes (**a**), detected *G1*-Like (G1L/ALOG) genes (**b**) and *STMADS11*-like MADS genes (**c**) in inflorescence meristems in transcripts per million (TPM).
+Expression of selected *STMADS11*-like MADS genes (**a**), detected *G1*-Like (G1L/ALOG) genes (**b**), *SPL* genes and a microRNA precursor (**c**) and cytokinin-related genes (**d**) in inflorescence meristems in transcripts per million (TPM).
 A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
+Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
+Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression.
+Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, recovered from cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/SBM and SM.
 *GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/SBM and SM than in RM and PBM.
-Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression.
-Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
 

### 2015-12-11 ms.Rmd
Introduction meristem section

@@
 Yield is directly dependent on the architecture of the inflorescence [@Ikeda:2004tk].
-The rice inflorescence (or panicle) is a branched structure composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
-Spikelets are short branches that form on other branches of the panicle and bear a single flower.
-They are the primary unit of grass inflorescence, with one spikelet bearing one floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo].
 
+The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
 During panicle development, the undifferentiated cells in the primary axis (*i.e.* the rachis meristem, RM) undergo transitions through a series of meristem identities.
-After the establishment of the rachis some RM cells differentiate into branch meristem (BM) for the formation and elongation of primary branches and the establishment of higher-order branches.
-The axillary BM cells differentiate into spikelet meristems (SM), which develop into floret meristems (FM) and produce flowers.
+After the establishment of the rachis some RM cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
+During elongation, the PBM is indeterminate and can give rise to axillary meristems, which may differentiate into spikelets or secondary and higher-order branches.
+The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM).
+Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo].
 After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].

### 2015-12-11 ms.Rmd
Comments & changes from martin and israr (from early Dec)

@@
 One step towards this goal is to describe the changes in gene expression between different cell types during development.
-In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of rice panicle development, using a sensitive laser microdissection and RNA sequencing approach.
+In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of rice panicle development, using a precise laser microdissection and RNA sequencing approach.
 
@@
 
-To investigate gene expression during the development of the rice inflorescence, laser microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem and secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
+To investigate gene expression during the development of the rice inflorescence, laser microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem with secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
 RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
-RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`; **Number of panicles used for each library?**).
-Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
+RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number (RIN) was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`; **Number of panicles used for each library**).
+Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA (RIN < 6.5) were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
 Following this, further samples were prepared to provide a total of three biological replicates with intact RNA (RIN ≥ 7) for each meristem type.
@@
 Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage.
-As a preliminary assessment of the results, the qualitative and quantitative expression of a selection of genes was compared with patterns that had previously been reported.
+As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of a selection of genes was compared with patterns that had previously been reported.
 Although there were several exceptions, most genes that were detected by *in situ* hybridisation were also detected in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
-Four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridisation (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
+Four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridization (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
 **How do in situ results compare to transcriptome?**
@@
 To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do].
-Detected genes were ranked by variance, and standardised, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` most-variable genes (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
+Detected genes were ranked by variance, and standardized, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` most-variable genes (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
@@
 Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
-Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
+G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
 *MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
 The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
-Of the G1L genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+Of the ALOG genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
 *g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff].
-This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*.
+This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*.
 Although the increase in secondary branches linked to *G1L5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.
@@
 *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
-In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of MADS genes in the SM.
+In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative of the overall enrichment of MADS genes in the SM.
 
@@
 In the eudicot *A.
-thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki].
+thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity.
+*AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*) to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki].
 In rice, functions similar to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br].
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; **ref Dreni 2011**].
 Two recent studies in *A.
@@
 thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems.
-*AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de].
+*agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de].
 The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems.
@@
 In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by @Han:2014dt].
-The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go].
+The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM and small panicles with small SAMs and branching defects [@Kurakawa:2007go].
 Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O.
@@
 The LABA1 protein present in *O.
-rufipogon*, which has an identical primary sequence to the protein from *O.
+rufipogon*, which shares an identical primary sequence with the protein from *O.
 sativa* ssp.
 *japonica* cv.
-Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O.
+Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele **I think this is correct, because the domesticated *laba1* gene product is non-functional.
+Can someone pls confirm?
+Tom.** from *O.
 rufipogon* into an *O.

### 2015-12-11 ms.Rmd
Abstract

@@
 # Summary
+
+In rice, inflorescence architecture is established at early stages of reproductive development by the activity of different meristem types, and contributes directly to grain yield potential.
+After induction of flowering, the shoot apical meristem is converted to an inflorescence meristem to form the rachis, which is the main axis of the inflorescence.
+The rachis meristem produces primary branch meristems, which themselves produce axillary meristems during elongation.
+The axillary meristems may become secondary or higher-order branch meristems, retaining their indeterminate state, or be converted directly to determinate spikelet meristems.
+After elongation, the primary and higher-order branch meristems differentiate into spikelet meristems, and the complexity of branching and number of spikelets is fixed.
+Although some of the genes involved in transitions in meristem identity in rice have been identified, understanding of the network of genes controlling this process is lacking.
+To address this, we used a precise laser microdissection and RNA sequencing approach in *Oryza sativa* ssp.
+*japonica* cv.
+Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types.
+A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of potential candidates were identified.
+This resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield that integrates gene expression with environmental and other epigenetic factors, as well as for comparative studies of the domestication process.
 
@@
 The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems.
-Expression of the putatative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage.
+Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage.
 In contrast, consistent with their functions in floral organ specification in rice and *A.
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`**: Differential expression
-of genes encoding putatative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*.
+of genes encoding putative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*.
 

### 2015-12-11 ms.Rmd
Specified number of differences between rnaseq and in situ and made some excuses in text (fixes #22). Removed date in SI (fixes #26)

@@
 Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage.
-As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of a selection of genes was compared with patterns that had previously been reported.
-Although there were several exceptions, most genes that were detected by *in situ* hybridisation were also detected in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
+As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with patterns that had previously been reported.
+`r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
 Four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridization (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
-**How do in situ results compare to transcriptome?**
 

### 2015-12-11 si.Rmd
Added intro
Abstract
Specified number of differences between rnaseq and in situ and made some excuses in text (fixes #22). Removed date in SI (fixes #26)

@@
 csl: csl/the-plant-journal.csl
+date: "\\vspace{0 mm}"
 header-includes:
@@
 st_libStats_x <- as.data.frame(st_libStats)
-# colnames(st_libStats_x) <- gsub("Uniquely mapped reads (M)",
-#                                 "Uniquely mapped reads (M)",
-#                                 colnames(st_libStats_x),
-#                                 fixed = TRUE)
 rownames(st_libStats_x) <- st_libStats_x$Library
@@
 
-# add more space to the longest colname
-#st_libStats_x_print <- sub("Uniquely mapped reads (M)", "Uniquely mapped reads (M) \\hspace{3 mm}", st_libStats_x_print, fixed = TRUE)
-# OR remove the first hline?
+# remove the first hline
 st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
@@
 cat("¹ ePBM/SBM")
+```
+
+```{r sf_isGenesTpm, fig.height = 9.535, fig.cap = paste0("\\protect\\label{fig:sf_isGenesTpm}","Comparison of the LMD RNA-sequencing dataset with published \\emph{in situ} results.
+Expression value in transcripts per million (TPM) is shown on the \\emph{y}-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below.
+Grey shading indicates a meristem type for which expression was detected by \\emph{in situ} hybridisation.
+See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
+Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).", "\\newline ", "Gene expression measured by RNAseq may differ from the qualitative patterns measured by RNA \\emph{in situ} hybridization for several reasons, including the relative sensitivity of the techniques, the specificity of the \\emph{in situ} probe, and differences in growth conditions (\\emph{e.g.
+HD5}) or the accession under study (\\emph{e.g.
+GN1A}).")}
+print(sf_isGenesTpm)
 ```
@@
 \end{figure*}
-
-```{r sf_isGenesTpm, fig.height = 9.535, fig.cap = paste0("\\protect\\label{fig:sf_isGenesTpm}","Comparison of the LMD RNA-sequencing dataset with published \\emph{in situ} results.
-Expression value in transcripts per million (TPM) is shown on the \\emph{y}-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below.
-Grey shading indicates a meristem type for which expression was detected by \\emph{in situ} hybridisation.
-See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
-Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).", "\\newline ", "\\textbf{Do we need to discuss the results of individual genes?}")}
-print(sf_isGenesTpm)
-```
 
@@
 
-```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putatative MIKC proteins in \\emph{A.
+```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putative MIKC proteins in \\emph{A.
 thaliana}, \\emph{Solanum lycopersicum} and \\emph{Oryza sativa} ssp.

### 2015-12-11 src/R/figures.R

@@ -198,25 +198,14 @@ genTpm[, c("xmin", "xmax") := .(
 
 # draw it
 setkey(genTpm, "plotLabel", "stage")
-# sf_isGenesTpm <- ggplot(genTpm, aes(x = stage, y = tpm, colour = Replicate, shape = isExpr)) +
-#   theme_minimal(base_size = 8, base_family = "Helvetica") +
-#   theme(strip.text = element_text(face = "italic")) +
-#   xlab(NULL) +
-#   scale_color_brewer(palette = "Set1") +
-#   guides(shape = FALSE, size = FALSE) +
-#   geom_rect(aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
-#             unique(genTpm[call.z == TRUE,]), colour = NA,  fill = "grey90", alpha = 0.75) +
-#   stat_smooth(aes(group = plotLabel), method = "loess", se = FALSE, colour = "grey", size = 0.5) + 
-#   geom_point(size = 1, alpha = 0.8, position = position_jitter(width = 0.2)) +
-#   facet_wrap(~ plotLabel, scales = "free_y", ncol = 4)
 
 # keep a subset of genes
 keep <- c("LOC_Os06g45460", "LOC_Os09g26999", "LOC_Os07g42410", "LOC_Os07g47330",
-          "LOC_Os10g33780", "LOC_Os03g60430", "LOC_Os01g61480", "LOC_Os03g11614",
-          "LOC_Os01g40630", "LOC_Os04g49150", "LOC_Os07g41370", "LOC_Os02g52340",
-          "LOC_Os09g32948", "LOC_Os06g11330", "LOC_Os08g41950", "LOC_Os02g45770",
+          "LOC_Os10g33780", "LOC_Os01g61480", "LOC_Os03g11614",
+          "LOC_Os01g40630", "LOC_Os07g41370", "LOC_Os02g52340",
+          "LOC_Os09g32948", "LOC_Os06g11330", "LOC_Os08g41950",
           "LOC_Os03g51690", "LOC_Os03g54170", "LOC_Os04g51000", "LOC_Os07g13170",
-          "LOC_Os08g39890")
+          "LOC_Os08g39890", "LOC_Os01g10110", "LOC_Os08g07740")
 
 genTpm <- genTpm[msuId %in% keep]
 
@@ -225,9 +214,6 @@ refTable <- refTable[msuId %in% keep]
 setkey(refTable, "msuId", "Reference")
 refTable <- unique(refTable)
[…]

### 2015-12-11 ms.Rmd
separated homeobox figure into panels (fixes #24)

@@
 *japonica* cv.
-Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele **I think this is correct, because the domesticated *laba1* gene product is non-functional.
+Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele [**I think "*Laba1*" is correct, because the gene product from the domesticated allele (called *laba1* in the paper) is non-functional.
 Can someone pls confirm?
-Tom.** from *O.
+Tom.**] from *O.
 rufipogon* into an *O.
@@
 *log1* mutants also appear to have reduced *HOMEOBOX 1* (*OSH1*) expression [@Kurakawa:2007go].
-*OSH1* encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor.
-The expression of class I *KNOX* genes in young inflorescences has been detected by *in situ* hybridisation [@Sentoku:1999ti].
+*OSH1* encodes a class I *Knotted1-like homeobox* (*KNOX*) homeodomain transcription factor, and the expression of class I *KNOX* genes in meristems in the developing inflorescence has been detected by *in situ* hybridisation [@Sentoku:1999ti].
 *osh1* mutants have defects in SAM maintenance during vegetative growth and lower expression of other *KNOX* genes, and expression of *KNOX* genes can be induced by cytokinin treatment during shoot regeneration [@Tsuda:2011gz].
@@
 sativa* cv.
-Nipponbare [@Sato:1998tp].
-OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone, and the higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity.
+Nipponbare [@Sato:1998tp], and a third, *OSH6*, has a peak in expression in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone.
+The higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity.
 In *A.
 thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [@Leibfried:2005hw; @Lindsay:2006bh; @Gordon:2009fp].
-The rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [@Nardmann:2006fd], was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+*wox1* plants carrying a likely null mutation in the rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [@Nardmann:2006fd], have defects in axillary meristem formation and lower expression of *OSH1* [@Tanaka:2015ko].
+*WOX1* was not detected in the meristem tissues described here, but another *WUSCHEL homeobox* (*WOX*) gene, *WOX8*, has a peak in expression in RM, and a third, *LOC_Os01g70810*, is predominantly expressed in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
 Although these results and previous studies support a role of *KNOX* and *WOX* genes in the maintenance of indeterminate meristem identity, possibly coordinated by phytohormone signalling, more work is required to elucidate this mechanism.
+
+**General HB introduction (link to a review), explain figure 5**.
 

### 2015-12-11 src/R/figures.R

@@ -658,15 +658,18 @@ f_hb <- ggplot() +
     axis.text.y = element_text(face = "italic", size = 5.5),
     #legend.title = element_text(size = 6),
     panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank()) +
+    panel.grid.minor = element_blank(),
+    strip.background = element_blank(),
+    strip.text = element_blank()) +
   xlab(NULL) + ylab(NULL) +
+  facet_grid(cut ~ ., space = "free", scales = "free") +
   scale_y_discrete(expand = c(0,0)) +
   scale_x_discrete(expand = c(0,0)) +
   geom_raster(aes(x = Stage, y = symbol, fill = `Scaled reads`),
               data = plotData.long) +
-  scale_fill_gradientn(colours = heatscale) +
-  geom_segment(aes(x = 0.255, y = y, xend = 0.255, yend = yend, colour = class),
+  geom_segment(aes(x = 0.255, y = yRel, xend = 0.255, yend = yRel + 1, colour = class),
                data = segData, size = 5) +
+  scale_fill_gradientn(colours = heatscale) +
   scale_colour_brewer(palette = "Set3", guide = guide_legend(title = NULL))
 
 figCount <- incCount(figCount, "f_hb")

### 2015-12-11 src/R/homeobox.R

@@ -67,9 +67,12 @@ hbVstScaled <- t(scale(t(hbVst), center = TRUE))
 dm <- "minkowski"
 hm <- "ward.D2"
 
+# cut the tree into 5 groups of expression
+hc <- hclust(dist(hbVstScaled, method = dm), method = hm)
+cuts <- cutree(hc, k = 5)
+
 # get the order for the x-axis
-geneOrder <- rownames(hbVstScaled)[
-  hclust(dist(hbVstScaled, method = dm), method = hm)$order]
+geneOrder <- rownames(hbVstScaled)[hc$order]
 
 # merge scaled vst to hb data
 hbVstScaled.table <- data.table(hbVstScaled, keep.rownames = TRUE, key = 'rn')
@@ -98,6 +101,14 @@ segData[, colour := plyr::mapvalues(
   from = unique(class),
   to = RColorBrewer::brewer.pal(length(unique(class)), "Set3")
   )]
+
+# add cuts to plotdata and segdata
+plotData.long[, cut := cuts[as.character(msuId)], by = msuId]
+segData[, cut := cuts[as.character(msuId)], by = msuId]
+
+# rescale y for segData relative to each facet
+segData[, minY := min(y), by = cut]
+segData[, yRel := y - minY + 0.5]
 
 # MAKE OUTPUT FOLDER
 outDir <- "output/homeobox"

### 2015-12-14 src/R/homeobox.R
add number of genes per class and number expressed to HB figure (#24)

@@ -30,7 +30,6 @@ tfdb <- readRDS(tfdbFile)
 
 # add tfdb Hbs missing from Jain et. al data
 missingHbs <- tfdb[Family == "HB", Protein.ID[which(!(Protein.ID %in% hbGenes[, msuId]))]]
-oryzr::LocToGeneName(missingHbs)
 if (length(missingHbs) >0 ){
   warning(paste(
     "The following msuIds were not found in the Jain et. al data:\n",
@@ -38,11 +37,15 @@ if (length(missingHbs) >0 ){
     "\nThey will be categorized as \"Not annotated\""))
   notAnn <- data.table(
     class = "Not annotated",
-    msuId = missingHbs
+    msuId = missingHbs,
+    
   )
   notAnn[, symbol := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
   hbGenes <- rbind(hbGenes, notAnn)
 }
+
+# number per class
+hbGenes[, nClass := length(unique(msuId)), by = class]
 
 # expressed genes
 expressedGenesAll <- readRDS(expressedGenesAllFile)
@@ -79,11 +82,16 @@ hbVstScaled.table <- data.table(hbVstScaled, keep.rownames = TRUE, key = 'rn')
 setnames(hbVstScaled.table, 'rn', 'msuId')
 setkey(expHb, 'msuId')
 plotData <- expHb[hbVstScaled.table, ]
+plotData[, nExprPerClass := length(unique(msuId)), by = class]
+plotData[, class := paste0(class, " (", nExprPerClass, "/", nClass, ")")]
 
 # set up plot
 plotData[!is.na(symbol), symbol := paste(symbol, msuId, sep = "\n")]
 plotData[is.na(symbol), symbol := msuId, by = msuId]
-plotData.long <- reshape2::melt(plotData, id.vars = c('msuId', 'symbol', 'class'), variable.name = 'Stage',
+plotData.long <- reshape2::melt(plotData,
+                                #id.vars = c('msuId', 'symbol', 'class', 'nClass'),
+                                measure.vars = c("RM", "PBM", "ePBM/SBM", "SM"),
+                                variable.name = 'Stage',
                                 value.name = 'Scaled reads')
 plotData.long[, msuId := factor(msuId, levels = geneOrder)]
 plotData.long[, symbol := factor(symbol, levels = plotData[geneOrder, symbol])]

### 2015-12-14 ms.Rmd
Class III HD-Zips

@@
 
+In *A.
+thaliana*, class III homeodomain-lucine-zipper (HD-Zip) genes are involved in meristem initiation and regulation (**ref Prigge et al 2005**).
+The class III HD-Zip gene *REVOLUTA* is required for axillary meristem formation (**ref Talbert 1995**; **ref Otsuga 2001**).
+Seven of the nine annotated class III HD-Zip genes in *O.
+sativa* were detected (`r I(pasteLabel("Figure ", figCount, "f_hb"))`), including two homologs of *REVOLUTA*, *HOX9* and *HOX10* (**ref Prigge and Clark 2006**), which are both highly expressed in the RM.
 Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were recovered in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems.
-A prominent group of homeobox genes follow a similar pattern, including five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1* (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+Five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1*, also follow this pattern (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
 *Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [@Javelle:2011dz; @DepegeFargeix:2011ds].

### 2015-12-14 ms.Rmd
Homeobox introduction in results section (#23)

@@
 
+Homeodomain genes are involved at various stages of plant development and in several hormone response pathways (**ref Chan 1998**; **ref Himmelbach 2002**; **ref Sawa 2002**).
+The expression of this family in inflorescence meristems was explored in greater detail using a heatmap of scaled, transformed read counts.
+Heirarchical clustering of the transformed counts recovered five prominent groups of expression,  with unequal distribution of subfamilies of homeodomain genes among the groups (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+
 # Discussion
@@
 *LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples.
-Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d).
 Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level.
@@
 Although these results and previous studies support a role of *KNOX* and *WOX* genes in the maintenance of indeterminate meristem identity, possibly coordinated by phytohormone signalling, more work is required to elucidate this mechanism.
-
-**General HB introduction (link to a review), explain figure 5**.
 
@@
 
+```{r f_hb, fig.height = 8.661}
+print(f_hb)
+```
+
+**`r I(pasteLabel("Figure ", figCount, "f_hb"))`**.
+Patterns of expression of homeobox genes in the meristem samples.
+The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [@Jain:2008kg].
+There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples.
+The former group contains *OSH1* (**function?**), whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes.
+
+----
+
 ```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614}
@@
 
-```{r f_hb, fig.height = 8.661}
-print(f_hb)
-```
-
-**`r I(pasteLabel("Figure ", figCount, "f_hb"))`**.
-Patterns of expression of homeobox genes in the meristem samples.
-The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [@Jain:2008kg].
-There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples.
-The former group contains *OSH1* (**function?**), whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes.
-
-----
 # `sessionInfo` (*hide*)

### 2015-12-14 src/R/figures.R

@@ -460,92 +460,6 @@ if (gsea[,any(showPval)]){
 
 figCount <- incCount(figCount, "f_gsea")
 
-############
-### ALOG ###
-############
-
-# HDZIP4 <- c('LOC_Os02g45250', 'LOC_Os04g48070', 'LOC_Os04g53540', 'LOC_Os08g04190', 'LOC_Os08g08820', 'LOC_Os08g19590', 'LOC_Os09g35760', 'LOC_Os10g42490')
-# HDZIP3 <- c('LOC_Os01g48170', 'LOC_Os03g01890', 'LOC_Os03g43930', 'LOC_Os05g48820', 'LOC_Os06g39906', 'LOC_Os10g33960', 'LOC_Os12g41860')
-# 
-# expTable <- data.table(msuId = HDZIP3, type = 'hb')
-
-svpMads <- c("LOC_Os02g52340", "LOC_Os03g08754", "LOC_Os06g11330")
-expG1s <- c("LOC_Os02g07030", "LOC_Os06g46030", "LOC_Os10g33780")
-ck <- c("LOC_Os01g51210","LOC_Os04g43840","LOC_Os01g40630", "LOC_Os01g10110")
-spl <- c("LOC_Os02g04680", "LOC_Os06g49010", "LOC_Os09g31438", "LOC_Os07g32170",
-         "LOC_Os08g39890", "LOC_Os01g46984")
-
-expTable <- rbind(data.table(msuId = svpMads, type = "mads"),
-                  data.table(msuId = expG1s, type = "g1l"),
-                  data.table(msuId = ck, type = 'ck'),
-                  data.table(msuId = spl, type = 'spl'))
-
-expTable[, symbol := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
-
-# add expression values
-tpm <- data.table(readRDS('output/tpm/tpm.Rds'), keep.rownames = TRUE)
-setkey(tpm, "rn")
-plotData.wide <- tpm[expTable]
-
-# convert to long
-plotData <- reshape2::melt(plotData.wide, id.vars = c("rn", "type", "symbol"),
-                           variable.name = "library", value.name = "Expression (TPM)")
-setkey(plotData, "rn", "library")
-
-# add expression calls
-expGenTT.wide <- data.table(readRDS('output/expressedGenes/expGenTT.Rds'), key = "id")
-expGenTT <- reshape2::melt(expGenTT.wide, id.vars = "id", variable = "library", value = "isExpr")
-setkey(expGenTT, "id", "library")
[…]

### 2015-12-14 figures.Rmd
Homeobox introduction in results section (#23)
f_hb legend (#23)

@@
 
+```{r f_hb, fig.height = 8.661, fig.cap = "Patterns of expression of homeobox genes that were detected in the meristem samples.
+The heatscale represents the mean scaled, VST-transformed read count, and the coloured bar next to the \\emph{y}-axis shows the homeobox subfamily for each gene (Jain \\emph{et al}., 2008).
+The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
+The expression patterns were separated into five groups using heirarchical clustering.
+The set of homeobox genes with lowest expression in the RM and highest expression in indeterminate axillary meristems contains \\emph{OSH1} and five of the eight expressed class IV HD-Zip genes."}
+print(f_hb)
+```
+
 ```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614, fig.cap = paste0("Expression of selected \\emph{STMADS11}-like MADS genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}), \\emph{SPL} genes and a microRNA precursor (\\textbf{c}) and cytokinin-related genes (\\textbf{d}) in inflorescence meristems in transcripts per million (TPM).
@@
 ```
-
-```{r f_hb, fig.height = 8.661, fig.cap = "Patterns of expression of homeobox genes in the meristem samples.
-The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the \\emph{y}-axis shows the homeobox clade for each gene [@Jain:2008kg].
-There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples.
-The former group contains \\emph{OSH1}, whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes."}
-print(f_hb)
-```

### 2015-12-14 ms.Rmd
f_hb legend (#23)

@@
 **`r I(pasteLabel("Figure ", figCount, "f_hb"))`**.
-Patterns of expression of homeobox genes in the meristem samples.
-The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [@Jain:2008kg].
-There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples.
-The former group contains *OSH1* (**function?**), whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes.
+Patterns of expression of homeobox genes that were detected in the meristem samples.
+The heatscale represents the mean scaled, VST-transformed read count, and the coloured bar next to the *y*-axis shows the homeobox subfamily for each gene [@Jain:2008kg].
+The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
+The expression patterns were separated into five groups using heirarchical clustering.
+The set of homeobox genes with lowest expression in the RM and highest expression in indeterminate axillary meristems contains *OSH1* and five of the eight expressed class IV HD-Zip genes.
 

### 2015-12-14 figures.Rmd
Changed SBM to AM (fixes #30)

@@
 \caption{Laser microdissection of meristem samples in the early stages of inflorescence development.
-Samples were collected from the rachis meristem (\textbf{a} and \textbf{e}), the primary branch meristem (\textbf{b} and \textbf{f}), the elongating primary branch meristem and secondary branch meristem (\textbf{c} and \textbf{g}) and the spikelet meristem (\textbf{d} and \textbf{h}).
+Samples were collected from the rachis meristem (\textbf{a} and \textbf{e}), the primary branch meristem (\textbf{b} and \textbf{f}), the elongating primary branch meristem and axillary meristem (\textbf{c} and \textbf{g}) and the spikelet meristem (\textbf{d} and \textbf{h}).
 Images show the samples before (\textbf{a}–\textbf{d}) and after (\textbf{e}–\textbf{h}) dissection.}
@@
 ```{r f_hb, fig.height = 8.661, fig.cap = "Patterns of expression of homeobox genes that were detected in the meristem samples.
-The heatscale represents the mean scaled, VST-transformed read count, and the coloured bar next to the \\emph{y}-axis shows the homeobox subfamily for each gene (Jain \\emph{et al}., 2008).
+The heatscale represents the geometric mean of scaled, VST-transformed read counts, and the coloured bar next to the \\emph{y}-axis shows the homeobox subfamily for each gene (Jain \\emph{et al}., 2008).
 The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
@@
 A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
-Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library.
+Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} (Yoshida \\emph{et al}., 2013), are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library.
 Only three G1L genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, but they share a similar pattern of expression.
-Several \\emph{SPL} genes and a co-regulated microRNA precursor, \\emph{MIR319A}, recovered from cluster ", getClusterName('small', 'dec'), ", are highly expressed in RM and PBM before decreasing in expression in ePBM/SBM and SM.
-\\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/SBM and SM than in RM and PBM.")}
+Several \\emph{SPL} genes and a co-regulated microRNA precursor, \\emph{MIR319A}, recovered from cluster ", getClusterName('small', 'dec')," are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.
+\\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/AM and SM than in RM and PBM.")}
 # set layout

### 2015-12-14 ms.Rmd
Changed SBM to AM (fixes #30)

@@
 After the establishment of the rachis some RM cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
-During elongation, the PBM is indeterminate and can give rise to axillary meristems, which may differentiate into spikelets or secondary and higher-order branches.
+During elongation, the PBM is indeterminate and can give rise to axillary meristems (AM), which may differentiate into spikelets or secondary and higher-order branches.
 The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM).
@@
 
-To investigate gene expression during the development of the rice inflorescence, laser microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem with secondary branch meristem (ePBM/SBM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
+To investigate gene expression during the development of the rice inflorescence, laser microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem with axillary meristem (ePBM/AM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
 RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
@@
 Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are more highly expressed in RM than in the other meristem types.
-Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.
+Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/AM samples but depleted or not enriched in the RM samples.
 
@@
 *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs].
-*LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples.
+*LABA1* was detected in the ePBM/AM and SM samples but not the RM or PBM samples.
 Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d).
@@
 Laser microdissection of meristem samples in the early stages of inflorescence development.
-Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and secondary branch meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
+Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and axillary meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
 Images show the samples before (**a**–**d**) and after (**e**–**h**) dissection.
@@
 Patterns of expression of homeobox genes that were detected in the meristem samples.
-The heatscale represents the mean scaled, VST-transformed read count, and the coloured bar next to the *y*-axis shows the homeobox subfamily for each gene [@Jain:2008kg].
+The heatscale represents the geometric mean of scaled, VST-transformed read counts, and the coloured bar next to the *y*-axis shows the homeobox subfamily for each gene [@Jain:2008kg].
 The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
@@
 Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression.
-Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, recovered from cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/SBM and SM.
-*GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/SBM and SM than in RM and PBM.
+Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, recovered from cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.
+*GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/AM and SM than in RM and PBM.
 

### 2015-12-14 si.Rmd
Changed SBM to AM (fixes #30)

@@
 st_libStats_x$Library <- NULL
-st_libStats_x$Sample <- sub("ePBM/SBM", "SBM¹", st_libStats_x$Sample, fixed = TRUE)
+st_libStats_x$Sample <- sub("ePBM/SBM", "AM¹", st_libStats_x$Sample, fixed = TRUE)
 st_libStats_x <- xtable::xtable(st_libStats_x)
@@
 cat(st_libStats_x_print)
-cat("¹ ePBM/SBM")
+cat("¹ ePBM/AM")
 ```

### 2015-12-14 src/R/figures.R

@@ -74,7 +74,7 @@ opIdx <- pcaPlotData[c('n2r3', 'n4r1'),'rIdx']
 
 # format legend
 pcaPlotData$Stage <- plyr::mapvalues(pcaPlotData$Stage, from = "ePBM/SBM",
-                                     to = "ePBM/\nSBM")
+                                     to = "ePBM/\nAM")
 
 sf_pca <- ggplot(data = pcaPlotData,
                  aes(x = PCA1, y = PCA2, colour = Stage, label = label)) +
@@ -153,7 +153,7 @@ setkey(tpm, "MSU identifier", "stage")
 genes <- compare[, .(msuId, stage, call.z)]
 setkey(genes, "msuId", "stage")
 genes <- unique(genes)
-genes[, stage := plyr::mapvalues(stage, "SBM", "ePBM/SBM")]
+genes[, stage := plyr::mapvalues(stage, "SBM", "ePBM/AM")]
 
 # dirty kludge, fix in compareVsInSitu.R
 genes <- genes[!stage == "FM"]
@@ -185,7 +185,7 @@ genTpm[, name := oryzr::LocToGeneName(msuId)$symbols, by = msuId]
 genTpm[, plotLabel := name]
 genTpm[!is.na(plotLabel), plotLabel := paste(plotLabel, "·", msuId)]
 genTpm[is.na(plotLabel), plotLabel := msuId]
-genTpm[, stage := plyr::mapvalues(stage, "ePBM/SBM", "ePBM/\nSBM")]
+genTpm[, stage := plyr::mapvalues(stage, "ePBM/AM", "ePBM/\nAM")]
 
 # order the plots
 setkey(genTpm, "name")
@@ -221,7 +221,8 @@ s_tableCount <- incCount(s_tableCount, "st_refTable")
 
 sf_isGenesTpm <- ggplot() +
   theme_minimal(base_size = 8, base_family = "Helvetica") +
-  theme(strip.text = element_text(face = "italic")) +
+  theme(strip.text = element_text(face = "italic"),
+        axis.text.x = element_text(vjust = 0.5)) +
   xlab(NULL) + ylab("Epxression (TPM)") +
   scale_color_brewer(palette = "Set1") +
   guides(shape = FALSE, size = FALSE) +
@@ -332,7 +333,7 @@ plotData <- reshape2::melt(plotData.wide,
                            value.name = "Scaled, transformed read counts")
 
 # fix stage label
[…]

### 2015-12-14 ms.Rmd
Updated m+m Squashed a bug in sf_isGenesTpm (fixes #30)

@@
 One step towards this goal is to describe the changes in gene expression between different cell types during development.
-In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of rice panicle development, using a precise laser microdissection and RNA sequencing approach.
+In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O.
+sativa* ssp.
+*japonica* cv.
+Nipponbare, using a precise laser microdissection and RNA sequencing approach.
 
@@
 
-Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (**filtered?**) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C.
+Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C.
 Embedding for dissection was performed as described by @Mantegazza:2014di.
@@
 RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN).
-Library preparation with **which kit** and 50-base, single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
+Library preparation with the Ovation Ultralow Library Sytem (NuGEN) and 50-base, single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).
+
+## RNA *in situ* hybridization
+
+Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by (**ref Huijser 1992**).
+Digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions.
+To generate the probes, cDNA was amplified using the following primers: *LOC_Os01g04670*, 5′-GTGTCAAGGCATCGCCAAC-3′ and 5′-CATCAGCTGGCTGCTTTACC-3′; *LOC_Os10g04270*, 5′-TCCCAGTTGACCGAGAACTG-3′ and 5′-TGAACTTCCGTCACGAACTCC-3′; *LOC_Os10g05990*, 5′-CCTCCGGCAAAGAACTGATG-3′ and 5′-CCGTCATTGGGACTAGTTTGTCAG-3′.
+For detection of *LOC_Os09g27730*, an Digoxigenin-labelled LNA probe with the sequence 5′-TCTG{A}CGACG{T}GCG{A}C{T}GGT-3′ was synthesised (Eurogentec, Liège, Belgium).
+Hybridization was performed as described by **ref Coen 1990** using NBT/BCIP (Roche) or VECTOR Blue Alkaline Phosphatase Substrate Kit (Vector Laboratories) for detection.
 
@@
 All plots were produced with the `ggplot2` package [@Wickham:2009tn].
-
-### Comparative analysis of MADS family
-
-The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
-An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn].
-To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed.
-To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible.
-Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`).
-L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
-The tree was plotted using the `ggtree` extension for `ggplot2` (**ref ggtree**).
 

### 2015-12-14 si.Rmd
Updated m+m Squashed a bug in sf_isGenesTpm (fixes #30)

@@
 \\emph{japonica}.
-The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences.
+The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1).
 Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
@@
 
+# Supporting experimental procedures
+
+## Methods S1: Comparative analysis of MADS family
+
+The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
+An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn].
+To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed.
+To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible.
+Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`).
+L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
+The tree was plotted using the `ggtree` extension for `ggplot2` ([@Wickham:2009tn]; https://github.com/GuangchuangYu/ggtree).
+
+\clearpage
+
 # References for supporting information

### 2015-12-14 src/R/figures.R

@@ -146,6 +146,7 @@ tpm <- reshape2::melt(tpm.wide, id.vars = "MSU identifier",
 colData <- data.table(as.data.frame(GenomicRanges::colData(
   readRDS('output/DESeq2/ddsLrt.Rds'))), keep.rownames = TRUE)
 tpm[, stage := colData[rn %in% as.character(library), stage], by = library]
+tpm[, stage := plyr::mapvalues(stage, "ePBM/SBM", "ePBM/AM")]
 
 setkey(tpm, "MSU identifier", "stage")
 

### 2015-12-14 ms.Rmd
Added our in situ results (related to #22)

@@
     number_sections: yes
+word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
+    toc: yes
 bibliography: bib/papersLib.bibtex
@@
 `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
-Four genes that were detected in specific meristem samples were selected for validation by *in situ* hybridization (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
+The expression of further four genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization.
+Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
 
@@
 
-**`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`**: Placeholder for the *in situ* results figure.
+**`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`**: RNA *in situ* hybridization analysis to confirm the specific expression patterns of four genes detected in the RNA sequencing dataset.
 

### 2015-12-14 si.Rmd
Added our in situ results (related to #22)

@@
 
+\clearpage
 \begin{figure*}
- \includegraphics{data/lmdFigure.pdf}
-\caption{Placeholder for the \emph{in situ} results figure.}
+  \includegraphics{data/isFigure.pdf}
 \end{figure*}
+\captionof{figure}{RNA \emph{in situ} hybridization analysis to confirm the specific expression patterns of four genes detected in the RNA sequencing dataset.
+Expression of \emph{LOC\_Os09g27730} (a--d), \emph{LOC\_Os01g04670} (e--h), \emph{LOC\_Os10g04270} (i--l) and \emph{LOC\_Os10g05990} (m--p) was analysed at the rachis meristem (a, e, i and m), primary branch meristem (b, f, j and n), elongating primary branch meristem plus axillary meristem (c, g, k and o) and spikelet meristem (d, h, l and p) stages.
+Each gene was detected in the same stage by \emph{in situ} hybridization as by RNA sequencing.}
 
@@
 L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
-The tree was plotted using the `ggtree` extension for `ggplot2` ([@Wickham:2009tn]; https://github.com/GuangchuangYu/ggtree).
+The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn].
 

### 2015-12-15 figures.Rmd
Stefan and Helene's comments from lab meeting Added acknowledgements (fixes #32) and more biology to the conclusion (fixes #31) Shorter title (fixes #25)

@@
  \includegraphics{data/lmdFigure.pdf}
-\caption{Laser microdissection of meristem samples in the early stages of inflorescence development.
+\end{figure*}
+\captionof{figure}{Laser microdissection of meristem samples in the early stages of inflorescence development.
 Samples were collected from the rachis meristem (\textbf{a} and \textbf{e}), the primary branch meristem (\textbf{b} and \textbf{f}), the elongating primary branch meristem and axillary meristem (\textbf{c} and \textbf{g}) and the spikelet meristem (\textbf{d} and \textbf{h}).
 Images show the samples before (\textbf{a}–\textbf{d}) and after (\textbf{e}–\textbf{h}) dissection.}
-\end{figure*}
+\clearpage
 
@@
 The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
-The expression patterns were separated into five groups using heirarchical clustering.
+The expression patterns were separated into five groups using hierarchical clustering.
 The set of homeobox genes with lowest expression in the RM and highest expression in indeterminate axillary meristems contains \\emph{OSH1} and five of the eight expressed class IV HD-Zip genes."}

### 2015-12-15 ms.Rmd
Stefan and Helene's comments from lab meeting Added acknowledgements (fixes #32) and more biology to the conclusion (fixes #31) Shorter title (fixes #25)

@@
 ---
-title: Changes in gene expression associated with transitions in reproductive meristem identity during early development of rice inflorescence.
+title: Gene expression profiling of reproductive meristem identities in rice inflorescence
 csl: csl/the-plant-journal.csl
@@
     number_sections: yes
-word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
-    toc: yes
 bibliography: bib/papersLib.bibtex
@@
 
+Expression profile of rice inflorescence meristems
+
 ## Keywords
@@
 After induction of flowering, the shoot apical meristem is converted to an inflorescence meristem to form the rachis, which is the main axis of the inflorescence.
-The rachis meristem produces primary branch meristems, which themselves produce axillary meristems during elongation.
+The rachis meristem produces primary branch meristems, which themselves produce further axillary meristems during elongation.
 The axillary meristems may become secondary or higher-order branch meristems, retaining their indeterminate state, or be converted directly to determinate spikelet meristems.
-After elongation, the primary and higher-order branch meristems differentiate into spikelet meristems, and the complexity of branching and number of spikelets is fixed.
+The timing of rachis meristem abortion specifies the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems conditions the complexity of branching and thereby the number of seeds on the panicle.
 Although some of the genes involved in transitions in meristem identity in rice have been identified, understanding of the network of genes controlling this process is lacking.
@@
 # Significance statement
+
+Rice inflorescence meristems undergo a series of transitions during development, and control of these transitions impacts the morphology of the adult inflorescence and hence grain yield.
+We use laser microdissection and RNA sequencing to describe the transcriptional changes that occur during this process.
 
@@
 After the establishment of the rachis some RM cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
-During elongation, the PBM is indeterminate and can give rise to axillary meristems (AM), which may differentiate into spikelets or secondary and higher-order branches.
+During primary branch elongation, the PBM is indeterminate and can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelets.
 The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM).
@@
 
-The landscape of gene expression is a characteristic factor that differentiates meristem cell types, and is involved in the control of transitions between meristem identities and therefore panicle phenotype.
+The landscape of gene expression is a characteristic that differentiates meristem types, and is involved in the control of transitions between meristem identities and therefore panicle branching.
 Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
@@
 Despite advances in characterization at the level of individual genes and interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia].
-One step towards this goal is to describe the changes in gene expression between different cell types during development.
+One step towards this goal is to describe the changes in gene expression between different meristem types during development.
 In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O.
@@
 `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
-The expression of further four genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization.
+The expression of a further four uncharacterized genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization.
 Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
+These results suggest that the tissues used for RNA sequencing were accurately dissected and represent the intended meristem types.
 
@@
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
-Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between the apical meristem (RM) and axillary meristem (PBM–SM) samples (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`) and genes that change expression gradually over the course of changes in axillary meristem identity (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
+Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between the apical meristem (RM) and axillary meristem (PBM to SM) samples (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`) and genes that change expression gradually over the course of changes in axillary meristem identity (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`).
 There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).
@@
 Unlike soft clustering, which was used to group genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
-Some  modes of expression shared between different families may suggest roles in inflorescence development.
 The most pronounced change in expression appears to occur between the RM and branch meristem samples (*i.e.* the change from apical to axillary meristem).
@@
 Homeodomain genes are involved at various stages of plant development and in several hormone response pathways (**ref Chan 1998**; **ref Himmelbach 2002**; **ref Sawa 2002**).
-The expression of this family in inflorescence meristems was explored in greater detail using a heatmap of scaled, transformed read counts.
-Heirarchical clustering of the transformed counts recovered five prominent groups of expression,  with unequal distribution of subfamilies of homeodomain genes among the groups (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+The expression of this family in inflorescence meristem types was explored in greater detail using a heatmap of scaled, transformed read counts.
+Hierarchical clustering of the transformed counts recovered five prominent groups of expression (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
+There was no clear relationship between homeodomain subfamily and expression pattern except for an apparent enrichment of class IV HD-Zip genes with a pattern of lowest expression in the RM, a peak in indeterminate axillary meristems and then a gradual decrease during axillary meristem transitions.
 
@@
 
-Two results are presented here that suggest significant changes in gene expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
+Significant changes in gene expression between apical and axillary meristems are supported by two observations: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
@@
 
-Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM.
+Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM, which corresponds to the acquisition of determinate fate by the axillary meristem.
 The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
@@
 Two recent studies in *A.
-thaliana* and the domesticated tomato, *Solanum lycopersicum*, used LMD and RNA sequencing to study inflorescence development [@Park:2012dg; @Mantegazza:2014di].
+thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di].
 Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
@@
 The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
-Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
+In *A.
+thaliana*, *miR319a* targets *TEOSINTE BRANCHED/CYCLOIDEA/PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development (**ref Palatnik 2003**; **ref Nag 2009**), but no direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice.
+Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
 WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O.
@@
 *japonica* varieties.
-Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf].
+Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*) [@Huang:2009kf].
 WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].
@@
 The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM and small panicles with small SAMs and branching defects [@Kurakawa:2007go].
-Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O.
+Similarly, reduced expression *GN1A*, which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O.
 sativa* ssp.
@@
 *japonica* cv.
-Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele [**I think "*Laba1*" is correct, because the gene product from the domesticated allele (called *laba1* in the paper) is non-functional.
-Can someone pls confirm?
-Tom.**] from *O.
+Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O.
 rufipogon* into an *O.
@@
 In *A.
-thaliana*, class III homeodomain-lucine-zipper (HD-Zip) genes are involved in meristem initiation and regulation (**ref Prigge et al 2005**).
+thaliana*, class III homeodomain-leucine-zipper (HD-Zip) genes are involved in meristem initiation and regulation (**ref Prigge et al 2005**).
 The class III HD-Zip gene *REVOLUTA* is required for axillary meristem formation (**ref Talbert 1995**; **ref Otsuga 2001**).
@@
 
-## Summary
+## Conclusion
 
 This report describes the changes in gene expression that occur during the early stages of rice panicle development.
-A switch in expression between apical and axillary meristems followed by gradual changes between PBM and SM were observed, based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling.
-These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
+Based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch in expression between apical and axillary meristems was evident followed by gradual changes during transitions between PBM and SM.
+This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SMs in a process that may be more comparable to the gradual meristem maturation during tomato inflorescence development than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem [@Park:2012dg].
+These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
 
@@
 
+All rice genes are referred to by either their locus identifier [MSUv7; @Ouyang:2007il] or their official gene symbol provided on Oryzabase [@McCouch:2008ef].
+
 # Acknowledgements
+
+We thank James Tregear for feedback and discussion, and Jean-Luc Verdeil and Geneviève Conejero for assistance with laser microdissection at the PHIV imaging platform (CIRAD, Monpellier, France).
+This project was jointly supported by the Agropolis Fondation (ANR-10-LABX-0001-01) and Fondazione Cariplo (project EVOREPRICE 1201-004).
 
@@
 The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
-The expression patterns were separated into five groups using heirarchical clustering.
+The expression patterns were separated into five groups using hierarchical clustering.
 The set of homeobox genes with lowest expression in the RM and highest expression in indeterminate axillary meristems contains *OSH1* and five of the eight expressed class IV HD-Zip genes.

### 2015-12-15 si.Rmd
Stefan and Helene's comments from lab meeting Added acknowledgements (fixes #32) and more biology to the conclusion (fixes #31) Shorter title (fixes #25)

@@
   \usepackage{rotating}
+  \usepackage{siunitx}
   \renewcommand{\thefigure}{S\arabic{figure}}
@@
 Expression of \emph{LOC\_Os09g27730} (a--d), \emph{LOC\_Os01g04670} (e--h), \emph{LOC\_Os10g04270} (i--l) and \emph{LOC\_Os10g05990} (m--p) was analysed at the rachis meristem (a, e, i and m), primary branch meristem (b, f, j and n), elongating primary branch meristem plus axillary meristem (c, g, k and o) and spikelet meristem (d, h, l and p) stages.
-Each gene was detected in the same stage by \emph{in situ} hybridization as by RNA sequencing.}
+Each gene was detected in the same stage by \emph{in situ} hybridization as by RNA sequencing.
+The scale bars represent \SI{100}{\micro\metre}.}
 
@@
 Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering \\citep{@Love:2014do}.", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
+Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering (Love \\emph{et al}., 2014).", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
 For \\emph{A.
-thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by \\citep{Mantegazza:2014di}.
-The sympodial inflorescence meristem and floret meristem samples generated by \\citep{Park:2012dg} were used to estimate L\\textsubscript{2}FC values for tomato genes.
-Clades are labelled following the nomenclature presented by \\citep{Arora:2007fa}.")}
+thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by Mantegazza \\emph{et al}.
+(2014).
+The sympodial inflorescence meristem and floret meristem samples generated by Park \\emph{et al}.
+(2012) were used to estimate L\\textsubscript{2}FC values for tomato genes.")}
 print(sf_madsTree)
@@
 
-# Supporting experimental procedures
+# Methods S1
 
-## Methods S1: Comparative analysis of MADS family
+## Comparative analysis of the MADS family
 
@@
 Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`).
-L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same method described above for the rice samples.
-The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn].
+L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same `STAR`–`DESeq2` pipeline described in the main text for the rice samples [@Dobin:2012fg; @Love:2014do].
+The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn], with clade labels following the nomenclature presented by [@Arora:2007fa].
 

### 2015-12-15 ms.Rmd
Added number of panicles to st_libStats

@@
 RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
-RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number (RIN) was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`; **Number of panicles used for each library**).
+RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number (RIN) was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
 Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA (RIN < 6.5) were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).

### 2015-12-15 si.Rmd
Added number of panicles to st_libStats

@@
 xtable::align(st_libStats_x) <- rep("r", ncol(st_libStats_x) + 1)
-xtable::digits(st_libStats_x) <- c(rep(0, 3), 3, rep(1, 11), 0)
+xtable::digits(st_libStats_x) <- c(rep(0, 4), 3, rep(1, 11), 0)
 st_libStats_x_print <- print(st_libStats_x,

### 2015-12-15 src/R/parseQuantStats.R

@@ -43,6 +43,10 @@ area <- c(727647,511191,300886,848070, # N1
           600176,2735798,394641,439534, # N2
           1563411,1574719,1109344, # N3
           975315,758919,2347647) # N4
+nb <- c(7, 7, 5, 4, # N1
+        5, 4, 2, 3, # N2
+        5, 5, 4, # N3
+        5, 4, 4) # N4
 
 # function to get required info from STAR output
 parseStarInfo <- function(x){
@@ -107,6 +111,7 @@ setkey(rnaStats, 'lib')
 libStats <- expGenPerLib[readsPerLib][rnaStats[starInfo]]
 libStats[, "Yield (ng)" := round(inputRNA, 1)][, RIN := RIN ]
 libStats[, "Dissected area (mm²)" := area/1000000] 
+libStats[, "Number of panicles" := nb]
 
 # make all columns except 'lib' numeric
 libStats <- libStats[, lapply(.SD, as.numeric), by = lib]
@@ -124,7 +129,8 @@ libStats[, Sample := levels(GenomicRanges::colData(dds)$stage)[
 libStats[, Replicate := as.integer(gsub(".*R(\\d+)", "\\1", lib))]
 
 # arrange table
-setcolorder(libStats, c('lib', "Sample", "Replicate", "Dissected area (mm²)", 'Yield (ng)', 'RIN',
+setcolorder(libStats, c('lib', "Sample", "Replicate", "Number of panicles",
+                        "Dissected area (mm²)", 'Yield (ng)', 'RIN',
                         'Reads (M)', 'rRNA reads (M)', 'rRNA (%)',
                         'tRNA reads (M)', 'tRNA (%)',
                         'Uniquely mapped reads (M)', 'Multimapped reads (M)',

### 2015-12-15 ms.Rmd
James's comments

@@
 A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of potential candidates were identified.
-This resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield that integrates gene expression with environmental and other epigenetic factors, as well as for comparative studies of the domestication process.
+The new resource generated here will be vital for a mechanistic understanding of the link between inflorescence development and grain yield that integrates gene expression patterns with environmental and epigenetic factors, as well as for comparative studies of the domestication process.
 
@@
 Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
-Despite advances in characterization at the level of individual genes and interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia].
-One step towards this goal is to describe the changes in gene expression between different meristem types during development.
+Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia].
+One step towards this goal is to describe the differences in gene expression between different meristem types during development.
 In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O.
@@
 For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TF genes, whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
-The enrichment of different TF families in different clusters suggests complex control of the landscape of gene expression during inflorescence development.
+The enrichment of different TF and transcriptional regulator families in different clusters suggests complex control of the landscape of gene expression during inflorescence development.
 
@@
 G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
-*MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
-The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes.
+In this study, *MADS22* and *MADS55* decrease in expression between PBM and SM, *MADS55* being in the same cluster as *G1L5*, whereas *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes that include novel signalling components.
 Of the ALOG genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
@@
 This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*.
-Although the increase in secondary branches linked to *G1L5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.
+Although the increase in secondary branches linked to *G1L5* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.
 
@@
 Nipponbare [@Sato:1998tp], and a third, *OSH6*, has a peak in expression in the RM and PBM (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
-OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to phytohormone.
+OSH1 binds to the upstream region of the *OSH15* locus, and *osh1 osh15* double-mutants have lower induction of *OSH6* after treatment with cytokinin [@Tsuda:2011gz], supporting a role of *KNOX* genes in the response to this phytohormone.
 The higher expression of *OSH1* and *OSH15* in indeterminate axillary meristems is compatible with a function in the promotion of BM identity.
 In *A.
-thaliana*, cytokinin increases the production of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [@Leibfried:2005hw; @Lindsay:2006bh; @Gordon:2009fp].
+thaliana*, cytokinin promotes the expression of the homeodomain gene *WUSCHEL*, which itself enhances cytokinin signalling [@Leibfried:2005hw; @Lindsay:2006bh; @Gordon:2009fp].
 *wox1* plants carrying a likely null mutation in the rice ortholog of *WUSCHEL*, *WUSCHEL-LIKE HOMEOBOX 1* (*WOX1*/*WUS*) [@Nardmann:2006fd], have defects in axillary meristem formation and lower expression of *OSH1* [@Tanaka:2015ko].

### 2015-12-15 figures.Rmd
Helene's comments on the figure legends

@@
 
-```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614, fig.cap = paste0("Expression of selected \\emph{STMADS11}-like MADS genes (\\textbf{a}), detected \\emph{G1}-Like (G1L/ALOG) genes (\\textbf{b}), \\emph{SPL} genes and a microRNA precursor (\\textbf{c}) and cytokinin-related genes (\\textbf{d}) in inflorescence meristems in transcripts per million (TPM).
-A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
-Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} (Yoshida \\emph{et al}., 2013), are also more strongly expressed in the RM than in other meristems, but the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library.
-Only three G1L genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, but they share a similar pattern of expression.
-Several \\emph{SPL} genes and a co-regulated microRNA precursor, \\emph{MIR319A}, recovered from cluster ", getClusterName('small', 'dec')," are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.
-\\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/AM and SM than in RM and PBM.")}
+```{r f_alogFamily, fig.width = 6.614, fig.height = 6.614, fig.cap = paste0("Expression of selected genes in the RNAseq dataset in transcripts per million (TPM).
+A red point indicates that the gene was above the detection cutoff, and blue indicates that the gene was not detected.
+(\\textbf{a}) Two reported \\emph{STMADS11}-like (\\emph{SVP}-like) targets of G1L5, \\emph{MADS22} and \\emph{MADS55} (Yoshida \\emph{et al}., 2013), are more strongly expressed in the RM than in other meristems, and the other \\emph{STMADS11}-like gene, \\emph{MADS47}, was only detected in one library.
+(\\textbf{b}) Only three ALOG genes (\\emph{G1L1}, \\emph{G1L2} and \\emph{G1L5}/\\emph{TAWAWA1}) were detected, and they share a similar pattern of expression.
+(\\textbf{c}) Several \\emph{SPL} genes and a co-regulated microRNA precursor, \\emph{MIR319A}, which was recovered in cluster ", getClusterName('small', 'dec')," are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.
+(\\textbf{d}) \\emph{GN1A} (\\emph{CKX2}) was not detected in the LMD dataset, but \\emph{LOG1} is expressed in all meristem types, and two other genes possibly related to cytokinin activation, \\emph{LABA1} (\\emph{LOGL6}) and \\emph{LOGL1}, are expressed more highly in ePBM/AM and SM than in RM and PBM.")}
 # set layout

### 2015-12-15 ms.Rmd
Helene's comments on the figure legends

@@
 This report describes the changes in gene expression that occur during the early stages of rice panicle development.
-Based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch in expression between apical and axillary meristems was evident followed by gradual changes during transitions between PBM and SM.
+Based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch in expression between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
 This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SMs in a process that may be more comparable to the gradual meristem maturation during tomato inflorescence development than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem [@Park:2012dg].
@@
 **`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
-Expression of selected *STMADS11*-like MADS genes (**a**), detected *G1*-Like (G1L/ALOG) genes (**b**), *SPL* genes and a microRNA precursor (**c**) and cytokinin-related genes (**d**) in inflorescence meristems in transcripts per million (TPM).
-A red point indicates that the gene was above the detection cutoff, and blue represents that the gene was not detected.
-Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are also more strongly expressed in the RM than in other meristems, but the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
-Only three G1L genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, but they share a similar pattern of expression.
-Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, recovered from cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.
-*GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/AM and SM than in RM and PBM.
+Expression of selected genes in the RNAseq dataset in transcripts per million (TPM).
+A red point indicates that the gene was above the detection cutoff, and blue indicates that the gene was not detected.
+(**a**) Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are more strongly expressed in the RM than in other meristems, and the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
+(**b**) Only three ALOG genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, and they share a similar pattern of expression.
+(**c**) Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, which was recovered in cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.
+(**d**) *GN1A* (*CKX2*) was not detected in the LMD dataset, but *LOG1* is expressed in all meristem types, and two other genes possibly related to cytokinin activation, *LABA1* (*LOGL6*) and *LOGL1*, are expressed more highly in ePBM/AM and SM than in RM and PBM.
 

### 2015-12-16 ms.Rmd
References

@@
     number_sections: yes
+word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
 bibliography: bib/papersLib.bibtex
@@
 
-Homeodomain genes are involved at various stages of plant development and in several hormone response pathways (**ref Chan 1998**; **ref Himmelbach 2002**; **ref Sawa 2002**).
+Homeodomain genes are involved at various stages of plant development and in several hormone response pathways [@Chan:1998wq; @Himmelbach:2002jr; @Sawa:2002vm].
 The expression of this family in inflorescence meristem types was explored in greater detail using a heatmap of scaled, transformed read counts.
@@
 In rice, functions similar to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; **ref Dreni 2011**].
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; @Dreni:2011bu].
 Two recent studies in *A.
@@
 *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems.
-These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [**ref Jeong 2011**; @Wang:2015jx].
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx].
 The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
 In *A.
-thaliana*, *miR319a* targets *TEOSINTE BRANCHED/CYCLOIDEA/PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development (**ref Palatnik 2003**; **ref Nag 2009**), but no direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice.
+thaliana*, *miR319a* targets *TEOSINTE BRANCHED/CYCLOIDEA/PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development [@Palatnik:2003kt; @Nag:2009fc], but no direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice.
 Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
@@
 In *A.
-thaliana*, class III homeodomain-leucine-zipper (HD-Zip) genes are involved in meristem initiation and regulation (**ref Prigge et al 2005**).
-The class III HD-Zip gene *REVOLUTA* is required for axillary meristem formation (**ref Talbert 1995**; **ref Otsuga 2001**).
+thaliana*, class III homeodomain-leucine-zipper (HD-Zip) genes are involved in meristem initiation and regulation [@Prigge:2005gj].
+The class III HD-Zip gene *REVOLUTA* is required for axillary meristem formation [@Talbert:1995wa; @Otsuga:2001us].
 Seven of the nine annotated class III HD-Zip genes in *O.
-sativa* were detected (`r I(pasteLabel("Figure ", figCount, "f_hb"))`), including two homologs of *REVOLUTA*, *HOX9* and *HOX10* (**ref Prigge and Clark 2006**), which are both highly expressed in the RM.
+sativa* were detected (`r I(pasteLabel("Figure ", figCount, "f_hb"))`), including two homologs of *REVOLUTA*, *HOX9* and *HOX10* [@Prigge:2006ky], which are both highly expressed in the RM.
 Three class IV HD-Zip genes, *RICE OUTMOST CELL-SPECIFIC GENE 1* (*ROC1*), *ROC3* and *LOC_Os09g35760*, were recovered in cluster `r getClusterName('inc', 'small')`, containing genes that had lowest expression in the RM and higher expression in the samples from axillary meristems.
@@
 
-Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by (**ref Huijser 1992**).
+Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by @Huijser:1992wz.
 Digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions.
@@
 For detection of *LOC_Os09g27730*, an Digoxigenin-labelled LNA probe with the sequence 5′-TCTG{A}CGACG{T}GCG{A}C{T}GGT-3′ was synthesised (Eurogentec, Liège, Belgium).
-Hybridization was performed as described by **ref Coen 1990** using NBT/BCIP (Roche) or VECTOR Blue Alkaline Phosphatase Substrate Kit (Vector Laboratories) for detection.
+Hybridization was performed as described by @Coen:1990ux using NBT/BCIP (Roche) or VECTOR Blue Alkaline Phosphatase Substrate Kit (Vector Laboratories) for detection.
 

### 2015-12-16 figures.Rmd
Figure 1 legend (fixes #33)

@@
 Samples were collected from the rachis meristem (\textbf{a} and \textbf{e}), the primary branch meristem (\textbf{b} and \textbf{f}), the elongating primary branch meristem and axillary meristem (\textbf{c} and \textbf{g}) and the spikelet meristem (\textbf{d} and \textbf{h}).
-Images show the samples before (\textbf{a}–\textbf{d}) and after (\textbf{e}–\textbf{h}) dissection.}
+Images show the samples before (\textbf{a}–\textbf{d}) and after (\textbf{e}–\textbf{h}) dissection.
+Scale bars represent \SI{200}{\micro\metre} (red, \textbf{a}--\textbf{c} and \textbf{e}--\textbf{g}) or \SI{320}{\micro\metre} (blue, \textbf{d} and \textbf{h}).}
 \clearpage

### 2015-12-16 ms.Rmd
Figure 1 legend (fixes #33)
Suppress titles in SI and figures pdfs Remove phantom osh1 reference Add keywords

@@
 
-## SRA details for libraries
+Inflorescence development, rice, reproductive meristems, RNA sequencing
+
+## **Sequencing library repository**
 
@@
 
-Rice inflorescence meristems undergo a series of transitions during development, and control of these transitions impacts the morphology of the adult inflorescence and hence grain yield.
+Rice inflorescence meristems undergo a series of transitions during differentiation, and control of these transitions impacts the morphology of the adult inflorescence and hence grain yield.
 We use laser microdissection and RNA sequencing to describe the transcriptional changes that occur during this process.
@@
 This report describes the changes in gene expression that occur during the early stages of rice panicle development.
-Based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch in expression between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
-This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SMs in a process that may be more comparable to the gradual meristem maturation during tomato inflorescence development than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem [@Park:2012dg].
+Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
+This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SM.
+The latter process may be comparable to the gradual meristem maturation during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem.
 These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
@@
 Images show the samples before (**a**–**d**) and after (**e**–**h**) dissection.
+Scale bars represent 200 µm (red, **a**–**c** and **e**–**g**) or 320 µm (blue, **d** and **h**).
 

### 2015-12-16 src/R/figures.R

@@ -119,6 +119,9 @@ citekeys <- structure(c("@Ikeda:2007ja", "@Li:2011es", "@Ren:2013jv", "@Suzaki:2
 
 compare[, Reference := citekeys[as.character(zhangRef)]]
 
+# remove phantom OSH1 reference
+compare <- compare[!Reference == "**m.Komatsu2009**"]
+
 # SI table
 st_reviewInSitu <- data.table(reshape2::dcast(compare, msuId + Reference ~ stage,
                                               value.var = 'compare'))

### 2015-12-16 ms.Rmd
knit word

@@
     number_sections: yes
-word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
 bibliography: bib/papersLib.bibtex

### 2015-12-18 figures.Rmd
Suppress titles in SI and figures pdfs Remove phantom osh1 reference Add keywords
Rejigged intro and results for morphology section Moved in situ figure to main article Added m&m for morphology

--- 2015-12-15 figures.Rmd
@@
 ---
-title: "Figures"
+title: ""
 date: "\\vspace{0 mm}"
@@
   \usepackage{rotating}
+  \usepackage{siunitx}
   \renewcommand{\thefigure}{\arabic{figure}}
@@
 \end{figure*}
-\captionof{figure}{Laser microdissection of meristem samples in the early stages of inflorescence development.
-Samples were collected from the rachis meristem (\textbf{a} and \textbf{e}), the primary branch meristem (\textbf{b} and \textbf{f}), the elongating primary branch meristem and axillary meristem (\textbf{c} and \textbf{g}) and the spikelet meristem (\textbf{d} and \textbf{h}).
-Images show the samples before (\textbf{a}–\textbf{d}) and after (\textbf{e}–\textbf{h}) dissection.}
+\captionof{figure}{Morphology of the early stages of inflorescence development and laser microdissection of meristem samples.
+Toluidine blue stained sections of developing panicles at rachis meristem (RM; \textbf{a}), primary branch meristem (PBM; \textbf{b}), elongating primary branch meristem with axillary merisem (ePBM/AM; \textbf{c}) and spikelet meristem (SM; \textbf{d}) stages of differentiation.
+Laser microdissection (LMD) samples were collected from RM (\textbf{e} and \textbf{i}), PBM (\textbf{f} and \textbf{j}), ePBM/AM (\textbf{g} and \textbf{k}) and SM (\textbf{h} and \textbf{l}).
+Images show the samples before (\textbf{e}--\textbf{f}) and after (\textbf{i}--\textbf{l}) dissection.
+Scale bars represent \SI{100}{\micro\metre} (black, \textbf{a}--\textbf{d}), \SI{200}{\micro\metre} (red, \textbf{e}--\textbf{g} and \textbf{i}--\textbf{k}) or \SI{320}{\micro\metre} (blue, \textbf{h} and \textbf{l}).}
 \clearpage
+
+\begin{figure*}
+  \includegraphics{data/isFigure.pdf}
+\end{figure*}
+\captionof{figure}{RNA \emph{in situ} hybridization analysis to confirm the specific expression patterns of four genes detected in the RNA sequencing dataset.
+Expression of \emph{LOC\_Os09g27730} (a--d), \emph{LOC\_Os01g04670} (e--h), \emph{LOC\_Os10g04270} (i--l) and \emph{LOC\_Os10g05990} (m--p) was analysed at the rachis meristem (a, e, i and m), primary branch meristem (b, f, j and n), elongating primary branch meristem plus axillary meristem (c, g, k and o) and spikelet meristem (d, h, l and p) stages.
+Each gene was detected in the same stage by \emph{in situ} hybridization as by RNA sequencing.
+The scale bars represent \SI{100}{\micro\metre}.}
 

### 2015-12-18 ms.Rmd
Rejigged intro and results for morphology section Moved in situ figure to main article Added m&m for morphology

@@
 The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
-During panicle development, the undifferentiated cells in the primary axis (*i.e.* the rachis meristem, RM) undergo transitions through a series of meristem identities.
-After the establishment of the rachis some RM cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
-During primary branch elongation, the PBM is indeterminate and can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelets.
-The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM).
-Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo].
-After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].
-
-The landscape of gene expression is a characteristic that differentiates meristem types, and is involved in the control of transitions between meristem identities and therefore panicle branching.
+Grain yield is determined by the complexity and arrangement of branches and spikelets, which is specified in the early stages of panicle development by the establishement and activity of apical and axillary meristems in two phases: the number of primary branches before the abortion of the rachis meristem, and branching complexity by the timing of a series of transitions in axillary meristem identity.
+After spikelet differentiation, branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].
+The landscape of gene expression is a characteristic that differentiates meristem types and is involved in the control of meristem identity transitions.
 Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
@@
 
+## Panicle morphology and sampling for laser microdissection
+
+In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformation events.
+After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
+During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelets.
+The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM).
+Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret [@Ikeda:2004tk].
+A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD).
+After differentiation of the flag leaf primordium the first bract primordium appeared opposite to the flag leaf and at the base of the bract primordium the first stage of reproductive meristem appeared as RM (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
+After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b).
+After the formation of primary branches the rachis meristem stopped growing, bract growth also ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
+Each PBM produced a number of AMs as they elongated.
+For this analysis, SM differentition was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d).
+The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which the florets will differentiate [@Ikeda:2004tk].
+
 ## Genome-wide expression analysis of inflorescence development in rice
 
-To investigate gene expression during the development of the rice inflorescence, laser microdissection (LMD) was used to collect meristematic tissues from the rachis meristem (RM), primary branch meristem (PBM), elongating primary branch meristem with axillary meristem (ePBM/AM) and spikelet meristem (SM) of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`).
+To investigate gene expression during the development of the rice inflorescence, LMD was used to collect meristematic tissues from the RM, PBM, ePBM/AM and SM of early panicles (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`e–l).
 RNA isolated from the meristem samples was amplified and used to produce cDNA libraries for sequencing.
@@
 The expression of a further four uncharacterized genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization.
-Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).
+Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure ", figCount, "f_inSitu"))`).
 These results suggest that the tissues used for RNA sequencing were accurately dissected and represent the intended meristem types.
@@
 To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
+Using the list of transcription factor and other transcriptional regulator families available in the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
 After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, has an enrichment of Homeobox and basic helix-loop-helix TF genes, whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) is enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
+For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, had an enrichment of Homeobox and basic helix-loop-helix TF genes, whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) was enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
 The enrichment of different TF and transcriptional regulator families in different clusters suggests complex control of the landscape of gene expression during inflorescence development.
@@
 Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are more highly expressed in RM than in the other meristem types.
-Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/AM samples but depleted or not enriched in the RM samples.
+Another group, including MYB and SBP TF genes and SET and PHD regulator genes, was enriched in one or both of the PBM and ePBM/AM samples but depleted or not enriched in the RM samples.
 
@@
 
-Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
-G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
-In this study, *MADS22* and *MADS55* decrease in expression between PBM and SM, *MADS55* being in the same cluster as *G1L5*, whereas *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
+TAW1 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
+In this study, *MADS22* and *MADS55* decrease in expression between PBM and SM, *MADS55* being in the same cluster as *TAW1*, whereas *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
 The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes that include novel signalling components.
-Of the ALOG genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
-*g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff].
-This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*.
-Although the increase in secondary branches linked to *G1L5* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.
+Of the ALOG genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+*taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff].
+This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one being affected by the RNAi construct used against *TAW1*.
+Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.
 
@@
 
+## Morphological analysis
+
+Panicles were harvested in FAA (formaldehyde 10%, acetic acid 5%, ethanol 50%), infiltrated under vacuum for 15 minutes, stored overnight at 4°C and embedded in paraffin as described by @Huijser:1992wz.
+Tissues were cut into 8 µm sections using an RM2155 microtome (Leica), mounted on glass slides and stained with 0.5% w/v toluidine blue for observation with an Axiophot D1 microscope (Zeiss) and image capture with an Axiocam MRc 5 camera (Zeiss).
+
 ## Tissue embedding, laser microdissection and sequencing
@@
 We thank James Tregear for feedback and discussion, and Jean-Luc Verdeil and Geneviève Conejero for assistance with laser microdissection at the PHIV imaging platform (CIRAD, Monpellier, France).
-This project was jointly supported by the Agropolis Fondation (ANR-10-LABX-0001-01) and Fondazione Cariplo (project EVOREPRICE 1201-004).
+This project was jointly supported by Fondation Agropolis (ANR-10-LABX-0001-01) and Fondazione Cariplo (project EVOREPRICE 1201-004).
 
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`**: Comparison of the LMD RNA-sequencing dataset with published *in situ* results.
-
-**`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`**: RNA *in situ* hybridization analysis to confirm the specific expression patterns of four genes detected in the RNA sequencing dataset.
 
@@
 **`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`**.
-Laser microdissection of meristem samples in the early stages of inflorescence development.
-Samples were collected from the rachis meristem (**a** and **e**), the primary branch meristem (**b** and **f**), the elongating primary branch meristem and axillary meristem (**c** and **g**) and the spikelet meristem (**d** and **h**).
-Images show the samples before (**a**–**d**) and after (**e**–**h**) dissection.
-Scale bars represent 200 µm (red, **a**–**c** and **e**–**g**) or 320 µm (blue, **d** and **h**).
+Morphology of the early stages of inflorescence development and laser microdissection of meristem samples.
+Toluidine blue stained sections of developing panicles at rachis meristem (RM; **a**), primary branch meristem (PBM; **b**), elongating primary branch meristem with axillary merisem (ePBM/AM; **c**) and spikelet meristem (SM; **d**) stages of differentiation.
+Laser microdissection (LMD) samples were collected from RM (**e** and **i**), PBM (**f** and **j**), ePBM/AM (**g** and **k**) and SM (**h** and **l**).
+Images show the samples before (**e**–**f**) and after (**i**–**l**) dissection.
+Scale bars represent 100 µm (black, **a**–**d**), 200 µm (red, **e**–**g** and **i**–**k**) or 320 µm (blue, **h** and **l**).
 
 ----
+
+![](data/isFigure.pdf)
+
+**`r I(pasteLabel("Figure ", figCount, "f_inSitu"))`**.
+RNA *in situ* hybridization analysis to confirm the specific expression patterns of four genes detected in the RNA sequencing dataset.
+Expression of *LOC_Os09g27730* (**a**–**d**), *LOC_Os01g04670* (**e**–**h**), *LOC_Os10g04270* (**i**–**l**) and *LOC_Os10g059908* (**m**–**p**) was analysed at the RM (**a**, **e**, **i** and **m**), PBM (**b**, **f**, **j** and **n**), ePBM/AM (**c**, **g**, **k** and **o**) and SM (**d**, **h**, **l** and **p**) stages.
+Each gene was detected in the same stage by *in situ* hybridization as by RNA sequencing.
+The scale bars represent 100 µm.
+
+____
 
@@
 A red point indicates that the gene was above the detection cutoff, and blue indicates that the gene was not detected.
-(**a**) Two reported *STMADS11*-like (*SVP*-like) targets of G1L5, *MADS22* and *MADS55* [@Yoshida:2013ff], are more strongly expressed in the RM than in other meristems, and the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
-(**b**) Only three ALOG genes (*G1L1*, *G1L2* and *G1L5*/*TAWAWA1*) were detected, and they share a similar pattern of expression.
+(**a**) Two reported *STMADS11*-like (*SVP*-like) targets of TAW1, *MADS22* and *MADS55* [@Yoshida:2013ff], are more strongly expressed in the RM than in other meristems, and the other *STMADS11*-like gene, *MADS47*, was only detected in one library.
+(**b**) Only three ALOG genes (*G1L1*, *G1L2* and *G1L5*/*TAW1*) were detected, and they share a similar pattern of expression.
 (**c**) Several *SPL* genes and a co-regulated microRNA precursor, *MIR319A*, which was recovered in cluster `r getClusterName('small', 'dec')`, are highly expressed in RM and PBM before decreasing in expression in ePBM/AM and SM.

### 2015-12-18 si.Rmd
Suppress titles in SI and figures pdfs Remove phantom osh1 reference Add keywords
Rejigged intro and results for morphology section Moved in situ figure to main article Added m&m for morphology

@@
 ---
-title: "Supporting Information"
+title: ""
 csl: csl/the-plant-journal.csl
@@
 \clearpage
-\begin{figure*}
-  \includegraphics{data/isFigure.pdf}
-\end{figure*}
-\captionof{figure}{RNA \emph{in situ} hybridization analysis to confirm the specific expression patterns of four genes detected in the RNA sequencing dataset.
-Expression of \emph{LOC\_Os09g27730} (a--d), \emph{LOC\_Os01g04670} (e--h), \emph{LOC\_Os10g04270} (i--l) and \emph{LOC\_Os10g05990} (m--p) was analysed at the rachis meristem (a, e, i and m), primary branch meristem (b, f, j and n), elongating primary branch meristem plus axillary meristem (c, g, k and o) and spikelet meristem (d, h, l and p) stages.
-Each gene was detected in the same stage by \emph{in situ} hybridization as by RNA sequencing.
-The scale bars represent \SI{100}{\micro\metre}.}
-
-\clearpage
 \captionof{table}{\protect\label{tab:refFrame_x}References for Figure \ref{fig:sf_isGenesTpm}.}

### 2015-12-18 src/R/figures.R

@@ -249,17 +249,17 @@ nIS <- genTpm[, length(unique(msuId))]
 nPos <- genTpm[, sum(isExpr) > 1, by = .(msuId, stage)][
   , any(V1), by = msuId][, sum(V1)]
 
-######################
-### IN SITU FIGURE ###
-######################
-
-s_figCount <- incCount(s_figCount, "sf_inSitu")
-
 ##################
 ### LMD FIGURE ###
 ##################
 
 figCount <- incCount(figCount, "f_lmdFigure")
+
+######################
+### IN SITU FIGURE ###
+######################
+
+figCount <- incCount(figCount, "f_inSitu")
 
 #############
 ### Mfuzz ###

### 2015-12-18 ms.Rmd
Initial submission TPJ

@@
 
-| ^§^ Institut de Recherche pour le Développement, 911 Avenue Agropolis, 34394 Montpellier, France.
+| ^§^ Institut de Recherche pour le Développement, UMR DIADE, 911 Avenue Agropolis, 34394 Montpellier, France.
 | ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
 
-## Running title
+**Running title**: Expression profile of rice inflorescence meristems
 
-Expression profile of rice inflorescence meristems
+**Keywords**: Inflorescence development, rice, reproductive meristems, RNA sequencing
 
-## Keywords
+**Data availability**: Unprocessed sequencing libraries have been uploaded to the NCBI Sequence Read Archive under accession SRP067488.
 
-Inflorescence development, rice, reproductive meristems, RNA sequencing
-
-## **Sequencing library repository**
-
-## Word count
+**Word count**: 6,242 words.
 
@@
 One step towards this goal is to describe the differences in gene expression between different meristem types during development.
-In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O.
-sativa* ssp.
+In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *Oryza sativa* ssp.
 *japonica* cv.
@@
 Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
-Axillary meristem initiation requires auxin synthesis and transport in *A.
-thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice.
+Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice.
 Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*.
@@
 
-*Oryza sativa* ssp.
+*O.
+sativa* ssp.
 *japonica* cv.
@@
 **`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`**: Differential expression
-of genes encoding putative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*.
+of genes encoding putative MIKC proteins in *A. thaliana*, *S. lycopersicum* and *O. sativa* ssp. *japonica*.
 

### 2015-12-18 si.Rmd
Initial submission TPJ

@@
 
+```{r sf_isGenesTpm, fig.height = 9.535, fig.cap = paste0("\\protect\\label{fig:sf_isGenesTpm}","Comparison of the LMD RNA-sequencing dataset with published \\emph{in situ} results.
+Expression value in transcripts per million (TPM) is shown on the \\emph{y}-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below.
+Grey shading indicates a meristem type for which expression was detected by \\emph{in situ} hybridisation.
+See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
+Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).", "\\newline ", "Gene expression measured by RNAseq may differ from the qualitative patterns measured by RNA \\emph{in situ} hybridization for several reasons, including the relative sensitivity of the techniques, the specificity of the \\emph{in situ} probe, and differences in growth conditions (\\emph{e.g.
+HD5}) or the accession under study (\\emph{e.g.
+GN1A}).")}
+print(sf_isGenesTpm)
+```
+
+```{r sf_mfuzzPcaCentroids, fig.height = 6.741, fig.cap = paste0("Assessment of clustering results.
+\\textbf{a}.
+Minimum centroid distance for 2–", maxClust, " cluster cores (\\emph{c}).
+Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8.
+\\textbf{b}.
+Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
+Genes are coloured by the cluster to which they were assigned.
+The separation of genes by cluster on the MDS plot supports the use of eight clusters.")}
+grid::grid.draw(sf_mfuzzPcaCentroids)
+```
+
+```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putative MIKC proteins in \\emph{A.
+thaliana}, \\emph{S.
+lycopersicum} and \\emph{O.
+sativa} ssp.
+\\emph{japonica}.
+The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1).
+Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
+Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering (Love \\emph{et al}., 2014).", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
+For \\emph{A.
+thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by Mantegazza \\emph{et al}.
+(2014).
+The sympodial inflorescence meristem and floret meristem samples generated by Park \\emph{et al}.
+(2012) were used to estimate L\\textsubscript{2}FC values for tomato genes.")}
+print(sf_madsTree)
+```
+
 \clearpage
@@
 # remove the first hline
-st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
+st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
 cat(st_libStats_x_print)
 cat("¹ ePBM/AM")
-```
-
-```{r sf_isGenesTpm, fig.height = 9.535, fig.cap = paste0("\\protect\\label{fig:sf_isGenesTpm}","Comparison of the LMD RNA-sequencing dataset with published \\emph{in situ} results.
-Expression value in transcripts per million (TPM) is shown on the \\emph{y}-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below.
-Grey shading indicates a meristem type for which expression was detected by \\emph{in situ} hybridisation.
-See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
-Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).", "\\newline ", "Gene expression measured by RNAseq may differ from the qualitative patterns measured by RNA \\emph{in situ} hybridization for several reasons, including the relative sensitivity of the techniques, the specificity of the \\emph{in situ} probe, and differences in growth conditions (\\emph{e.g.
-HD5}) or the accession under study (\\emph{e.g.
-GN1A}).")}
-print(sf_isGenesTpm)
 ```
@@
 
-```{r sf_mfuzzPcaCentroids, fig.height = 6.741, fig.cap = paste0("Assessment of clustering results.
-\\textbf{a}.
-Minimum centroid distance for 2–", maxClust, " cluster cores (\\emph{c}).
-Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8.
-\\textbf{b}.
-Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
-Genes are coloured by the cluster to which they were assigned.
-The separation of genes by cluster on the MDS plot supports the use of eight clusters.")}
-grid::grid.draw(sf_mfuzzPcaCentroids)
-```
+# Methods S1: Comparative analysis of the MADS family
 
-```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putative MIKC proteins in \\emph{A.
-thaliana}, \\emph{Solanum lycopersicum} and \\emph{Oryza sativa} ssp.
-\\emph{japonica}.
-The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1).
-Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering (Love \\emph{et al}., 2014).", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
-For \\emph{A.
-thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by Mantegazza \\emph{et al}.
-(2014).
-The sympodial inflorescence meristem and floret meristem samples generated by Park \\emph{et al}.
-(2012) were used to estimate L\\textsubscript{2}FC values for tomato genes.")}
-print(sf_madsTree)
-```
-
-\clearpage
-
-# Methods S1
-
-## Comparative analysis of the MADS family
-
-The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
+The primary sequences of *A.
+thaliana*, *S.
+lycopersicum* and *O.
+sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
 An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn].
@@
 L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same `STAR`–`DESeq2` pipeline described in the main text for the rice samples [@Dobin:2012fg; @Love:2014do].
-The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn], with clade labels following the nomenclature presented by [@Arora:2007fa].
-
-\clearpage
+The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn], with clade labels following the nomenclature presented by @Arora:2007fa.
 

### 2016-01-21 ms.Rmd
Reviewer comments

@@
     number_sections: yes
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
+
 bibliography: bib/papersLib.bibtex
@@
 
-**Data availability**: Unprocessed sequencing libraries have been uploaded to the NCBI Sequence Read Archive under accession SRP067488.
+**Data availability**: Sequence data from this article have been deposited with the National Center for Biotechnology Information Sequence Read Archive under accession number SRP067488.
 
-**Word count**: 6,242 words.
+**Word count**:
 
@@
 
-In rice, inflorescence architecture is established at early stages of reproductive development by the activity of different meristem types, and contributes directly to grain yield potential.
-After induction of flowering, the shoot apical meristem is converted to an inflorescence meristem to form the rachis, which is the main axis of the inflorescence.
-The rachis meristem produces primary branch meristems, which themselves produce further axillary meristems during elongation.
-The axillary meristems may become secondary or higher-order branch meristems, retaining their indeterminate state, or be converted directly to determinate spikelet meristems.
-The timing of rachis meristem abortion specifies the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems conditions the complexity of branching and thereby the number of seeds on the panicle.
-Although some of the genes involved in transitions in meristem identity in rice have been identified, understanding of the network of genes controlling this process is lacking.
+In rice, inflorescence architecture is established at early stages of reproductive development and contributes directly to grain yield potential.
+After induction of flowering, the complexity of branching and therefore the number of seeds on the panicle are determined by the activity of different meristem types and the timing of transitions between them.
+Although some of the genes involved in these transitions have been identified, understanding of the network of transcriptional regulators controlling this process is lacking.
 To address this, we used a precise laser microdissection and RNA sequencing approach in *Oryza sativa* ssp.
@@
 Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types.
-A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of potential candidates were identified.
-The new resource generated here will be vital for a mechanistic understanding of the link between inflorescence development and grain yield that integrates gene expression patterns with environmental and epigenetic factors, as well as for comparative studies of the domestication process.
+A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of candidate genes were identified that may be involved in branching.
+The new resource generated here will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.
 
@@
 The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
-Grain yield is determined by the complexity and arrangement of branches and spikelets, which is specified in the early stages of panicle development by the establishement and activity of apical and axillary meristems in two phases: the number of primary branches before the abortion of the rachis meristem, and branching complexity by the timing of a series of transitions in axillary meristem identity.
+Grain yield is determined by the complexity and arrangement of branches and spikelets, which is specified in the early stages of panicle development by the establishement and activity of apical and axillary meristems in two phases.
+The timing of rachis meristem abortion specifies the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems conditions the complexity of branching and thereby the number of seeds on the panicle.
 After spikelet differentiation, branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].
@@
 
-In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformation events.
+In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations.
 After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
@@
 A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD).
-After differentiation of the flag leaf primordium the first bract primordium appeared opposite to the flag leaf and at the base of the bract primordium the first stage of reproductive meristem appeared as RM (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
+After differentiation of the flag leaf primordium the first bract primordium appeared opposite to the flag leaf, and the first stage of RM appeared at the base of the bract primordium (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
 After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b).
-After the formation of primary branches the rachis meristem stopped growing, bract growth also ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
+After the formation of primary branches, the rachis meristem stopped growing, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
 Each PBM produced a number of AMs as they elongated.
-For this analysis, SM differentition was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d).
+For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d).
 The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which the florets will differentiate [@Ikeda:2004tk].
@@
 RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number (RIN) was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
+A single sequencing library was produced from each biological replicate.
 Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA (RIN < 6.5) were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
@@
 At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
-Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage.
+Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage (Data S1).
 As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with patterns that had previously been reported.
@@
 To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do].
-Detected genes were ranked by variance, and standardized, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` most-variable genes (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
+Detected genes were ranked by variance, and standardized, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` genes with the highest variance (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
 To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
@@
 
-Examination of the clustered genes suggested that transcription factor (TF) genes were overrepresented in most clusters (Data S1).
-To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests.
-Using the list of transcription factor and other transcriptional regulator families available in the Plant Transcription Factor Database [@PerezRodriguez:2010co], a hypergeometric test was carried out if there were at least two members of one family in a given cluster.
-After adjustment for multiple comparisons, most clusters had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`).
-For example, cluster `r getClusterName('inc', 'small')`, containing genes that are more highly expressed in the axillary meristem (PBM–SM) samples than in the RM sample, had an enrichment of Homeobox and basic helix-loop-helix TF genes, whereas the cluster with the complementary pattern (cluster `r getClusterName('dec', 'small')`) was enriched for ABI3VP1 TFs and Aux/IAA auxin response genes.
-The enrichment of different TF and transcriptional regulator families in different clusters suggests complex control of the landscape of gene expression during inflorescence development.
-
 ## Dynamic expression of transcription factor families
 
-Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
+Examination of the clustered genes suggested that transcription factor (TF) genes were overrepresented in most clusters (Data S2).
+Using lists available in the Plant Transcription Factor Database [@PerezRodriguez:2010co], geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
 Unlike soft clustering, which was used to group genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members.
@@
 
+Other transcriptomic studies using whole inflorescences [*e.g.* @Furutani:2006hx], including those at later developmental stages than described here [*e.g.* @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im], have highlighted the importance of transcription factors in rice panicle development.
+LMD has also been used for precise control of developmental stage during the collection of whole panicle sections [@Kobayashi:2012fi].
+This microarray study identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in plants with defects in inflorescence development [@Kobayashi:2012fi].
+All four were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems.
+In contrast to these studies, only the meristematic regions of young panicles were collected by LMD in the RNAseq dataset described in this article, resulting in a narrow-scale transcriptomic analysis describing expression in these tissues alone.
+
 ## Switch in gene expression between apical and axillary meristems
 
-Significant changes in gene expression between apical and axillary meristems are supported by two observations: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM.
-Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM).
-Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice.
-Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*.
-In *A.
-thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db].
-GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling.
-Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
+Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM to SM).
+Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], and excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], but these processes are not as well understood in rice.
+Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters (*ALS1* and *MDR12*), while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*.
+Transgenic rice plants presumably expressing *GH3-8* in the endogenous pattern of *FLO-LFY HOMOLOG OF RICE* (*RFL*) [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co] have smaller panicles with fewer branches [@Yadav:2011db], and GH3-8 conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling.
+Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type.
+Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
+Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM both support significant changes in gene expression between apical and axillary meristems.
 
-Overexpression of the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
-TAW1 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff].
-In this study, *MADS22* and *MADS55* decrease in expression between PBM and SM, *MADS55* being in the same cluster as *TAW1*, whereas *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
-The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes that include novel signalling components.
-Of the ALOG genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+*MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* was not reported [@Yoshida:2013ff].
+Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
+*TAW1* was recovered from the same cluster as *MADS55*, and the presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components.
+Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset.
+Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), whilst the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
 *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff].
@@
 For example, the test statistic for the whole MADS family was higher in the SM than in the RM, but there are several MADS genes that have lower expression in the SM than in the RM, such as *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
-*PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
-In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative of the overall enrichment of MADS genes in the SM.
+*pap2* mutants have altered primary and secondary branch number, suggesting a role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
+*PAP2* also acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz].
+*LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative of the overall enrichment of MADS genes in the SM.
 
-Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM, which corresponds to the acquisition of determinate fate by the axillary meristem.
-The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants.
+Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM `r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`, which corresponds to the acquisition of determinate fate by the axillary meristem.
+The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants `r I(pasteLabel("Figure ", figCount, "f_gsea"))`.
 In the eudicot *A.
@@
 In rice, functions similar to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; @Dreni:2011bu].
+thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy].
+Stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; @Dreni:2011bu].
 Two recent studies in *A.
 thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di].
-Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
+Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^-type MADS genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
 *A.
@@
 *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems.
-These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx].
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx].
+*SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx].
 The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
 In *A.
-thaliana*, *miR319a* targets *TEOSINTE BRANCHED/CYCLOIDEA/PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development [@Palatnik:2003kt; @Nag:2009fc], but no direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice.
+thaliana*, *miR319a* targets *TEOSINTE BRANCHED*/*CYCLOIDEA*/*PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development [@Palatnik:2003kt; @Nag:2009fc].
+No direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice.
 Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
@@
 Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*) [@Huang:2009kf].
-WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].
+WFP also binds to the promoters of genes that are involved in the determination of inflorescence architecture, such as *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].
 
@@
 To generate the probes, cDNA was amplified using the following primers: *LOC_Os01g04670*, 5′-GTGTCAAGGCATCGCCAAC-3′ and 5′-CATCAGCTGGCTGCTTTACC-3′; *LOC_Os10g04270*, 5′-TCCCAGTTGACCGAGAACTG-3′ and 5′-TGAACTTCCGTCACGAACTCC-3′; *LOC_Os10g05990*, 5′-CCTCCGGCAAAGAACTGATG-3′ and 5′-CCGTCATTGGGACTAGTTTGTCAG-3′.
-For detection of *LOC_Os09g27730*, an Digoxigenin-labelled LNA probe with the sequence 5′-TCTG{A}CGACG{T}GCG{A}C{T}GGT-3′ was synthesised (Eurogentec, Liège, Belgium).
+For detection of *LOC_Os09g27730*, a Digoxigenin-labelled LNA probe with the sequence 5′-TCTG{A}CGACG{T}GCG{A}C{T}GGT-3′ was synthesised (Eurogentec, Liège, Belgium).
 Hybridization was performed as described by @Coen:1990ux using NBT/BCIP (Roche) or VECTOR Blue Alkaline Phosphatase Substrate Kit (Vector Laboratories) for detection.
@@
 
-**Data S1**: List of clustered genes with annotations.
+**Data S1**: Expression values for all genes in transcripts per million with annotations.
 
-# Tables
-
-```{r t_hypergeom, results = "asis"}
-pander::set.alignment('center', row.names = 'right')
-pander::set.caption(paste0(
-  I(pasteLabel("**Table ", tableCount, "t_hypergeom")),
-  "**: Transcription factor and transcriptional regulator family enrichment for families that had at least two members in any one cluster, including the number of family members (*n*), number expressed in the whole dataset (*n*~expr~) and number and adjusted *p*-value for each cluster."))
-pander::pandoc.table(t_hypergeom, split.tables = Inf, emphasize.rownames = FALSE)
-```
+**Data S2**: List of clustered genes with annotations.
 
@@
 The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.
-Although some test statistics had low *p*-values none were lower than 0.10 after Benjamini-Hochberg correction.
 
@@
 **`r I(pasteLabel("Figure ", figCount, "f_hb"))`**.
-Patterns of expression of homeobox genes that were detected in the meristem samples.
-The heatscale represents the geometric mean of scaled, VST-transformed read counts, and the coloured bar next to the *y*-axis shows the homeobox subfamily for each gene [@Jain:2008kg].
-The number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
-The expression patterns were separated into five groups using hierarchical clustering.
-The set of homeobox genes with lowest expression in the RM and highest expression in indeterminate axillary meristems contains *OSH1* and five of the eight expressed class IV HD-Zip genes.
+Five prominent patterns of expression of homeobox genes in meristems.
+Scaled, transformed read counts for the homeobox genes that were detected in the dataset are represented by the continuous heat scale.
+The colour of the *y*-axis denotes the homeobox subfamily for each gene [@Jain:2008kg], and the number of genes detected and the total number of genes for each subfamily is given in brackets after the subfamily name in the legend.
+The expression patterns were separated into five groups using hierarchical clustering, including a set of homeobox genes with lowest expression in the RM and highest expression in indeterminate axillary meristems, which contains *OSH1* and five of the eight expressed class IV HD-Zip genes.
 

### 2016-01-21 src/py/pipeline.py

@@ -625,6 +625,20 @@ def f_alogFamily(inputFiles, outputFiles):
 def f_hb(inputFiles, outputFiles):
     touch(outputFiles)
 
+#---------------------------------------------------------------
+# data s1
+#
+@merge([calculateTpm_R, downloadGenomes_sh], "ruffus/datas1")
+def datas1_R(inputFiles, outputFiles):
+    jobScript = 'src/R/datas1.R'
+    ntasks = '1'
+    cpus_per_task = '1'
+    job_name = 'datas1_R'
+    jobId = submit_job(jobScript, ntasks, cpus_per_task, job_name)
+    # update ruffus flag
+    print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
+    touch(outputFiles)
+
 
 
 # options for visualising

### 2016-01-21 ms.Rmd
Proof read

@@
     number_sections: yes
-  word_document:
-    fig_caption: yes
-    reference_docx: csl/wordStyles.docx
-
 bibliography: bib/papersLib.bibtex
@@
 Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types.
-A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of candidate genes were identified that may be involved in branching.
-The new resource generated here will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.
+A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed, and a number of candidate genes were identified that may be involved in branching.
+This new resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.
 
@@
 Agricultural development is essential to ensure food production and security for a growing population [@Borlaug:2007cp].
-Rice is a staple food for over half the world's population, including many developing countries, and a sustainable increase in the production of rice under the constraints of a changing climate and diminishing water and land availability will require plants with improved grain output, making the establishment of high-yield rice varieties a goal of modern breeding programs [@Peng:2008jb].
-Rice yield is a complex trait influenced by genetic and epigenetic factors, and is progressively defined during the life cycle of the plant, first during the vegetative phase, where the number of fertile tillers is established, and then during the reproductive phase and grain-filling phase.
-Yield is directly dependent on the architecture of the inflorescence [@Ikeda:2004tk].
+Rice is a staple food for over half the world's population, including many developing countries.
+A sustainable increase in the production of rice under the constraints of a changing climate and diminishing water and land availability will require plants with improved grain output, making the establishment of high-yield rice varieties a goal of modern breeding programs [@Peng:2008jb].
+Rice yield is a complex trait influenced by genetic and epigenetic factors, and is progressively defined during the life cycle of the plant, first during the vegetative phase, where the number of fertile tillers is established, and then during the reproductive phase and grain-filling phase [@Ikeda:2004tk].
 
@@
 The landscape of gene expression is a characteristic that differentiates meristem types and is involved in the control of meristem identity transitions.
-Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
+Mutant analyses and mapping of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
 Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq].
@@
 
-Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
+Understanding the events during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
 Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia].
@@
 For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d).
-The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which the florets will differentiate [@Ikeda:2004tk].
+The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which florets will differentiate [@Ikeda:2004tk].
 
@@
 RNA input for amplification was between `r min(st_libStats$"Yield (ng)")` and `r max(st_libStats$"Yield (ng)")` ng and the average RNA integrity number (RIN) was `r round(mean(st_libStats$"RIN"), 1)` (`r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`).
-A single sequencing library was produced from each biological replicate.
 Initially, two biological replicates were prepared for each stage, but after performing a principal components analysis (PCA) on transformed expression values, two libraries that were produced from degraded RNA (RIN < 6.5) were excluded from the analysis (`r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`).
 Following this, further samples were prepared to provide a total of three biological replicates with intact RNA (RIN ≥ 7) for each meristem type.
+A single sequencing library was produced from each biological replicate.
 
-At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA.
+At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million uniquely mapped reads within genes, after excluding reads likely to have originated from rRNA and tRNA.
 Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage (Data S1).
@@
 
-To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do].
+To recover common expression patterns, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do].
 Detected genes were ranked by variance, and standardized, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` genes with the highest variance (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw].
@@
 Hierarchical clustering of the transformed counts recovered five prominent groups of expression (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
-There was no clear relationship between homeodomain subfamily and expression pattern except for an apparent enrichment of class IV HD-Zip genes with a pattern of lowest expression in the RM, a peak in indeterminate axillary meristems and then a gradual decrease during axillary meristem transitions.
+In general, there was no clear relationship between homeodomain subfamily and expression pattern.
+However, there was an apparent enrichment of class IV HD-Zip genes with lowest expression in the RM, a peak in indeterminate axillary meristems and then a gradual decrease during axillary meristem transitions.
 
@@
 
-Other transcriptomic studies using whole inflorescences [*e.g.* @Furutani:2006hx], including those at later developmental stages than described here [*e.g.* @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im], have highlighted the importance of transcription factors in rice panicle development.
+Other transcriptomic studies using whole inflorescences [*e.g.* @Furutani:2006hx], including inflorescences at later developmental stages than described here [*e.g.* @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im], have highlighted the importance of transcription factors in rice panicle development.
 LMD has also been used for precise control of developmental stage during the collection of whole panicle sections [@Kobayashi:2012fi].
@@
 Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
-*TAW1* was recovered from the same cluster as *MADS55*, and the presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components.
+*TAW1* was recovered from the same cluster as *MADS55*.
+The presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components.
 Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset.
@@
 *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff].
-This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one being affected by the RNAi construct used against *TAW1*.
+This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one affected by the RNAi construct used against *TAW1*.
 Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.
@@
 The patterns of differential expression of TF genes presented in `r I(pasteLabel("Figure ", figCount, "f_gsea"))` are not intended to represent each member of each gene family.
-For example, the test statistic for the whole MADS family was higher in the SM than in the RM, but there are several MADS genes that have lower expression in the SM than in the RM, such as *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM.
+For example, the test statistic for the whole MADS family was higher in the SM than in the RM, but there are several MADS genes that have lower expression in the SM than in the RM.
+*PAP2* was strongly expressed at all stages with higher expression in the RM and PBM (Data S1).
 *pap2* mutants have altered primary and secondary branch number, suggesting a role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
@@
 
-Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM `r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`, which corresponds to the acquisition of determinate fate by the axillary meristem.
-The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants `r I(pasteLabel("Figure ", figCount, "f_gsea"))`.
+Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
+This transition corresponds to the acquisition of determinate fate by the axillary meristem.
+The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
 In the eudicot *A.
 thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity.
-*AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*) to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki].
+*AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki].
 In rice, functions similar to those provided by *A.
@@
 The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM and small panicles with small SAMs and branching defects [@Kurakawa:2007go].
-Similarly, reduced expression *GN1A*, which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O.
+Similarly, reduced expression of *GN1A*, which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O.
 sativa* ssp.
 *indica* varieties [@Ashikari:2005eg].
-Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role.
+Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity.
+However, the function of the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role.
 The LABA1 protein present in *O.
@@
 *japonica* cv.
-Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O.
+Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs].
+Introduction of the *Laba1* allele from *O.
 rufipogon* into an *O.
-sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy].
+sativa* ssp *indica* background, which contains a non-functional *laba1* allele, results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy].
 *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs].
@@
 Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d).
-Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level.
+Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is not clear how this would influence meristem identity or panicle branching at the molecular level.
 A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.
@@
 The core values for each cluster are plotted in black.
-The ordering of the panels in the plot was rearranged to enable side-by-side comparison of complementary clusters.
+The ordering of the panels in the plot was arranged to enable side-by-side comparison of complementary clusters.
 
@@
 **`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**.
-Geneset enrichment analysis (GSEA) for transcription factor and other transcriptional regulator families.
+Geneset enrichment analysis (GSEA) of transcription factor and other transcriptional regulator families.
 Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset.

### 2016-01-27 ms.Rmd
Round 2 revisions Fix clade labelling for new version of figtree Use oryzr to annotate cluster genes

@@
 ---
-title: Gene expression profiling of reproductive meristem identities in rice inflorescence
+title: Gene expression profiling of reproductive meristem identities in rice inflorescence by laser microdissection and RNA sequencing.
 csl: csl/the-plant-journal.csl
@@
 *japonica* cv.
-Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types.
-A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed, and a number of candidate genes were identified that may be involved in branching.
-This new resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.
+Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types: the rachis meristem, the primary branch meristem, the elongating primary branch meristem (including axillary meristems) and the spikelet meristem.
+A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed, and a number of genes potentially involved in branching were identified.
+This resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.
 
@@
 
-The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
-Grain yield is determined by the complexity and arrangement of branches and spikelets, which is specified in the early stages of panicle development by the establishement and activity of apical and axillary meristems in two phases.
-The timing of rachis meristem abortion specifies the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems conditions the complexity of branching and thereby the number of seeds on the panicle.
+The branched inflorescence of rice is a compound raceme, classified as a panicle, and is composed of a rachis (the main axis), primary branches, higher order branches, and spikelets.
+The number of seeds on the panicle is influenced by the complexity and arrangement of branches and spikelets.
+The establishement and activity of apical and axillary meristems conditions branching in two phases during the early stages of panicle development: the timing of rachis meristem abortion determines the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems specifies the complexity of branching.
 After spikelet differentiation, branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].
@@
 Mutant analyses and mapping of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq].
-Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq].
+Some of these genes are involved in the patterning of axillary meristems and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq].
 A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], some of which may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].
 
-Understanding the events during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
+Understanding the events during development that determine panicle characteristics such as branching complexity and its plasticity will be vital for sustainable improvement of rice yield potential using targeted breeding programs.
 Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia].
@@
 After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
-During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelets.
-The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM).
-Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret [@Ikeda:2004tk].
+During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelet meristem (SM).
+The PBM and secondary branch meristem (SBM) are both converted to a terminal SM.
+Each terminal or axillary SM produces one floret meristem (FM), which differentiates into a single floret [@Ikeda:2004tk].
 A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD).
-After differentiation of the flag leaf primordium the first bract primordium appeared opposite to the flag leaf, and the first stage of RM appeared at the base of the bract primordium (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
+After differentiation of the flag leaf, the first stage of reproductive development is the conversion of the shoot apical meristem (SAM) to the RM.
+The first bract primordium was produced opposite the flag leaf (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
 After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b).
-After the formation of primary branches, the rachis meristem stopped growing, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
+After the formation of primary branches, RM activity was lost, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
 Each PBM produced a number of AMs as they elongated.
@@
 Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage (Data S1).
-As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with patterns that had previously been reported.
+As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with previously reported patterns.
 `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`).
@@
 
-Other transcriptomic studies using whole inflorescences [*e.g.* @Furutani:2006hx], including inflorescences at later developmental stages than described here [*e.g.* @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im], have highlighted the importance of transcription factors in rice panicle development.
-LMD has also been used for precise control of developmental stage during the collection of whole panicle sections [@Kobayashi:2012fi].
-This microarray study identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in plants with defects in inflorescence development [@Kobayashi:2012fi].
-All four were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems.
-In contrast to these studies, only the meristematic regions of young panicles were collected by LMD in the RNAseq dataset described in this article, resulting in a narrow-scale transcriptomic analysis describing expression in these tissues alone.
+Several previous transcriptomic studies have measured gene expression in whole rice inflorescences at later stages of development than those described here [*e.g.* @Wang:2010im; @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im].
+Another study used microarray experiments at earlier stages to highlight the importance of transcription factors in panicle development [@Furutani:2006hx].
+LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi].
+This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi].
+All four genes were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems.
+In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article to restrict the measurement of gene expression to those tissues.
 
@@
 Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
-Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM both support significant changes in gene expression between apical and axillary meristems.
+Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM indicate a significant change in gene expression between apical and axillary meristems.
 
-Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
-*MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* was not reported [@Yoshida:2013ff].
+Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) was not detected (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a).
+*MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* has not been reported [@Yoshida:2013ff].
 Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff].
-*TAW1* was recovered from the same cluster as *MADS55*.
-The presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components.
+*TAW1* was recovered from the same cluster as *MADS55*, and the presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components.
 Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset.
-Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), whilst the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
+Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), and the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b).
 *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff].
@@
 thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di].
-Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^-type MADS genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
+These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type subgroup of MADS genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
 *A.
@@
 *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems.
-These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx].
-*SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx].
+These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], and *SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx].
 The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c).
@@
 thaliana*, *miR319a* targets *TEOSINTE BRANCHED*/*CYCLOIDEA*/*PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development [@Palatnik:2003kt; @Nag:2009fc].
-No direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice.
+No direct interaction with *SPL* genes or role in reproductive development has been reported for *miR319a* in rice.
 Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it].
@@
 Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*) [@Huang:2009kf].
-WFP also binds to the promoters of genes that are involved in the determination of inflorescence architecture, such as *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].
+WFP also binds to the promoters of genes that are involved in the determination of inflorescence architecture via the cytokinin, auxin and gibberelic acid pathways, including *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1* respectively, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].
 
@@
 Five of the eight detected class IV HD-Zip genes and several genes from other homeobox subclasses, including *OSH1*, also follow this pattern (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).
-*Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and overexpression of *OUTER CELL LAYER 1* delays flowering in transgenic plants [@Javelle:2011dz; @DepegeFargeix:2011ds].
+*Zea mays* class IV HD-Zip genes are expressed in the outer cell layer of the SAM and transgenic overexpression of *OUTER CELL LAYER 1* delays flowering [@Javelle:2011dz; @DepegeFargeix:2011ds].
 In rice, several class IV HD-Zip genes have been detected in the vegetative SAM by *in situ* hybridisation [@Ito:2003ta], but their role in axillary meristems in the developing inflorescence has not been explained.
@@
 
-This report describes the changes in gene expression that occur during the early stages of rice panicle development.
+This report describes changes in gene expression during the early stages of rice panicle development.
 Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
@@
 Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package [@Kumar:2007uw] and geneset enrichment analysis with the `gage` package [@Luo:2009kp] respectively.
+Gene lists were analysed using annotations from several databases, including TIGR [@Kawahara:2013gh], Oryzabase [@Kurata:2006gb] and OGRO [@Yamamoto:2012kw].
 All plots were produced with the `ggplot2` package [@Wickham:2009tn].
@@
 # Figure legends
-
-----
 
@@
 Geneset enrichment analysis (GSEA) of transcription factor and other transcriptional regulator families.
-Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset.
+Test statistics were calculated from the Log~2~-fold change between the stage of interest and all other stages for the expressed genes in each geneset.
 The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.
@@
 **`r I(pasteLabel("Figure ", figCount, "f_hb"))`**.
-Five prominent patterns of expression of homeobox genes in meristems.
+Five prominent patterns of expression of homeobox genes in reproductive meristems.
 Scaled, transformed read counts for the homeobox genes that were detected in the dataset are represented by the continuous heat scale.
@@
 **`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`**.
-Expression of selected genes in the RNAseq dataset in transcripts per million (TPM).
+Expression of selected genes  in transcripts per million (TPM).
 A red point indicates that the gene was above the detection cutoff, and blue indicates that the gene was not detected.
@@
 
-----
-
-# `sessionInfo` (*hide*)
-
-```{r sessionInfo, results = "asis"}
-cat(paste0("**branch:** _", system("git rev-parse --abbrev-ref HEAD", intern = TRUE), "_\n\n"))
-cat(paste0("**hash:** _", system("git rev-parse HEAD", intern = TRUE), "_\n\n"))
-pander::pander(sessionInfo())
-```
-
-# References (move to before **Tables**)
+# References (move to before **Figure Legends**)

### 2016-01-27 src/R/figures.R

@@ -130,6 +130,8 @@ st_reviewInSitu <- st_reviewInSitu[, .(
   Reference = paste(Reference, collapse = ", "))
   , by = c("msuId", "RM", "PBM", "SBM", "SM", "FM")]
 
+### THIS IS FAILING ###
+
 st_reviewInSitu[, `Gene symbol` :=
                   oryzr::LocToGeneName(msuId)$symbols,
                 by = msuId]
@@ -508,35 +510,58 @@ sf_madsTree <- sf_madsTree +
              label.r = unit(0.05, "lines")) +
   ggplot2::geom_text(size= 1.5, na.rm = TRUE, hjust = -0.01)
 
-# annotate clades
+# annotate the clades
 sf_madsTree <- sf_madsTree +
-  geom_cladelabel(node = 145, "AGL2-like",
-                  offset = 0.10, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 159, "AGL6-like",
-                  offset = 0.06, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 132, "TM3-like",
-                  offset = 0.06, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 168, "AG-like",
-                  offset = 0.06, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 165, "AGL12-like",
-                  offset = 0.06, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 178, "SQUA-like",
-                  offset = 0.08, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 192, "STMADS11-like",
-                  offset = 0.08, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 200, "AGL17-like",
-                  offset = 0.09, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 208, "FLC-like",
-                  offset = 0.06, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 214, "GLO-like",
-                  offset = 0.03, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 218, "DEF-like",
-                  offset = 0.04, fontsize = 1.5, barsize = 0.5) +
-  geom_cladelabel(node = 114, "MIKC*",
[…]

### 2016-01-27 src/py/pipeline.py

@@ -639,8 +639,6 @@ def datas1_R(inputFiles, outputFiles):
     print("[", print_now(), ": Job " + job_name + " run with JobID " + jobId + " ]")
     touch(outputFiles)
 
-
-
 # options for visualising
 pipeline_printout()
 pipeline_printout_graph("ruffus/flowchart." + slurm_jobid + ".pdf", "pdf")

### 2016-02-01 ms.Rmd
Removed mads guff Reverted title Switched to pdf (from cairo) for cmyk

@@
 ---
-title: Gene expression profiling of reproductive meristem identities in rice inflorescence by laser microdissection and RNA sequencing.
+title: Gene expression profiling of reproductive meristem identities in rice inflorescence
 csl: csl/the-plant-journal.csl
@@
                cache.path='cache/',
-               dev=c("png", "cairo_pdf"),
+               dev=c("png", "pdf"),
+               dev.args=list(
+                 pdf = list(colormodel = 'cmyk'),
+                 png = list()),
                fig.width=3.150,
@@
 All four genes were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems.
-In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article to restrict the measurement of gene expression to those tissues.
+In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article, to restrict the measurement of gene expression to those tissues.
 
@@
 
-The patterns of differential expression of TF genes presented in `r I(pasteLabel("Figure ", figCount, "f_gsea"))` are not intended to represent each member of each gene family.
-For example, the test statistic for the whole MADS family was higher in the SM than in the RM, but there are several MADS genes that have lower expression in the SM than in the RM.
-*PAP2* was strongly expressed at all stages with higher expression in the RM and PBM (Data S1).
-*pap2* mutants have altered primary and secondary branch number, suggesting a role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
-*PAP2* also acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz].
-*LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative of the overall enrichment of MADS genes in the SM.
-
 Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`).
 This transition corresponds to the acquisition of determinate fate by the axillary meristem.
-The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`).
-In the eudicot *A.
-thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity.
-*AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki].
-In rice, functions similar to those provided by *A.
-thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy].
-Stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; @Dreni:2011bu].
+The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants [`r I(pasteLabel("Figure ", figCount, "f_gsea"))`; *e.g.* @Kyozuka:2002vm; @Nagasawa:2003gu;  @Yamaguchi:2006br; @Yao:2008dy; @Dreni:2011bu; the role of MADS genes in Arabidopsis flower development was reviewed by @Prunet:2013ki].
 Two recent studies in *A.
 thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di].
-These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type subgroup of MADS genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
+These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type class of MADS-box genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
 *A.
@@
 Toluidine blue stained sections of developing panicles at rachis meristem (RM; **a**), primary branch meristem (PBM; **b**), elongating primary branch meristem with axillary merisem (ePBM/AM; **c**) and spikelet meristem (SM; **d**) stages of differentiation.
+The position of the first bract primordium (bp) and flag leaf (fl) are indicated in **a**.
 Laser microdissection (LMD) samples were collected from RM (**e** and **i**), PBM (**f** and **j**), ePBM/AM (**g** and **k**) and SM (**h** and **l**).
 Images show the samples before (**e**–**f**) and after (**i**–**l**) dissection.
-Scale bars represent 100 µm (black, **a**–**d**), 200 µm (red, **e**–**g** and **i**–**k**) or 320 µm (blue, **h** and **l**).
+Scale bars represent 50 µm (**a**–**c**), 100 µm (**d**), 200 µm (**e**–**g** and **i**–**k**) or 320 µm (**h** and **l**).
 

### 2016-02-01 ms.Rmd
New conclusion

@@
 
-This report describes changes in gene expression during the early stages of rice panicle development.
-Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
+The molecular mechanisms controlling the meristematic activities that guide panicle development in rice remain largely uncharacterized.
+The dataset presented here addresses this by describing the transcriptome of specific meristem identities and the changes in gene expression during the early stages of panicle development.
+A switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
 This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SM.
 The latter process may be comparable to the gradual meristem maturation during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem.
-These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
+Promising uncharacterized genes were identified, including transcription factors with expression patterns similar to other regulators of panicle architecture, and others potentially related to hormone signalling such as the auxin-responsive genes in clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')`.
+Functional studies of these genes will reveal the mechanics of the regulatory networks that establish gene expression in meristems and their control by hormone signalling.
+Genes with expression specific to certain meristem types were also identified, which may additionally serve as valuable marker genes, for example in developmental studies of mutants affected in determination of reproductive meristem identity.
+Combined with current technology for the creation of mutants, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, which will be vital for the sustainable increase in grain yield required by population growth.
 

### 2016-02-03 ms.Rmd
typo
Comments stefan helene james

--- 2016-02-01 ms.Rmd
@@
 ---
-title: Gene expression profiling of reproductive meristem identities in rice inflorescence
+title: Gene expression profiling of reproductive meristem types in early rice inflorescences by laser microdissection
 csl: csl/the-plant-journal.csl
@@
 
-In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations.
-After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches.
+In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations [described by @Ikeda:2004tk].
+A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD).
+After differentiation of the flag leaf, the first stage of reproductive development is the conversion of the shoot apical meristem (SAM) to the rachis meristem (RM).
+The first bract primordium was produced opposite the flag leaf (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
+After the establishment of the RM, some cells differentiate into primary branch meristem (PBM) in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b).
+After the formation of primary branches, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
 During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelet meristem (SM).
+For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d).
 The PBM and secondary branch meristem (SBM) are both converted to a terminal SM.
-Each terminal or axillary SM produces one floret meristem (FM), which differentiates into a single floret [@Ikeda:2004tk].
-A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD).
-After differentiation of the flag leaf, the first stage of reproductive development is the conversion of the shoot apical meristem (SAM) to the RM.
-The first bract primordium was produced opposite the flag leaf (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a).
-After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b).
-After the formation of primary branches, RM activity was lost, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c).
-Each PBM produced a number of AMs as they elongated.
-For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d).
+Each terminal or axillary SM produces one floret meristem (FM), which differentiates into a single floret.
 The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which florets will differentiate [@Ikeda:2004tk].
@@
 LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi].
-This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi].
-All four genes were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems.
+This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi].
+All four genes were detected in all stages in the RNAseq dataset (Data S1), confirming their expression in reproductive meristems.
 In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article, to restrict the measurement of gene expression to those tissues.
@@
 Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.
-Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM indicate a significant change in gene expression between apical and axillary meristems.
 
@@
 Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.
+The difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM indicate a significant change in gene expression between apical and axillary meristems.
 
@@
 thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.
+*PAP2*, which was strongly expressed at all stages with higher expression in the RM and PBM (Data S1), acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz].
+*pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role is spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn].
+*LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was enriched in the SM (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
 
@@
 
-This report describes changes in gene expression during the early stages of rice panicle development.
-Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM.
-This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SM.
-The latter process may be comparable to the gradual meristem maturation during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem.
-These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
+The molecular mechanisms controlling the meristematic activities that guide panicle development in rice remain largely uncharacterized.
+The dataset presented here addresses this by describing both the transcriptome profiles associated with specific meristem identities and the changes in gene expression that occur during the early stages of panicle development.
+A marked expression switch was evident between apical and axillary meristems, followed by gradual changes during transitions between PBM and SM.
+This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which undergo transition to determinate SMs.
+The latter process may be comparable to the gradual meristem maturation observed during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems characteristic of the Arabidopsis inflorescence meristem.
+Promising uncharacterized genes were identified, notable examples including transcription factor genes with expression patterns similar to known regulators of panicle architecture and putative hormone-related such as the auxin-responsive genes.
+Functional studies of these genes will reveal the mechanics of the regulatory networks that establish gene expression patterns in meristems and their control by hormone signalling.
+Genes with expression specific to certain meristem types were also identified, which may prove valuable as marker genes, for example in developmental studies of mutants affected in determination of reproductive meristem identity.
+Combined with current technology for the creation of mutants, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, a vital component of efforts to achieve the sustainable increase in grain yield required to address population growth.
 
@@
 Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by @Huijser:1992wz.
-Digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions.
+digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions.
 To generate the probes, cDNA was amplified using the following primers: *LOC_Os01g04670*, 5′-GTGTCAAGGCATCGCCAAC-3′ and 5′-CATCAGCTGGCTGCTTTACC-3′; *LOC_Os10g04270*, 5′-TCCCAGTTGACCGAGAACTG-3′ and 5′-TGAACTTCCGTCACGAACTCC-3′; *LOC_Os10g05990*, 5′-CCTCCGGCAAAGAACTGATG-3′ and 5′-CCGTCATTGGGACTAGTTTGTCAG-3′.
@@
 Analysis of the sequenced libraries was carried out with `bash` and `R` scripts [@RALanguageandEn:2015wf], which were arranged into a pipeline using the `ruffus` package for `python3` [@Goodstadt:2010jp].
-All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.
+All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and parameters of the software used, are available with revision history in a public GitHub repository at [https://github.com/evoreprice/lmdPaper](https://github.com/evoreprice/lmdPaper).
 
@@
 The `'quantMode'` argument of `STAR` was used for read counting.
-rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` (https://github.com/lh3/wgsim) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
+rRNA and tRNA contamination was estimated using `htseq-count` [@Anders:2014gf] to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB [@Ohyanagi:2006go] or to regions where reads simulated with `wgsim` ([https://github.com/lh3/wgsim](https://github.com/lh3/wgsim)) from rRNA sequences in the TIGR Plant Repeats Database [@Ouyang:2004ia] also mapped.
 Differential expression analysis to calculate Log~2~-fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package [@Love:2014do].
@@
 We thank James Tregear for feedback and discussion, and Jean-Luc Verdeil and Geneviève Conejero for assistance with laser microdissection at the PHIV imaging platform (CIRAD, Monpellier, France).
-This project was jointly supported by Fondation Agropolis (ANR-10-LABX-0001-01) and Fondazione Cariplo (project EVOREPRICE 1201-004).
+This project was jointly supported by Fondation Agropolis through the *Investissements d'avenir* programme (ANR-10-LABX-0001-01) and Fondazione Cariplo under the EVOREPRICE 1201-004 project.
 

### 2016-02-03 ms.Rmd
Comments Martin Vero Israr

@@
 
-Thomas W.R. Harrop^§,1^, Israr Ud Din^¶,^[^1], Veronica Gregis^¶^, Michela Osnato^¶^, Stefan Jouannic^§^, Hélène Adam^§,2^ and Martin Kater^¶,^[^2]
+Thomas W.R. Harrop^§,1^, Israr Ud Din^¶,†,^[^1], Veronica Gregis^¶^, Michela Osnato^¶,‡^, Stefan Jouannic^§^, Hélène Adam^§,2^ and Martin Kater^¶,†,^[^2]
 
@@
 | ^§^ Institut de Recherche pour le Développement, UMR DIADE, 911 Avenue Agropolis, 34394 Montpellier, France.
-| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
+| ^¶^ Dipartimento di Bioscienze, Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
+| ^†^ Dipartimento di Scienze Farmacologiche e Biomolecolari, Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
+| ^‡^ Present address: Centre for Research in Agricultural Genomics, Campus UAB, 08193 Cerdanyola, Spain.
 
@@
 This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi].
-All four genes were detected in all stages in the RNAseq dataset (Data S1), confirming their expression in reproductive meristems.
-In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article, to restrict the measurement of gene expression to those tissues.
+All four genes were detected in all stages in the RNAseq dataset presented in this article (Data S1), confirming their expression in reproductive meristems.
+In contrast to previous transcriptomic datasets, only the meristematic regions of young panicles were collected by LMD for the RNAseq analysis described here, to restrict the measurement of gene expression to those tissues.
 
@@
 The latter process may be comparable to the gradual meristem maturation observed during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems characteristic of the Arabidopsis inflorescence meristem.
-Promising uncharacterized genes were identified, notable examples including transcription factor genes with expression patterns similar to known regulators of panicle architecture and putative hormone-related such as the auxin-responsive genes.
+Promising uncharacterized genes were identified, notable examples including transcription factor genes with expression patterns similar to known regulators of panicle architecture, and putative hormone-related genes such as the auxin-responsive genes.
 Functional studies of these genes will reveal the mechanics of the regulatory networks that establish gene expression patterns in meristems and their control by hormone signalling.
 Genes with expression specific to certain meristem types were also identified, which may prove valuable as marker genes, for example in developmental studies of mutants affected in determination of reproductive meristem identity.
-Combined with current technology for the creation of mutants, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, a vital component of efforts to achieve the sustainable increase in grain yield required to address population growth.
+Combined with current technology for the creation of mutants such as CRISPR/Cas9 genome editing, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, which is a vital for efforts to achieve the sustainable increase in grain yield required to address population growth.
 

### 2016-02-03 ms.Rmd
Revert capitalisation error

@@
 Samples harvested from the main stem from four developmental stages of panicle development were embedded in Paraplast X-TRA (Sigma-Aldrich) as described by @Huijser:1992wz.
-digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions.
+Digoxigenin-labelled antisense and sense RNA probes were generated with the DIG RNA Labeling Kit SP6/T7 (Roche) according to the manufacturer's instructions.
 To generate the probes, cDNA was amplified using the following primers: *LOC_Os01g04670*, 5′-GTGTCAAGGCATCGCCAAC-3′ and 5′-CATCAGCTGGCTGCTTTACC-3′; *LOC_Os10g04270*, 5′-TCCCAGTTGACCGAGAACTG-3′ and 5′-TGAACTTCCGTCACGAACTCC-3′; *LOC_Os10g05990*, 5′-CCTCCGGCAAAGAACTGATG-3′ and 5′-CCGTCATTGGGACTAGTTTGTCAG-3′.

### 2016-02-09 Legends for Supporting Information.Rmd
Illogical layout required by TPJ

---
title: ""
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
bibliography: bib/papersLib.bibtex
---

```{r Setup, include=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               fig.align = 'center', 
               dpi=600,
               fig.show='hold',
               fig.height = 4.213,
               fig.width = 4.213,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

source('src/R/figures.R')
```

# `r I(pasteLabel("Figure S", s_figCount, "sf_pca"))`

Principal components analysis of transformed read counts for each library. The two libraries made from degraded RNA did not cluster with other libraries (see `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`), suggesting that the use of degraded RNA affected quantification of expression. For this reason these two libraries were not included in subsequent analyses.

# `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`

Comparison of the LMD RNA-sequencing dataset with published *in situ* results. Expression value in transcripts per million (TPM) is shown on the *y*-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below. Grey shading indicates a meristem type for which expression was detected by *in situ* hybridisation. See `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))` for references for the *in situ* expression patterns. Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).

Gene expression measured by RNAseq may differ from the qualitative patterns measured by RNA *in situ* hybridization for several reasons, including the relative sensitivity of the techniques, the specificity of the *in situ* probe, and differences in growth conditions (*e.g. HD5*) or the accession under study (*e.g. GN1A*).

# `r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`

Assessment of clustering results. **a**. Minimum centroid distance for 2–`r maxClust` cluster cores (*c*). Eight cluster cores were indicated because the rate of change of the minimum centroid distance over *c* decreased after *c* = 8. **b**. Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes. Genes are coloured by the cluster to which they were assigned. The separation of genes by cluster on the MDS plot supports the use of eight clusters.

# `r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`

Differential expression of genes encoding putative MIKC proteins in *A. thaliana*, *S. lycopersicum* and *O. sativa* ssp. *japonica*. The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1). Named proteins starting with AT_ or OS_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.

The colour of the label represents the Log~2~-fold change (L~2~FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the adjusted *p*-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by `DESeq2`, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering [@Love:2014do].

For rice, L~2~FC was estimated from the RM and SM samples reported here. For *A. thaliana*, L~2~FC was estimated using the inflorescence meristem and floret meristem samples reported by @Mantegazza:2014di. The sympodial inflorescence meristem and floret meristem samples generated by @Park:2012dg were used to estimate L~2~FC values for tomato genes.

# `r I(pasteLabel("Table S", s_tableCount, "st_libStats"))`

Library and sequencing statistics.

# `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`

References for `r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))`.

```{r insituRefs, results = "asis"}
pander::pandoc.table(st_refTable, emphasize.cols = c(1,2))
```

# Data S1

Expression values for all genes in transcripts per million with annotations.

# Data S2

List of clustered genes with annotations.

# References for Supporting Information

### 2016-02-09 Methods S1.Rmd
Illogical layout required by TPJ

--- 2015-12-18 si.Rmd
@@
 csl: csl/the-plant-journal.csl
-date: "\\vspace{0 mm}"
 header-includes:
   \usepackage{graphicx}
-  \usepackage{caption}
-  \usepackage{fixltx2e}
-  \usepackage{rotating}
-  \usepackage{siunitx}
-  \renewcommand{\thefigure}{S\arabic{figure}}
-  \renewcommand{\thetable}{S\arabic{table}}
-documentclass: tufte-handout
-classoption: a4paper
+  \usepackage[utf8]{inputenc}
+  \DeclareUnicodeCharacter{2265}{$\geq$}
 output:
-  rmarkdown::tufte_handout:
-    dev: CairoPDF
-    keep_tex: yes
-    fig_crop: no
+  word_document:
+    fig_caption: yes
+    reference_docx: csl/wordStyles.docx
 bibliography: bib/papersLib.bibtex
@@
 
-```{r sf_pca, fig.cap = paste0("Principal components analysis of transformed read counts for each library.
-The two libraries made from degraded RNA did not cluster with other libraries (see Table \\ref{tab:st_libstats}), suggesting that the use of degraded RNA affected quantification of expression.
-For this reason these two libraries were not included in subsequent analyses.")}
-print(sf_pca)
-```
-
-```{r sf_isGenesTpm, fig.height = 9.535, fig.cap = paste0("\\protect\\label{fig:sf_isGenesTpm}","Comparison of the LMD RNA-sequencing dataset with published \\emph{in situ} results.
-Expression value in transcripts per million (TPM) is shown on the \\emph{y}-axis, and points are plotted as a triangle if the expression value was above the cutoff for that library or a circle if below.
-Grey shading indicates a meristem type for which expression was detected by \\emph{in situ} hybridisation.
-See Table \\ref{tab:refFrame_x} for references for the \\emph{in situ} expression patterns.
-Although four replicates were produced from each of the RM and PBM, only three replicates for each stage are included in the analysis (see Results).", "\\newline ", "Gene expression measured by RNAseq may differ from the qualitative patterns measured by RNA \\emph{in situ} hybridization for several reasons, including the relative sensitivity of the techniques, the specificity of the \\emph{in situ} probe, and differences in growth conditions (\\emph{e.g.
-HD5}) or the accession under study (\\emph{e.g.
-GN1A}).")}
-print(sf_isGenesTpm)
-```
-
-```{r sf_mfuzzPcaCentroids, fig.height = 6.741, fig.cap = paste0("Assessment of clustering results.
-\\textbf{a}.
-Minimum centroid distance for 2–", maxClust, " cluster cores (\\emph{c}).
-Eight cluster cores were indicated because the rate of change of the minimum centroid distance over \\emph{c} decreases after \\emph{c} = 8.
-\\textbf{b}.
-Classical multidimensional scaling (MDS) of Euclidean distances between transformed read counts for clustered genes.
-Genes are coloured by the cluster to which they were assigned.
-The separation of genes by cluster on the MDS plot supports the use of eight clusters.")}
-grid::grid.draw(sf_mfuzzPcaCentroids)
-```
-
-```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putative MIKC proteins in \\emph{A.
-thaliana}, \\emph{S.
-lycopersicum} and \\emph{O.
-sativa} ssp.
-\\emph{japonica}.
-The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1).
-Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems.
-Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering (Love \\emph{et al}., 2014).", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here.
-For \\emph{A.
-thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by Mantegazza \\emph{et al}.
-(2014).
-The sympodial inflorescence meristem and floret meristem samples generated by Park \\emph{et al}.
-(2012) were used to estimate L\\textsubscript{2}FC values for tomato genes.")}
-print(sf_madsTree)
-```
-
-\clearpage
-\captionof{table}{\protect\label{tab:st_libstats}Library and sequencing statistics.}
-
-``` {r st_libstats, results = "asis"}
-st_libStats_x <- as.data.frame(st_libStats)
-rownames(st_libStats_x) <- st_libStats_x$Library
-st_libStats_x$Library <- NULL
-st_libStats_x$Sample <- sub("ePBM/SBM", "AM¹", st_libStats_x$Sample, fixed = TRUE)
-st_libStats_x <- xtable::xtable(st_libStats_x)
-xtable::align(st_libStats_x) <- rep("r", ncol(st_libStats_x) + 1)
-xtable::digits(st_libStats_x) <- c(rep(0, 4), 3, rep(1, 11), 0)
-st_libStats_x_print <- print(st_libStats_x,
-      caption.placement = "top",
-      comment = FALSE,
-      rotate.colnames = TRUE,
-      floating = TRUE,
-      floating.environment = "table*",
-      print.results = FALSE)
-
-# remove the first hline
-st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
-cat(st_libStats_x_print)
-cat("¹ ePBM/AM")
-```
-
-\clearpage
-\captionof{table}{\protect\label{tab:refFrame_x}References for Figure \ref{fig:sf_isGenesTpm}.}
-
-```{r insituRefs, results = "asis"}
-pander::pandoc.table(st_refTable, emphasize.cols = c(1,2))
-```
-
 # Methods S1: Comparative analysis of the MADS family
@@
 
-# References for supporting information
+# References for Methods S1

### 2016-02-09 ms.Rmd
Illogical layout required by TPJ

@@
 thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di].
-These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type class of MADS-box genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`).
+These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type class of MADS-box genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`; Methods S1).
 *A.

### 2016-02-09 st_refTable.Rmd
Illogical layout required by TPJ

--- 2016-02-09 Methods S1.Rmd
@@
 
-# Methods S1: Comparative analysis of the MADS family
-
-The primary sequences of *A.
-thaliana*, *S.
-lycopersicum* and *O.
-sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki].
-An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn].
-To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed.
-To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible.
-Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`).
-L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same `STAR`–`DESeq2` pipeline described in the main text for the rice samples [@Dobin:2012fg; @Love:2014do].
-The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn], with clade labels following the nomenclature presented by @Arora:2007fa.
-
-# References for Methods S1
+```{r insituRefs, results = "asis"}
+data.table::setnames(st_refTable, "MSU name", "MSU ID")
+pander::pandoc.table(st_refTable, emphasize.cols = c(1,2))
+```

### 2016-02-09 src/R/figures.R

@@ -48,6 +48,12 @@ capwords <- function(s, strict = FALSE) {
 
 st_libStats <- readRDS('output/quantStats/libStats.Rds')
 s_tableCount <- incCount(s_tableCount, "st_libStats")
+# write as tsv for tpj
+st_libStats[Sample == "ePBM/SBM", Sample := "ePBM/AM"]
+file.name <- I(pasteLabel("Table S", s_tableCount, "st_libStats"))
+write.table(st_libStats, file = paste0(
+  "xlsx/", file.name, ".tsv"
+), quote = FALSE, sep = "\t", row.names = FALSE, na = "")
 
 ################
 ### PCA PLOT ###
@@ -131,8 +137,6 @@ st_reviewInSitu <- data.table(reshape2::dcast(compare,
 st_reviewInSitu <- st_reviewInSitu[, .(
   Reference = paste(Reference, collapse = ", "))
   , by = c("msuId", "RM", "PBM", "SBM", "SM", "FM")]
-
-### THIS IS FAILING ###
 
 st_reviewInSitu[, `Gene symbol` :=
                   oryzr::LocToGeneName(msuId)$symbols,

### 2016-03-15 src/R/figures.R
use env to find rscript in figures.R

@@ -1,4 +1,4 @@
-#!/usr/bin/Rscript
+#!/usr/bin/env Rscript
 
 library(data.table)
 library(ggplot2)
